### 2022-01-12 doc/university-covid.tex
first commit?

\documentclass[12pt]{article}
\usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}

\usepackage{graphicx}
%\usepackage{adjustbox}

\newcommand{\eref}[1]{(\ref{eq:#1})}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\Fref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\sref}[1]{Sec.~\ref{#1}}
\newcommand{\frange}[2]{Fig.~\ref{fig:#1}--\ref{fig:#2}}
\newcommand{\tref}[1]{Table~\ref{tab:#1}}
\newcommand{\tlab}[1]{\label{tab:#1}}
\newcommand{\seminar}{SE\mbox{$^m$}I\mbox{$^n$}R}

\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}

\usepackage{lineno}
%\linenumbers

\usepackage[pdfencoding=auto, psdextra]{hyperref}

\bibliographystyle{chicago}
\usepackage{natbib}
\date{\today}

\usepackage{xspace}
\newcommand*{\ie}{i.e.\@\xspace}

\usepackage{color}

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}} 
\newcommand{\Ro}{\Rx{0}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}}
\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
\newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}

\newcommand{\jdnew}{\jd{NEW}}
\newcommand{\jddel}[1]{\jd{DELETE: #1}}

\begin{document}

\begin{flushleft}{
	\Large
	\textbf\newline{
		Modeling the spread of SARS-CoV-2 on a university campus: lessons from Princeton University
	}
}
\newline
\\

\end{flushleft}

\pagebreak

\section{Introduction}

Predicting and controlling the spread of SARS-CoV-2 remains a critical question as new variants continue to emerge and become dominant.
Throughout the pandemic, testing of both asymptomatic and symptomatic individuals has

University campuses provide unique test cases for understanding the

- Usual pandemic start

- modeling, understanding control
- testing frequency is important
- universities provide a particularly test case; asymptomatic infection and testing is very high

- little review of what other universities have done

- epideimc in relatively well controlled documented setting
- really well until omicron
- perhaps because of immunity waning

\section{Methods}

\subsection{Data descriptions}

\begin{figure}[!th]
\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton.pdf}
\caption{
\textbf{Dynamics of SARS-CoV-2 outbreaks in Princeton University}
(A--C) Epidemic trajectories across three semesters: Fall 2020 (A), Spring 2020 (B), and Fall 20201 (C).
Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in Princeton University.
Red lines represent the weekly number of cases in Mercer County.
Number of cases in Mercery County is obtained from \url{https://github.com/nytimes/covid-19-data}.
(D--F) Correlations between the weekly number of cases in Princeton University and in Mercer County.
Solid lines and shaded areas represent the estimated lienar regression lines and the associated 95\% CIs.
\label{fig:princeton}
}
\end{figure}

We analyze time series of COVID-19 case reports from Princeton University (\fref{princeton}A--C).
Princeton University is located in Mercer County, New Jersey, USA and consists of roughly 5000 undergraduate students, 3000 graduate students, and 7000 staffs and faculties.
All data are publicly available on Princeton University COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.
We also analyze time series of COVID-19 case reports from Mercer County, New Jersey, USA to compare the levels of community (\fref{princeton}D--F);
these data are taken from the COVID-19 data repository provided by New York Times (\url{https://github.com/nytimes/covid-19-data}).
We aggregate data at a weekly level to remove changes in testing patterns throughout a week.

For simplicity, we divide the time series into three time periods representing three semesters: Fall 2020 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020 (January 16, 2021--May 14, 2021; \fref{princeton}B), and Fall 2021 (August 14, 2021--December 24, 2021; \fref{princeton}C).
Throughout the study period, all students, staff and faculty members who were physically present for more than 8 hours on campus per week were required to asymptomatic testing with varying frequencies.
Asymptomatic individuals submitted self-collected saliva samples, from which the presence of SARS-CoV-2 was tested using Polymerase Chain Reaction (PCR) \swp{regular PCR? RT-PCR? what are we using?}.  
Asymptomatic individuals who tested positive were required to isolate for 10 days.
Symptomatic individuals were eligible for rapid PCR tests and had to isolate \swp{self-isolate?} until they received their test results---those who tested positive were required to isolate for at least 10 days after symptom onset ond were release 3 days after symptoms resolved.

During fall 2020, roughly 1000 grad students and 2000 staffs and faculties were present on campus and were subject to asymptomatic testing. 
Undergraduate students were not allowed to return back to campus, and so their numbers were minimal (<300).  
Both undergraduate students and graduate students were required to get tested twice a week, whereas staffs and faculties were required to get tested once a week.
The number of cases remained relatively low throughout the semester with peak occuring in early December (\fref{princeton}A), coinciding with the epidemic trajectory in Mercer County (\fref{princeton}D).
We note that a sudden decrease in the number of cases around Thanksgiving may reflect reduced number of tests (3852 and 2972 asymptomatic tests performed on the week ending November 20th and 27th, respectively) as well as actual transmission.
The highest number of cases was reported among staffs and faculties (169), followed by graduate students (41), and undergraduate students (4).

In the beginning of spring 2020, the number of cases suddenly increased before classes began (\fref{princeton}B), reflecting $\approx 3000$ undergraduate students returning to campus.
The classes remained virtual and testing protocol did not change (twice a week for undergraduate and graduate students, and once a week for staffs and faculties).
The number of cases persisted at similar levels as the fall semester and eventually decreased as classes ended and students went back home.
The highest number of cases was reported among staffs and faculties (111), followed by undergraduate students (101), and graduate students (29).

In fall 2021, all students, staffs, and faculties were required to be vaccinated \swp{with two doses?} with a very few exceptions for medical and religious reasons. 
By the beginning of the semesters, ??\% of students and ??\% of staffs and faculties were vaccinated.
As of January 11th, 99\% of undergraduate students, 98\% of graduate students, and 96\% of staffs and faculties have been vaccinated.
Vaccinated individuals were required to be tested once a week, while unvaccinated individuals were required to be tested twice a week.
In-person teaching and activities were allowed on campus.
Everyone was required to wear masks indoors with a few exceptions (e.g., when eating or drinking, or when teaching a small class).
The number of cases remained similar to previous semesters until November when the number of cases began to increase primarily among undergraduate students around Thanksgiving (\fref{princeton}C).  
On November 27th, 2021, Princeton University announced undergraduate students to be tested twice a week and limited the size of non-academic gatherings to 20 people.
The number of cases decreased slightly as classes ended but soon increased back up as the Omicron variant was introduced.

\subsection{Mathematical model}

Throughout the pandemic, mathematical models were used to inform policy decisions, including the frequency of asymptomatic tests.

\section{Results}

\section{Discussion}

\pagebreak

\begin{figure}[!th]
\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_simulation.pdf}
\caption{
\textbf{Dynamics of SARS-CoV-2 outbreaks in Princeton University}
(A, D, G) Time series comparisons of model predictions with observed data across ranges of paramters: basic reproduction number $\mathcal R_0$, scaling parameter for community transmission $\theta$, and the initial proportion of infected individuals $I_0$.
For each paramter combination, we simulate the model 100 times and calculate the sum of squared differences (SSQ) between the reported number of positive cases and the model-predicted number of positive cases. 
Heatmaps represent medians of the logged sum of squared differences.
(B, E, H) Model predictions. 
Solid lines represent median predictions.
Shaded areas represent 90\% quantiles for the best matching parameter set.
Points represent the observed data.
(C, F, I) Correlations between model predictions with observed data.
Colored solid lines and shaded areas represent the estimated lienar regression lines and the associated 95\% CIs.
Dashed lines represent the one-to-one line.
}
\end{figure}

\pagebreak

\begin{figure}[!th]
\includegraphics[width=\textwidth]{../figure_omicron/figure_omicron.pdf}
\caption{
\textbf{Scenario projects of the spread of omicron variant on university campus}
(A) The predicted daily number of new infections across a wide ranges assumptions about testing frequencies, basic reproduction number $\mathcal R_0$, and the number of booster shots given per day.
(B) The predicted effective susceptibility over time. 
Effective susceptibility is calculated by taking the mean of individual-level susceptibility of each student.
Solid lines and shaded areas represent the median and the corresponding 90\% quantiles across 100 simulations.
All other parameters are the same as in Figure 2.
}
\end{figure}

\pagebreak

\begin{figure}[!th]
\includegraphics[width=\textwidth]{../figure_omicron/figure_omicron_limit.pdf}
\caption{
\textbf{Scenario projects of the spread of omicron variant on university campus with limited isolation beds}
(A) The predicted numbers of individuals in isolation on a given day.
(B) The predicted daily number of new infections.
Solid lines and shaded areas represent the median and the corresponding 90\% quantiles across 100 simulations.
We assume $\mathcal R_0 = 8$, 30 booster shots given per day, and testing frequency of 3 days for these simulations.
All other parameters are the same as in Figure 3.
}
\end{figure}

\end{document}

### 2022-01-26 doc/university-covid.tex
ms

@@
 
-\section{Introduction}
+Predicting and controlling the spread of SARS-CoV-2 remains a critical public health and scientific question.
+Rapid, asymptomatic transmission of SARS-CoV-2 has hindered intervention efforts such as contact tracing.
+Stringent measures to limit social contacts has been effective in preventing transmission but are unsustainable.
+The recent development of vaccines have provided a safe means of reopening the society, but uncertainty remains in their long-term effectiveness as well as .
+In addition, as new variants continues to emerge and
 
-Predicting and controlling the spread of SARS-CoV-2 remains a critical question as new variants continue to emerge and become dominant.
-Throughout the pandemic, testing of both asymptomatic and symptomatic individuals has
+the emergence of new variants, which can evade immunity and transmit better, continues to pose additional challenges.
 
-University campuses provide unique test cases for understanding the
+Since the beginning of the pandemic, mathematical models have played a key role in evaluating intervention measures and guiding pandemic respones.
+
+testing has played a key role in monitoring and controlling the spread.
+In addition, a frequent testing of a large fraction of the population, followed by rapid isolation and contact tracing, has proven effective in preventing chains of transmission and reducing the spread in large congregate settings.
+
+University campuses provide unique test case in a relatively well-controlled setting for evaluating the feasibility of safe reopening.
+High density environemtns and large fraction of asymptomatic infections (due to young age of university students) can readily permit rapid transmission.
+
+.
 
@@
 
-\section{Methods}
-
-\subsection{Data descriptions}
+In this study, we analyze the spread of SARS-CoV-2 on Princeton University campus (\fref{princeton}).
+Princeton University is located in Mercer County, New Jersey, USA and consists of roughly 5000 undergraduate students, 3000 graduate students, and 7000 staffs and faculties.
+For simplicity, we divide the epidemic into three time periods representing three semesters: Fall 2020 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020 (January 16, 2021--May 14, 2021; \fref{princeton}B), and Fall 2021 (August 14, 2021--December 24, 2021; \fref{princeton}C).
+Throughout the study period, all students, staff and faculty members who were physically present for more than 8 hours on campus per week were required to participate in asymptomatic testing with varying frequencies.
+Asymptomatic individuals submitted self-collected saliva samples, from which the presence of SARS-CoV-2 was tested using Polymerase Chain Reaction (PCR) \swp{regular PCR?
+RT-PCR?
+what are we using?}.
+Asymptomatic individuals who tested positive were required to isolate for 10 days.
+Symptomatic individuals were eligible for rapid PCR tests and had to isolate \swp{self-isolate?} until they received their test results---those who tested positive were required to isolate for at least 10 days after symptom onset ond were release 3 days after symptoms resolved.
+Throughout the study period, contact tracing was also performed to quarantine close contacts of confirmed cases.
+Close contacts are defined as those who spent \swp{???}.
 
@@
 
-We analyze time series of COVID-19 case reports from Princeton University (\fref{princeton}A--C).
-Princeton University is located in Mercer County, New Jersey, USA and consists of roughly 5000 undergraduate students, 3000 graduate students, and 7000 staffs and faculties.
-All data are publicly available on Princeton University COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.
-We also analyze time series of COVID-19 case reports from Mercer County, New Jersey, USA to compare the levels of community (\fref{princeton}D--F);
-these data are taken from the COVID-19 data repository provided by New York Times (\url{https://github.com/nytimes/covid-19-data}).
-We aggregate data at a weekly level to remove changes in testing patterns throughout a week.
-
-For simplicity, we divide the time series into three time periods representing three semesters: Fall 2020 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020 (January 16, 2021--May 14, 2021; \fref{princeton}B), and Fall 2021 (August 14, 2021--December 24, 2021; \fref{princeton}C).
-Throughout the study period, all students, staff and faculty members who were physically present for more than 8 hours on campus per week were required to asymptomatic testing with varying frequencies.
-Asymptomatic individuals submitted self-collected saliva samples, from which the presence of SARS-CoV-2 was tested using Polymerase Chain Reaction (PCR) \swp{regular PCR?
-RT-PCR?
-what are we using?}.
-Asymptomatic individuals who tested positive were required to isolate for 10 days.
-Symptomatic individuals were eligible for rapid PCR tests and had to isolate \swp{self-isolate?} until they received their test results---those who tested positive were required to isolate for at least 10 days after symptom onset ond were release 3 days after symptoms resolved.
-
 During fall 2020, roughly 1000 grad students and 2000 staffs and faculties were present on campus and were subject to asymptomatic testing. 
@@
 Both undergraduate students and graduate students were required to get tested twice a week, whereas staffs and faculties were required to get tested once a week.
-The number of cases remained relatively low throughout the semester with peak occuring in early December (\fref{princeton}A), coinciding with the epidemic trajectory in Mercer County (\fref{princeton}D).
+The number of cases remained relatively low throughout the semester with peak occuring in early December, coinciding with the epidemic trajectory in Mercer County (\fref{princeton}A).
 We note that a sudden decrease in the number of cases around Thanksgiving may reflect reduced number of tests (3852 and 2972 asymptomatic tests performed on the week ending November 20th and 27th, respectively) as well as actual transmission.
@@
 
-In the beginning of spring 2020, the number of cases suddenly increased before classes began (\fref{princeton}B), reflecting $\approx 3000$ undergraduate students returning to campus.
+In the beginning of spring 2020, the number of cases suddenly increased before classes began (\fref{princeton}B), reflecting $\approx 3000$ undergraduate students returning to campus and bringing back infections.
+Returning students were required to be tested and quarantine \swp{for how long} if they were coming back from CDC-designated states \swp{not sure if this is the right wording...
+but I remember we had a rule like this...}.
 The classes remained virtual and testing protocol did not change (twice a week for undergraduate and graduate students, and once a week for staffs and faculties).
-The number of cases persisted at similar levels as the fall semester and eventually decreased as classes ended and students went back home.
+The number of cases persisted at similar levels as the fall semester and eventually decreased as classes ended and students went back home---the decrease in the number of cases in Princeton University also coincided with the decrease in the number of cases in Mercer County.
 The highest number of cases was reported among staffs and faculties (111), followed by undergraduate students (101), and graduate students (29).
@@
 On November 27th, 2021, Princeton University announced undergraduate students to be tested twice a week and limited the size of non-academic gatherings to 20 people.
-The number of cases decreased slightly as classes ended but soon increased back up as the Omicron variant was introduced.
+The number of cases decreased slightly as classes ended but soon increased back up as the Omicron variant was introduced on campus as well as in Mercer County.
 
-\subsection{Mathematical model}
+Across all three semesters, we find strong and significant correlations between the weekly logged numbers of cases from Princeton University and those from Mercer County:
+Fall 2020 ($\rho = 0.81$  (95\% CI: 0.55--0.92); $p < 0.001$; \fref{princeton}D); Spring 2020 ($\rho = 0.80$  (95\% CI: 0.51--0.92); $p < 0.001$; \fref{princeton}E); and Fall 2021 ($\rho = 0.90$  (95\% CI: 0.78--0.96); $p < 0.001$; \fref{princeton}F).
+Although we cannot infer causation from correlations, these strong correlations are likely indicative of community imports driving epidemic patterns on campus.
+Indeed, contact tracing analyses suggest that the majority of cases were linked with community transmission until fall 2021 when transmission clusters were identified on campus (e.g., holiday parties).
+\swp{Can we provide more quantitiative information than this?
+Or is IRB going to hate us?}
 
-Throughout the pandemic, mathematical models were used to inform policy decisions, including the frequency of asymptomatic tests.
+\section*{Mathematical model}
 
-\section{Results}
+We use a discrete-time, individual-based model to simulate the spread of SARS-CoV-2 in Princeton University and to evaluate effects of intervention measures, such as testing.
+The model consists of five main components that are simulated on a daily time scale: (1) transmission and infection dynamics, (2) sampling and testing protocols, (3) contact tracing, (4) quarantine and isolation protocols, and (5) vaccination dynamics, including waning immunity and booster shots.
+Here, we describe each component briefly (see Supplementary files for details).
 
-\section{Discussion}
+Infection processes are modeled based on standard compartmental models.
+Once infected, susceptible individuals enter the exposed stage, during which they cannot transmit or test positive. 
+After $D_e$ days on average, exposed individuals then enter the presymptomatic, during which they can test positive and transmit infections at rate $\beta_p$ for $D_p$ days on average.
+Presymptomatic individuals can then either remain asymptomatic with probability $p_a$ or develop symptoms with the remaining probability of $1-p_a$; both asymptomatic and symptomatic individuals are assumed to have the same duration of infection ($D_s$) and transmission rates ($\beta_s$) for simplicity.
+Recovered individuals are assumed to be immune to reinfections throughout a semester.
+Transmissions are modeled using a negative binomial distribution with an overdispersion parameter of $k=0.1$ to account for the possibility of super-spreading events.
+External (community) transmission is modeled using a Poisson distribution with a time-varying mean calculated by scaling the daily number of cases by $\theta$ and shifting it by 1 week to account for reporting delays.
 
-\pagebreak
+All individuals on campus are assumed follow a pre-determined asymptomatic testing plan at a fixed frequency for simplicity---
+for example, under weekly testing, one individual can get sampled on days 1, 8, and 15, whereas another inidividual get sampled on days 2, 9, and 16.
+We assume that test results come back after one day.
+Symptomatic individuals can choose to take rapid PCR tests (with results returning on the same day) with a given probability on each day until their symptoms resolve---this probability is set to 1 for simulations presented in the main text.
+We further assume that symptomatic individuals would isolate immediately when they submit their samples until they receive negative results.
+All individuals who test positive are required to isolate (following the same isolation rule as before) and are exempt from asymptomatic testing for 90 days.
+Isolated individuals are assuemd to no longer transmit infections.
+We assume that PCR tests have 95\% sensitivity and 100\% specificity.
+
+When an individual tests positive, contact tracing is performed to identify and quarantine their contacts.
+Contact tracing depend on two parameter: tracing efficacy, which we define as the proportion of secondary infections, and tracing delay, which represents the time between when an individual tests positive and when their contacts are quarantined.
+Following Princeton \swp{New Jersey? CDC?} guidelines for contact tracing, contacts are traced up to two days before symptom onset or two days before the positive test (for asymptomatic cases).
+Quarantined individuals are released after 14 days.
+We do not model quarantining of non-infectious contacts as they do not affect epidemic dynamics.
+For simplicity, we do not model contact tracing during parameter matching.
+
+As most students, staffs, and faculties had received two doses of vaccination in the beginning of fall 2021, we do not distinguish the first and second doses.
+Instead, we assume that vaccines reduce susceptibility as well as transmissibility of all vaccinated individuals by an equal amount for all vaccinated individuals.
+Vaccine efficacies against both infection and transmission are allowed to exponentially wane from 90\% to 50\% in 20 weeks (and continues to wane at the same rate) for each vaccinated individuals.
+
+This model was developed and used throughout the pandemic to inform policy decisions in Princeton University, including the frequency of asymptomatic tests and the number of isolation beds required.
+We continuously updated the model to reflect changes in school settings (e.g., students returning back to campus in spring 2020) as well as intervention measures (e.g., vaccination in fall 2021 and booster shots with the emergence of the Omicron variant).
+The model we use here is a generic version that encompasses all of the necessary details to model the spread of SARS-CoV-2 in Princeton University.
+Here, we use this model to retrospectively analyze past outbreaks and further predict what future epidemic trajectories might look like as the new Omicron variant spreads on campus.
+Simulation details are provided in the supplementary materials.
+
+First, we try to match our model to epidemic patterns seen on campus by varying the basic reprodcution number $\mathcal R_0$ and the amount of community transmission $\theta$ and holding all other parameters constant.
+For each parameter combination, we simulate 100 epidemic trajectories and calculate the sum of squared differences between the weekly numbers of the observed and predicted positive cases.
+The population size and testing frequencies (with twice weekly testing modeled as testing every 3 days) are set to previously discussed values to reflect realistic campus settings.
 
@@
 Dashed lines represent the one-to-one line.
+\label{fig:matching}
 }
@@
 
-\pagebreak
+We find that a low value of $\mathcal R_0=0.5$ and a small amount of community transmission $\theta=0.015$ is most consistent with the observed epidemic dynamics in fall 2020 (\fref{matching}A).
+With these parameters, the model is able to capture the rise and fall in the number of cases with the exception of sudden decrease in the number of cases around Thanksgiving, which we do not model explicitly (\fref{matching}B).
+The median predictions are positively correlated with the observed dynamics ($\rho = 0.83$; 95\% CI: 0.61--0.93; \fref{matching}C).
+Although a wide range of assumptions about the levels of community transmission $\theta$ are consistent with the observed dynamics, our simulations preclude high $\mathcal R_0 > 2$ (SUPP).
+Distancing measures on campus, including virtual classes, likely contributed to lowering $\mathcal R_0$.
+
+A similar set of parameters can capture the observed dynamics in spring 2020.
+The best matching parameter predicts a slightly higher levels of community transmission $\theta=0.02$ (\fref{matching}D), but a wide range of parameters are consistent with the observed dynamics as before (SUPP).
+Simulations also preclude high $\mathcal R_0 > 2$ again, suggesting that the contact levels between students likely remained low even though they had returned to campus---low $\mathcal R_0$ may be attributed in part to the absence of in-person teaching.
+We note that the introduction of infections from returning students are necessary to capture the initial rise in the number of cases (\fref{matching}E).
+Once again, we find a positive (but slightly weaker) correlation between the predicted and the observed numbers of cases ($\rho = 0.62$; 95\% CI: 0.20--0.85; \fref{matching}F).
+
+For fall 2021, we limited our model comparison to November 26th before the Omicron variant was introduced on campus.
+We assume that $>95\%$ of the school population is vaccinated at the beginning of the semester and do not model additional doses throughout the semester for simplicity.
+We also allow vaccine-derived immunity to wane over time to ask whether the increase in the number of cases around November is consistent with the dynamics predicted by immunity waning.
+
+Even though the numbers of cases during fall 2021 (before a large outbreak) were similar to those during previous semesters, we find that considerably higher levels of community transmission $\theta$ (~10 fold higher) is required to explain the observed dynamics due to a decreased susceptibility from vaccination (\fref{matching}G).
+We note that the parameter $\theta$ necessarily depends on our assumed vaccine efficacy, and $\theta$ would decrease if we assume a lower vaccine efficacy.
+Nonetheless, the amount of community transmission would still need to be higher than previous semesters as long as vaccine provides some protection against infection.
+
+We also find that the predicted dynamics are largely insensitive to $\mathcal R_0$ until November 26th (\fref{matching}H)---
+in addition, all simulations shown in \fref{matching}H are positively correlated with the observed numbers of cases (\fref{matching}G). 
+An increasing pattern in the sum of squared errors with $\mathcal R_0$ shown in \fref{matching}G is primarily driven by the discrepancy around fall break (week ending October 26th) when the number of cases decreased suddenly, which we do not model explicitly.
+As nearly all school population was vaccinated, transmission between individuals on campus would have been limited, making $\mathcal R_0$ difficult to estimate.
+These simulations also suggest that an increase in the number of cases in November can be explained by a combination of waning immunity alone without requiring additional changes in transmission dynamics (note we do not allow $\theta$ or $\mathcal R_0$ to vary over time).
+
+Projecting the model beyond November 26th shows that we would have seen a similarly growth in the number of cases if conditions remained constant even without the introduction of the Omicron variant.
+In other words, the Delta strain would have continued to spread on campus at a similar rate if the semester were to (hypothetically) continue until January without additional interventions (\fref{matching}H).
+In reality, the situation was more complex:  testing frequencies increased and social gatherings were limited in response to an increase in the number of cases.
+These interventions---as well as students returning back home as classes ended---likely would have reduced contact rates (and therefore transmission of the Delta variant).
+This reduction in transmission was likely counterbalanced by the introduction of the Omicron variant and its immune evasion, leading to similar and persistent growth in the number of cases.
+
+We then project epidemic trajectories for the spread of the Omicron variant among 4000 students for the spring semester of 2021 across a wide range of basic reproduction number ($\mathcal R_0 = 2--16$) and testing frequencies (every 2, 4, and 7 days).
+We assume a two-dose efficacy 10\% and a three-dose efficacy of 70\% against the infection and transmission of the Omicron variant. 
+We further assume that immunity from the third dose takes 14 days to develop and can wane at the same rate as previously assumed (in this case, 70\% to 39\% in 20 weeks).
+For simplicity, we assume a constant amount of external infections.
+
+\fref{omicron} shows predicted trajectories of new infections per day (A) and the mean susceptibility over time (B).
+Across a wide range of $\mathcal R_0$, we find that frequent testing alone is not sufficient to prevent an outbreak---while testing and isolation can prevent onward transmission once infected, it cannot prevent infections from happening especially if they are coming from outisde of the population.
+Administering booster shots at a faster rate (compare 30 vs 60 booster shots per day in \fref{omicron}A) can help reduce the size of the epidemic peak and cause the epidemic to decay faster, their overall impact is limited due to delays in developing immunity and imperfect protection against the Omicron variant.
+Comparing changes in the mean susceptibility further reveal that administering booster shots at a faster rate can also lead to earlier immunity waning and higher mean susceptibility by the end of the semester, which may increase chances of an outbreak in the future \fref{omicron}.
 
@@
 All other parameters are the same as in Figure 2.
+\label{fig:omicron}
 }
@@
 
-\pagebreak
+So far we have assumed that all individuals who test positive are isolated for 10 days;
+in practice,  however, isolation of infected individuals can be limited by the isolation bed capacity.
+As the Omicron variant to spread rapidly on university campuses (\fref{omicron}), universities participating in testing plans can expect to face difficulties in isolating infected individuals.
+Here, we explore how limitintg the number of isolation beds on university campus can affect epidemic dynamics by assuming that individuals who test positive will not be isolated and will continue to transmit when isolation beds are full.
+In doing so, we also consider the effects of shorter isolation periods (5, 7, and 10 days).
 
@@
 All other parameters are the same as in Figure 3.
+\label{fig:isolation}
 }
@@
 
+When there are no limits to the number of isolation beds on campus, shortening isolation periods results in a slightly higher number of total infections (\fref{isolation}A): .
+The epidemic trajectory remains similar across different values of isolation periods (\fref{isolation}B), but the number of students in isolation becomes considerably lower with shorter isolation periods (\fref{isolation}C).
+For example, the maximum number of students in isolation decreases from ??? to ???.
+
+When the number of isolation beds is moderately limited ($=200$), shortening isolation periods can decrease the total numbers of infections (\fref{isolation}A) and considerably reduce the size of the epidemic peak (\fref{isolation}B).
+Even when the number of isolation beds is severly limited ($=100$), shortening isolation periods can still reduce the size of the epidemic peak (\fref{isolation}B) because we are able to isolate a greater proportion of infected individuals and prevent onward transmission.
+Longer isolation can prevent a greater amount of onward transmission \emph{per individual} but permits a lower number of individuals in isolation, making it less effectivess when the number of isolation beds are limited.
+
+\section{Discussion}
+
+Here, we present the analysis of SARS-CoV-2 outbreaks on Princeton University campus between fall 2020 and winter 2022.
+Our analysis demonstrates strong spatiotemporal correlations between the patterns of spread of SARS-CoV-2 on university campus and those from surrounding communities.
+These correlations decrease with distance from Mercer County in fall 2021, likely reflecting contact and commuting patterns as university re-opened.
+Mathematical modeling further suggests limited transmission between the university population during fall and spring semesters of 2020 and an increased amount of community contact during the fall semester of 2021 compared previous semesters.
+An increase in the number of cases by the end of November 2021 is consistent with the increase in the amount of community contacts and waning immunity.
+Finally, we predict the spread of Omicron variant to likely cause a large outbreak in the spring semester of 2021 even under more stringent testing protocols.
+A rapid administration of booster shots and shortening isolation periods can help reduce the size of an outbreak, but uncertainty remains for future semesters.
+
+\section{Methods}
+
+We analyze time series of COVID-19 case reports from Princeton University (\fref{princeton}A--C).
+Princeton University is located in Mercer County, New Jersey, USA and consists of roughly 5000 undergraduate students, 3000 graduate students, and 7000 staffs and faculties.
+All data are publicly available on Princeton University COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.
+We also analyze time series of COVID-19 case reports from Mercer County, New Jersey, USA to compare the levels of community (\fref{princeton}D--F);
+these data are taken from the COVID-19 data repository provided by New York Times (\url{https://github.com/nytimes/covid-19-data}).
+We aggregate data at a weekly level to remove changes in testing patterns throughout a week.
+
 \end{document}

### 2022-02-04 doc/university-covid.tex
lots and lots of edits

--- 2022-01-12 doc/university-covid.tex
@@
 	\textbf\newline{
-		Modeling the spread of SARS-CoV-2 on a university campus: lessons from Princeton University
+		The role of community transmission in driving the spread of SARS-CoV-2 on a university campus: lessons from Princeton University
 	}
@@
 
-\section{Introduction}
+Predicting and controlling the spread of SARS-CoV-2 remains a critical public health and scientific question.
+Rapid, asymptomatic transmission of SARS-CoV-2 has hindered intervention efforts such as contact tracing.
+Stringent measures to limit social contacts have been effective in preventing transmission in the beginning of the pandemic but are unsustainable.
+The recent development of vaccines have provided a safe means of reopening the society, but uncertainty remains in their long-term effectiveness in protecting infection and transmission.
+In addition, as new variants continues to emerge and
 
-Predicting and controlling the spread of SARS-CoV-2 remains a critical question as new variants continue to emerge and become dominant.
-Throughout the pandemic, testing of both asymptomatic and symptomatic individuals has
+the emergence of new variants, which can evade immunity and transmit better, continues to pose additional challenges.
 
-University campuses provide unique test cases for understanding the
+Since the beginning of the pandemic, mathematical models have played a key role in evaluating intervention measures and guiding pandemic response.
+
+testing has played a key role in monitoring and controlling the spread.
+In addition, a frequent testing of a large fraction of the population, followed by rapid isolation and contact tracing, has proven effective in preventing chains of transmission and reducing the spread in large congregate settings.
+
+University campuses provide unique test case in a relatively well-controlled setting for evaluating the feasibility of safe reopening.
+High density environments and large fraction of asymptomatic infections (due to young age of university students) can readily permit rapid transmission.
 
@@
 
-\section{Methods}
-
-\subsection{Data descriptions}
+In this study, we analyze the spread of SARS-CoV-2 on Princeton University campus (\fref{princeton}).
+Princeton University is located in Mercer County, New Jersey, USA and consists of roughly 5000 undergraduate students, 3000 graduate students, and 7000 faculty and staff members.
+For simplicity, we divide the epidemic into three time periods representing four semesters: Fall 2020 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020 (January 16, 2021--May 14, 2021; \fref{princeton}B), and Fall and Winter 2021 (August 14, 2021--January 14, 2021; \fref{princeton}C).
+Throughout the study period, all students, faculty and staff members who were physically present for more than 8 hours on campus per week were required to participate in asymptomatic testing with varying frequencies.
+Asymptomatic individuals submitted self-collected saliva samples, from which the presence of SARS-CoV-2 was tested using Reverse Transcription Polymerase Chain Reaction (RT-PCR).  
+Asymptomatic individuals who tested positive were required to isolate for 10 days.
+Symptomatic individuals were eligible for rapid PCR tests and had to isolate \swp{self-isolate?} until they received their test results---those who tested positive were required to isolate for at least 10 days after symptom onset and were release 3 days after symptoms resolved.
+Those who tested positive were exempt from asymptomatic testing for 90 days.
+Throughout the study period, contact tracing was also performed for positive cases to quarantine their close contacts, which refer to individuals who have been within 6 feet for at least 15 minutes, cumulatively within a 24-hour period.
+All data used in this analysis are publicly available on Princeton University COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.
 
@@
 Red lines represent the weekly number of cases in Mercer County.
-Number of cases in Mercery County is obtained from \url{https://github.com/nytimes/covid-19-data}.
+Number of cases in Mercer County is obtained from \url{https://github.com/nytimes/covid-19-data}.
 (D--F) Correlations between the weekly number of cases in Princeton University and in Mercer County.
-Solid lines and shaded areas represent the estimated lienar regression lines and the associated 95\% CIs.
+Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
 \label{fig:princeton}
@@
 
-We analyze time series of COVID-19 case reports from Princeton University (\fref{princeton}A--C).
-Princeton University is located in Mercer County, New Jersey, USA and consists of roughly 5000 undergraduate students, 3000 graduate students, and 7000 staffs and faculties.
-All data are publicly available on Princeton University COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.
-We also analyze time series of COVID-19 case reports from Mercer County, New Jersey, USA to compare the levels of community (\fref{princeton}D--F);
-these data are taken from the COVID-19 data repository provided by New York Times (\url{https://github.com/nytimes/covid-19-data}).
-We aggregate data at a weekly level to remove changes in testing patterns throughout a week.
-
-For simplicity, we divide the time series into three time periods representing three semesters: Fall 2020 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020 (January 16, 2021--May 14, 2021; \fref{princeton}B), and Fall 2021 (August 14, 2021--December 24, 2021; \fref{princeton}C).
-Throughout the study period, all students, staff and faculty members who were physically present for more than 8 hours on campus per week were required to asymptomatic testing with varying frequencies.
-Asymptomatic individuals submitted self-collected saliva samples, from which the presence of SARS-CoV-2 was tested using Polymerase Chain Reaction (PCR) \swp{regular PCR?
-RT-PCR?
-what are we using?}.
-Asymptomatic individuals who tested positive were required to isolate for 10 days.
-Symptomatic individuals were eligible for rapid PCR tests and had to isolate \swp{self-isolate?} until they received their test results---those who tested positive were required to isolate for at least 10 days after symptom onset ond were release 3 days after symptoms resolved.
-
-During fall 2020, roughly 1000 grad students and 2000 staffs and faculties were present on campus and were subject to asymptomatic testing. 
-Undergraduate students were not allowed to return back to campus, and so their numbers were minimal (<300).  
-Both undergraduate students and graduate students were required to get tested twice a week, whereas staffs and faculties were required to get tested once a week.
-The number of cases remained relatively low throughout the semester with peak occuring in early December (\fref{princeton}A), coinciding with the epidemic trajectory in Mercer County (\fref{princeton}D).
-We note that a sudden decrease in the number of cases around Thanksgiving may reflect reduced number of tests (3852 and 2972 asymptomatic tests performed on the week ending November 20th and 27th, respectively) as well as actual transmission.
-The highest number of cases was reported among staffs and faculties (169), followed by graduate students (41), and undergraduate students (4).
+During fall 2020, roughly 1000 grad students and 2000 faculty and staff members were present on campus and participated in asymptomatic testing. 
+Undergraduate students were not allowed to return back to campus as all classes were held virtually, and so their numbers were minimal on campus ($<300$).  
+Both undergraduate students and graduate students were required to get tested twice a week, whereas faculty and staff members were required to get tested once a week.
+The number of cases remained relatively low throughout the semester with a peak occurring in early December, coinciding with the epidemic trajectory in Mercer County (\fref{princeton}A).  
+A sudden decrease in the number of cases around Thanksgiving partly reflects the reduced number of tests (3852 and 2972 asymptomatic tests performed on the week ending November 20th and 27th, respectively).
+The highest number of cases was reported among faculty and staff members (169), followed by graduate students (41), and undergraduate students (4).
 
 In the beginning of spring 2020, the number of cases suddenly increased before classes began (\fref{princeton}B), reflecting $\approx 3000$ undergraduate students returning to campus.
-The classes remained virtual and testing protocol did not change (twice a week for undergraduate and graduate students, and once a week for staffs and faculties).
-The number of cases persisted at similar levels as the fall semester and eventually decreased as classes ended and students went back home.
-The highest number of cases was reported among staffs and faculties (111), followed by undergraduate students (101), and graduate students (29).
+Returning students were required to be tested and quarantine \swp{for how long} if they were coming back from CDC-designated states \swp{not sure if this is the right wording...
+but I remember we had a rule like this...
+we can ask Irini and Robin for this}.
+All classes remained virtual, and the testing protocol did not change (twice a week for undergraduate and graduate students, and once a week for faculty and staff members).
+The number of cases persisted at similar levels as the fall semester and eventually decreased as classes ended and students went back home---the decrease in the number of cases in Princeton University also coincided with the decrease in the number of cases in Mercer County.
+The highest number of cases was reported among faculty and staff members (111), followed by undergraduate students (101), and graduate students (29).
 
-In fall 2021, all students, staffs, and faculties were required to be vaccinated \swp{with two doses?} with a very few exceptions for medical and religious reasons. 
-By the beginning of the semesters, ??\% of students and ??\% of staffs and faculties were vaccinated.
-As of January 11th, 99\% of undergraduate students, 98\% of graduate students, and 96\% of staffs and faculties have been vaccinated.
+In fall 2021, all students and faculty and staff members were required to be vaccinated \swp{with two doses? need to ask Robin} with a very few medical and religious exemptions.
+By the beginning of the semesters, ??\% of students and ??\% of faculty and staff members were vaccinated \swp{ask Robin}.
+As of January 11th, 99\% of undergraduate students, 98\% of graduate students, and 96\% of faculty and staff members were vaccinated.
 Vaccinated individuals were required to be tested once a week, while unvaccinated individuals were required to be tested twice a week.
-In-person teaching and activities were allowed on campus.
-Everyone was required to wear masks indoors with a few exceptions (e.g., when eating or drinking, or when teaching a small class).
+In-person classes and social events fully resumed on campus while all individuals were required to wear masks indoors with a few exceptions (e.g., when eating or drinking, or when teaching a small class).
 The number of cases remained similar to previous semesters until November when the number of cases began to increase primarily among undergraduate students around Thanksgiving (\fref{princeton}C).  
-On November 27th, 2021, Princeton University announced undergraduate students to be tested twice a week and limited the size of non-academic gatherings to 20 people.
-The number of cases decreased slightly as classes ended but soon increased back up as the Omicron variant was introduced.
+In order to prevent transmission, testing frequency was increased to twice a week for undergraduate students on November 27th, 2021; the size of non-academic gatherings were also limited to 20 people.
+The number of cases decreased slightly as classes ended but soon increased back up as the Omicron variant began to spread on campus and in Mercer county.
 
-\subsection{Mathematical model}
+Across all three semesters, we find strong and significant correlations between the weekly logged numbers of cases from Princeton University and those from Mercer county:
+fall 2020 ($\rho = 0.81$ (95\% CI: 0.55--0.92; $p < 0.001$), \fref{princeton}D); spring 2020 ($\rho = 0.80$ (95\% CI: 0.51--0.92; $p < 0.001$), \fref{princeton}E); and fall 2021 ($\rho = 0.90$ (95\% CI: 0.78--0.96; $p < 0.001$), \fref{princeton}F).
+These correlations are robust even when we stratify cases by the population, except for undergraduate students during Fall 2020 when most of them were not physically present on campus (Supplementary Figure S1).
+We also find strong correlations between the weekly logged numbers of cases from Princeton University and those from other counties in New Jersey (Supplementary Figure S2)---these correlations significantly decreased with distance from Mercer county in both spring 2020 ($\rho=-0.48$ (95\% CI: $-0.75$--$-0.06$; $p = 0.03$) for all cases and $\rho=-0.51$ (95\% CI: $-0.77$--$-0.10$; $p = 0.02$) for faculty and staff cases) and fall 2021 ($\rho=-0.68$ (95\% CI: $-0.86$--$-0.35$, $p < 0.001$) for all cases and $\rho=-0.74$ (95\% CI: $-0.89$--$-0.46$, $p < 0.001$) for faculty and staff cases).
 
-Throughout the pandemic, mathematical models were used to inform policy decisions, including the frequency of asymptomatic tests.
+These correlations likely reflect commuting and contact patterns, and therefore we expect SARS-CoV-2 dynamics on campus to be correlated with those from nearby large cities as well. 
+We find similarly strong correlations with New York City: fall 2020 ($\rho = 0.74$ (95\% CI: 0.44--0.90; $p < 0.001$), Supplementary Figure S3A); spring 2020 ($\rho = 0.84$ (95\% CI: 0.61--0.94; $p < 0.001$), Supplementary Figure S3B); and fall 2021 ($\rho = 0.86$ (95\% CI: 0.70--0.94; $p < 0.001$), Supplementary Figure S3C).
+We also find similarly strong correlations with Philadelphia except for spring 2020: fall 2020 ($\rho = 0.83$ (95\% CI: 0.59--0.93; $p < 0.001$), Supplementary Figure S4A); spring 2020 ($\rho = 0.29$ (95\% CI: $-0.22$--0.68; $p = 0.25$), Supplementary Figure S4B); and fall 2021 ($\rho = 0.86$ (95\% CI: 0.68--0.94; $p < 0.001$), Supplementary Figure S4C).
+Including counties from New York and Pennsylvania states into the spatial correlation analysis yields additional insights (Supplementary Figure S5):
+epidemic dynamics were highly synchronized across all counties in fall 2020 and became less synchronized over time. 
+These correlations significantly decreased with distance in spring 2020 ($\rho = -0.36$ (95\% CI: $-0.49$--$-0.21$; $p < 0.001$)) and fall 20201 ($\rho = -0.58$ (95\% CI: $-0.68$--$-0.46$; $p < 0.001$)).
+These variations likely reflect differences in vaccination levels and the timing of the introduction of the Omicron variant.
 
-\section{Results}
+Finally, we calculate the ratio between the weekly numbers of cases from Princeton and those from Mercer county---
+we expect this ratio to remain constant over time if the majority of the infections in Princeton University is caused by community transmission provided that testing patterns remain constant in both places.
+We do not find significant changes in the ratio between the weekly numbers of cases from Princeton and those from Mercer county (\fref{princeton}G).
+Instead, the mean ratios remained similar across three semesters despite changes in campus populations (e.g., undergraduate student returning in spring 2020) and vaccination levels: 0.034 (95\% CI: 0.021--0.047) for fall 2020, 0.019 (95\% CI: 0.015--0.023) for spring 2020, and 0.046 (95\% CI: 0.035--0.057) for fall 2021.
+However, this does not necessarily imply that the contact or transmission levels stayed constant over time: given extremely high vaccination coverage in Princeton University during fall 2021, the university population would been subject to a much higher forces of infection (e.g., a greater amount of mixing with the community population) to obtain similar levels of cases.
+We also find that variations in the ratio around the mean can be explained by holiday effects.
+For example, sudden decreases in the number of cases around Thanksgiving of 2020, fall recess of 2021, and end of classes of 2021 correspond to a reduction in the ratio between university and community cases.
+Similarly, a sudden increase in the number of cases during Thanksgiving of 2021 caused the ratio to increase, reflecting decoupling of epidemic dynamics due to on-campus transmission.
 
-\section{Discussion}
+\section*{Mathematical model}
 
-\pagebreak
+We use a discrete-time, individual-based model to simulate the spread of SARS-CoV-2 in Princeton University.
+This model was initially developed and used throughout the pandemic to inform policy decisions in Princeton University, including the frequency of asymptomatic tests and the number of isolation beds required.
+We continuously updated the model to reflect changes in school settings (e.g., students returning back to campus after a virtual semester) as well as intervention measures (e.g., vaccination in fall 2021 and booster shots with the emergence of the Omicron variant).
+Here, we present a generic version that encompasses all of the necessary details to characterize the spread of SARS-CoV-2 in Princeton University.
+The model consists of four main components that are simulated on a daily time scale: (1) transmission and infection dynamics, (2) sampling and testing protocols, (3) isolation protocols, and (4) vaccination dynamics, including waning immunity and booster shots.
+Previous versions of the model included contact tracing, but we exclude it in this model for simplicity.
+Here, we describe each component briefly. 
+Mathematical details are provided in Supplementary Materials.
+
+Infection processes are modeled based on standard compartmental structures.
+Once infected, susceptible individuals enter the exposed stage, during which they cannot transmit or test positive. 
+After $D_e$ days on average, exposed individuals then enter the presymptomatic stage, during which they can test positive and transmit infections at rate $\beta_p$ for $D_p$ days on average.
+Presymptomatic individuals can then either remain asymptomatic with probability $p_a$ or develop symptoms with the remaining probability of $1-p_a$; both asymptomatic and symptomatic individuals are assumed to have the same duration of infectiousness ($D_s$) and transmission rates ($\beta_s$) for simplicity.
+Recovered individuals are assumed to be immune to reinfections throughout a semester.
+Transmissions are modeled using a negative binomial distribution with an overdispersion parameter of $k=0.1$ to account for the possibility of super-spreading events.
+External (community) transmission is modeled using a Poisson distribution with a time-varying mean calculated by scaling the daily number of cases by $\theta$ and shifting it by 1 week to account for reporting delays.
+
+All individuals on campus are assumed follow a pre-determined asymptomatic testing plan at a fixed frequency---
+for example, under weekly testing, one individual can get sampled on days 1, 8, 15, and so forth, while another individual get sampled on days 2, 9, 16, and so forth.
+We assume that test results come back after one day.
+Symptomatic individuals can choose to take rapid PCR tests (with results returning on the same day) with a given probability on each day until their symptoms resolve---this probability is set to 1 for simulations presented in the main text.
+We further assume that symptomatic individuals would isolate immediately when they submit their samples until they receive negative results.
+All individuals who test positive are required to isolate (following the same isolation rule as before) and are exempt from asymptomatic testing for 90 days.
+Isolated individuals are assumed to no longer transmit infections.
+We assume that PCR tests have 95\% sensitivity and 100\% specificity.
+
+As most students as well as faculty and staff members had received two doses of vaccination in the beginning of fall 2021, we do not distinguish the first and second doses.
+Instead, we assume that vaccines reduce susceptibility as well as transmissibility of all vaccinated individuals by an equal amount for all vaccinated individuals.
+Vaccine efficacies are allowed to exponentially wane from 90\% to 50\% in 20 weeks (and continues to wane at the same rate) for each vaccinated individuals.
+While 90\% efficacy against susceptibility is consistent with recent estimates, 90\% efficacy against transmission is likely optimistic.
+For example, \cite{prunas2022vaccination} recently estimated that vaccination with BNT162b2 reduces susceptibility by 89.4\% (95\% CI: 88.7\%--90.0\%) and infectiousness by 23.0\% (95\% CI: $-11.3\%$--46.7\%) against the Delta strain.
+In Supplementary Materials, we present the same set of simulations, which assume that vaccines do not reduce transmissibility.
+
+Here, we use this model to retrospectively analyze past outbreaks and further predict what future epidemic trajectories might look like as the new Omicron variant spreads on campus (see Supplementary Materials for simulation details).
+First, we try to match our model to epidemic patterns seen on campus by varying the effective reproduction number $\mathcal R_e$ and the amount of community transmission $\theta$ and holding all other parameters constant.
+Here, the effective reproduction number implicitly accounts for all intervention measures that we do not model explicitly, such as social distancing and contact tracing---in other words, this effective reproduction number does not account for asymptomatic testing or vaccination, which are modeled separately.
+For each parameter combination, we simulate 100 epidemic trajectories and calculate the sum of squared differences between the weekly numbers of the observed and predicted positive cases.
+The population size and testing frequencies (with twice weekly testing modeled as testing every 3 days) are set to previously discussed values to reflect realistic campus settings.
 
@@
 \textbf{Dynamics of SARS-CoV-2 outbreaks in Princeton University}
-(A, D, G) Time series comparisons of model predictions with observed data across ranges of paramters: basic reproduction number $\mathcal R_0$, scaling parameter for community transmission $\theta$, and the initial proportion of infected individuals $I_0$.
-For each paramter combination, we simulate the model 100 times and calculate the sum of squared differences (SSQ) between the reported number of positive cases and the model-predicted number of positive cases.
-Heatmaps represent medians of the logged sum of squared differences.
+(A, D, G) Time series comparisons of model predictions with observed data across ranges of effective reproduction number $\mathcal R_e$ and scaling parameter for community transmission $\theta$.
+For each parameter combination, we simulate the model 100 times and calculate the sum of squared differences (SSQ) between the reported number of positive cases and the model-predicted number of positive cases.
+Heat maps represent medians of the logged sum of squared differences.
 (B, E, H) Model predictions. 
@@
 (C, F, I) Correlations between model predictions with observed data.
-Colored solid lines and shaded areas represent the estimated lienar regression lines and the associated 95\% CIs.
+Colored solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
 Dashed lines represent the one-to-one line.
+\label{fig:matching}
 }
@@
 
-\pagebreak
+We find that a low value of $\mathcal R_e=0.5$ and a small amount of community transmission $\theta=0.015$ is most consistent with the observed epidemic dynamics in fall 2020 (\fref{matching}A).
+With these parameters, the model is able to capture the rise and fall in the number of cases with the exception of sudden decrease in the number of cases around Thanksgiving, which we do not model explicitly (\fref{matching}B).
+The median predictions are positively correlated with the observed dynamics ($\rho = 0.83$; 95\% CI: 0.61--0.93; \fref{matching}C).
+Although a wide range of assumptions about the levels of community transmission $\theta$ are consistent with the observed dynamics, our simulations preclude high $\mathcal R_e > 2$ (SUPP).
+Distancing measures on campus and contact tracing efforts likely contributed to lowering $\mathcal R_e$.
+
+A similar set of parameters can capture the observed dynamics in spring 2020.
+The best matching parameter predicts a slightly higher levels of community transmission $\theta=0.02$ (\fref{matching}D), but a wide range of parameters are consistent with the observed dynamics as before (SUPP).
+Simulations also preclude high $\mathcal R_e > 2$ again, suggesting that the contact levels between students likely remained low even though they had returned to campus---the absence of in-person teaching is likely to have contributed to lowering $\mathcal R_e$.
+Once again, we find a positive (but slightly weaker) correlation between the predicted and the observed numbers of cases ($\rho = 0.62$; 95\% CI: 0.20--0.85; \fref{matching}F).
+
+For fall 2021, we limited our model comparison to November 26th before the Omicron variant was introduced on campus.
+We assume that $>95\%$ of the school population is vaccinated at the beginning of the semester and do not model additional doses throughout the semester for simplicity.
+We also allow vaccine-derived immunity to wane over time to ask whether the increase in the number of cases around November is consistent with the dynamics predicted by immunity waning.
+
+Even though the numbers of cases during fall 2021 (before a large outbreak) were similar to those during previous semesters, we find that considerably higher levels of community transmission $\theta$ (~10 fold higher) is required to explain the observed dynamics due to a decreased susceptibility from vaccination (\fref{matching}G).
+We note that the parameter $\theta$ necessarily depends on our assumed vaccine efficacy, and $\theta$ would decrease if we assume a lower vaccine efficacy.
+Nonetheless, the amount of community transmission would still need to be higher than previous semesters as long as vaccine provides some protection against infection.
+
+We also find that the predicted dynamics are largely insensitive to $\mathcal R_e$ until November 26th (\fref{matching}H)---
+in addition, all simulations shown in \fref{matching}H are positively correlated with the observed numbers of cases (\fref{matching}G). 
+An increasing pattern in the sum of squared errors with $\mathcal R_e$ shown in \fref{matching}G is primarily driven by the discrepancy around fall break (week ending October 26th) when the number of cases decreased suddenly, which we do not model explicitly.
+As nearly all school population was vaccinated, transmission between individuals on campus would have been limited, making $\mathcal R_e$ difficult to estimate.
+These simulations also suggest that an increase in the number of cases in November can be explained by a combination of waning immunity alone without requiring additional changes in transmission dynamics (note we do not allow $\theta$ or $\mathcal R_e$ to vary over time).
+
+Projecting the model beyond November 26th shows that we would have seen a similarly growth in the number of cases if conditions remained constant even without the introduction of the Omicron variant.
+In other words, the Delta strain would have continued to spread on campus at a similar rate if the semester were to (hypothetically) continue until January without additional interventions (\fref{matching}H).
+In reality, the situation was more complex:  testing frequencies increased and social gatherings were limited in response to an increase in the number of cases.
+These interventions---as well as students returning back home as classes ended---likely would have reduced contact rates (and therefore transmission of the Delta variant).
+This reduction in transmission was likely counterbalanced by the introduction of the Omicron variant and its immune evasion, leading to similar and persistent growth in the number of cases.
+
+We then project epidemic trajectories for the spread of the Omicron variant among 4000 students for the spring semester of 2021 across a wide range of effective reproduction number ($\mathcal R_e = 2--16$) and testing frequencies (every 2, 4, and 7 days).
+We assume a two-dose efficacy 10\% and a three-dose efficacy of 70\% against the infection and transmission of the Omicron variant. 
+We further assume that immunity from the third dose takes 14 days to develop and can wane at the same rate as previously assumed (in this case, 70\% to 39\% in 20 weeks).
+For simplicity, we assume a constant amount of external infections.
+
+\fref{omicron} shows predicted trajectories of new infections per day (A) and the mean susceptibility over time (B).
+Across a wide range of $\mathcal R_e$, we find that a frequent testing alone is not sufficient to prevent an outbreak---while testing and isolation can prevent onward transmission once infected, it cannot prevent infections from happening especially if they are coming from outside of the population.
+Administering booster shots at a faster rate (compare 30 vs 60 booster shots per day in \fref{omicron}A) can help reduce the size of the epidemic peak and cause the epidemic to decay faster, their overall impact is limited due to delays in developing immunity and imperfect protection against the Omicron variant.
+Comparing changes in the mean susceptibility further reveal that administering booster shots at a faster rate can also lead to earlier immunity waning and higher mean susceptibility by the end of the semester, which may increase chances of an outbreak in the future \fref{omicron}.
 
@@
 \textbf{Scenario projects of the spread of omicron variant on university campus}
-(A) The predicted daily number of new infections across a wide ranges assumptions about testing frequencies, basic reproduction number $\mathcal R_0$, and the number of booster shots given per day.
+(A) The predicted daily number of new infections across a wide ranges assumptions about testing frequencies, effective reproduction number $\mathcal R_e$, and the number of booster shots given per day.
 (B) The predicted effective susceptibility over time. 
@@
 
-\pagebreak
+So far we have assumed that all individuals who test positive are isolated for 10 days;
+in practice, however, isolation of infected individuals can be limited by the isolation bed capacity.
+As the Omicron variant is expected to spread rapidly on university campuses (\fref{omicron}), universities participating in testing plans can expect to face difficulties in isolating infected individuals.
+Here, we explore how limiting the number of isolation beds on university campus can affect epidemic dynamics by assuming that individuals who test positive will not be isolated and will continue to transmit when isolation beds are full.
+In doing so, we also consider the effects of shorter isolation periods (5, 7, and 10 days).
 
@@
 
+When there are no limits to the number of isolation beds on campus, shortening isolation periods results in a slightly higher number of total infections (\fref{isolation}A): .
+The epidemic trajectory remains similar across different values of isolation periods (\fref{isolation}B), but the number of students in isolation becomes considerably lower with shorter isolation periods (\fref{isolation}C).
+For example, the maximum number of students in isolation decreases from ??? to ???.
+
+When the number of isolation beds is moderately limited ($=200$), shortening isolation periods can decrease the total numbers of infections (\fref{isolation}A) and considerably reduce the size of the epidemic peak (\fref{isolation}B).
+Even when the number of isolation beds is severely limited ($=100$), shortening isolation periods can still reduce the size of the epidemic peak (\fref{isolation}B) because we are able to isolate a greater proportion of infected individuals and prevent onward transmission.
+Longer isolation can prevent a greater amount of onward transmission \emph{per individual} but permits a lower number of individuals in isolation, making it less effectiveness when the number of isolation beds are limited.
+
+\section*{Discussion}
+
+Here, we present the analysis of SARS-CoV-2 outbreaks on Princeton University campus between fall 2020 and early 2022.
+Our analysis demonstrates strong spatiotemporal correlations between the patterns of spread of SARS-CoV-2 on university campus and those from surrounding communities.
+These correlations decreased with distance from Mercer County in fall 2021, likely reflecting contact and commuting patterns as university re-opened.
+Mathematical modeling further suggests limited transmission between the university population during fall and spring semesters of 2020 and an increased amount of community contacts during the fall semester of 2021 compared previous semesters.
+An increase in the number of cases by the end of November 2021 is consistent with the increase in the levels of community cases and waning immunity.
+Finally, we predict the spread of Omicron variant to likely cause a large outbreak in the spring semester of 2021 even under more stringent testing protocols.
+A rapid administration of booster shots and shortening isolation periods can help reduce the size of an outbreak, but uncertainty remains for future semesters.
+
+Previous outbreak reports from other universities primarily focused on transmission on university campuses \citep{wilson2020multiple,currie2021interventions} and provided limited insights into the importance of community transmission in shaping epidemic dynamics.
+For example, extensive modeling efforts from Cornell University also demonstrated increase in the amount of transmission from outside the university campus and found that community transmissions are the biggest risk for faculty and staff members \citep{frazier2022modeling};
+but they did not demonstrate explicit correlations between university and community cases and further assumed that the amount of community transmission does not change over a semester.
+To our knowledge, our analysis is the first to demonstrate a strong spatiotemporal correlation in the spread of SARS-CoV-2 between university campuses and surrounding communities.
+Our analysis further indicate that comparing the ratios between the cases on university campuses and neighboring communities can provide a useful measure for how well a university campus might be doing in controlling the epidemic---we note that this ratio need to be interpreted with caution as it is sensitive to changes in testing patterns.
+
+The recent recommendations by the Centers for Disease Control and Prevention to reduce isolation periods from 10 to 5 days \citep{cdcisolation} have generated much discussion regarding its individual- and population-level implications \citep{soljak2022reducing}.
+Given uncertainty in the duration of infectiousness of the Omicron variant, shortening the duration of isolation can increase the chances of transmission after infected individuals are released from their isolation---as we also show here, this can lead to an increase in the final size of an outbreak.
+When the number of isolation beds are limited, however, shorter isolation periods can help mitigate spread by allowing a greater number of individuals to be isolated.
+We note that the net benefits of shortening isolation periods are sensitive to many factors, such as the duration of infectiousness, isolation bed capacity, and time from infection to isolation, which will vary across different universities.
+Therefore, caution is needed in implementing these changes in other places.
+Incorporating lateral flow tests at the end of isolation can also help prevent excess transmission \citep{quilty2022test}.
+
+There are several limitations to our analysis.
+While our analysis demonstrates strong spatiotemporal correlation in the spread of SARS-CoV-2, we are not able to infer the direction of causality---that is, our analysis does not rule out the possibility that transmission on campus drove infections in nearby communities (as opposed to community transmission driving on-campus infections).
+Nonetheless, intervention measures on campus (e.g., frequent asymptomatic testing, contact tracing, and virtual classes during fall and spring semesters of 2020) likely limited onward transmission on campus, and therefore, transmission from community to campus is likely to have had greater effects.
+Decreasing patterns in epidemic correlations with distance further highlight the role of spatial spread in driving dynamics of SARS-CoV-2---such patterns are consistent with spatial spread of many other respiratory pathogens \citep{grenfell2001travelling, viboud2006synchrony, baker2019epidemic}.
+
+Our mathematical model relies on simplifying assumptions.
+For example, we assume that the entire university populations mix homogeneously and have identical campus and community contact rates (captured by $\mathcal R_e$ and $\theta$, respectively).
+In reality, increases in cases were often associated with specific transmission clusters, suggesting heterogeneity in transmission patterns.
+Contact levels also likely differ between different groups;
+for example, faculty and staff members are more likely to interact with community members than undergraduate students and are therefore at a higher risk for community infections \citep{frazier2022modeling}.
+We also do not account for changes in behavior. 
+Instead, we assume a constant $\theta$ and $\mathcal R_e$ throughout a semester (note this $\mathcal R_e$ implicitly accounts for some interventions, but not for testing or changes in susceptibility driven by vaccination or infection).
+We also do not explore parameter uncertainty, which can lead to underestimation of overall uncertainty \citep{elderd2006uncertainty}. 
+We also note that intervention measures that were introduced to Princeton University may not necessarily be applicable in other universities.
+
+Despite the simplicity of the analysis, our study provides important lessons for controlling SARS-CoV-2 outbreaks on university campuses in general.
+First, the safe reopening of a university campus must consider the spread of SARS-CoV-2 within the surrounding community as they can both drive transmission in each other.
+Second, combining multiple interventions measures, such as social distancing, mask wearing, frequent testing, and vaccination, can help provide safe means of re-opening university campuses---
+the extent to which these interventions are implemented will necessarily depend on resource availability
+Finally, intervention measures placed on campuses must continue to adapt and change to reflect changes in epidemiological conditions.
+
+\bibliography{university-covid}
+
+\pagebreak
+
+\section*{Supplementary Figures}
+
+\begin{figure}[!th]
+\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_correlation2.pdf}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!th]
+\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_map.pdf}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!th]
+\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_newyork.pdf}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!th]
+\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_phil.pdf}
+\end{figure}
+
 \end{document}

### 2022-02-21 doc/university-covid.tex
done?

@@
 
+\usepackage{setspace}
+\linespread{2}
+
 \usepackage[pdfencoding=auto, psdextra]{hyperref}
@@
 \newcommand{\Ro}{\Rx{0}}
+\newcommand{\Rc}{\Rx{\textrm{\tiny{contact}}}}
 \newcommand{\RR}{\ensuremath{{\mathcal R}}}
@@
 	\textbf\newline{
-		The role of community transmission in driving the spread of SARS-CoV-2 on a university campus: lessons from Princeton University
+		Relative role of community transmission and campus contagion in driving the spread of SARS-CoV-2: lessons from Princeton University
 	}
@@
 \\ 
+\end{flushleft}
 
-\end{flushleft}
+\section*{Abstract}
+
+Mathematical models have played a crucial role in guiding pandemic responses, especially for reopening university campuses.
+Here, we present descriptive and modeling analyses of SARS-CoV-2 outbreaks on the Princeton University campus---this model was used throughout the pandemic to inform policy decisions in Princeton University.
+We demonstrate strong spatiotemporal correlations in epidemic patterns between Princeton University and surrounding communities.
+These findings are corroborated by our model predictions, which indicate that the amount of on-campus transmission was likely limited;
+mathematical modeling further illustrates that an increase in the number of cases in November 2021 is consistent with waning immunity.
+Predicting future outbreaks and the effectiveness of intervention measures on campus is expected to be sensitive to underlying epidemiological conditions, including community prevalence and immunity landscape.
 
@@
 
-Predicting and controlling the spread of SARS-CoV-2 remains a critical public health and scientific question.
-Rapid, asymptomatic transmission of SARS-CoV-2 has hindered intervention efforts such as contact tracing.
-Stringent measures to limit social contacts have been effective in preventing transmission in the beginning of the pandemic but are unsustainable.
-The recent development of vaccines have provided a safe means of reopening the society, but uncertainty remains in their long-term effectiveness in protecting infection and transmission.
-In addition, as new variants continues to emerge and
+Predicting and controlling the spread of SARS-CoV-2 has remained a critical public health and scientific question throughout the ongoing SARS-CoV-2 pandemic \citep{baker2021limits}.
+Rapid, asymptomatic transmission of SARS-CoV-2 has hindered intervention efforts, such as contact tracing \citep{hellewell2020feasibility}.
+Social distancing measures have played major roles in preventing transmission but can be difficult to maintain for a prolonged period \citep{galanti2021social}.
+The development of vaccines has provided a safe means of reopening society, but uncertainty remains in their long-term effectiveness in protecting infection and transmission, especially in the face of new viral variants.
 
-the emergence of new variants, which can evade immunity and transmit better, continues to pose additional challenges.
+Mathematical models have played a central role in guiding these pandemic responses and devising control strategies \citep{metcalf2020mathematical}.
+Models can help monitor key parameters that govern epidemic dynamics \citep{kraemer2021monitoring} and retrospectively estimate the impact of intervention measures in reducing transmission \citep{flaxman2020estimating}.
+These estimates can further inform projections of future scenarios and allow us to explore the endemicity of SARS-CoV-2 \citep{kissler2020projecting,saad2020immune,lavine2021immunological,saad2021epidemiological}.
+% While models are useful for exploring such counterfactuals, they must be paired with real-life implementations to properly assess the effectiveness of intervention measures.
 
-Since the beginning of the pandemic, mathematical models have played a key role in evaluating intervention measures and guiding pandemic response.
+Mathematical models have been particularly useful in planning campus reopenings.
+Researchers from various institutions---including Cornell \citep{frazier2022modeling}, Emory \citep{lopman2020model}, Georgia Institute of Technology \citep{gibson2021surveillance}, and UC Berkeley \citep{brook2021optimizing}---modeled the feasibility of controlling the epidemic on their campuses and considered mass asymptomatic testing as their main intervention.
+These modeling efforts helped identify key parameters for control, such as the testing turnaround time, and provided support for implementing similar measures at other institutions.
+Coupling modeling efforts with real-life implementations in university campuses further provided unique opportunities to directly test model-based predictions of intervention effects in preventing the transmission of SARS-CoV-2 \citep{frazier2022modeling}---
+each university campus offers a relatively well-controlled epidemic setting with a behaviorally homogeneous population (especially among undergraduate students).
+Mass asymptomatic testing also provides near-perfect ascertainment for epidemic sizes, allowing for accurate understanding of epidemic patterns.
 
-testing has played a key role in monitoring and controlling the spread.
-In addition, a frequent testing of a large fraction of the population, followed by rapid isolation and contact tracing, has proven effective in preventing chains of transmission and reducing the spread in large congregate settings.
-
-University campuses provide unique test case in a relatively well-controlled setting for evaluating the feasibility of safe reopening.
-High density environments and large fraction of asymptomatic infections (due to young age of university students) can readily permit rapid transmission.
-
-- Usual pandemic start
-
-- modeling, understanding control
-- testing frequency is important
-- universities provide a particularly test case; asymptomatic infection and testing is very high
-
-- little review of what other universities have done
-
-- epideimc in relatively well controlled documented setting
-- really well until omicron
-- perhaps because of immunity waning
+University campuses also present unique challenges to controlling an outbreak.
+High-density environments, especially within dorms, and a large fraction of asymptomatic infections (due to the young age of university students) can readily permit rapid transmission.
+These kinds of contacts are inherently difficult to trace, making contact tracing less effective.
+The impact of intervention measures is also expected to vary across different university campuses, reflecting heterogeneity in campus settings such as compliance, resources, community prevalence, as well as effects of other interventions present on campuses.
+For example, Duke and Harvard Universities experienced moderate outbreaks at the beginning of the fall semester in 2021 when in-person classes were allowed, despite high vaccination rates and weekly testing protocols \citep{dukeoutbreak,harvardoutbreak}, whereas the number of cases remained low in Princeton University during the same time period despite similar levels of testing and vaccination.
+Here, we choose to focus on the spread of SARS-CoV-2 on Princeton University campus alone to eliminate heterogeneities inherent to such comparisons.
 
 In this study, we analyze the spread of SARS-CoV-2 on Princeton University campus (\fref{princeton}).
-Princeton University is located in Mercer County, New Jersey, USA and consists of roughly 5000 undergraduate students, 3000 graduate students, and 7000 faculty and staff members.
-For simplicity, we divide the epidemic into three time periods representing four semesters: Fall 2020 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020 (January 16, 2021--May 14, 2021; \fref{princeton}B), and Fall and Winter 2021 (August 14, 2021--January 14, 2021; \fref{princeton}C).
+We first begin with a descriptive analysis of the outbreak and then present modeling analyses of the past outbreaks and future scenarios.
+Princeton University is located in Mercer County, New Jersey, USA and consists of 5267 undergraduate students, 2946 graduate students, and 7000 faculty and staff members.
+For simplicity, we divide the epidemic into three time periods representing three semesters: Fall 2020 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020 (January 16, 2021--May 14, 2021; \fref{princeton}B), and Fall 2021 (August 14, 2021--January 14, 2021; \fref{princeton}C).
 Throughout the study period, all students, faculty and staff members who were physically present for more than 8 hours on campus per week were required to participate in asymptomatic testing with varying frequencies.
@@
 Symptomatic individuals were eligible for rapid PCR tests and had to isolate \swp{self-isolate?} until they received their test results---those who tested positive were required to isolate for at least 10 days after symptom onset and were release 3 days after symptoms resolved.
-Those who tested positive were exempt from asymptomatic testing for 90 days.
-Throughout the study period, contact tracing was also performed for positive cases to quarantine their close contacts, which refer to individuals who have been within 6 feet for at least 15 minutes, cumulatively within a 24-hour period.
-All data used in this analysis are publicly available on Princeton University COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.
+PCR positives were exempt from asymptomatic testing for 90 days.
+Throughout the study period, contact tracing was also performed for positive cases to quarantine their close contacts, defined as individuals who have been within 6 feet for at least 15 minutes, cumulatively within a 24-hour period.
+All data used in this analysis are publicly available on the Princeton University COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.
 
@@
 \caption{
-\textbf{Dynamics of SARS-CoV-2 outbreaks in Princeton University}
+\textbf{Dynamics of SARS-CoV-2 outbreaks in Princeton University.}
 (A--C) Epidemic trajectories across three semesters: Fall 2020 (A), Spring 2020 (B), and Fall 20201 (C).
@@
 Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
+(G) Ratios between weekly numbers of cases reported from Princeton University and Mercer County.
+Solid lines represent the observed ratios.
+Dotted lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
 \label{fig:princeton}
@@
 During fall 2020, roughly 1000 grad students and 2000 faculty and staff members were present on campus and participated in asymptomatic testing. 
-Undergraduate students were not allowed to return back to campus as all classes were held virtually, and so their numbers were minimal on campus ($<300$).  
-Both undergraduate students and graduate students were required to get tested twice a week, whereas faculty and staff members were required to get tested once a week.
+Undergraduate students were not allowed to return to campus as all classes were held virtually, and so their numbers were minimal on campus ($<300$).  
+Both undergraduate and graduate students were required to get tested twice a week, whereas faculty and staff members were required to get tested once a week.
 The number of cases remained relatively low throughout the semester with a peak occurring in early December, coinciding with the epidemic trajectory in Mercer County (\fref{princeton}A).  
 A sudden decrease in the number of cases around Thanksgiving partly reflects the reduced number of tests (3852 and 2972 asymptomatic tests performed on the week ending November 20th and 27th, respectively).
-The highest number of cases was reported among faculty and staff members (169), followed by graduate students (41), and undergraduate students (4).
+The highest number of cases was reported among faculty and staff members ($=169$), followed by graduate students ($=41$), and undergraduate students ($=4$).
 
 In the beginning of spring 2020, the number of cases suddenly increased before classes began (\fref{princeton}B), reflecting $\approx 3000$ undergraduate students returning to campus.
-Returning students were required to be tested and quarantine \swp{for how long} if they were coming back from CDC-designated states \swp{not sure if this is the right wording...
-but I remember we had a rule like this...
-we can ask Irini and Robin for this}.
+Returning students were required to be tested and quarantine \swp{for how long} if they were coming back from CDC-designated states and countries \swp{ask Irini and Robin for details}.
 All classes remained virtual, and the testing protocol did not change (twice a week for undergraduate and graduate students, and once a week for faculty and staff members).
 The number of cases persisted at similar levels as the fall semester and eventually decreased as classes ended and students went back home---the decrease in the number of cases in Princeton University also coincided with the decrease in the number of cases in Mercer County.
-The highest number of cases was reported among faculty and staff members (111), followed by undergraduate students (101), and graduate students (29).
+The highest number of cases was reported among faculty and staff members ($=111$), followed by undergraduate students ($=101$), and graduate students ($=29$).
 
@@
 As of January 11th, 99\% of undergraduate students, 98\% of graduate students, and 96\% of faculty and staff members were vaccinated.
-Vaccinated individuals were required to be tested once a week, while unvaccinated individuals were required to be tested twice a week.
+Vaccinees were required to be tested once a week, while unvaccinated individuals were required to be tested twice a week.
 In-person classes and social events fully resumed on campus while all individuals were required to wear masks indoors with a few exceptions (e.g., when eating or drinking, or when teaching a small class).
@@
 We find similarly strong correlations with New York City: fall 2020 ($\rho = 0.74$ (95\% CI: 0.44--0.90; $p < 0.001$), Supplementary Figure S3A); spring 2020 ($\rho = 0.84$ (95\% CI: 0.61--0.94; $p < 0.001$), Supplementary Figure S3B); and fall 2021 ($\rho = 0.86$ (95\% CI: 0.70--0.94; $p < 0.001$), Supplementary Figure S3C).
-We also find similarly strong correlations with Philadelphia except for spring 2020: fall 2020 ($\rho = 0.83$ (95\% CI: 0.59--0.93; $p < 0.001$), Supplementary Figure S4A); spring 2020 ($\rho = 0.29$ (95\% CI: $-0.22$--0.68; $p = 0.25$), Supplementary Figure S4B); and fall 2021 ($\rho = 0.86$ (95\% CI: 0.68--0.94; $p < 0.001$), Supplementary Figure S4C).
+The same picture emerges for Philadelphia except for spring 2020: fall 2020 ($\rho = 0.83$ (95\% CI: 0.59--0.93; $p < 0.001$), Supplementary Figure S4A); spring 2020 ($\rho = 0.29$ (95\% CI: $-0.22$--0.68; $p = 0.25$), Supplementary Figure S4B); and fall 2021 ($\rho = 0.86$ (95\% CI: 0.68--0.94; $p < 0.001$), Supplementary Figure S4C).
 Including counties from New York and Pennsylvania states into the spatial correlation analysis yields additional insights (Supplementary Figure S5):
@@
 
-Finally, we calculate the ratio between the weekly numbers of cases from Princeton and those from Mercer county---
-we expect this ratio to remain constant over time if the majority of the infections in Princeton University is caused by community transmission provided that testing patterns remain constant in both places.
+Finally, mass testing allows us to infer the ratio between the weekly numbers of cases from Princeton and those from Mercer county---
+we expect this ratio to remain constant over time if the majority of infections on campus is caused by community transmission provided that testing patterns remain constant in both places.
 We do not find significant changes in the ratio between the weekly numbers of cases from Princeton and those from Mercer county (\fref{princeton}G).
@@
 However, this does not necessarily imply that the contact or transmission levels stayed constant over time: given extremely high vaccination coverage in Princeton University during fall 2021, the university population would been subject to a much higher forces of infection (e.g., a greater amount of mixing with the community population) to obtain similar levels of cases.
-We also find that variations in the ratio around the mean can be explained by holiday effects.
+We also find that some variations in the ratio around the mean can be explained by holiday effects.
 For example, sudden decreases in the number of cases around Thanksgiving of 2020, fall recess of 2021, and end of classes of 2021 correspond to a reduction in the ratio between university and community cases.
@@
 
-We use a discrete-time, individual-based model to simulate the spread of SARS-CoV-2 in Princeton University.
+We use a discrete-time, individual-based model to simulate the spread of SARS-CoV-2 on the Princeton University campus.
 This model was initially developed and used throughout the pandemic to inform policy decisions in Princeton University, including the frequency of asymptomatic tests and the number of isolation beds required.
 We continuously updated the model to reflect changes in school settings (e.g., students returning back to campus after a virtual semester) as well as intervention measures (e.g., vaccination in fall 2021 and booster shots with the emergence of the Omicron variant).
-Here, we present a generic version that encompasses all of the necessary details to characterize the spread of SARS-CoV-2 in Princeton University.
-The model consists of four main components that are simulated on a daily time scale: (1) transmission and infection dynamics, (2) sampling and testing protocols, (3) isolation protocols, and (4) vaccination dynamics, including waning immunity and booster shots.
+Here, we present a generic and parsimonious version that encompasses sufficient details to characterize the overall spread of SARS-CoV-2 in Princeton University without an over-proliferation of parameters.
+The model consists of four main components that are simulated on a daily time scale: (1) infection and transmission dynamics, (2) sampling and testing protocols, (3) isolation protocols, and (4) vaccination dynamics, including waning immunity and booster shots.
 Previous versions of the model included contact tracing, but we exclude it in this model for simplicity.
-Here, we describe each component briefly. 
-Mathematical details are provided in Supplementary Materials.
 
 Infection processes are modeled based on standard compartmental structures.
-Once infected, susceptible individuals enter the exposed stage, during which they cannot transmit or test positive. 
-After $D_e$ days on average, exposed individuals then enter the presymptomatic stage, during which they can test positive and transmit infections at rate $\beta_p$ for $D_p$ days on average.
-Presymptomatic individuals can then either remain asymptomatic with probability $p_a$ or develop symptoms with the remaining probability of $1-p_a$; both asymptomatic and symptomatic individuals are assumed to have the same duration of infectiousness ($D_s$) and transmission rates ($\beta_s$) for simplicity.
+Once infected, susceptible individuals remain in the exposed stage for $D_e = 2$ days on average, during which they cannot transmit or test positive. 
+Exposed individuals then enter the presymptomatic stage, during which they can test positive and transmit infections for $D_p = 3$ days on average.
+Presymptomatic individuals can then either remain asymptomatic with probability $p_a = 0.4$ or develop symptoms with the remaining probability of $1-p_a = 0.6$; both asymptomatic and symptomatic individuals are assumed to have the same duration of infectiousness ($D_s=3$) and equal transmission rates.
 Recovered individuals are assumed to be immune to reinfections throughout a semester.
-Transmissions are modeled using a negative binomial distribution with an overdispersion parameter of $k=0.1$ to account for the possibility of super-spreading events.
-External (community) transmission is modeled using a Poisson distribution with a time-varying mean calculated by scaling the daily number of cases by $\theta$ and shifting it by 1 week to account for reporting delays.
+Presymptomatic, symptomatic, and asymptomatic infection stages are further divided into two subcompartments to allow for more realistic and narrower distributions than the exponential distribution \citep{brett2020transmission}.
+Transitions between each (sub)compartments are modeled using a Bernoulli process with probabilities that match the assumed means \citep{he2010plug}.
+More specifically, transition probabilities are equal to $1 - \exp(-\delta_x)$, where $\delta_x = -log(1-n/D_x)$ represent the transition rate from stage $X$ and $n$ represents the number of subcompartments.
+
+Transmission processes are modeled by first setting the contact reproduction number $\Rc$, which we define as the average number of infectious contacts an infected individual would make throughout the course of their infection;
+here, infectious contacts refer to contacts that would result in infection when the contactee is susceptible to infection.
+The contact reproduction number implicitly accounts for all intervention measures that we do not model explicitly, such as social distancing and contact tracing---in other words, this contact reproduction number does not account for asymptomatic testing or vaccination, which are modeled separately.
+We further decompose $\Rc$ into pre-symptomatic $\mathcal R_p = \beta_p D_p$ and (a)symptomatic $\mathcal R_s = \beta_s D_s$ reproduction numbers, where $\beta_p$ and $\beta_s$ represent the corresponding infectious contact rates during pre-symptomatic and (a)symptomatic stages, respectively.
+Pre-symptomatic and (a)symptomatic reproduction numbers are calculated based on the assumed value of the proportion of presymptomatic transmission $p_p = 0.5$: $\mathcal R_p/\mathcal R_s = p_p/(1 - p_p)$.
+On each day, all infected individuals who have not yet been isolated then make infectious contacts at random to anyone on campus; the number of infectious contacts are drawn from a negative binomial distribution with a mean of either $\beta_p$ or $\beta_s$ and an overdispersion parameter of $k=0.1$ to account for the possibility of super-spreading events.
+Infectious contacts from the community can be also made at random to anyone on campus; these contacts are modeled using a Poisson distribution with a time-varying mean, which is calculated by scaling the daily number of cases by $\theta$ and shifting it by 1 week to account for reporting delays.
+Infectious contacts, whether made by individuals on campus or from outside, result in infection only when the contacted individuals are susceptible; when the contacted individuals are vaccinated, and therefore partially susceptible to infection, contactees have a reduced probability of infection corresponding to their susceptibility (discussed later).
 
@@
 Symptomatic individuals can choose to take rapid PCR tests (with results returning on the same day) with a given probability on each day until their symptoms resolve---this probability is set to 1 for simulations presented in the main text.
-We further assume that symptomatic individuals would isolate immediately when they submit their samples until they receive negative results.
-All individuals who test positive are required to isolate (following the same isolation rule as before) and are exempt from asymptomatic testing for 90 days.
+We further assume that symptomatic individuals are isolated immediately when they submit their samples until they receive negative results.
+All individuals who test positive are required to isolate (following the same isolation rule as described earlier) and are exempt from asymptomatic testing for 90 days.
 Isolated individuals are assumed to no longer transmit infections.
@@
 As most students as well as faculty and staff members had received two doses of vaccination in the beginning of fall 2021, we do not distinguish the first and second doses.
-Instead, we assume that vaccines reduce susceptibility as well as transmissibility of all vaccinated individuals by an equal amount for all vaccinated individuals.
-Vaccine efficacies are allowed to exponentially wane from 90\% to 50\% in 20 weeks (and continues to wane at the same rate) for each vaccinated individuals.
-While 90\% efficacy against susceptibility is consistent with recent estimates, 90\% efficacy against transmission is likely optimistic.
-For example, \cite{prunas2022vaccination} recently estimated that vaccination with BNT162b2 reduces susceptibility by 89.4\% (95\% CI: 88.7\%--90.0\%) and infectiousness by 23.0\% (95\% CI: $-11.3\%$--46.7\%) against the Delta strain.
-In Supplementary Materials, we present the same set of simulations, which assume that vaccines do not reduce transmissibility.
+Instead, we assume that all vaccinated individuals have 90\% reduced susceptibility and 20\% reduced transmissibility at the beginning of the semester---these assumptions are consistent with recent estimates by \citep{prunas2022vaccination} that vaccination with BNT162b2 reduces susceptibility by 89.4\% (95\% CI: 88.7\%--90.0\%) and infectiousness by 23.0\% (95\% CI: $-11.3\%$--46.7\%) against the Delta strain.
+Based on \citep{tartof2021effectiveness}, vaccine efficacy against susceptibility is allowed to exponentially wane from 90\% to 50\% in 20 weeks (and continues to wane at the same rate) for each vaccinated individual;
+vaccine efficacy against transmissibility is also allowed to wane at the same rate (i.e., from 20\% to 11\% in 20 weeks).
 
-Here, we use this model to retrospectively analyze past outbreaks and further predict what future epidemic trajectories might look like as the new Omicron variant spreads on campus (see Supplementary Materials for simulation details).
-First, we try to match our model to epidemic patterns seen on campus by varying the effective reproduction number $\mathcal R_e$ and the amount of community transmission $\theta$ and holding all other parameters constant.
-Here, the effective reproduction number implicitly accounts for all intervention measures that we do not model explicitly, such as social distancing and contact tracing---in other words, this effective reproduction number does not account for asymptomatic testing or vaccination, which are modeled separately.
+In this study, we use this model to retrospectively analyze past outbreaks and further predict future epidemic trajectories for the Omicron variant.
+First, we try to match our model to epidemic patterns seen on campus by varying the contact reproduction number $\Rc$ and the amount of community transmission $\theta$ and holding all other parameters constant.
 For each parameter combination, we simulate 100 epidemic trajectories and calculate the sum of squared differences between the weekly numbers of the observed and predicted positive cases.
-The population size and testing frequencies (with twice weekly testing modeled as testing every 3 days) are set to previously discussed values to reflect realistic campus settings.
+The population size and testing frequencies (with twice weekly testing modeled as testing every 3 days) are set to reflect realistic campus settings.
+Although we account for heterogeneity in the number of individuals in each population group on campus (i.e., undergraduate students, graduate students, and faculty and staff members) and their respective testing patterns (e.g., twice a week for undergraduate and graduate students and once a week for faculty and staff members during fall and spring, 2020), we assume that all other parameters are equal across different groups for simplicity.
+We further assume that the population mixes homogeneously.
+In practice, epidemiological parameters, such as contact rates and the proportion of asymptomatic infections, likely differ across groups, and mixing patterns likely exhibit assortativity (e.g., undergraduate students are more likely to infect undergraduate students).
+Therefore, our model parameters describe average dynamics across different groups and must be interpreted with care.
 
@@
 \caption{
-\textbf{Dynamics of SARS-CoV-2 outbreaks in Princeton University}
-(A, D, G) Time series comparisons of model predictions with observed data across ranges of effective reproduction number $\mathcal R_e$ and scaling parameter for community transmission $\theta$.
+\textbf{Retrospective analysis of past SARS-CoV-2 outbreaks on Princeton University.}
+(A, D, G) Time series comparisons of model predictions with observed data across ranges of contact reproduction number $\Rc$ and scaling parameter for community transmission $\theta$.
 For each parameter combination, we simulate the model 100 times and calculate the sum of squared differences (SSQ) between the reported number of positive cases and the model-predicted number of positive cases.
@@
 
-We find that a low value of $\mathcal R_e=0.5$ and a small amount of community transmission $\theta=0.015$ is most consistent with the observed epidemic dynamics in fall 2020 (\fref{matching}A).
-With these parameters, the model is able to capture the rise and fall in the number of cases with the exception of sudden decrease in the number of cases around Thanksgiving, which we do not model explicitly (\fref{matching}B).
+For fall 2020, we simulate the model assuming 3000 individuals (1000 graduate students and 2000 faculty and staff members) on campus with 1000 of them participating in asymptomatic testing twice a week.
+We find that a low level of contacts $\Rc=0.5$ and a small amount of community transmission $\theta=0.015$ is most consistent with the observed epidemic dynamics in fall 2020 (\fref{matching}A).
+With these parameters, the model is able to capture the rise and fall in the number of cases with the exception of a sudden decrease in the number of cases around Thanksgiving, which we do not model explicitly (\fref{matching}B).
 The median predictions are positively correlated with the observed dynamics ($\rho = 0.83$; 95\% CI: 0.61--0.93; \fref{matching}C).
-Although a wide range of assumptions about the levels of community transmission $\theta$ are consistent with the observed dynamics, our simulations preclude high $\mathcal R_e > 2$ (SUPP).
-Distancing measures on campus and contact tracing efforts likely contributed to lowering $\mathcal R_e$.
+Although a wide range of assumptions about the levels of community transmission $\theta$ are consistent with the observed dynamics, our simulations preclude high levels of contact, $\Rc > 2$ (Supplementary Figure S6).
+Distancing measures on campus and contact tracing efforts likely contributed to lowering contact levels $\Rc$.
+
+For spring 2020, we simulate the model assuming 8000 individuals (3000 undergraduate students, 2000 graduate students, and 3000 faculty and staff members) on campus with 5000 of them participating in asymptomatic testing twice a week.
+We further assume that 4000 individuals (3000 undergraduate students and 1000 graduate students) returned to campus over 14 days (January 16, 2021--January 29, 2021);
+all returning individuals are assumed to be quarantined for 14 days and tested upon returning. 
+Finally, to match the initial influx of cases, we assume that 1\% of both returning and on-campus populations are infected at the beginning of simulation (January 16, 2021).
 
 A similar set of parameters can capture the observed dynamics in spring 2020.
-The best matching parameter predicts a slightly higher levels of community transmission $\theta=0.02$ (\fref{matching}D), but a wide range of parameters are consistent with the observed dynamics as before (SUPP).
-Simulations also preclude high $\mathcal R_e > 2$ again, suggesting that the contact levels between students likely remained low even though they had returned to campus---the absence of in-person teaching is likely to have contributed to lowering $\mathcal R_e$.
-Once again, we find a positive (but slightly weaker) correlation between the predicted and the observed numbers of cases ($\rho = 0.62$; 95\% CI: 0.20--0.85; \fref{matching}F).
+The best matching parameter predicts a slightly higher levels of community transmission $\theta=0.02$ (\fref{matching}D), but a wide range of parameters are consistent with the observed dynamics as before (Supplementary Figure S7).
+Simulations also preclude high $\Rc > 2$ again, suggesting that transmission between students were likely limited even though they had returned to campus---the absence of in-person teaching is likely to have contributed to lowering $\Rc$.
+We also find that initial infections are required to match relatively high levels of cases in the beginning of semester (\fref{matching}E). 
+Once again, the predicted and the observed numbers of cases are positively correlated ($\rho = 0.62$; 95\% CI: 0.20--0.85; \fref{matching}F).
 
-For fall 2021, we limited our model comparison to November 26th before the Omicron variant was introduced on campus.
-We assume that $>95\%$ of the school population is vaccinated at the beginning of the semester and do not model additional doses throughout the semester for simplicity.
-We also allow vaccine-derived immunity to wane over time to ask whether the increase in the number of cases around November is consistent with the dynamics predicted by immunity waning.
+For fall 2021, we assume 13000 individuals are present on campus (5000 undergraduate students, 2000 graduate students, and 6000 staff and faculty members) with 98\% of them vaccinated---here, vaccine-derived immunity is allowed to wane over time to ask whether the increase in the number of cases around November is consistent with the dynamics predicted by immunity waning.
+Vaccinated individuals are tested every week, whereas unvaccinated individuals are tested every 3 days. 
+We further assume 5000 undergraduate students returned to campus over 16 days (August 14, 2021--August 29, 2021);
+all returning students are assumed to be tested upon return but are not quarantined.
+Finally, we assume that 0.5\% of both returning and on-campus populations are infected at the beginning of simulation (August 14, 2021).
+We limit our model comparison to November 26th before the Omicron variant was introduced on campus.
 
-Even though the numbers of cases during fall 2021 (before a large outbreak) were similar to those during previous semesters, we find that considerably higher levels of community transmission $\theta$ (~10 fold higher) is required to explain the observed dynamics due to a decreased susceptibility from vaccination (\fref{matching}G).
-We note that the parameter $\theta$ necessarily depends on our assumed vaccine efficacy, and $\theta$ would decrease if we assume a lower vaccine efficacy.
-Nonetheless, the amount of community transmission would still need to be higher than previous semesters as long as vaccine provides some protection against infection.
+Even though the numbers of cases during fall 2021 (before a large outbreak) were similar to those during previous semesters, we find that considerably higher levels of community contact $\theta$ ($\approx 10$ fold higher) are required to explain the observed dynamics due to a decreased susceptibility from vaccination (\fref{matching}G).
+We note that the parameter $\theta$ necessarily depends on our assumed vaccine efficacy against susceptibility, and $\theta$ would decrease if we assume a lower vaccine efficacy.
+Nonetheless, the amount of community contact would still need to be higher than previous semesters as long as vaccine provides some protection against infection.
 
-We also find that the predicted dynamics are largely insensitive to $\mathcal R_e$ until November 26th (\fref{matching}H)---
-in addition, all simulations shown in \fref{matching}H are positively correlated with the observed numbers of cases (\fref{matching}G). 
-An increasing pattern in the sum of squared errors with $\mathcal R_e$ shown in \fref{matching}G is primarily driven by the discrepancy around fall break (week ending October 26th) when the number of cases decreased suddenly, which we do not model explicitly.
-As nearly all school population was vaccinated, transmission between individuals on campus would have been limited, making $\mathcal R_e$ difficult to estimate.
-These simulations also suggest that an increase in the number of cases in November can be explained by a combination of waning immunity alone without requiring additional changes in transmission dynamics (note we do not allow $\theta$ or $\mathcal R_e$ to vary over time).
+While $\theta = 0.15$ and $\Rc = 0.5$ gives the best matching parameter set with a median logged sum of squared errors of 8.88 (95\% quantile: 6.55--12.6), other parameter sets also give nearly identical fits (\fref{matching}H; Supplementary Figure S8):
+for example, $\theta = 0.1$ and $\Rc = 1$ gives a median logged sum of squared errors of 8.9 (95\% quantile: 5.79--13.6).
+Comparing simulations across a wide range of $\Rc$ (0.5--8) with $\theta = 0.1$ further illustrates that the predicted dynamics are largely insensitive to $\Rc$ until November 26th (\fref{matching}H).
+All simulations shown in \fref{matching}H, except for the $\Rc=8$ scenario, are similarly correlated with the observed numbers of cases (\fref{matching}G). 
+An increasing pattern in the logged sum of squared errors with $\Rc$ shown in \fref{matching}G is likely driven by the discrepancy around fall break (week ending October 26th) when the number of cases decreased suddenly, rather than a lack of fits---we did not explicitly model holiday effects for simplicity.
+Extremely high vaccination rates and frequent testing likely limited transmission on campus, making epidemic dynamics largely insensitive to $\Rc$ even at a reasonably high value of $\Rc = 4$.
+These simulations also suggest that an increase in the number of cases in November can be explained by a combination of waning immunity alone without requiring additional changes in transmission dynamics (note we do not allow $\theta$ or $\Rc$ to vary over time).
+In Supplementary Figure S9, \swp{TODO}.
 
@@
 
-We then project epidemic trajectories for the spread of the Omicron variant among 4000 students for the spring semester of 2021 across a wide range of effective reproduction number ($\mathcal R_e = 2--16$) and testing frequencies (every 2, 4, and 7 days).
-We assume a two-dose efficacy 10\% and a three-dose efficacy of 70\% against the infection and transmission of the Omicron variant. 
-We further assume that immunity from the third dose takes 14 days to develop and can wane at the same rate as previously assumed (in this case, 70\% to 39\% in 20 weeks).
-For simplicity, we assume a constant amount of external infections.
+We project epidemic trajectories for the spread of the Omicron variant among 4000 students for the spring semester of 2021.
+In doing so, we assume the following as a nominal parameter set: $\Rc = 4$; testing every 3 days; 60\% boosted (and 99\% having received two doses) by the beginning of semesters with 30 students additionally boosted per day; 10 infectious community contacts per day; and 20\% infected with the Omicron variant before the beginning of the semester.
+We then ask how model predictions are sensitive to these assumptions about underlying epidemiological and intervention conditions by varying each parameter separately.
+In particular, to test for the effects of community transmission, we vary the baseline daily number of infectious contacts from community (5--20) and further allow the number of infectious contacts to exponentially decrease over time from these baseline values (between $r=-0.1/\textrm{day}$--$0/\textrm{day}$), reflecting a recent decrease in cases in New Jersey.
+% We also consider the impact of shortening isolation periods given the recent changes by the Centers for Disease Control and Prevention to reduce isolation periods from 10 to 5 days \citep{cdcisolation}.
 
-\fref{omicron} shows predicted trajectories of new infections per day (A) and the mean susceptibility over time (B).
-Across a wide range of $\mathcal R_e$, we find that a frequent testing alone is not sufficient to prevent an outbreak---while testing and isolation can prevent onward transmission once infected, it cannot prevent infections from happening especially if they are coming from outside of the population.
-Administering booster shots at a faster rate (compare 30 vs 60 booster shots per day in \fref{omicron}A) can help reduce the size of the epidemic peak and cause the epidemic to decay faster, their overall impact is limited due to delays in developing immunity and imperfect protection against the Omicron variant.
-Comparing changes in the mean susceptibility further reveal that administering booster shots at a faster rate can also lead to earlier immunity waning and higher mean susceptibility by the end of the semester, which may increase chances of an outbreak in the future \fref{omicron}.
+Throughout these simulations, we make additional assumptions about the vaccine efficacy against the Omicron variant.
+Based on \cite{ferguson2021report}, we assume that two and three doses of vaccines reduce susceptibility against Omicron by 10\% and 70\%, respectively.
+We further assume that the transmissibility of Omicron is reduced proportionally following previously assumed 90-to-20 ratio for the Delta variant---in other words, two and three doses of vaccines reduce transmissibility by 2.2\% and 15.6\%, respectively.
+Finally, we assume that immunity from the third dose takes 14 days to develop and wanes at the same rate as previously assumed (in this case, 70\% to 39\% in 20 weeks).
+For simplicity, we do not explore the sensitivity of our projections to these assumptions.
 
-\begin{figure}[!th]
-\includegraphics[width=\textwidth]{../figure_omicron/figure_omicron.pdf}
+\begin{figure}[!ht]
+\includegraphics[width=\textwidth]{../figure_omicron/figure_omicron_sensitivity.pdf}
 \caption{
-\textbf{Scenario projects of the spread of omicron variant on university campus}
-(A) The predicted daily number of new infections across a wide ranges assumptions about testing frequencies, effective reproduction number $\mathcal R_e$, and the number of booster shots given per day.
-(B) The predicted effective susceptibility over time. 
-Effective susceptibility is calculated by taking the mean of individual-level susceptibility of each student.
-Solid lines and shaded areas represent the median and the corresponding 90\% quantiles across 100 simulations.
+\textbf{Sensitivity of the spread of Omicron on campus to underlying epidemiological and intervention conditions.}
+(A, C, E, G, I, K) Predicted numbers of Omicron infections per week on campus. 
+Colored lines represent median predictions.
+Shaded areas represent 90\% quantiles across 100 simulations.
+We vary one parameter at a time while holding all other parameters constant at their nominal values.
+(B, D, F, H, J, L) Predicted numbers of total Omicron infections on campus.
 All other parameters are the same as in Figure 2.
-\label{fig:omicron}
+\label{fig:sensitivity}
 }
@@
 
+Intervention measures, such as reducing contact rates $\Rc$ (\fref{sensitivity}A, B), increasing testing frequencies (\fref{sensitivity}C, D), and vaccinating students (\fref{sensitivity}E, F), can help reduce the size of the epidemic peak and the final size of an epidemic but have limited impact.
+While social distancing and mass asymptomatic testing can be effective in preventing onward transmission, they cannot prevent community transmission.
+Vaccines can help reduce community transmission by reducing susceptibility, but current vaccines are likely imperfect against the Omicron variant \citep{ferguson2021report}.
+Nonetheless, we note that vaccinating students throughout the semester is just as important as vaccinating them before the beginning of the semester---for example, providing 60 booster shots per day with 40\% coverage at the beginning of the semester yield similar outcomes as providing 30 booster shots per day with 60\% coverage at the beginning of the semester despite 20\% differences in the initial vaccine coverage (\fref{sensitivity}F).
+
+Instead, epidemic dynamics are more sensitive to underlying epidemiological factors that cannot be controlled by intervention measures on campus, such as levels of community transmission (\fref{sensitivity}G, H) and previous infection histories (\fref{sensitivity}I, J).
+In particular, the epidemic on campus dies out much quicker when community cases are also decreasing---when community infectious contacts decrease at rate $r=-0.1/\textrm{day}$ starting from 10 infectious contacts at the beginning of semester, the median predicted number of total infections is 85 (90\% quantile: 54--143).
+Such rapid decrease in the number of infections was not achievable with intervention measures that we considered.  
+The proportion of students who have already had Omicron infections before the beginning of semester also play an important role in determining the size of an outbreak (\fref{sensitivity}I, J)---we note that this result depends on our assumption that natural infections provide stronger immunity than vaccinal immunity.
+
 So far we have assumed that all individuals who test positive are isolated for 10 days;
 in practice, however, isolation of infected individuals can be limited by the isolation bed capacity.
-As the Omicron variant is expected to spread rapidly on university campuses (\fref{omicron}), universities participating in testing plans can expect to face difficulties in isolating infected individuals.
 Here, we explore how limiting the number of isolation beds on university campus can affect epidemic dynamics by assuming that individuals who test positive will not be isolated and will continue to transmit when isolation beds are full.
@@
 
+When there are no limits to the number of isolation beds on campus, shortening isolation periods results in a slightly higher number of total infections due to residual transmission after the isolation periods are over (\fref{sensitivity}L); but the impact on overall epidemic dynamics under our nominal parameter assumptions (\fref{sensitivity}K).
+When the number of isolation beds is moderately limited ($=100$), shortening isolation periods can decrease the total numbers of infections (compare 10 vs 7 day in \fref{sensitivity}L), but shortening isolation periods too much can increase the total numbers of infections again (compare 7 vs 5 day in \fref{sensitivity}L).
+When the number of isolation beds is severely limited ($=50$), the benefits of shortening isolation periods can further outweigh the risk of transmitting after isolation by allowing a greater proportion of infected individuals to be isolated.
+
+Finally, as the number of COVID-19 cases is decreasing in the US, many institutions, including Princeton University, participating asymptomatic testing programs may be considering reducing testing frequencies.
+Here, we test whether switching testing frequencies from twice a week to once a week in the middle of a semester will result in larger outbreaks across a wide range of contact $\Rc$ and immunity levels (i.e., proportion of students who have already had Omicron infections before the beginning of semester).
+To present a challenging scenario, we assume that the number of infectious community contacts decrease exponentially at rate $r=-0.1/\textrm{day}$ until March 1, 2022, when the testing protocol (hypothetically) changes, and allow them to increase again exponentially at rate $r=0.1/\textrm{day}$.
+
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{../figure_omicron/figure_omicron_limit.pdf}
+\includegraphics[width=\textwidth]{../figure_omicron/figure_omicron_switch.pdf}
 \caption{
-\textbf{Scenario projects of the spread of omicron variant on university campus with limited isolation beds}
-(A) The predicted numbers of individuals in isolation on a given day.
-(B) The predicted daily number of new infections.
-Solid lines and shaded areas represent the median and the corresponding 90\% quantiles across 100 simulations.
-We assume $\mathcal R_0 = 8$, 30 booster shots given per day, and testing frequency of 3 days for these simulations.
-All other parameters are the same as in Figure 3.
-\label{fig:isolation}
+\textbf{Effects of reducing testing frequencies in the middle of a semester}
+Solid lines represent median predictions.
+Shaded areas represent 90\% quantiles across 100 simulations.
+Vertical dashed lines represent March 1, 2022, when testing protocol changes and infectious community contacts begins increasing exponentially.
+All other parameters are set to previously assumed values in \fref{sensitivity}.
+\label{fig:switch}
 }
@@
 
-When there are no limits to the number of isolation beds on campus, shortening isolation periods results in a slightly higher number of total infections (\fref{isolation}A): .
-The epidemic trajectory remains similar across different values of isolation periods (\fref{isolation}B), but the number of students in isolation becomes considerably lower with shorter isolation periods (\fref{isolation}C).
-For example, the maximum number of students in isolation decreases from ??? to ???.
-
-When the number of isolation beds is moderately limited ($=200$), shortening isolation periods can decrease the total numbers of infections (\fref{isolation}A) and considerably reduce the size of the epidemic peak (\fref{isolation}B).
-Even when the number of isolation beds is severely limited ($=100$), shortening isolation periods can still reduce the size of the epidemic peak (\fref{isolation}B) because we are able to isolate a greater proportion of infected individuals and prevent onward transmission.
-Longer isolation can prevent a greater amount of onward transmission \emph{per individual} but permits a lower number of individuals in isolation, making it less effectiveness when the number of isolation beds are limited.
+The impact of changing testing frequencies depends on the relative amount of community transmission and campus contagion.
+When contact levels are low or a large fraction of students have already had Omicron infections, switching from twice-a-week to once-a-week testing has little impact on epidemic dynamics because most infections are caused by community contacts \fref{switch}.
+Increasing contact levels or decreasing immunity levels permits a greater amount of campus transmission, causing greater discrepancies between two scenarios, when testing frequencies are held constant and when testing frequencies are allowed to change;
+in these cases, the exact effects of decreasing testing frequencies are still sensitive to the underlying immunity levels.
 
@@
 
-Here, we present the analysis of SARS-CoV-2 outbreaks on Princeton University campus between fall 2020 and early 2022.
-Our analysis demonstrates strong spatiotemporal correlations between the patterns of spread of SARS-CoV-2 on university campus and those from surrounding communities.
-These correlations decreased with distance from Mercer County in fall 2021, likely reflecting contact and commuting patterns as university re-opened.
+Here, we present the analysis of SARS-CoV-2 outbreaks on the Princeton University campus between fall 2020 and early 2022.
+Our analysis demonstrates strong spatiotemporal correlations between the patterns of spread of SARS-CoV-2 on campus and those from surrounding communities.
+These correlations decreased with distance from Mercer County in fall 2021, likely reflecting contact and commuting patterns as university reopened.
 Mathematical modeling further suggests limited transmission between the university population during fall and spring semesters of 2020 and an increased amount of community contacts during the fall semester of 2021 compared previous semesters.
 An increase in the number of cases by the end of November 2021 is consistent with the increase in the levels of community cases and waning immunity.
-Finally, we predict the spread of Omicron variant to likely cause a large outbreak in the spring semester of 2021 even under more stringent testing protocols.
-A rapid administration of booster shots and shortening isolation periods can help reduce the size of an outbreak, but uncertainty remains for future semesters.
+Finally, predicting future SARS-CoV-2 dynamics and changes in policy (e.g., decreasing testing frequency) will be sensitive to other intervention measures present on campus and underlying epidemiological conditions, such as community transmission and immunity levels.
 
-Previous outbreak reports from other universities primarily focused on transmission on university campuses \citep{wilson2020multiple,currie2021interventions} and provided limited insights into the importance of community transmission in shaping epidemic dynamics.
-For example, extensive modeling efforts from Cornell University also demonstrated increase in the amount of transmission from outside the university campus and found that community transmissions are the biggest risk for faculty and staff members \citep{frazier2022modeling};
-but they did not demonstrate explicit correlations between university and community cases and further assumed that the amount of community transmission does not change over a semester.
-To our knowledge, our analysis is the first to demonstrate a strong spatiotemporal correlation in the spread of SARS-CoV-2 between university campuses and surrounding communities.
-Our analysis further indicate that comparing the ratios between the cases on university campuses and neighboring communities can provide a useful measure for how well a university campus might be doing in controlling the epidemic---we note that this ratio need to be interpreted with caution as it is sensitive to changes in testing patterns.
+Although previous outbreak reports from other universities primarily focused on transmission on university campuses \citep{wilson2020multiple,currie2021interventions}, a few studies identified off-campus transmission as an important source of transmission \citep{fox2021response,hamer2021assessment}.
+For example, extensive modeling efforts from Cornell University demonstrated an increase in the amount of transmission from outside the university campus during fall 2021 and found that community transmissions are the biggest risk for faculty and staff members \citep{frazier2022modeling}.
+Our study further extends these findings in demonstrating a strong spatiotemporal correlation in the spread of SARS-CoV-2 between university campuses and surrounding communities.
+Our analysis suggest that comparing the ratios between the cases on university campuses and neighboring communities can also provide a useful measure for how well a university campus is controlling the epidemic; however, we note that this ratio need to be interpreted with caution as it is sensitive to changes in testing patterns.
 
@@
 While our analysis demonstrates strong spatiotemporal correlation in the spread of SARS-CoV-2, we are not able to infer the direction of causality---that is, our analysis does not rule out the possibility that transmission on campus drove infections in nearby communities (as opposed to community transmission driving on-campus infections).
-Nonetheless, intervention measures on campus (e.g., frequent asymptomatic testing, contact tracing, and virtual classes during fall and spring semesters of 2020) likely limited onward transmission on campus, and therefore, transmission from community to campus is likely to have had greater effects.
+However, seeding from campus is unlikely: 
+intervention measures on campus (e.g., frequent asymptomatic testing, contact tracing, and virtual classes during fall and spring semesters of 2020) likely limited onward transmission on campus.
 Decreasing patterns in epidemic correlations with distance further highlight the role of spatial spread in driving dynamics of SARS-CoV-2---such patterns are consistent with spatial spread of many other respiratory pathogens \citep{grenfell2001travelling, viboud2006synchrony, baker2019epidemic}.
@@
 Our mathematical model relies on simplifying assumptions.
-For example, we assume that the entire university populations mix homogeneously and have identical campus and community contact rates (captured by $\mathcal R_e$ and $\theta$, respectively).
+For example, we assume conservatively that the entire university populations mix homogeneously and have identical campus and community contact rates (captured by $\Rc$ and $\theta$, respectively).
 In reality, increases in cases were often associated with specific transmission clusters, suggesting heterogeneity in transmission patterns.
@@
 for example, faculty and staff members are more likely to interact with community members than undergraduate students and are therefore at a higher risk for community infections \citep{frazier2022modeling}.
-We also do not account for changes in behavior. 
-Instead, we assume a constant $\theta$ and $\mathcal R_e$ throughout a semester (note this $\mathcal R_e$ implicitly accounts for some interventions, but not for testing or changes in susceptibility driven by vaccination or infection).
+We also do not account for changes in behavior; 
+instead, we assume constant values for $\theta$ and $\Rc$ throughout a semester.
 We also do not explore parameter uncertainty, which can lead to underestimation of overall uncertainty \citep{elderd2006uncertainty}. 
-We also note that intervention measures that were introduced to Princeton University may not necessarily be applicable in other universities.
+We also note that intervention measures that were introduced to Princeton University may not necessarily be applicable in other institutions.
 
 Despite the simplicity of the analysis, our study provides important lessons for controlling SARS-CoV-2 outbreaks on university campuses in general.
-First, the safe reopening of a university campus must consider the spread of SARS-CoV-2 within the surrounding community as they can both drive transmission in each other.
-Second, combining multiple interventions measures, such as social distancing, mask wearing, frequent testing, and vaccination, can help provide safe means of re-opening university campuses---
-the extent to which these interventions are implemented will necessarily depend on resource availability
+First, our analysis highlights the power of mass testing for epidemic measurement and planning.
+Combining other interventions measures, such as social distancing, mask wearing, and vaccination, can help provide a safe means of reopening university campuses---but the extent to which these interventions are implemented will necessarily depend on resource availability.
+Second, the safe reopening of a university campus must consider the spread of SARS-CoV-2 within the surrounding community as they can both drive transmission in each other.
 Finally, intervention measures placed on campuses must continue to adapt and change to reflect changes in epidemiological conditions.
+
+The emergence of new variants---in particular, their ability to evade prior immunity and transmit better---continues to add uncertainty to the future controllability of the ongoing SARS-CoV-2 pandemic.
+Nonetheless, as population-level immunity increases (either due to infection or vaccination), we are transitioning to an endemic phase, during which COVID-19 is expected to become less severe \citep{lavine2021immunological}.
+Many institutions have so far sought to minimize infections on their campuses, thereby implementing as many interventions as possible---but such measures can be difficult to maintain for prolonged periods both economically and societally.
+Instead, the expectations for reopening campuses (e.g., whether to minimize infections on campuses) need to be re-evaluated as we transition to next phases of the pandemic, accounting not only for changes in epidemic dynamics but also for our perception of the pathogen.
+The answers to these questions will ultimately lie in the landscape of SARS-CoV-2 immunity \citep{saad2020immune,baker2021limits}.
 
@@
 
-\begin{figure}[!th]
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
+
+\begin{figure}[!htp]
 \includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_correlation2.pdf}
+\caption{
+\textbf{Correlations between the weekly number of cases in Princeton University and in Mercer County stratified by subpopulations.}
+Points represent the number of reported cases.
+Solid lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
+}
 \end{figure}
@@
 
-\begin{figure}[!th]
-\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_map.pdf}
+\begin{figure}[!htp]
+\begin{center}
+\includegraphics[width=0.7\textwidth]{../figure_princeton/figure_princeton_map.pdf}
+\end{center}
+\caption{
+\textbf{Correlations between the weekly number of cases in Princeton University and in counties in New Jersey.}
+(A--C) Map of correlations between the weekly number of cases in Princeton University and in counties in New Jersey.
+(D--F) Relationship between case correlations and distance from Mercer County.
+Points represent the estimated correlation coefficients.
+Solid lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
+Black points and lines represent correlations based on all cases in Princeton University.
+Red points and lines represent correlations based on cases among faculty and staff members in Princeton University.
+}
 \end{figure}
@@
 
-\begin{figure}[!th]
+\begin{figure}[!htp]
 \includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_newyork.pdf}
+\caption{
+\textbf{Dynamics of SARS-CoV-2 outbreaks in Princeton University and New York City}
+(A--C) Epidemic trajectories across three semesters: Fall 2020 (A), Spring 2020 (B), and Fall 20201 (C).
+Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in Princeton University.
+Red lines represent the weekly number of cases in New York City.
+(D--F) Correlations between the weekly number of cases in Princeton University and in New York City.
+Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
+}
 \end{figure}
@@
 
-\begin{figure}[!th]
+\begin{figure}[!htp]
 \includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_phil.pdf}
+\caption{
+\textbf{Dynamics of SARS-CoV-2 outbreaks in Princeton University and Philadelphia}
+(A--C) Epidemic trajectories across three semesters: Fall 2020 (A), Spring 2020 (B), and Fall 20201 (C).
+Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in Princeton University.
+Red lines represent the weekly number of cases in Philadelphia.
+(D--F) Correlations between the weekly number of cases in Princeton University and in Philadelphia.
+Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
+}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!htp]
+\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_map_extended.pdf}
+\caption{
+\textbf{Correlations between the weekly number of cases in Princeton University and in counties in New Jersey, Pennsylvania, and New York State.}
+(A--C) Map of correlations between the weekly number of cases in Princeton University and in counties in New Jersey, Pennsylvania, and New York State.
+(D--F) Relationship between case correlations and distance from Mercer County.
+Points represent the estimated correlation coefficients.
+Solid lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
+Black points and lines represent correlations based on all cases in Princeton University.
+Red points and lines represent correlations based on cases among faculty and staff members in Princeton University.
+New York City is excluded from this analysis as the data provided by New York Times are not further stratified by county levels.
+}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!htp]
+\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_simulation_fall_2020_all.pdf}
+\caption{
+\textbf{Comparisons between model predictions and the observed numbers of cases for fall 2020}
+Points represent the weekly number of reported cases in Princeton University.
+Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
+}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!htp]
+\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_simulation_spring_2020_all.pdf}
+\caption{
+\textbf{Comparisons between model predictions and the observed numbers of cases for spring 2020}
+Points represent the weekly number of reported cases in Princeton University.
+Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
+}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!htp]
+\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_simulation_fall_2021_all.pdf}
+\caption{
+\textbf{Comparisons between model predictions and the observed numbers of cases for fall 2021}
+Points represent the weekly number of reported cases in Princeton University.
+Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
+}
 \end{figure}

### 2022-02-22 doc/university-covid.tex
done?

@@
 
-Mathematical models have played a crucial role in guiding pandemic responses, especially for reopening university campuses.
-Here, we present descriptive and modeling analyses of SARS-CoV-2 outbreaks on the Princeton University campus---this model was used throughout the pandemic to inform policy decisions in Princeton University.
+Mathematical models have played a crucial role in guiding pandemic responses; university campuses provide a particularly well documented case for institutional outbreaks.
+Here, we present descriptive and modeling analyses of SARS-CoV-2 outbreaks on the Princeton University campus---this model was used throughout the pandemic to inform policy decisions for the Princeton University campus.
 We demonstrate strong spatiotemporal correlations in epidemic patterns between Princeton University and surrounding communities.
-These findings are corroborated by our model predictions, which indicate that the amount of on-campus transmission was likely limited;
-mathematical modeling further illustrates that an increase in the number of cases in November 2021 is consistent with waning immunity.
-Predicting future outbreaks and the effectiveness of intervention measures on campus is expected to be sensitive to underlying epidemiological conditions, including community prevalence and immunity landscape.
+These findings are corroborated by our model predictions, which indicate that the amount of on-campus transmission was likely limited during much of the wider pandemic.
+Mathematical modeling further illustrates that an increase in the number of cases in November 2021 is consistent with waning immunity.
+Predicting future outbreaks, and the effectiveness of intervention measures on campus, is then expected to be sensitive to underlying epidemiological conditions, including community prevalence and immunity landscape.
 
@@
 
-Mathematical models have been particularly useful in planning campus reopenings.
-Researchers from various institutions---including Cornell \citep{frazier2022modeling}, Emory \citep{lopman2020model}, Georgia Institute of Technology \citep{gibson2021surveillance}, and UC Berkeley \citep{brook2021optimizing}---modeled the feasibility of controlling the epidemic on their campuses and considered mass asymptomatic testing as their main intervention.
+Mathematical models have also been widely deployed in planning campus reopenings.
+Researchers from various institutions in the US---including Cornell \citep{frazier2022modeling}, Emory \citep{lopman2020model}, Georgia Institute of Technology \citep{gibson2021surveillance}, and UC Berkeley \citep{brook2021optimizing}---modeled the feasibility of controlling the epidemic on their campuses and considered mass asymptomatic testing as their main intervention.
 These modeling efforts helped identify key parameters for control, such as the testing turnaround time, and provided support for implementing similar measures at other institutions.
@@
 The impact of intervention measures is also expected to vary across different university campuses, reflecting heterogeneity in campus settings such as compliance, resources, community prevalence, as well as effects of other interventions present on campuses.
-For example, Duke and Harvard Universities experienced moderate outbreaks at the beginning of the fall semester in 2021 when in-person classes were allowed, despite high vaccination rates and weekly testing protocols \citep{dukeoutbreak,harvardoutbreak}, whereas the number of cases remained low in Princeton University during the same time period despite similar levels of testing and vaccination.
-Here, we choose to focus on the spread of SARS-CoV-2 on Princeton University campus alone to eliminate heterogeneities inherent to such comparisons.
+For example, Duke and Harvard Universities experienced moderate outbreaks at the beginning of the fall semester in 2021 when in-person classes were allowed, despite high vaccination rates and weekly testing protocols \citep{dukeoutbreak,harvardoutbreak}, whereas the number of cases remained low in Princeton University (PU) during the same time period despite similar levels of testing and vaccination.
+Here, we choose to focus on the spread of SARS-CoV-2 on PU campus alone to eliminate heterogeneities inherent to such comparisons.
 
-In this study, we analyze the spread of SARS-CoV-2 on Princeton University campus (\fref{princeton}).
+In this study, we analyze the spread of SARS-CoV-2 on PU campus (\fref{princeton}).
 We first begin with a descriptive analysis of the outbreak and then present modeling analyses of the past outbreaks and future scenarios.
-Princeton University is located in Mercer County, New Jersey, USA and consists of 5267 undergraduate students, 2946 graduate students, and 7000 faculty and staff members.
+PU is located in Mercer County, New Jersey, USA;
+the population comprises of 5267 undergraduate students, 2946 graduate students, and 7000 faculty and staff members.
 For simplicity, we divide the epidemic into three time periods representing three semesters: Fall 2020 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020 (January 16, 2021--May 14, 2021; \fref{princeton}B), and Fall 2021 (August 14, 2021--January 14, 2021; \fref{princeton}C).
@@
 Throughout the study period, contact tracing was also performed for positive cases to quarantine their close contacts, defined as individuals who have been within 6 feet for at least 15 minutes, cumulatively within a 24-hour period.
-All data used in this analysis are publicly available on the Princeton University COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.
+All data used in this analysis are publicly available on the PU COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.
 
@@
 \caption{
-\textbf{Dynamics of SARS-CoV-2 outbreaks in Princeton University.}
+\textbf{Dynamics of SARS-CoV-2 outbreaks in Princeton University (PU).}
 (A--C) Epidemic trajectories across three semesters: Fall 2020 (A), Spring 2020 (B), and Fall 20201 (C).
-Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in Princeton University.
+Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in PU.
 Red lines represent the weekly number of cases in Mercer County.
 Number of cases in Mercer County is obtained from \url{https://github.com/nytimes/covid-19-data}.
-(D--F) Correlations between the weekly number of cases in Princeton University and in Mercer County.
+(D--F) Correlations between the weekly number of cases in PU and in Mercer County.
 Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
-(G) Ratios between weekly numbers of cases reported from Princeton University and Mercer County.
+(G) Ratios between weekly numbers of cases reported from PU and Mercer County.
 Solid lines represent the observed ratios.
@@
 All classes remained virtual, and the testing protocol did not change (twice a week for undergraduate and graduate students, and once a week for faculty and staff members).
-The number of cases persisted at similar levels as the fall semester and eventually decreased as classes ended and students went back home---the decrease in the number of cases in Princeton University also coincided with the decrease in the number of cases in Mercer County.
+The number of cases persisted at similar levels as the fall semester and eventually decreased as classes ended and students went back home---the decrease in the number of cases in PU also coincided with the decrease in the number of cases in Mercer County.
 The highest number of cases was reported among faculty and staff members ($=111$), followed by undergraduate students ($=101$), and graduate students ($=29$).
@@
 
-Across all three semesters, we find strong and significant correlations between the weekly logged numbers of cases from Princeton University and those from Mercer county:
+Across all three semesters, we find strong and significant correlations between the weekly logged numbers of cases from PU and those from Mercer county:
 fall 2020 ($\rho = 0.81$ (95\% CI: 0.55--0.92; $p < 0.001$), \fref{princeton}D); spring 2020 ($\rho = 0.80$ (95\% CI: 0.51--0.92; $p < 0.001$), \fref{princeton}E); and fall 2021 ($\rho = 0.90$ (95\% CI: 0.78--0.96; $p < 0.001$), \fref{princeton}F).
 These correlations are robust even when we stratify cases by the population, except for undergraduate students during Fall 2020 when most of them were not physically present on campus (Supplementary Figure S1).
-We also find strong correlations between the weekly logged numbers of cases from Princeton University and those from other counties in New Jersey (Supplementary Figure S2)---these correlations significantly decreased with distance from Mercer county in both spring 2020 ($\rho=-0.48$ (95\% CI: $-0.75$--$-0.06$; $p = 0.03$) for all cases and $\rho=-0.51$ (95\% CI: $-0.77$--$-0.10$; $p = 0.02$) for faculty and staff cases) and fall 2021 ($\rho=-0.68$ (95\% CI: $-0.86$--$-0.35$, $p < 0.001$) for all cases and $\rho=-0.74$ (95\% CI: $-0.89$--$-0.46$, $p < 0.001$) for faculty and staff cases).
+We also find strong correlations between the weekly logged numbers of cases from PU and those from other counties in New Jersey (Supplementary Figure S2)---these correlations significantly decreased with distance from Mercer county in both spring 2020 ($\rho=-0.48$ (95\% CI: $-0.75$--$-0.06$; $p = 0.03$) for all cases and $\rho=-0.51$ (95\% CI: $-0.77$--$-0.10$; $p = 0.02$) for faculty and staff cases) and fall 2021 ($\rho=-0.68$ (95\% CI: $-0.86$--$-0.35$, $p < 0.001$) for all cases and $\rho=-0.74$ (95\% CI: $-0.89$--$-0.46$, $p < 0.001$) for faculty and staff cases).
 
@@
 Instead, the mean ratios remained similar across three semesters despite changes in campus populations (e.g., undergraduate student returning in spring 2020) and vaccination levels: 0.034 (95\% CI: 0.021--0.047) for fall 2020, 0.019 (95\% CI: 0.015--0.023) for spring 2020, and 0.046 (95\% CI: 0.035--0.057) for fall 2021.
-However, this does not necessarily imply that the contact or transmission levels stayed constant over time: given extremely high vaccination coverage in Princeton University during fall 2021, the university population would been subject to a much higher forces of infection (e.g., a greater amount of mixing with the community population) to obtain similar levels of cases.
+However, this does not necessarily imply that the contact or transmission levels stayed constant over time: given extremely high vaccination coverage in PU during fall 2021, the university population would been subject to a much higher forces of infection (e.g., a greater amount of mixing with the community population) to obtain similar levels of cases.
 We also find that some variations in the ratio around the mean can be explained by holiday effects.
@@
 
-\section*{Mathematical model}
+\section*{Mathematical modeling of past outbreaks}
 
-We use a discrete-time, individual-based model to simulate the spread of SARS-CoV-2 on the Princeton University campus.
-This model was initially developed and used throughout the pandemic to inform policy decisions in Princeton University, including the frequency of asymptomatic tests and the number of isolation beds required.
+We use a discrete-time, individual-based model to simulate the spread of SARS-CoV-2 on the PU campus.
+This model was initially developed and used throughout the pandemic to inform policy decisions in PU, including the frequency of asymptomatic tests and the number of isolation beds required.
 We continuously updated the model to reflect changes in school settings (e.g., students returning back to campus after a virtual semester) as well as intervention measures (e.g., vaccination in fall 2021 and booster shots with the emergence of the Omicron variant).
-Here, we present a generic and parsimonious version that encompasses sufficient details to characterize the overall spread of SARS-CoV-2 in Princeton University without an over-proliferation of parameters.
+Here, we present a generic and parsimonious version that encompasses sufficient details to characterize the overall spread of SARS-CoV-2 in PU without an over-proliferation of parameters.
 The model consists of four main components that are simulated on a daily time scale: (1) infection and transmission dynamics, (2) sampling and testing protocols, (3) isolation protocols, and (4) vaccination dynamics, including waning immunity and booster shots.
@@
 
-Infection processes are modeled based on standard compartmental structures.
+Infection processes are modeled based on standard compartmental structures (Supplementary Figure S6).
 Once infected, susceptible individuals remain in the exposed stage for $D_e = 2$ days on average, during which they cannot transmit or test positive. 
@@
 Presymptomatic, symptomatic, and asymptomatic infection stages are further divided into two subcompartments to allow for more realistic and narrower distributions than the exponential distribution \citep{brett2020transmission}.
-Transitions between each (sub)compartments are modeled using a Bernoulli process with probabilities that match the assumed means \citep{he2010plug}.
-More specifically, transition probabilities are equal to $1 - \exp(-\delta_x)$, where $\delta_x = -log(1-n/D_x)$ represent the transition rate from stage $X$ and $n$ represents the number of subcompartments.
+Transitions between each (sub)compartments are modeled using a Bernoulli process with probabilities that match the assumed means \citep{he2010plug}:
+more specifically, transition probabilities are equal to $1 - \exp(-\delta_x)$, where $\delta_x = -log(1-n/D_x)$ represent the transition rate from stage $X$ and $n$ represents the number of subcompartments.
+Assumed parameters are broadly consistent with other modeling approaches  \citep{brett2020transmission,lavezzo2020suppression}.
 
@@
 Pre-symptomatic and (a)symptomatic reproduction numbers are calculated based on the assumed value of the proportion of presymptomatic transmission $p_p = 0.5$: $\mathcal R_p/\mathcal R_s = p_p/(1 - p_p)$.
-On each day, all infected individuals who have not yet been isolated then make infectious contacts at random to anyone on campus; the number of infectious contacts are drawn from a negative binomial distribution with a mean of either $\beta_p$ or $\beta_s$ and an overdispersion parameter of $k=0.1$ to account for the possibility of super-spreading events.
+On each day, all infected individuals who have not yet been isolated then make infectious contacts at random to anyone on campus; the number of infectious contacts are drawn from a negative binomial distribution with a mean of either $\beta_p$ or $\beta_s$ and an overdispersion parameter of $k=0.1$ to account for the possibility of super-spreading events \citep{10.12688/wellcomeopenres.15842.3}.
 Infectious contacts from the community can be also made at random to anyone on campus; these contacts are modeled using a Poisson distribution with a time-varying mean, which is calculated by scaling the daily number of cases by $\theta$ and shifting it by 1 week to account for reporting delays.
@@
 Isolated individuals are assumed to no longer transmit infections.
-We assume that PCR tests have 95\% sensitivity and 100\% specificity.
+We assume that PCR tests can detect infections from individuals who are in pre-symptomatic, symptomatic, and asymptomatic stages with 95\% sensitivity and 100\% specificity.
 
@@
 We further assume that the population mixes homogeneously.
-In practice, epidemiological parameters, such as contact rates and the proportion of asymptomatic infections, likely differ across groups, and mixing patterns likely exhibit assortativity (e.g., undergraduate students are more likely to infect undergraduate students).
+While these assumptions are most conservative, epidemiological parameters  and mixing patterns likely differ across groups (e.g., undergraduate students are more likely to infect undergraduate students and also remain asymptomatic).
 Therefore, our model parameters describe average dynamics across different groups and must be interpreted with care.
@@
 \caption{
-\textbf{Retrospective analysis of past SARS-CoV-2 outbreaks on Princeton University.}
+\textbf{Retrospective analysis of past SARS-CoV-2 outbreaks on Princeton University (PU).}
 (A, D, G) Time series comparisons of model predictions with observed data across ranges of contact reproduction number $\Rc$ and scaling parameter for community transmission $\theta$.
@@
 The median predictions are positively correlated with the observed dynamics ($\rho = 0.83$; 95\% CI: 0.61--0.93; \fref{matching}C).
-Although a wide range of assumptions about the levels of community transmission $\theta$ are consistent with the observed dynamics, our simulations preclude high levels of contact, $\Rc > 2$ (Supplementary Figure S6).
+Although a wide range of assumptions about the levels of community transmission $\theta$ are consistent with the observed dynamics, our simulations preclude high levels of contact, $\Rc > 2$ (Supplementary Figure S7).
 Distancing measures on campus and contact tracing efforts likely contributed to lowering contact levels $\Rc$.
@@
 A similar set of parameters can capture the observed dynamics in spring 2020.
-The best matching parameter predicts a slightly higher levels of community transmission $\theta=0.02$ (\fref{matching}D), but a wide range of parameters are consistent with the observed dynamics as before (Supplementary Figure S7).
+The best matching parameter predicts a slightly higher levels of community transmission $\theta=0.02$ (\fref{matching}D), but a wide range of parameters are consistent with the observed dynamics as before (Supplementary Figure S8).
 Simulations also preclude high $\Rc > 2$ again, suggesting that transmission between students were likely limited even though they had returned to campus---the absence of in-person teaching is likely to have contributed to lowering $\Rc$.
@@
 
-While $\theta = 0.15$ and $\Rc = 0.5$ gives the best matching parameter set with a median logged sum of squared errors of 8.88 (95\% quantile: 6.55--12.6), other parameter sets also give nearly identical fits (\fref{matching}H; Supplementary Figure S8):
+While $\theta = 0.15$ and $\Rc = 0.5$ gives the best matching parameter set with a median logged sum of squared errors of 8.88 (95\% quantile: 6.55--12.6), other parameter sets also give nearly identical fits (\fref{matching}H; Supplementary Figure S9):
 for example, $\theta = 0.1$ and $\Rc = 1$ gives a median logged sum of squared errors of 8.9 (95\% quantile: 5.79--13.6).
@@
 All simulations shown in \fref{matching}H, except for the $\Rc=8$ scenario, are similarly correlated with the observed numbers of cases (\fref{matching}G). 
-An increasing pattern in the logged sum of squared errors with $\Rc$ shown in \fref{matching}G is likely driven by the discrepancy around fall break (week ending October 26th) when the number of cases decreased suddenly, rather than a lack of fits---we did not explicitly model holiday effects for simplicity.
+While the logged sum of squared errors increases with $\Rc$ (\fref{matching}G), these patterns are likely driven by the discrepancy around fall break (week ending October 26th) when the number of cases decreased suddenly, rather than a lack of fit---we did not explicitly model holiday effects for simplicity.
 Extremely high vaccination rates and frequent testing likely limited transmission on campus, making epidemic dynamics largely insensitive to $\Rc$ even at a reasonably high value of $\Rc = 4$.
-These simulations also suggest that an increase in the number of cases in November can be explained by a combination of waning immunity alone without requiring additional changes in transmission dynamics (note we do not allow $\theta$ or $\Rc$ to vary over time).
-In Supplementary Figure S9, \swp{TODO}.
 
-Projecting the model beyond November 26th shows that we would have seen a similarly growth in the number of cases if conditions remained constant even without the introduction of the Omicron variant.
+In summary, these simulations also suggest that an increase in the number of cases in November can be explained by a combination of waning immunity alone without requiring additional changes in transmission dynamics (note we do not allow $\theta$ or $\Rc$ to vary over time).
+When we exclude immunity waning from the model, predicted epidemic dynamics exhibit slower growth and require even higher values of campus and community contact rates ($\Rc$ and $\theta$) to qualitatively match the observed dynamics (Supplementary Figure S10)---even so, the logged sum of squared differences are generally higher (with median logged sum of squared differences ranging from 6.9 to 41.3 for the same parameter regime).
+Thus, combining some amount of immune waning and high campus and community contact rates likely best explain the epidemic growth near the end of the semester.
+
+Projecting the model beyond November 26th implies that we would have seen a similarly growth in the number of cases if conditions remained constant even without the introduction of the Omicron variant.
 In other words, the Delta strain would have continued to spread on campus at a similar rate if the semester were to (hypothetically) continue until January without additional interventions (\fref{matching}H).
@@
 
-We project epidemic trajectories for the spread of the Omicron variant among 4000 students for the spring semester of 2021.
+\section*{Future projections of the Omicron variant}
+
+We project epidemic trajectories for the spread of the Omicron variant among 4000 students for the spring semester of the 2021--2022 academic year.
 In doing so, we assume the following as a nominal parameter set: $\Rc = 4$; testing every 3 days; 60\% boosted (and 99\% having received two doses) by the beginning of semesters with 30 students additionally boosted per day; 10 infectious community contacts per day; and 20\% infected with the Omicron variant before the beginning of the semester.
@@
 \textbf{Sensitivity of the spread of Omicron on campus to underlying epidemiological and intervention conditions.}
+We perform sensitivity analysis for potential Omicron outbreaks on campus by varying contact reproduction number $\Rc$ (A--B), asymptomatic testing frequency (C--D), vaccination strategy (E--F), community contacts (G--H), prevailing immunity (I--J), and isolation strategy (K--L).
 (A, C, E, G, I, K) Predicted numbers of Omicron infections per week on campus. 
@@
 We vary one parameter at a time while holding all other parameters constant at their nominal values.
+We present representative simulations in these panels; all other simulations can be found in Supplementary Figures S11--S14.
 (B, D, F, H, J, L) Predicted numbers of total Omicron infections on campus.
@@
 
-Intervention measures, such as reducing contact rates $\Rc$ (\fref{sensitivity}A, B), increasing testing frequencies (\fref{sensitivity}C, D), and vaccinating students (\fref{sensitivity}E, F), can help reduce the size of the epidemic peak and the final size of an epidemic but have limited impact.
+Intervention measures, such as reducing contact rates $\Rc$ (\fref{sensitivity}A, B), increasing testing frequencies (\fref{sensitivity}C, D, Supplementary Figure S11), and vaccinating students (\fref{sensitivity}E, F, Supplementary Figure S12), can help reduce the size of the epidemic peak and the final size of an epidemic but have limited impact.
 While social distancing and mass asymptomatic testing can be effective in preventing onward transmission, they cannot prevent community transmission.
-Vaccines can help reduce community transmission by reducing susceptibility, but current vaccines are likely imperfect against the Omicron variant \citep{ferguson2021report}.
+Vaccines can help reduce community transmission by reducing susceptibility, but current vaccines can still allow reinfections with the Omicron variant \citep{ferguson2021report}.
 Nonetheless, we note that vaccinating students throughout the semester is just as important as vaccinating them before the beginning of the semester---for example, providing 60 booster shots per day with 40\% coverage at the beginning of the semester yield similar outcomes as providing 30 booster shots per day with 60\% coverage at the beginning of the semester despite 20\% differences in the initial vaccine coverage (\fref{sensitivity}F).
+Faster deliveries of vaccines throughout the semester can also help outbreaks to decay faster (Supplementary Figure S12).
 
-Instead, epidemic dynamics are more sensitive to underlying epidemiological factors that cannot be controlled by intervention measures on campus, such as levels of community transmission (\fref{sensitivity}G, H) and previous infection histories (\fref{sensitivity}I, J).
-In particular, the epidemic on campus dies out much quicker when community cases are also decreasing---when community infectious contacts decrease at rate $r=-0.1/\textrm{day}$ starting from 10 infectious contacts at the beginning of semester, the median predicted number of total infections is 85 (90\% quantile: 54--143).
-Such rapid decrease in the number of infections was not achievable with intervention measures that we considered.  
+Instead, epidemic dynamics are more sensitive to underlying epidemiological factors that cannot be controlled by intervention measures on campus, such as levels of community transmission (\fref{sensitivity}G, H, Supplementary Figure S13) and prevailing immunity (\fref{sensitivity}I, J.
+In particular, the epidemic on campus dies out much quicker when community cases are also decreasing---such rapid decrease in the number of infections was not achievable with intervention measures that we considered when community infections are assumed to be constant.
 The proportion of students who have already had Omicron infections before the beginning of semester also play an important role in determining the size of an outbreak (\fref{sensitivity}I, J)---we note that this result depends on our assumption that natural infections provide stronger immunity than vaccinal immunity.
+
+\section*{Isolation}
 
@@
 
-When there are no limits to the number of isolation beds on campus, shortening isolation periods results in a slightly higher number of total infections due to residual transmission after the isolation periods are over (\fref{sensitivity}L); but the impact on overall epidemic dynamics under our nominal parameter assumptions (\fref{sensitivity}K).
-When the number of isolation beds is moderately limited ($=100$), shortening isolation periods can decrease the total numbers of infections (compare 10 vs 7 day in \fref{sensitivity}L), but shortening isolation periods too much can increase the total numbers of infections again (compare 7 vs 5 day in \fref{sensitivity}L).
+When there are no limits to the number of isolation beds on campus, shortening isolation periods results in a slightly higher number of total infections due to residual transmission after the isolation periods are over (\fref{sensitivity}L).
+The impact on overall epidemic dynamics under our nominal parameter assumptions (\fref{sensitivity}K), but the number of students in isolation is greatly reduced with shorter isolation periods (Supplementary Figure S14).
+When the number of isolation beds is moderately limited ($=100$), shortening isolation periods can decrease the total numbers of infections (compare 10 vs 7 day in \fref{sensitivity}L and Supplementary Figure S14), but shortening isolation periods too much can increase the total numbers of infections again (compare 7 vs 5 day in \fref{sensitivity}L and Supplementary Figure S14).
 When the number of isolation beds is severely limited ($=50$), the benefits of shortening isolation periods can further outweigh the risk of transmitting after isolation by allowing a greater proportion of infected individuals to be isolated.
 
-Finally, as the number of COVID-19 cases is decreasing in the US, many institutions, including Princeton University, participating asymptomatic testing programs may be considering reducing testing frequencies.
+\section*{Changes in asymptomatic testing frequency}
+
+Finally, as the number of COVID-19 cases is decreasing in the US, many institutions, including PU, may be considering reducing asymptomatic testing frequencies.
 Here, we test whether switching testing frequencies from twice a week to once a week in the middle of a semester will result in larger outbreaks across a wide range of contact $\Rc$ and immunity levels (i.e., proportion of students who have already had Omicron infections before the beginning of semester).
@@
 \caption{
-\textbf{Effects of reducing testing frequencies in the middle of a semester}
+\textbf{Effects of reducing testing frequencies in the middle of a semester.}
 Solid lines represent median predictions.
 Shaded areas represent 90\% quantiles across 100 simulations.
-Vertical dashed lines represent March 1, 2022, when testing protocol changes and infectious community contacts begins increasing exponentially.
+Vertical dashed lines represent March 1, 2022, when testing protocol changes and infectious community contacts begin increasing exponentially.
 All other parameters are set to previously assumed values in \fref{sensitivity}.
@@
 
-Here, we present the analysis of SARS-CoV-2 outbreaks on the Princeton University campus between fall 2020 and early 2022.
+Here, we present the analysis of SARS-CoV-2 outbreaks on the PU campus between fall 2020 and early 2022.
 Our analysis demonstrates strong spatiotemporal correlations between the patterns of spread of SARS-CoV-2 on campus and those from surrounding communities.
@@
 
-Although previous outbreak reports from other universities primarily focused on transmission on university campuses \citep{wilson2020multiple,currie2021interventions}, a few studies identified off-campus transmission as an important source of transmission \citep{fox2021response,hamer2021assessment}.
+Although previous outbreak reports from other universities primarily focused on transmission on university campuses \citep{wilson2020multiple,currie2021interventions}, a few studies identified off-campus infections as an important source of transmission \citep{fox2021response,hamer2021assessment}.
 For example, extensive modeling efforts from Cornell University demonstrated an increase in the amount of transmission from outside the university campus during fall 2021 and found that community transmissions are the biggest risk for faculty and staff members \citep{frazier2022modeling}.
 Our study further extends these findings in demonstrating a strong spatiotemporal correlation in the spread of SARS-CoV-2 between university campuses and surrounding communities.
-Our analysis suggest that comparing the ratios between the cases on university campuses and neighboring communities can also provide a useful measure for how well a university campus is controlling the epidemic; however, we note that this ratio need to be interpreted with caution as it is sensitive to changes in testing patterns.
+Our analysis suggest that comparing the ratios between the cases on university campuses and neighboring communities can also provide a useful measure for how well a university campus is controlling the epidemic;
+however, we note that this ratio needs to be interpreted with caution as it is sensitive to changes in testing patterns.
+Future studies may combine phylogenetic data to better understand spatial patterns of SARS-CoV-2 on campus.
 
@@
 There are several limitations to our analysis.
-While our analysis demonstrates strong spatiotemporal correlation in the spread of SARS-CoV-2, we are not able to infer the direction of causality---that is, our analysis does not rule out the possibility that transmission on campus drove infections in nearby communities (as opposed to community transmission driving on-campus infections).
+While we demonstrate strong spatiotemporal correlation in the spread of SARS-CoV-2, we are not able to infer the direction of causality---that is, our analysis does not rule out the possibility that transmission on campus drove infections in nearby communities (as opposed to community transmission driving on-campus infections).
 However, seeding from campus is unlikely: 
@@
 We also do not explore parameter uncertainty, which can lead to underestimation of overall uncertainty \citep{elderd2006uncertainty}. 
-We also note that intervention measures that were introduced to Princeton University may not necessarily be applicable in other institutions.
+We also note that intervention measures that were introduced to PU may not necessarily be applicable in other institutions.
 
@@
 \caption{
-\textbf{Correlations between the weekly number of cases in Princeton University and in Mercer County stratified by subpopulations.}
+\textbf{Correlations between the weekly number of cases in PU and in Mercer County stratified by subpopulations.}
 Points represent the number of reported cases.
@@
 \caption{
-\textbf{Correlations between the weekly number of cases in Princeton University and in counties in New Jersey.}
-(A--C) Map of correlations between the weekly number of cases in Princeton University and in counties in New Jersey.
+\textbf{Correlations between the weekly number of cases in PU and in counties in New Jersey.}
+(A--C) Map of correlations between the weekly number of cases in PU and in counties in New Jersey.
 (D--F) Relationship between case correlations and distance from Mercer County.
@@
 Solid lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
-Black points and lines represent correlations based on all cases in Princeton University.
-Red points and lines represent correlations based on cases among faculty and staff members in Princeton University.
+Black points and lines represent correlations based on all cases in PU.
+Red points and lines represent correlations based on cases among faculty and staff members in PU.
 }
@@
 \caption{
-\textbf{Dynamics of SARS-CoV-2 outbreaks in Princeton University and New York City}
+\textbf{Dynamics of SARS-CoV-2 outbreaks in PU and New York City.}
 (A--C) Epidemic trajectories across three semesters: Fall 2020 (A), Spring 2020 (B), and Fall 20201 (C).
-Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in Princeton University.
+Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in PU.
 Red lines represent the weekly number of cases in New York City.
-(D--F) Correlations between the weekly number of cases in Princeton University and in New York City.
+(D--F) Correlations between the weekly number of cases in PU and in New York City.
 Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
@@
 \caption{
-\textbf{Dynamics of SARS-CoV-2 outbreaks in Princeton University and Philadelphia}
+\textbf{Dynamics of SARS-CoV-2 outbreaks in PU and Philadelphia.}
 (A--C) Epidemic trajectories across three semesters: Fall 2020 (A), Spring 2020 (B), and Fall 20201 (C).
-Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in Princeton University.
+Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in PU.
 Red lines represent the weekly number of cases in Philadelphia.
-(D--F) Correlations between the weekly number of cases in Princeton University and in Philadelphia.
+(D--F) Correlations between the weekly number of cases in PU and in Philadelphia.
 Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
@@
 \caption{
-\textbf{Correlations between the weekly number of cases in Princeton University and in counties in New Jersey, Pennsylvania, and New York State.}
-(A--C) Map of correlations between the weekly number of cases in Princeton University and in counties in New Jersey, Pennsylvania, and New York State.
+\textbf{Correlations between the weekly number of cases in PU and in counties in New Jersey, Pennsylvania, and New York State.}
+(A--C) Map of correlations between the weekly number of cases in PU and in counties in New Jersey, Pennsylvania, and New York State.
 (D--F) Relationship between case correlations and distance from Mercer County.
@@
 Solid lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
-Black points and lines represent correlations based on all cases in Princeton University.
-Red points and lines represent correlations based on cases among faculty and staff members in Princeton University.
+Black points and lines represent correlations based on all cases in PU.
+Red points and lines represent correlations based on cases among faculty and staff members in PU.
 New York City is excluded from this analysis as the data provided by New York Times are not further stratified by county levels.
+}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!htp]
+\includegraphics[width=\textwidth]{../figure_diagram/diagram.pdf}
+\caption{
+\textbf{Compartmental diagram of the individual-based model.}
+Each compartment represents a stage of infection: susceptible $S$, exposed $S$, pre-symptomatic $I_p$, symptomatic $I_s$, asymptomatic $I_a$, and recovered $R$.
+Pre-symptomatic, symptomatic, and asymptomatic stages are further divided into two subcompartments.
+Individuals in pre-symptomatic, symptomatic, and asymptomatic stages can test positive with 95\% sensitivity.
 }
@@
 \caption{
-\textbf{Comparisons between model predictions and the observed numbers of cases for fall 2020}
-Points represent the weekly number of reported cases in Princeton University.
+\textbf{Comparisons between model predictions and the observed numbers of cases for fall 2020.}
+Points represent the weekly number of reported cases in PU.
 Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
+See figure 2 in the main text for details.
 }
@@
 \caption{
-\textbf{Comparisons between model predictions and the observed numbers of cases for spring 2020}
-Points represent the weekly number of reported cases in Princeton University.
+\textbf{Comparisons between model predictions and the observed numbers of cases for spring 2020.}
+Points represent the weekly number of reported cases in PU.
 Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
+See figure 2 in the main text for details.
 }
@@
 \caption{
-\textbf{Comparisons between model predictions and the observed numbers of cases for fall 2021}
-Points represent the weekly number of reported cases in Princeton University.
+\textbf{Comparisons between model predictions and the observed numbers of cases for fall 2021.}
+Points represent the weekly number of reported cases in PU.
 Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
+See figure 2 in the main text for details.
+}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!htp]
+\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_simulation_nowaning.pdf}
+\caption{
+\textbf{Comparisons between model predictions and the observed numbers of cases for fall 2021 without immunity waning.}
+Points represent the weekly number of reported cases in PU.
+Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
+See figure 2 in the main text for details.
+}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!htp]
+\includegraphics[width=\textwidth]{../figure_omicron/figure_omicron_sensitivity_testfreq.pdf}
+\caption{
+\textbf{Sensitivity of the spread of Omicron on campus to asymptomatic testing frequencies.}
+Colored lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
+See figure 3 in the main text for details.
+}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!htp]
+\includegraphics[width=\textwidth]{../figure_omicron/figure_omicron_sensitivity_boost_daily.pdf}
+\caption{
+\textbf{Sensitivity of the spread of Omicron on campus to vaccination strategies.}
+Colored lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
+See figure 3 in the main text for details.
+}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!htp]
+\includegraphics[width=\textwidth]{../figure_omicron/figure_omicron_sensitivity_thetabase.pdf}
+\caption{
+\textbf{Sensitivity of the spread of Omicron on campus to community transmission.}
+Colored lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
+See figure 3 in the main text for details.
+}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!htp]
+\includegraphics[width=\textwidth]{../figure_omicron/figure_omicron_sensitivity_isolation.pdf}
+\caption{
+\textbf{Sensitivity of the spread of Omicron on campus to isolation strategies.}
+Colored lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
+See figure 3 in the main text for details.
 }

### 2022-04-17 doc/university-covid.tex
bunch of edits

@@
 These findings are corroborated by our model predictions, which indicate that the amount of on-campus transmission was likely limited during much of the wider pandemic.
-Mathematical modeling further illustrates that an increase in the number of cases in November 2021 is consistent with waning immunity.
-Predicting future outbreaks, and the effectiveness of intervention measures on campus, is then expected to be sensitive to underlying epidemiological conditions, including community prevalence and immunity landscape.
 
@@
 
-University campuses also present unique challenges to controlling an outbreak.
+University campuses present unique challenges to controlling an outbreak.
 High-density environments, especially within dorms, and a large fraction of asymptomatic infections (due to the young age of university students) can readily permit rapid transmission.
-These kinds of contacts are inherently difficult to trace, making contact tracing less effective.
+These kinds of contacts are inherently difficult to keep track of, making contact tracing less effective.
 The impact of intervention measures is also expected to vary across different university campuses, reflecting heterogeneity in campus settings such as compliance, resources, community prevalence, as well as effects of other interventions present on campuses.
-For example, Duke and Harvard Universities experienced moderate outbreaks at the beginning of the fall semester in 2021 when in-person classes were allowed, despite high vaccination rates and weekly testing protocols \citep{dukeoutbreak,harvardoutbreak}, whereas the number of cases remained low in Princeton University (PU) during the same time period despite similar levels of testing and vaccination.
-Here, we choose to focus on the spread of SARS-CoV-2 on PU campus alone to eliminate heterogeneities inherent to such comparisons.
+For example, Duke and Harvard Universities experienced moderate outbreaks at the beginning of the fall semester in 2021 when in-person classes were allowed, despite high vaccination rates and weekly asymptomatic testing protocols \citep{dukeoutbreak,harvardoutbreak}, whereas the number of cases remained low in Princeton University (PU) during the same time period with similar levels of testing and vaccination.
+Here, we focused on the spread of SARS-CoV-2 on PU campus alone to eliminate heterogeneities inherent to such comparisons.
 
-In this study, we analyze the spread of SARS-CoV-2 on PU campus (\fref{princeton}).
+In this study, we analyzed the spread of SARS-CoV-2 on PU campus (\fref{princeton}).
 We first begin with a descriptive analysis of the outbreak and then present modeling analyses of the past outbreaks and future scenarios.
@@
 the population comprises of 5267 undergraduate students, 2946 graduate students, and 7000 faculty and staff members.
-For simplicity, we divide the epidemic into three time periods representing three semesters: Fall 2020 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020 (January 16, 2021--May 14, 2021; \fref{princeton}B), and Fall 2021 (August 14, 2021--January 14, 2021; \fref{princeton}C).
-Throughout the study period, all students, faculty and staff members who were physically present for more than 8 hours on campus per week were required to participate in asymptomatic testing with varying frequencies.
+For simplicity, we divided the epidemic into four time periods representing four semesters across two academic years: Fall 2020--2021 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020--2021 (January 16, 2021--May 14, 2021; \fref{princeton}B), Fall 2021--2022 (August 14, 2021--December 31, 2021; \fref{princeton}C), and Spring 2021--2022 (January 1, 2022--March 18, 2022; \fref{princeton}D).
+Throughout the majority of the study period, all students, faculty and staff members who were physically present for more than 8 hours on campus per week were required to participate in asymptomatic testing with varying frequencies.
 Asymptomatic individuals submitted self-collected saliva samples, from which the presence of SARS-CoV-2 was tested using Reverse Transcription Polymerase Chain Reaction (RT-PCR).  
 Asymptomatic individuals who tested positive were required to isolate for 10 days.
-Symptomatic individuals were eligible for rapid PCR tests and had to isolate \swp{self-isolate?} until they received their test results---those who tested positive were required to isolate for at least 10 days after symptom onset and were release 3 days after symptoms resolved.
+Symptomatic individuals were eligible for rapid PCR tests and had to isolate \swp{Robin and Irini: is self-isolate a better term?} until they received their test results---those who tested positive were required to isolate for at least 10 days after symptom onset and were release 3 days after symptoms resolved.
 PCR positives were exempt from asymptomatic testing for 90 days.
+Since March 7, 2022, asymptomatic testing frequencies decreased to once a month from once a week for individuals whose vaccine statuses are up-to-date, which in turn likely reduced the accuracy of surveillance; therefore, we chose to focus on epidemic patterns before this change was implemented.
 Throughout the study period, contact tracing was also performed for positive cases to quarantine their close contacts, defined as individuals who have been within 6 feet for at least 15 minutes, cumulatively within a 24-hour period.
@@
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton.pdf}
+\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_new.pdf}
 \caption{
 \textbf{Dynamics of SARS-CoV-2 outbreaks in Princeton University (PU).}
-(A--C) Epidemic trajectories across three semesters: Fall 2020 (A), Spring 2020 (B), and Fall 20201 (C).
+(A--D) Epidemic trajectories across four semesters: Fall 2020--2021 (A), Spring 2020--2021 (B), Fall 2021--2022 (C), and Spring 2021--2022 (D).
 Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in PU.
@@
 Number of cases in Mercer County is obtained from \url{https://github.com/nytimes/covid-19-data}.
-(D--F) Correlations between the weekly number of cases in PU and in Mercer County.
+(E--H) Correlations between the weekly number of cases in PU and in Mercer County.
 Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
-(G) Ratios between weekly numbers of cases reported from PU and Mercer County.
-Solid lines represent the observed ratios.
-Dotted lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
+(I--L) Ratios between weekly numbers of cases reported from PU and Mercer County.
+The shaded orange area covers the region between 1:100 and 1:25 ratios.
 \label{fig:princeton}
@@
 
-During fall 2020, roughly 1000 grad students and 2000 faculty and staff members were present on campus and participated in asymptomatic testing. 
+During the fall semester of the 2020--2021 academic year, roughly 1000 grad students and 2000 faculty and staff members were present on campus and participated in asymptomatic testing. 
 Undergraduate students were not allowed to return to campus as all classes were held virtually, and so their numbers were minimal on campus ($<300$).  
@@
 
-In the beginning of spring 2020, the number of cases suddenly increased before classes began (\fref{princeton}B), reflecting $\approx 3000$ undergraduate students returning to campus.
-Returning students were required to be tested and quarantine \swp{for how long} if they were coming back from CDC-designated states and countries \swp{ask Irini and Robin for details}.
+In the beginning of the spring semester of the 2020--2021 academic year, the number of cases suddenly increased before classes started (\fref{princeton}B), reflecting $\approx 3000$ undergraduate students returning to campus.
+Returning students were required to be tested and quarantine \swp{Robin and Irini: for how long} if they were coming back from CDC-designated states and countries \swp{ask Irini and Robin for details}.
 All classes remained virtual, and the testing protocol did not change (twice a week for undergraduate and graduate students, and once a week for faculty and staff members).
@@
 
-In fall 2021, all students and faculty and staff members were required to be vaccinated \swp{with two doses? need to ask Robin} with a very few medical and religious exemptions.
-By the beginning of the semesters, ??\% of students and ??\% of faculty and staff members were vaccinated \swp{ask Robin}.
-As of January 11th, 99\% of undergraduate students, 98\% of graduate students, and 96\% of faculty and staff members were vaccinated.
+For the fall semester of the 2021--2022 academic year, all students and faculty and staff members were required to be vaccinated \swp{with two doses?
+need to ask Robin and Irini} with a very few medical and religious exemptions.
+By the beginning of the semesters, ??\% of students and ??\% of faculty and staff members were vaccinated \swp{ask Robin and Irini}.
+% As of January 11th, 99\% of undergraduate students, 98\% of graduate students, and 96\% of faculty and staff members were vaccinated.
 Vaccinees were required to be tested once a week, while unvaccinated individuals were required to be tested twice a week.
@@
 In order to prevent transmission, testing frequency was increased to twice a week for undergraduate students on November 27th, 2021; the size of non-academic gatherings were also limited to 20 people.
-The number of cases decreased slightly as classes ended but soon increased back up as the Omicron variant began to spread on campus and in Mercer county.
+The number of cases decreased slightly as classes ended but soon increased back up as the Omicron (BA.1) variant began to spread on campus and in Mercer county.
 
-Across all three semesters, we find strong and significant correlations between the weekly logged numbers of cases from PU and those from Mercer county:
-fall 2020 ($\rho = 0.81$ (95\% CI: 0.55--0.92; $p < 0.001$), \fref{princeton}D); spring 2020 ($\rho = 0.80$ (95\% CI: 0.51--0.92; $p < 0.001$), \fref{princeton}E); and fall 2021 ($\rho = 0.90$ (95\% CI: 0.78--0.96; $p < 0.001$), \fref{princeton}F).
+For the spring semester of the 2021--2022 academic year, all eligible students and faculty and staff members were required to be boosted.
+By the beginning of the semesters, ??\% of students and ??\% of faculty and staff members were boosted \swp{ask Robin and Irini}.
+Undergraduate students were still required to be tested twice a week to prevent the additional spread of the Omicron variant.
+The number of cases remained high before classes began but decreased over time, following epidemic patterns in Mercer County (\fref{princeton}D).
+As the campus and local cases numbers decreased, the gathering policy was updated to allow food in events were no longer limited to 20 people;
+in addition, the testing frequency was reduced a once a week.
+Following the policy change, a large outbreak occurred on campus with high case numbers persisting until the spring break.
+On March 7, 2022, mask mandates were lifted and testing frequency was reduced to once a month.
+The timing of this outbreak also coincided with a rapid turnover of the Omicron subvariant BA.2---the proportion of BA.2 subvariant reached ??\% compared to ??\% from the previous week \swp{Ask Dan}.
+
+Across the first three semesters, we find strong and significant correlations between the weekly logged numbers of cases from PU and those from Mercer county:
+fall 2020--2021 ($\rho = 0.81$ (95\% CI: 0.55--0.92; $p < 0.001$), \fref{princeton}E); spring 2020--2021 ($\rho = 0.80$ (95\% CI: 0.51--0.92; $p < 0.001$), \fref{princeton}F); and fall 2021--2022 ($\rho = 0.86$ (95\% CI: 0.67--0.94; $p < 0.001$), \fref{princeton}G).
 These correlations are robust even when we stratify cases by the population, except for undergraduate students during Fall 2020 when most of them were not physically present on campus (Supplementary Figure S1).
-We also find strong correlations between the weekly logged numbers of cases from PU and those from other counties in New Jersey (Supplementary Figure S2)---these correlations significantly decreased with distance from Mercer county in both spring 2020 ($\rho=-0.48$ (95\% CI: $-0.75$--$-0.06$; $p = 0.03$) for all cases and $\rho=-0.51$ (95\% CI: $-0.77$--$-0.10$; $p = 0.02$) for faculty and staff cases) and fall 2021 ($\rho=-0.68$ (95\% CI: $-0.86$--$-0.35$, $p < 0.001$) for all cases and $\rho=-0.74$ (95\% CI: $-0.89$--$-0.46$, $p < 0.001$) for faculty and staff cases).
+However, the case patterns in PU were decoupled from those in Mercer county for the spring semester of the 2021--2022 academic year with unclear correlations (\fref{princeton}H): $\rho = 0.32$ (95\% CI: -0.34--0.77; $p = 0.33$).
+Stratifying cases by the population shows that case patterns in graduate students and faculty and staff members were still strongly correlated with case patterns in Mercer county, meaning that the campus transmission was limited to undergraduate students (Supplementary Figure S1).
+We also find strong correlations between the weekly logged numbers of cases from PU and those from other counties in New Jersey (Supplementary Figure S2)---these correlations significantly decreased with distance from Mercer county in both spring 2020--2021 ($\rho=-0.48$ (95\% CI: $-0.75$--$-0.06$; $p = 0.03$) for all cases and $\rho=-0.51$ (95\% CI: $-0.77$--$-0.10$; $p = 0.02$) for faculty and staff cases) and fall 2021--2022 ($\rho=-0.68$ (95\% CI: $-0.86$--$-0.35$, $p < 0.001$) for all cases and $\rho=-0.74$ (95\% CI: $-0.89$--$-0.46$, $p < 0.001$) for faculty and staff cases).
+Across the first three semesters, both the total cases and faculty and staff cases showed similar levels of correlations with local cases. 
+For the spring semester of the 2021--2022 academic year, we still find high correlations between faculty and staff cases and local cases throughout other countries ($\rho > 0.8$ across all counties in New Jersey); however, the total cases exhibit considerably weaker correlations due to student-to-student transmission on campus (Supplementary Figure S2).
 
 These correlations likely reflect commuting and contact patterns, and therefore we expect SARS-CoV-2 dynamics on campus to be correlated with those from nearby large cities as well. 
-We find similarly strong correlations with New York City: fall 2020 ($\rho = 0.74$ (95\% CI: 0.44--0.90; $p < 0.001$), Supplementary Figure S3A); spring 2020 ($\rho = 0.84$ (95\% CI: 0.61--0.94; $p < 0.001$), Supplementary Figure S3B); and fall 2021 ($\rho = 0.86$ (95\% CI: 0.70--0.94; $p < 0.001$), Supplementary Figure S3C).
-The same picture emerges for Philadelphia except for spring 2020: fall 2020 ($\rho = 0.83$ (95\% CI: 0.59--0.93; $p < 0.001$), Supplementary Figure S4A); spring 2020 ($\rho = 0.29$ (95\% CI: $-0.22$--0.68; $p = 0.25$), Supplementary Figure S4B); and fall 2021 ($\rho = 0.86$ (95\% CI: 0.68--0.94; $p < 0.001$), Supplementary Figure S4C).
+We find similarly strong correlations with New York City for the first three semesters: fall 2020--2021 ($\rho = 0.74$ (95\% CI: 0.44--0.90; $p < 0.001$), Supplementary Figure S3A); spring 2020--2021 ($\rho = 0.84$ (95\% CI: 0.61--0.94; $p < 0.001$), Supplementary Figure S3B); fall 2021--2022 ($\rho = 0.78$ (95\% CI: 0.51--0.91; $p < 0.001$), Supplementary Figure S3C); and spring 2021--2022 ($\rho = 0.33$ (95\% CI: $-0.34$--0.78; $p=0.3$), Supplementary Figure S3D).
+
+The same picture emerges for Philadelphia except for spring 2020: fall 2020--2021 ($\rho = 0.83$ (95\% CI: 0.59--0.93; $p < 0.001$), Supplementary Figure S4A); spring 2020--2021 ($\rho = 0.29$ (95\% CI: $-0.22$--0.68; $p = 0.25$), Supplementary Figure S4B); fall 2021--2022 ($\rho = 0.78$ (95\% CI: 0.51--0.91; $p < 0.001$), Supplementary Figure S4C);
+and spring 2021--2022 ($\rho=0.37$ (95\% CI: $-0.30$--0.80; $p=0.27$), Supplementary Figure S4D).
 Including counties from New York and Pennsylvania states into the spatial correlation analysis yields additional insights (Supplementary Figure S5):
-epidemic dynamics were highly synchronized across all counties in fall 2020 and became less synchronized over time. 
-These correlations significantly decreased with distance in spring 2020 ($\rho = -0.36$ (95\% CI: $-0.49$--$-0.21$; $p < 0.001$)) and fall 20201 ($\rho = -0.58$ (95\% CI: $-0.68$--$-0.46$; $p < 0.001$)).
+epidemic dynamics were highly synchronized across all counties in fall 2020--2021 and became less synchronized over time. 
+These correlations significantly decreased with distance in spring 2020--2021 ($\rho = -0.36$ (95\% CI: $-0.49$--$-0.21$; $p < 0.001$)) and fall 2021--2022 ($\rho = -0.58$ (95\% CI: $-0.68$--$-0.46$; $p < 0.001$)).
 These variations likely reflect differences in vaccination levels and the timing of the introduction of the Omicron variant.
@@
 we expect this ratio to remain constant over time if the majority of infections on campus is caused by community transmission provided that testing patterns remain constant in both places.
-We do not find significant changes in the ratio between the weekly numbers of cases from Princeton and those from Mercer county (\fref{princeton}G).
-Instead, the mean ratios remained similar across three semesters despite changes in campus populations (e.g., undergraduate student returning in spring 2020) and vaccination levels: 0.034 (95\% CI: 0.021--0.047) for fall 2020, 0.019 (95\% CI: 0.015--0.023) for spring 2020, and 0.046 (95\% CI: 0.035--0.057) for fall 2021.
-However, this does not necessarily imply that the contact or transmission levels stayed constant over time: given extremely high vaccination coverage in PU during fall 2021, the university population would been subject to a much higher forces of infection (e.g., a greater amount of mixing with the community population) to obtain similar levels of cases.
-We also find that some variations in the ratio around the mean can be explained by holiday effects.
-For example, sudden decreases in the number of cases around Thanksgiving of 2020, fall recess of 2021, and end of classes of 2021 correspond to a reduction in the ratio between university and community cases.
-Similarly, a sudden increase in the number of cases during Thanksgiving of 2021 caused the ratio to increase, reflecting decoupling of epidemic dynamics due to on-campus transmission.
+We find that the ratio between the weekly numbers of cases from Princeton and those from Mercer county stayed between 1:100 and 1:25 until the end of the fall semester of the 2021--2022 academic year (\fref{princeton}I--K).
+An increase in this ratio at the end of November 2021 was associated with the campus outbreak before Thanksgiving followed by an introduction of the Omicron variant in December---this deviation indicates an increase in the amount of transmission on campus.
+During the spring semester of the 2021--2022 academic year, the ratio between PU cases and Mercer county cases increased above one, meaning that more cases were reported in PU than elsewhere in Mercer county (\fref{princeton}L);
+notably, we did not see an increase in Mercer county cases, meaning that there was little-to-no transmission from campus to local community.
+
+% We do not find significant changes in the ratio between the weekly numbers of cases from Princeton and those from Mercer county (\fref{princeton}G).
+% Instead, the mean ratios remained similar across three semesters despite changes in campus populations (e.g., undergraduate student returning in spring 2020) and vaccination levels: 0.034 (95\% CI: 0.021--0.047) for fall 2020, 0.019 (95\% CI: 0.015--0.023) for spring 2020, and 0.046 (95\% CI: 0.035--0.057) for fall 2021.
+% However, this does not necessarily imply that the contact or transmission levels stayed constant over time: given extremely high vaccination coverage in PU during fall 2021, the university population would been subject to a much higher forces of infection (e.g., a greater amount of mixing with the community population) to obtain similar levels of cases.
+% We also find that some variations in the ratio around the mean can be explained by holiday effects.
+% For example, sudden decreases in the number of cases around Thanksgiving of 2020, fall recess of 2021, and end of classes of 2021 correspond to a reduction in the ratio between university and community cases.
+% Similarly, a sudden increase in the number of cases during Thanksgiving of 2021 caused the ratio to increase, reflecting decoupling of epidemic dynamics due to on-campus transmission.
 
@@
 The model consists of four main components that are simulated on a daily time scale: (1) infection and transmission dynamics, (2) sampling and testing protocols, (3) isolation protocols, and (4) vaccination dynamics, including waning immunity and booster shots.
-Previous versions of the model included contact tracing, but we exclude it in this model for simplicity.
+Previous versions of the model included contact tracing, but we exclude it in this study for simplicity.
 
@@
 more specifically, transition probabilities are equal to $1 - \exp(-\delta_x)$, where $\delta_x = -log(1-n/D_x)$ represent the transition rate from stage $X$ and $n$ represents the number of subcompartments.
-Assumed parameters are broadly consistent with other modeling approaches  \citep{brett2020transmission,lavezzo2020suppression}.
+Assumed parameters are broadly consistent with other models of SARS-CoV-2 \citep{brett2020transmission,lavezzo2020suppression}.
 
 Transmission processes are modeled by first setting the contact reproduction number $\Rc$, which we define as the average number of infectious contacts an infected individual would make throughout the course of their infection;
-here, infectious contacts refer to contacts that would result in infection when the contactee is susceptible to infection.
+here, infectious contacts refer to contacts that would result in infection when the contactee (i.e., the contacted individual) is susceptible to infection.
 The contact reproduction number implicitly accounts for all intervention measures that we do not model explicitly, such as social distancing and contact tracing---in other words, this contact reproduction number does not account for asymptomatic testing or vaccination, which are modeled separately.
@@
 On each day, all infected individuals who have not yet been isolated then make infectious contacts at random to anyone on campus; the number of infectious contacts are drawn from a negative binomial distribution with a mean of either $\beta_p$ or $\beta_s$ and an overdispersion parameter of $k=0.1$ to account for the possibility of super-spreading events \citep{10.12688/wellcomeopenres.15842.3}.
-Infectious contacts from the community can be also made at random to anyone on campus; these contacts are modeled using a Poisson distribution with a time-varying mean, which is calculated by scaling the daily number of cases by $\theta$ and shifting it by 1 week to account for reporting delays.
-Infectious contacts, whether made by individuals on campus or from outside, result in infection only when the contacted individuals are susceptible; when the contacted individuals are vaccinated, and therefore partially susceptible to infection, contactees have a reduced probability of infection corresponding to their susceptibility (discussed later).
+Infectious contacts from local community can be also made at random to anyone on campus; these contacts are modeled using a Poisson distribution with a time-varying mean, which is calculated by scaling the daily number of cases by $\theta$ and shifting it by 1 week to account for reporting delays.
+Infectious contacts, whether made by individuals on campus or from outside, result in infection only when the contacted individuals are susceptible; when the contacted individuals are vaccinated, and therefore partially susceptible to infection, they have a reduced probability of infection corresponding to their susceptibility (discussed later).
 
@@
 
-In this study, we use this model to retrospectively analyze past outbreaks and further predict future epidemic trajectories for the Omicron variant.
-First, we try to match our model to epidemic patterns seen on campus by varying the contact reproduction number $\Rc$ and the amount of community transmission $\theta$ and holding all other parameters constant.
+In this study, we use this model to retrospectively analyze past outbreaks.
+First, we try to match our model to epidemic patterns seen on campus for the first three semesters, during which there was limited campus transmission, by varying the contact reproduction number $\Rc$ and the amount of community transmission $\theta$ and holding all other parameters constant.
 For each parameter combination, we simulate 100 epidemic trajectories and calculate the sum of squared differences between the weekly numbers of the observed and predicted positive cases.
@@
 
-For fall 2020, we simulate the model assuming 3000 individuals (1000 graduate students and 2000 faculty and staff members) on campus with 1000 of them participating in asymptomatic testing twice a week.
+For fall 2020--2021, we simulate the model assuming 3000 individuals (1000 graduate students and 2000 faculty and staff members) on campus with 1000 of them participating in asymptomatic testing twice a week.
 We find that a low level of contacts $\Rc=0.5$ and a small amount of community transmission $\theta=0.015$ is most consistent with the observed epidemic dynamics in fall 2020 (\fref{matching}A).
@@
 
-For spring 2020, we simulate the model assuming 8000 individuals (3000 undergraduate students, 2000 graduate students, and 3000 faculty and staff members) on campus with 5000 of them participating in asymptomatic testing twice a week.
+For spring 2020--2021, we simulate the model assuming 8000 individuals (3000 undergraduate students, 2000 graduate students, and 3000 faculty and staff members) on campus with 5000 of them participating in asymptomatic testing twice a week.
 We further assume that 4000 individuals (3000 undergraduate students and 1000 graduate students) returned to campus over 14 days (January 16, 2021--January 29, 2021);
@@
 
-A similar set of parameters can capture the observed dynamics in spring 2020.
+A similar set of parameters can capture the observed dynamics in spring 2020--2021.
 The best matching parameter predicts a slightly higher levels of community transmission $\theta=0.02$ (\fref{matching}D), but a wide range of parameters are consistent with the observed dynamics as before (Supplementary Figure S8).
@@
 
-For fall 2021, we assume 13000 individuals are present on campus (5000 undergraduate students, 2000 graduate students, and 6000 staff and faculty members) with 98\% of them vaccinated---here, vaccine-derived immunity is allowed to wane over time to ask whether the increase in the number of cases around November is consistent with the dynamics predicted by immunity waning.
+For fall 2021--2022, we assume 13000 individuals are present on campus (5000 undergraduate students, 2000 graduate students, and 6000 staff and faculty members) with 98\% of them vaccinated---here, vaccine-derived immunity is allowed to wane over time to ask whether the increase in the number of cases around November is consistent with the dynamics predicted by immunity waning.
 Vaccinated individuals are tested every week, whereas unvaccinated individuals are tested every 3 days. 
@@
 
-In summary, these simulations also suggest that an increase in the number of cases in November can be explained by a combination of waning immunity alone without requiring additional changes in transmission dynamics (note we do not allow $\theta$ or $\Rc$ to vary over time).
+These simulations suggest that an increase in the number of cases in November can be explained by a combination of waning immunity alone without requiring additional changes in transmission dynamics (note we do not allow $\theta$ or $\Rc$ to vary over time).
 When we exclude immunity waning from the model, predicted epidemic dynamics exhibit slower growth and require even higher values of campus and community contact rates ($\Rc$ and $\theta$) to qualitatively match the observed dynamics (Supplementary Figure S10)---even so, the logged sum of squared differences are generally higher (with median logged sum of squared differences ranging from 6.9 to 41.3 for the same parameter regime).
-Thus, combining some amount of immune waning and high campus and community contact rates likely best explain the epidemic growth near the end of the semester.
+Thus, combining some amount of immune waning and high campus and community contact rates likely best explains the epidemic growth near the end of the semester.
 
@@
 
-\section*{Future projections of the Omicron variant}
+\section*{The spread of the Omicron variant on campus}
 
-We project epidemic trajectories for the spread of the Omicron variant among 4000 students for the spring semester of the 2021--2022 academic year.
-In doing so, we assume the following as a nominal parameter set: $\Rc = 4$; testing every 3 days; 60\% boosted (and 99\% having received two doses) by the beginning of semesters with 30 students additionally boosted per day; 10 infectious community contacts per day; and 20\% infected with the Omicron variant before the beginning of the semester.
-We then ask how model predictions are sensitive to these assumptions about underlying epidemiological and intervention conditions by varying each parameter separately.
-In particular, to test for the effects of community transmission, we vary the baseline daily number of infectious contacts from community (5--20) and further allow the number of infectious contacts to exponentially decrease over time from these baseline values (between $r=-0.1/\textrm{day}$--$0/\textrm{day}$), reflecting a recent decrease in cases in New Jersey.
-% We also consider the impact of shortening isolation periods given the recent changes by the Centers for Disease Control and Prevention to reduce isolation periods from 10 to 5 days \citep{cdcisolation}.
+Epidemiological and interventions conditions changed throughout the spring semester of the 2021--2022 academic year.
+We therefore extend to model to account for more realistic settings and focus on the outbreak patterns among undergraduate students.
+First, based on \citep{ferguson2021report}, we assume that two and three doses of vaccines reduce susceptibility against the Omicron variant by 10\% and 70\%, respectively. 
+We also assume that the transmissibility of Omicron is reduced proportionally following previously assumed 90--to--20 ratio for the Delta variant;
+in other words, two and three doses of  vaccines reduce transmissibility by 2.2\% and 15.6\%, respectively.
+The immunity from the third dose is assumed to take 7 days to develop and wane at the same rate as before (in this case, 70\% to 39\% in 20 weeks).
+Finally, the isolation period is reduced to 5 days.
 
-Throughout these simulations, we make additional assumptions about the vaccine efficacy against the Omicron variant.
-Based on \cite{ferguson2021report}, we assume that two and three doses of vaccines reduce susceptibility against Omicron by 10\% and 70\%, respectively.
-We further assume that the transmissibility of Omicron is reduced proportionally following previously assumed 90-to-20 ratio for the Delta variant---in other words, two and three doses of vaccines reduce transmissibility by 2.2\% and 15.6\%, respectively.
-Finally, we assume that immunity from the third dose takes 14 days to develop and wanes at the same rate as previously assumed (in this case, 70\% to 39\% in 20 weeks).
-For simplicity, we do not explore the sensitivity of our projections to these assumptions.
+Here, we use the extended model to try to understand the drivers of a large campus outbreak that happened on the week ending February 18, 2022 (\fref{princeton}D).
+First, we ask whether changes in testing frequency from biweekly to weekly and an increased reproduction number can explain the outbreak. 
+The increase in the reproduction number can reflect increased contact rates following changes in distancing policy as well as increased transmissibility of the BA.2 subvariant---we do not explicitly distinguish the cause of the increase in the reproduction number.
+We do so by simulating the model forward across a range of contact reproduction numbers that are consistent with previous estimates ($\Rc=2--6$) and introducing a 20\%--100\% increase in the contact reproduction number on February 8, 2022, with changes in the testing frequency.
+To match the realistic campus setting, we assume that 700 students are present on campus as of January 1, 2022, and the remaining 4300 students come back to campus across 28 days.
+We assume that 99\% of students are vaccinated with 60\% of them being boosted---we further allow 70 booster shots on each day such that most students will be boosted by the time everyone is back on campus.
+To match the high numbers of cases on the week ending January 7, 2022, we assume 14\% of the students present on campus are infected as of January 7, 2022 (roughly 100/700).
+To account for students who were infected with the Omicron variant during the fall semester, we assume that 100 students are already immune to Omicron infection at the beginning of the spring semester---this roughly corresponds to the number of PU cases that were reported in December.
+Finally, we take the best matching $\theta$ value for the previous semester and scale it by the number of undergraduate students relative to the entire population (therefore using $\theta = 0.15 \times 5000/13000$ throughout).
 
-\begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../figure_omicron/figure_omicron_sensitivity.pdf}
+In the absence of changes in testing frequency or an increased reproduction number, the model predicts the number of cases among undergraduate students to continue to decrease over time (\fref{omicron1}).
+Changes in testing frequency alone have negligible impact on the overall dynamics; when the baseline contact reproduction number $\Rc$ is sufficiently high ($\Rc = 6$), changing testing frequency from biweekly to weekly causes the weekly case numbers to stay at a constant level (instead of decreasing).
+Additional increases in the reproduction number (alongside the changes in testing frequency) can cause the case numbers to further increase, but we are unable to match the observed dynamics even with a 100\% increase in the reproduction number.
+Indeed, a $>10$-fold increase in the numbers of cases between the weeks ending February 11 and 18, 2022, will require an unrealistically high increase in the reproduction number to explain.
+These simulations indicate that changes in distancing and testing policies and the increased transmissibility of the BA.2 subvariant alone are unlikely to be the direct causes of the outbreak.
+
+\begin{figure}[!thp]
+\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_omicron_spring.pdf}
 \caption{
-\textbf{Sensitivity of the spread of Omicron on campus to underlying epidemiological and intervention conditions.}
-We perform sensitivity analysis for potential Omicron outbreaks on campus by varying contact reproduction number $\Rc$ (A--B), asymptomatic testing frequency (C--D), vaccination strategy (E--F), community contacts (G--H), prevailing immunity (I--J), and isolation strategy (K--L).
-(A, C, E, G, I, K) Predicted numbers of Omicron infections per week on campus. 
-Colored lines represent median predictions.
+\textbf{The impact of changes in testing frequency and an increased reproduction number on the spread of the Omicron variant.}
+Solid lines represent median predictions.
 Shaded areas represent 90\% quantiles across 100 simulations.
-We vary one parameter at a time while holding all other parameters constant at their nominal values.
-We present representative simulations in these panels; all other simulations can be found in Supplementary Figures S11--S14.
-(B, D, F, H, J, L) Predicted numbers of total Omicron infections on campus.
-All other parameters are the same as in Figure 2.
-\label{fig:sensitivity}
+Points represent the observed data.
+Vertical dashed lines represent the week including February 8, 2022, when distancing and testing policies were updated on PU campus.
 }
+\label{fig:omicron1}
 \end{figure}
 
-Intervention measures, such as reducing contact rates $\Rc$ (\fref{sensitivity}A, B), increasing testing frequencies (\fref{sensitivity}C, D, Supplementary Figure S11), and vaccinating students (\fref{sensitivity}E, F, Supplementary Figure S12), can help reduce the size of the epidemic peak and the final size of an epidemic but have limited impact.
-While social distancing and mass asymptomatic testing can be effective in preventing onward transmission, they cannot prevent community transmission.
-Vaccines can help reduce community transmission by reducing susceptibility, but current vaccines can still allow reinfections with the Omicron variant \citep{ferguson2021report}.
-Nonetheless, we note that vaccinating students throughout the semester is just as important as vaccinating them before the beginning of the semester---for example, providing 60 booster shots per day with 40\% coverage at the beginning of the semester yield similar outcomes as providing 30 booster shots per day with 60\% coverage at the beginning of the semester despite 20\% differences in the initial vaccine coverage (\fref{sensitivity}F).
-Faster deliveries of vaccines throughout the semester can also help outbreaks to decay faster (Supplementary Figure S12).
+Instead, we consider the role of super-spreading events in driving a large Omicron outbreak by simulating 100--300 infections happening on the same day (February 12, 2022, the weekend following the policy change).
+We still include changes in testing to reflect realistic settings on campus but do not model the increase in the reproduction number to test the sole effects of super-spreading events.
+In contrast to previous simulations (\fref{omicron1}), which showed persistent growth in cases following the increase in the reproduction number, an epidemic driven by a super-spreading event plateaus and decays quickly (\fref{omicron2}).
+In this case, moderate values of baseline reproduction numbers permit a small amount of onward transmission, which can sustain the epidemic for a few weeks, but the reproduction number is not high enough to cause the epidemic to keep growing.
+Overall, the observed patterns in cases are more consistent with the epidemic dynamics driven by super-spreading events.
 
-Instead, epidemic dynamics are more sensitive to underlying epidemiological factors that cannot be controlled by intervention measures on campus, such as levels of community transmission (\fref{sensitivity}G, H, Supplementary Figure S13) and prevailing immunity (\fref{sensitivity}I, J.
-In particular, the epidemic on campus dies out much quicker when community cases are also decreasing---such rapid decrease in the number of infections was not achievable with intervention measures that we considered when community infections are assumed to be constant.
-The proportion of students who have already had Omicron infections before the beginning of semester also play an important role in determining the size of an outbreak (\fref{sensitivity}I, J)---we note that this result depends on our assumption that natural infections provide stronger immunity than vaccinal immunity.
-
-\section*{Isolation}
-
-So far we have assumed that all individuals who test positive are isolated for 10 days;
-in practice, however, isolation of infected individuals can be limited by the isolation bed capacity.
-Here, we explore how limiting the number of isolation beds on university campus can affect epidemic dynamics by assuming that individuals who test positive will not be isolated and will continue to transmit when isolation beds are full.
-In doing so, we also consider the effects of shorter isolation periods (5, 7, and 10 days).
-
-When there are no limits to the number of isolation beds on campus, shortening isolation periods results in a slightly higher number of total infections due to residual transmission after the isolation periods are over (\fref{sensitivity}L).
-The impact on overall epidemic dynamics under our nominal parameter assumptions (\fref{sensitivity}K), but the number of students in isolation is greatly reduced with shorter isolation periods (Supplementary Figure S14).
-When the number of isolation beds is moderately limited ($=100$), shortening isolation periods can decrease the total numbers of infections (compare 10 vs 7 day in \fref{sensitivity}L and Supplementary Figure S14), but shortening isolation periods too much can increase the total numbers of infections again (compare 7 vs 5 day in \fref{sensitivity}L and Supplementary Figure S14).
-When the number of isolation beds is severely limited ($=50$), the benefits of shortening isolation periods can further outweigh the risk of transmitting after isolation by allowing a greater proportion of infected individuals to be isolated.
-
-\section*{Changes in asymptomatic testing frequency}
-
-Finally, as the number of COVID-19 cases is decreasing in the US, many institutions, including PU, may be considering reducing asymptomatic testing frequencies.
-Here, we test whether switching testing frequencies from twice a week to once a week in the middle of a semester will result in larger outbreaks across a wide range of contact $\Rc$ and immunity levels (i.e., proportion of students who have already had Omicron infections before the beginning of semester).
-To present a challenging scenario, we assume that the number of infectious community contacts decrease exponentially at rate $r=-0.1/\textrm{day}$ until March 1, 2022, when the testing protocol (hypothetically) changes, and allow them to increase again exponentially at rate $r=0.1/\textrm{day}$.
-
-\begin{figure}[!th]
-\includegraphics[width=\textwidth]{../figure_omicron/figure_omicron_switch.pdf}
+\begin{figure}[!thp]
+\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_omicron_outbreak.pdf}
 \caption{
-\textbf{Effects of reducing testing frequencies in the middle of a semester.}
+\textbf{The impact of large super spreading events on the spread of the Omicron variant.}
 Solid lines represent median predictions.
 Shaded areas represent 90\% quantiles across 100 simulations.
-Vertical dashed lines represent March 1, 2022, when testing protocol changes and infectious community contacts begin increasing exponentially.
-All other parameters are set to previously assumed values in \fref{sensitivity}.
-\label{fig:switch}
+Points represent the observed data.
+Vertical dashed lines represent the week including February 8, 2022, when distancing and testing policies were updated on PU campus.
 }
+\label{fig:omicron2}
 \end{figure}
-
-The impact of changing testing frequencies depends on the relative amount of community transmission and campus contagion.
-When contact levels are low or a large fraction of students have already had Omicron infections, switching from twice-a-week to once-a-week testing has little impact on epidemic dynamics because most infections are caused by community contacts \fref{switch}.
-Increasing contact levels or decreasing immunity levels permits a greater amount of campus transmission, causing greater discrepancies between two scenarios, when testing frequencies are held constant and when testing frequencies are allowed to change;
-in these cases, the exact effects of decreasing testing frequencies are still sensitive to the underlying immunity levels.
 
@@
 Our analysis demonstrates strong spatiotemporal correlations between the patterns of spread of SARS-CoV-2 on campus and those from surrounding communities.
-These correlations decreased with distance from Mercer County in fall 2021, likely reflecting contact and commuting patterns as university reopened.
-Mathematical modeling further suggests limited transmission between the university population during fall and spring semesters of 2020 and an increased amount of community contacts during the fall semester of 2021 compared previous semesters.
+These correlations decreased with distance from Mercer County in fall 2021--2022, likely reflecting contact and commuting patterns as the university campus reopened.
+Mathematical modeling further suggests limited transmission between the university population during fall and spring semesters of the 2020--2021 academic year and an increased amount of community contacts during the fall semester of the 2021--2022 academic year compared previous semesters.
 An increase in the number of cases by the end of November 2021 is consistent with the increase in the levels of community cases and waning immunity.
-Finally, predicting future SARS-CoV-2 dynamics and changes in policy (e.g., decreasing testing frequency) will be sensitive to other intervention measures present on campus and underlying epidemiological conditions, such as community transmission and immunity levels.
+Finally, our analysis highlights the role of super-spreading events in driving the spread of the Omicron variant on PU campus.
 
@@
 
-The recent recommendations by the Centers for Disease Control and Prevention to reduce isolation periods from 10 to 5 days \citep{cdcisolation} have generated much discussion regarding its individual- and population-level implications \citep{soljak2022reducing}.
-Given uncertainty in the duration of infectiousness of the Omicron variant, shortening the duration of isolation can increase the chances of transmission after infected individuals are released from their isolation---as we also show here, this can lead to an increase in the final size of an outbreak.
-When the number of isolation beds are limited, however, shorter isolation periods can help mitigate spread by allowing a greater number of individuals to be isolated.
-We note that the net benefits of shortening isolation periods are sensitive to many factors, such as the duration of infectiousness, isolation bed capacity, and time from infection to isolation, which will vary across different universities.
-Therefore, caution is needed in implementing these changes in other places.
-Incorporating lateral flow tests at the end of isolation can also help prevent excess transmission \citep{quilty2022test}.
-
 There are several limitations to our analysis.
@@
 intervention measures on campus (e.g., frequent asymptomatic testing, contact tracing, and virtual classes during fall and spring semesters of 2020) likely limited onward transmission on campus.
+In addition, even during periods of large Omicron outbreaks on campus in early 2022, the number of COVID-19 cases in Mercer County remained low, implying limited transmission from campus to community.
 Decreasing patterns in epidemic correlations with distance further highlight the role of spatial spread in driving dynamics of SARS-CoV-2---such patterns are consistent with spatial spread of many other respiratory pathogens \citep{grenfell2001travelling, viboud2006synchrony, baker2019epidemic}.
@@
 Instead, the expectations for reopening campuses (e.g., whether to minimize infections on campuses) need to be re-evaluated as we transition to next phases of the pandemic, accounting not only for changes in epidemic dynamics but also for our perception of the pathogen.
-The answers to these questions will ultimately lie in the landscape of SARS-CoV-2 immunity \citep{saad2020immune,baker2021limits}.
+The answers to these questions ultimately depend on the landscape of SARS-CoV-2 immunity \citep{saad2020immune,baker2021limits}.
 
@@
 \begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_simulation_fall_2020_all.pdf}
+\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_fall_2020_all.pdf}
 \caption{
@@
 \begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_simulation_spring_2020_all.pdf}
+\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_spring_2020_all.pdf}
 \caption{
@@
 \begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_simulation_fall_2021_all.pdf}
+\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_fall_2021_all.pdf}
 \caption{
@@
 
-\pagebreak
-
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_omicron/figure_omicron_sensitivity_testfreq.pdf}
-\caption{
-\textbf{Sensitivity of the spread of Omicron on campus to asymptomatic testing frequencies.}
-Colored lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
-See figure 3 in the main text for details.
-}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_omicron/figure_omicron_sensitivity_boost_daily.pdf}
-\caption{
-\textbf{Sensitivity of the spread of Omicron on campus to vaccination strategies.}
-Colored lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
-See figure 3 in the main text for details.
-}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_omicron/figure_omicron_sensitivity_thetabase.pdf}
-\caption{
-\textbf{Sensitivity of the spread of Omicron on campus to community transmission.}
-Colored lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
-See figure 3 in the main text for details.
-}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_omicron/figure_omicron_sensitivity_isolation.pdf}
-\caption{
-\textbf{Sensitivity of the spread of Omicron on campus to isolation strategies.}
-Colored lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
-See figure 3 in the main text for details.
-}
-\end{figure}
-
 \end{document}

### 2022-04-18 doc/university-covid.tex
done?

@@
 \\
+Sang Woo Park\textsuperscript{1}
+Irini Daskalaki\textsuperscript{?}
+Robin M. Izzo\textsuperscript{?}
+C. Jessica E. Metcalf\textsuperscript{1}
+Daniel A. Notterman\textsuperscript{?}
+Bryan T.\ Grenfell\textsuperscript{1}
+\\
+\bigskip
+\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
+\\
+\bigskip
+
+Corresponding author: ???
+\bigskip
 \end{flushleft}
@@
 
-Mathematical models have played a crucial role in guiding pandemic responses; university campuses provide a particularly well documented case for institutional outbreaks.
+Mathematical models have played a crucial role in guiding pandemic responses, espeicially among higher education institutions.
+University campuses further present a particularly well-documented case for institutional outbreaks, thereby providing a unique opportunity to understand detailed patterns of pathogen spread.
 Here, we present descriptive and modeling analyses of SARS-CoV-2 outbreaks on the Princeton University campus---this model was used throughout the pandemic to inform policy decisions for the Princeton University campus.
-We demonstrate strong spatiotemporal correlations in epidemic patterns between Princeton University and surrounding communities.
-These findings are corroborated by our model predictions, which indicate that the amount of on-campus transmission was likely limited during much of the wider pandemic.
+We demonstrate strong spatiotemporal correlations in epidemic patterns between the Princeton University campus and surrounding communities.
+These findings are corroborated by our model predictions, which indicate that the amount of on-campus transmission was likely limited during much of the wider pandemic until the end of 2021.
+Finally, we find that a super-spreading event played a major role in driving the recent outbreak on the Princeton University campus during the spring semester of the 2021--2022 academic year.
+Despite large numbers of cases on campus, case levels in surrounding communities remained low, suggesting that there was little to no spillover transmission from campus to the local community.
 
@@
 Asymptomatic individuals submitted self-collected saliva samples, from which the presence of SARS-CoV-2 was tested using Reverse Transcription Polymerase Chain Reaction (RT-PCR).  
-Asymptomatic individuals who tested positive were required to isolate for 10 days.
-Symptomatic individuals were eligible for rapid PCR tests and had to isolate \swp{Robin and Irini: is self-isolate a better term?} until they received their test results---those who tested positive were required to isolate for at least 10 days after symptom onset and were release 3 days after symptoms resolved.
+Those who tested positive were required to isolate for at least 10 days after symptom onset or test date (whichever was longer) and were released when they had been at least 48 hours with improving or resolving symptoms as per New Jersey Department of Health guidance.
 PCR positives were exempt from asymptomatic testing for 90 days.
 Since March 7, 2022, asymptomatic testing frequencies decreased to once a month from once a week for individuals whose vaccine statuses are up-to-date, which in turn likely reduced the accuracy of surveillance; therefore, we chose to focus on epidemic patterns before this change was implemented.
-Throughout the study period, contact tracing was also performed for positive cases to quarantine their close contacts, defined as individuals who have been within 6 feet for at least 15 minutes, cumulatively within a 24-hour period.
+Throughout the study period, contact tracing was also performed for positive cases to alert their close contacts to either quarantine or test more frequently according to the close contacts' vaccination status, and to gather data that could help uncover clusters of transmission or superspreader events.
 All data used in this analysis are publicly available on the PU COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.
@@
 In the beginning of the spring semester of the 2020--2021 academic year, the number of cases suddenly increased before classes started (\fref{princeton}B), reflecting $\approx 3000$ undergraduate students returning to campus.
-Returning students were required to be tested and quarantine \swp{Robin and Irini: for how long} if they were coming back from CDC-designated states and countries \swp{ask Irini and Robin for details}.
+Returning students were required to be tested and quarantine for 7 days regardless of their returning location.
 All classes remained virtual, and the testing protocol did not change (twice a week for undergraduate and graduate students, and once a week for faculty and staff members).
@@
 
-For the fall semester of the 2021--2022 academic year, all students and faculty and staff members were required to be vaccinated \swp{with two doses?
-need to ask Robin and Irini} with a very few medical and religious exemptions.
+For the fall semester of the 2021--2022 academic year, all students and faculty and staff members were required to be vaccinated with a very few medical and religious exemptions.
 By the beginning of the semesters, ??\% of students and ??\% of faculty and staff members were vaccinated \swp{ask Robin and Irini}.
@@
 Transmission processes are modeled by first setting the contact reproduction number $\Rc$, which we define as the average number of infectious contacts an infected individual would make throughout the course of their infection;
-here, infectious contacts refer to contacts that would result in infection when the contactee (i.e., the contacted individual) is susceptible to infection.
+here, infectious contacts refer to contacts that would result in infection when the contacted individual is susceptible to infection.
 The contact reproduction number implicitly accounts for all intervention measures that we do not model explicitly, such as social distancing and contact tracing---in other words, this contact reproduction number does not account for asymptomatic testing or vaccination, which are modeled separately.
@@
 in other words, two and three doses of vaccines reduce transmissibility by 2.2\% and 15.6\%, respectively.
-The immunity from the third dose is assumed to take 7 days to develop and wane at the same rate as before (in this case, 70\% to 39\% in 20 weeks).
+The immunity from the third dose is assumed to take 7 days to develop \citep{moreira2022safety} and wane at the same rate as before (in this case, 70\% to 39\% in 20 weeks).
 Finally, the isolation period is reduced to 5 days.
@@
 
-\begin{figure}[!thp]
+\begin{figure}[!ht]
 \includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_omicron_spring.pdf}
@@
 Our analysis suggest that comparing the ratios between the cases on university campuses and neighboring communities can also provide a useful measure for how well a university campus is controlling the epidemic;
-however, we note that this ratio needs to be interpreted with caution as it is sensitive to changes in testing patterns.
+however, we note that this ratio needs to be interpreted with caution as it is sensitive to changes in testing patterns as well as the numbers of students on campus.
+For example, the ratios of cases can suddenly change during holidays when students are away from campus.
 Future studies may combine phylogenetic data to better understand spatial patterns of SARS-CoV-2 on campus.
@@
 \begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_correlation2.pdf}
+\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_correlation2.pdf}
 \caption{
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth]{../figure_princeton/figure_princeton_map.pdf}
+\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_map_new.pdf}
 \end{center}
@@
 \begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_newyork.pdf}
+\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_newyork_new.pdf}
 \caption{
@@
 \begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_phil.pdf}
+\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_philly_new.pdf}
 \caption{
@@
 \begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_map_extended.pdf}
+\includegraphics[width=0.8\textwidth]{../figure_princeton_new/figure_princeton_map_extended.pdf}
 \caption{
@@
 \begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_simulation_nowaning.pdf}
+\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_nowaning.pdf}
 \caption{

### 2022-06-15 doc/university-covid.tex
done

@@
 
-Mathematical models have played a crucial role in guiding pandemic responses, espeicially among higher education institutions.
-University campuses further present a particularly well-documented case for institutional outbreaks, thereby providing a unique opportunity to understand detailed patterns of pathogen spread.
-Here, we present descriptive and modeling analyses of SARS-CoV-2 outbreaks on the Princeton University campus---this model was used throughout the pandemic to inform policy decisions for the Princeton University campus.
-We demonstrate strong spatiotemporal correlations in epidemic patterns between the Princeton University campus and surrounding communities.
+Mathematical models have played a crucial role in guiding pandemic responses. 
+University campuses present a particularly well-documented case for institutional outbreaks, thereby providing a unique opportunity to understand detailed patterns of pathogen spread.
+Here, we present descriptive and modeling analyses of SARS-CoV-2 transmission on the Princeton University campus---this model was used throughout the pandemic to inform policy decisions and operational guidelines for the university campus.
+We demonstrate strong spatiotemporal correlations in epidemic patterns between the university campus and surrounding communities.
 These findings are corroborated by our model predictions, which indicate that the amount of on-campus transmission was likely limited during much of the wider pandemic until the end of 2021.
-Finally, we find that a super-spreading event played a major role in driving the recent outbreak on the Princeton University campus during the spring semester of the 2021--2022 academic year.
-Despite large numbers of cases on campus, case levels in surrounding communities remained low, suggesting that there was little to no spillover transmission from campus to the local community.
+Finally, we find that a super-spreading event likely played a major role in driving the recent outbreak on the Princeton University campus during the spring semester of a 2021--2022 academic year.
+Despite large numbers of cases on campus, case levels in surrounding communities remained low, suggesting that there was little spillover transmission from campus to the local community.
 
@@
 Rapid, asymptomatic transmission of SARS-CoV-2 has hindered intervention efforts, such as contact tracing \citep{hellewell2020feasibility}.
-Social distancing measures have played major roles in preventing transmission but can be difficult to maintain for a prolonged period \citep{galanti2021social}.
-The development of vaccines has provided a safe means of reopening society, but uncertainty remains in their long-term effectiveness in protecting infection and transmission, especially in the face of new viral variants.
+Social distancing measures have played major roles in preventing transmission, but can be difficult to maintain for a prolonged period \citep{galanti2021social}.
+The development of vaccines has provided a safe means of reopening society, but uncertainty remains in their long-term effectiveness in preventing infection and transmission, especially in the face of new viral variants.
 
-Mathematical models have played a central role in guiding these pandemic responses and devising control strategies \citep{metcalf2020mathematical}.
+Mathematical models have played a significant role in guiding these pandemic responses and devising control strategies \citep{cobey2020modeling,holmdahl2020wrong,metcalf2020mathematical,koelle2022changing}.
 Models can help monitor key parameters that govern epidemic dynamics \citep{kraemer2021monitoring} and retrospectively estimate the impact of intervention measures in reducing transmission \citep{flaxman2020estimating}.
@@
 Coupling modeling efforts with real-life implementations in university campuses further provided unique opportunities to directly test model-based predictions of intervention effects in preventing the transmission of SARS-CoV-2 \citep{frazier2022modeling}---
-each university campus offers a relatively well-controlled epidemic setting with a behaviorally homogeneous population (especially among undergraduate students).
-Mass asymptomatic testing also provides near-perfect ascertainment for epidemic sizes, allowing for accurate understanding of epidemic patterns.
+each university campus offers a relatively well-controlled epidemic setting with a relatively homogeneously behaving population (especially among undergraduate students).
+Campuses can also offer strong opportunities for control by non-pharmaceutical interventions, such as isolation and mask-wearing;
+mass asymptomatic testing further provides robust ascertainment for epidemic sizes, allowing for accurate understanding of epidemic patterns.
 
-University campuses present unique challenges to controlling an outbreak.
-High-density environments, especially within dorms, and a large fraction of asymptomatic infections (due to the young age of university students) can readily permit rapid transmission.
+On the other hand, university campuses also present unique challenges to controlling an outbreak.
+A large fraction of asymptomatic infections (due to the young age of university students) and high-density interactions---such as eating in large dining halls and various social activities---can readily permit rapid transmission.
 These kinds of contacts are inherently difficult to keep track of, making contact tracing less effective.
-The impact of intervention measures is also expected to vary across different university campuses, reflecting heterogeneity in campus settings such as compliance, resources, community prevalence, as well as effects of other interventions present on campuses.
+The impact of intervention measures is expected to vary across different university campuses, reflecting heterogeneity in campus settings such as compliance, resources, community prevalence, as well as effects of other interventions present on campuses.
 For example, Duke and Harvard Universities experienced moderate outbreaks at the beginning of the fall semester in 2021 when in-person classes were allowed, despite high vaccination rates and weekly asymptomatic testing protocols \citep{dukeoutbreak,harvardoutbreak}, whereas the number of cases remained low in Princeton University (PU) during the same time period with similar levels of testing and vaccination.
-Here, we focused on the spread of SARS-CoV-2 on PU campus alone to eliminate heterogeneities inherent to such comparisons.
+Here, we focused on the dynamics of SARS-CoV-2 on the PU campus alone to eliminate heterogeneities inherent to such comparisons.
 
-In this study, we analyzed the spread of SARS-CoV-2 on PU campus (\fref{princeton}).
-We first begin with a descriptive analysis of the outbreak and then present modeling analyses of the past outbreaks and future scenarios.
+We analyzed the spread of SARS-CoV-2 on campus (\fref{princeton}).
+We begin with a descriptive analysis of the outbreak and then present modeling analyses of the various outbreaks during 2020--2022.
 PU is located in Mercer County, New Jersey, USA;
-the population comprises of 5267 undergraduate students, 2946 graduate students, and 7000 faculty and staff members.
+the population comprises of 5267 undergraduate students, 2946 graduate students, and around 7000 faculty and staff members.
 For simplicity, we divided the epidemic into four time periods representing four semesters across two academic years: Fall 2020--2021 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020--2021 (January 16, 2021--May 14, 2021; \fref{princeton}B), Fall 2021--2022 (August 14, 2021--December 31, 2021; \fref{princeton}C), and Spring 2021--2022 (January 1, 2022--March 18, 2022; \fref{princeton}D).
@@
 PCR positives were exempt from asymptomatic testing for 90 days.
-Since March 7, 2022, asymptomatic testing frequencies decreased to once a month from once a week for individuals whose vaccine statuses are up-to-date, which in turn likely reduced the accuracy of surveillance; therefore, we chose to focus on epidemic patterns before this change was implemented.
+Since March 7, 2022, asymptomatic testing frequencies decreased to once a month from once a week for individuals whose vaccine status is up-to-date.
+This in turn likely reduced the accuracy of surveillance; therefore, we chose to focus on epidemic patterns before this change was implemented.
 Throughout the study period, contact tracing was also performed for positive cases to alert their close contacts to either quarantine or test more frequently according to the close contacts' vaccination status, and to gather data that could help uncover clusters of transmission or superspreader events.
+Changes in testing frequency and other intervention measures throughout the study period reflected various factors, including the impact of COVID-19 cases on continuity of operations or continuity of teaching; on severity of disease on campus; the capacity of testing and the healthcare system; and hospitalization rates on campus and in the area.
 All data used in this analysis are publicly available on the PU COVID-19 Dashboard website: \url{https://covid.princeton.edu/dashboard}.
@@
 During the fall semester of the 2020--2021 academic year, roughly 1000 grad students and 2000 faculty and staff members were present on campus and participated in asymptomatic testing. 
-Undergraduate students were not allowed to return to campus as all classes were held virtually, and so their numbers were minimal on campus ($<300$).  
+All classes were held virtually, and so only a few undergraduate students remained on campus ($<300$).  
 Both undergraduate and graduate students were required to get tested twice a week, whereas faculty and staff members were required to get tested once a week.
@@
 Returning students were required to be tested and quarantine for 7 days regardless of their returning location.
-All classes remained virtual, and the testing protocol did not change (twice a week for undergraduate and graduate students, and once a week for faculty and staff members).
-The number of cases persisted at similar levels as the fall semester and eventually decreased as classes ended and students went back home---the decrease in the number of cases in PU also coincided with the decrease in the number of cases in Mercer County.
+Most classes remained virtual, and the testing protocol did not change (twice a week for undergraduate and graduate students, and once a week for faculty and staff members).
+Some smaller classes were held in-person, but required social distancing (thereby limiting the size of the class) and mask-wearing at all times.
+The number of cases persisted at similar levels to the fall semester and eventually decreased as classes ended and students went home---the decrease in the number of cases in PU also coincided with the decrease in the number of cases in Mercer County.
 The highest number of cases was reported among faculty and staff members ($=111$), followed by undergraduate students ($=101$), and graduate students ($=29$).
 
-For the fall semester of the 2021--2022 academic year, all students and faculty and staff members were required to be vaccinated with a very few medical and religious exemptions.
-By the beginning of the semesters, ??\% of students and ??\% of faculty and staff members were vaccinated \swp{ask Robin and Irini}.
-% As of January 11th, 99\% of undergraduate students, 98\% of graduate students, and 96\% of faculty and staff members were vaccinated.
+For the fall semester of the 2021--2022 academic year, all students and faculty and staff members were required to be vaccinated, with very few medical and religious exemptions.
+By the beginning of the semesters, 97\% of undergraduate students, 96\% of graduate students, and 94\% of faculty and staff members were vaccinated.
 Vaccinees were required to be tested once a week, while unvaccinated individuals were required to be tested twice a week.
-In-person classes and social events fully resumed on campus while all individuals were required to wear masks indoors with a few exceptions (e.g., when eating or drinking, or when teaching a small class).
-The number of cases remained similar to previous semesters until November when the number of cases began to increase primarily among undergraduate students around Thanksgiving (\fref{princeton}C).  
+In-person classes and social events fully resumed on campus, though all individuals were required to wear masks indoors with a few exceptions (e.g., when eating or drinking, or when teaching a small class).
+The number of cases remained similar to previous semesters until November when cases began to increase, primarily among undergraduate students around Thanksgiving (\fref{princeton}C).  
 In order to prevent transmission, testing frequency was increased to twice a week for undergraduate students on November 27th, 2021; the size of non-academic gatherings were also limited to 20 people.
-The number of cases decreased slightly as classes ended but soon increased back up as the Omicron (BA.1) variant began to spread on campus and in Mercer county.
+The number of cases decreased slightly as classes ended but soon increased again as the Omicron (BA.1) variant began to spread on campus and in Mercer county.
 
 For the spring semester of the 2021--2022 academic year, all eligible students and faculty and staff members were required to be boosted.
-By the beginning of the semesters, ??\% of students and ??\% of faculty and staff members were boosted \swp{ask Robin and Irini}.
+By the beginning of the semesters, 65\% of undergraduate students, 71\% of graduate students, and 82\% of faculty and staff members were boosted.
 Undergraduate students were still required to be tested twice a week to prevent the additional spread of the Omicron variant.
 The number of cases remained high before classes began but decreased over time, following epidemic patterns in Mercer County (\fref{princeton}D).
-As the campus and local cases numbers decreased, the gathering policy was updated to allow food in events were no longer limited to 20 people;
-in addition, the testing frequency was reduced a once a week.
-Following the policy change, a large outbreak occurred on campus with high case numbers persisting until the spring break.
+Coinciding with the decrease in the campus and local cases numbers, the gathering policy was updated to allow food in events were no longer limited to 20 people;
+in addition, the testing frequency was reduced to once a week.
+Following the policy change, a large outbreak occurred on campus with high case numbers persisting until Spring Break (March 5th, 2022).
+The timing of this outbreak also coincided with a rapid turnover of the Omicron subvariant BA.2---the proportion of BA.2 subvariant reached 93.5\% (372/398) compared to 26.9\% (14/52) from the previous week.
 On March 7, 2022, mask mandates were lifted and testing frequency was reduced to once a month.
-The timing of this outbreak also coincided with a rapid turnover of the Omicron subvariant BA.2---the proportion of BA.2 subvariant reached ??\% compared to ??\% from the previous week \swp{Ask Dan}.
+
+\subsection{Comparison with community transmission}
 
@@
 fall 2020--2021 ($\rho = 0.81$ (95\% CI: 0.55--0.92; $p < 0.001$), \fref{princeton}E); spring 2020--2021 ($\rho = 0.80$ (95\% CI: 0.51--0.92; $p < 0.001$), \fref{princeton}F); and fall 2021--2022 ($\rho = 0.86$ (95\% CI: 0.67--0.94; $p < 0.001$), \fref{princeton}G).
-These correlations are robust even when we stratify cases by the population, except for undergraduate students during Fall 2020 when most of them were not physically present on campus (Supplementary Figure S1).
-However, the case patterns in PU were decoupled from those in Mercer county for the spring semester of the 2021--2022 academic year with unclear correlations (\fref{princeton}H): $\rho = 0.32$ (95\% CI: -0.34--0.77; $p = 0.33$).
-Stratifying cases by the population shows that case patterns in graduate students and faculty and staff members were still strongly correlated with case patterns in Mercer county, meaning that the campus transmission was limited to undergraduate students (Supplementary Figure S1).
+These correlations are robust even when we stratify cases by the population, except for undergraduate students during Fall 2020, when most were not physically present on campus (Supplementary Figure S1).
+However, the case patterns in PU were decoupled from those in Mercer county for the spring semester of the 2021--2022 academic year with unclear correlations (\fref{princeton}H): $\rho = 0.32$ (95\% CI: $-0.34$--0.77; $p = 0.33$).
+Stratifying cases by subpopulation shows that case patterns in graduate students and faculty and staff members were still strongly correlated with case patterns in Mercer county, meaning that the campus transmission was limited to undergraduate students (Supplementary Figure S1).
 We also find strong correlations between the weekly logged numbers of cases from PU and those from other counties in New Jersey (Supplementary Figure S2)---these correlations significantly decreased with distance from Mercer county in both spring 2020--2021 ($\rho=-0.48$ (95\% CI: $-0.75$--$-0.06$; $p = 0.03$) for all cases and $\rho=-0.51$ (95\% CI: $-0.77$--$-0.10$; $p = 0.02$) for faculty and staff cases) and fall 2021--2022 ($\rho=-0.68$ (95\% CI: $-0.86$--$-0.35$, $p < 0.001$) for all cases and $\rho=-0.74$ (95\% CI: $-0.89$--$-0.46$, $p < 0.001$) for faculty and staff cases).
@@
 Finally, mass testing allows us to infer the ratio between the weekly numbers of cases from Princeton and those from Mercer county---
-we expect this ratio to remain constant over time if the majority of infections on campus is caused by community transmission provided that testing patterns remain constant in both places.
+we expect this ratio to remain constant over time if the majority of infections on campus is caused by community transmission, provided that testing patterns remain roughly constant in both places.
 We find that the ratio between the weekly numbers of cases from Princeton and those from Mercer county stayed between 1:100 and 1:25 until the end of the fall semester of the 2021--2022 academic year (\fref{princeton}I--K).
@@
 Here, we present a generic and parsimonious version that encompasses sufficient details to characterize the overall spread of SARS-CoV-2 in PU without an over-proliferation of parameters.
-The model consists of four main components that are simulated on a daily time scale: (1) infection and transmission dynamics, (2) sampling and testing protocols, (3) isolation protocols, and (4) vaccination dynamics, including waning immunity and booster shots.
-Previous versions of the model included contact tracing, but we exclude it in this study for simplicity.
+The model consists of four main components simulated on a daily time scale: (1) infection and transmission dynamics, (2) sampling and testing protocols, (3) isolation protocols, and (4) vaccination dynamics, including waning immunity and booster shots.
+Previous versions of the model included contact tracing, but we exclude it in this model for simplicity.
 
@@
 On each day, all infected individuals who have not yet been isolated then make infectious contacts at random to anyone on campus; the number of infectious contacts are drawn from a negative binomial distribution with a mean of either $\beta_p$ or $\beta_s$ and an overdispersion parameter of $k=0.1$ to account for the possibility of super-spreading events \citep{10.12688/wellcomeopenres.15842.3}.
-Infectious contacts from local community can be also made at random to anyone on campus; these contacts are modeled using a Poisson distribution with a time-varying mean, which is calculated by scaling the daily number of cases by $\theta$ and shifting it by 1 week to account for reporting delays.
+We also rely on cases from Mercer County to crudely capture community dynamics.
+In particular, we assume that infectious contacts from local or regional community can be made at random to anyone on campus; these contacts are modeled using a Poisson distribution with a time-varying mean, which is calculated by scaling the daily number of cases by $\theta$ and shifting it by 1 week to account for reporting delays.
 Infectious contacts, whether made by individuals on campus or from outside, result in infection only when the contacted individuals are susceptible; when the contacted individuals are vaccinated, and therefore partially susceptible to infection, they have a reduced probability of infection corresponding to their susceptibility (discussed later).
 
-All individuals on campus are assumed follow a pre-determined asymptomatic testing plan at a fixed frequency---
+All individuals on campus are assumed to follow a pre-determined asymptomatic testing plan at a fixed frequency---
 for example, under weekly testing, one individual can get sampled on days 1, 8, 15, and so forth, while another individual get sampled on days 2, 9, 16, and so forth.
@@
 
-As most students as well as faculty and staff members had received two doses of vaccination in the beginning of fall 2021, we do not distinguish the first and second doses.
-Instead, we assume that all vaccinated individuals have 90\% reduced susceptibility and 20\% reduced transmissibility at the beginning of the semester---these assumptions are consistent with recent estimates by \citep{prunas2022vaccination} that vaccination with BNT162b2 reduces susceptibility by 89.4\% (95\% CI: 88.7\%--90.0\%) and infectiousness by 23.0\% (95\% CI: $-11.3\%$--46.7\%) against the Delta strain.
-Based on \citep{tartof2021effectiveness}, vaccine efficacy against susceptibility is allowed to exponentially wane from 90\% to 50\% in 20 weeks (and continues to wane at the same rate) for each vaccinated individual;
+As most students, as well as faculty and staff members, had received two doses of vaccination in the beginning of fall 2021, we do not distinguish the first and second doses.
+Instead, we assume that all vaccinated individuals have 90\% reduced susceptibility and 20\% reduced transmissibility at the beginning of the semester---these assumptions are consistent with recent estimates by \cite{prunas2022vaccination} that vaccination with BNT162b2 reduces susceptibility by 89.4\% (95\% CI: 88.7\%--90.0\%) and infectiousness by 23.0\% (95\% CI: $-11.3\%$--46.7\%) against the Delta strain.
+Based on \cite{tartof2021effectiveness}, vaccine efficacy against susceptibility is allowed to exponentially wane from 90\% to 50\% in 20 weeks (and continues to wane at the same rate) for each vaccinated individual;
 vaccine efficacy against transmissibility is also allowed to wane at the same rate (i.e., from 20\% to 11\% in 20 weeks).
@@
 We further assume that the population mixes homogeneously.
-While these assumptions are most conservative, epidemiological parameters  and mixing patterns likely differ across groups (e.g., undergraduate students are more likely to infect undergraduate students and also remain asymptomatic).
+While these assumptions are most parsimonious, epidemiological parameters and mixing patterns likely differ across groups (e.g., undergraduate students are more likely to infect undergraduate students and also remain asymptomatic).
 Therefore, our model parameters describe average dynamics across different groups and must be interpreted with care.
@@
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{../figure_princeton/figure_princeton_simulation.pdf}
+\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation.pdf}
 \caption{
-\textbf{Retrospective analysis of past SARS-CoV-2 outbreaks on Princeton University (PU).}
+\textbf{Retrospective analysis of past SARS-CoV-2 outbreaks on Princeton University (PU) campus.}
 (A, D, G) Time series comparisons of model predictions with observed data across ranges of contact reproduction number $\Rc$ and scaling parameter for community transmission $\theta$.
@@
 
-For fall 2020--2021, we simulate the model assuming 3000 individuals (1000 graduate students and 2000 faculty and staff members) on campus with 1000 of them participating in asymptomatic testing twice a week.
+For \textbf{fall 2020--2021}, we simulate the model assuming 3000 individuals (1000 graduate students and 2000 faculty and staff members) on campus with 1000 of them participating in asymptomatic testing twice a week.
 We find that a low level of contacts $\Rc=0.5$ and a small amount of community transmission $\theta=0.015$ is most consistent with the observed epidemic dynamics in fall 2020 (\fref{matching}A).
@@
 
-For spring 2020--2021, we simulate the model assuming 8000 individuals (3000 undergraduate students, 2000 graduate students, and 3000 faculty and staff members) on campus with 5000 of them participating in asymptomatic testing twice a week.
+For \textbf{spring 2020--2021}, we simulate the model assuming 8000 individuals (3000 undergraduate students, 2000 graduate students, and 3000 faculty and staff members) on campus with 5000 of them participating in asymptomatic testing twice a week.
 We further assume that 4000 individuals (3000 undergraduate students and 1000 graduate students) returned to campus over 14 days (January 16, 2021--January 29, 2021);
@@
 Simulations also preclude high $\Rc > 2$ again, suggesting that transmission between students were likely limited even though they had returned to campus---the absence of in-person teaching is likely to have contributed to lowering $\Rc$.
-We also find that initial infections are required to match relatively high levels of cases in the beginning of semester (\fref{matching}E). 
+We also find that initial infections (e.g., from returning students) are required to match relatively high levels of cases in the beginning of semester (\fref{matching}E). 
 Once again, the predicted and the observed numbers of cases are positively correlated ($\rho = 0.62$; 95\% CI: 0.20--0.85; \fref{matching}F).
 
-For fall 2021--2022, we assume 13000 individuals are present on campus (5000 undergraduate students, 2000 graduate students, and 6000 staff and faculty members) with 98\% of them vaccinated---here, vaccine-derived immunity is allowed to wane over time to ask whether the increase in the number of cases around November is consistent with the dynamics predicted by immunity waning.
+For \textbf{fall 2021--2022}, we assume 13000 individuals are present on campus (5000 undergraduate students, 2000 graduate students, and 6000 staff and faculty members) with 98\% of them vaccinated---here, vaccine-derived immunity is allowed to wane over time to ask whether the increase in the number of cases around November is consistent with the dynamics predicted by immunity waning.
 Vaccinated individuals are tested every week, whereas unvaccinated individuals are tested every 3 days. 
-We further assume 5000 undergraduate students returned to campus over 16 days (August 14, 2021--August 29, 2021);
-all returning students are assumed to be tested upon return but are not quarantined.
+We further assume 5000 undergraduate students returned to campus over 16 days (August 14, 2021--August 29, 2021).
+All students were required to test upon return and quarantine until they received a negative test result;
+for simplicity, we only model the testing process in our simulation (without quarantine) given a short testing delay.
 Finally, we assume that 0.5\% of both returning and on-campus populations are infected at the beginning of simulation (August 14, 2021).
@@
 
-Even though the numbers of cases during fall 2021 (before a large outbreak) were similar to those during previous semesters, we find that considerably higher levels of community contact $\theta$ ($\approx 10$ fold higher) are required to explain the observed dynamics due to a decreased susceptibility from vaccination (\fref{matching}G).
+Even though the numbers of cases during fall 2021 (before a large outbreak) were similar to those during previous semesters, we find that considerably higher levels of community contact $\theta$ ($\approx 10$ fold higher) are required to explain the observed dynamics due to a decreased susceptibility derived from vaccination (\fref{matching}G).
 We note that the parameter $\theta$ necessarily depends on our assumed vaccine efficacy against susceptibility, and $\theta$ would decrease if we assume a lower vaccine efficacy.
-Nonetheless, the amount of community contact would still need to be higher than previous semesters as long as vaccine provides some protection against infection.
+Nonetheless, the amount of community contact would still need to be higher than previous semesters as long as the vaccine provides some protection against infection and onward transmission.
 
@@
 These simulations suggest that an increase in the number of cases in November can be explained by a combination of waning immunity alone without requiring additional changes in transmission dynamics (note we do not allow $\theta$ or $\Rc$ to vary over time).
-When we exclude immunity waning from the model, predicted epidemic dynamics exhibit slower growth and require even higher values of campus and community contact rates ($\Rc$ and $\theta$) to qualitatively match the observed dynamics (Supplementary Figure S10)---even so, the logged sum of squared differences are generally higher (with median logged sum of squared differences ranging from 6.9 to 41.3 for the same parameter regime).
+When we exclude immune waning from the model, predicted epidemic dynamics exhibit slower growth and require even higher values of campus and community contact rates ($\Rc$ and $\theta$) to qualitatively match the observed dynamics (Supplementary Figure S10)---even so, the logged sum of squared differences are generally higher (with median logged sum of squared differences ranging from 6.9 to 41.3 for the same parameter regime).
 Thus, combining some amount of immune waning and high campus and community contact rates likely best explains the epidemic growth near the end of the semester.
 
-Projecting the model beyond November 26th implies that we would have seen a similarly growth in the number of cases if conditions remained constant even without the introduction of the Omicron variant.
-In other words, the Delta strain would have continued to spread on campus at a similar rate if the semester were to (hypothetically) continue until January without additional interventions (\fref{matching}H).
+Projecting the model beyond November 26th implies that we would have seen a similar growth in the number of cases if conditions remained constant even without the introduction of the Omicron variant.
+In other words, the Delta strain would have continued to spread on campus at a similar rate if the semester were to (hypothetically) continue until January without additional interventions due to immune waning and growing cases in the community (\fref{matching}H).
 In reality, the situation was more complex: testing frequencies increased and social gatherings were limited in response to an increase in the number of cases.
@@
 
-Epidemiological and interventions conditions changed throughout the spring semester of the 2021--2022 academic year.
-We therefore extend to model to account for more realistic settings and focus on the outbreak patterns among undergraduate students.
+Epidemiological conditions and intervention measures changed throughout the spring semester of the 2021--2022 academic year.
+We therefore extend to model to account for these alterations and focus on the outbreak patterns among undergraduate students.
 First, based on \citep{ferguson2021report}, we assume that two and three doses of vaccines reduce susceptibility against the Omicron variant by 10\% and 70\%, respectively. 
-We also assume that the transmissibility of Omicron is reduced proportionally following previously assumed 90--to--20 ratio for the Delta variant;
+We also assume that the transmissibility of Omicron is reduced proportionally following the previously assumed 90--to--20 ratio for the Delta variant;
 in other words, two and three doses of vaccines reduce transmissibility by 2.2\% and 15.6\%, respectively.
@@
 The increase in the reproduction number can reflect increased contact rates following changes in distancing policy as well as increased transmissibility of the BA.2 subvariant---we do not explicitly distinguish the cause of the increase in the reproduction number.
-We do so by simulating the model forward across a range of contact reproduction numbers that are consistent with previous estimates ($\Rc=2--6$) and introducing a 20\%--100\% increase in the contact reproduction number on February 8, 2022, with changes in the testing frequency.
+We do so by simulating the model forward across a range of contact reproduction numbers that are consistent with previous estimates ($\Rc$=2--6) and introducing a 20\%--100\% increase in the contact reproduction number on February 8, 2022, with changes in the testing frequency.
 To match the realistic campus setting, we assume that 700 students are present on campus as of January 1, 2022, and the remaining 4300 students come back to campus across 28 days.
@@
 Additional increases in the reproduction number (alongside the changes in testing frequency) can cause the case numbers to further increase, but we are unable to match the observed dynamics even with a 100\% increase in the reproduction number.
-Indeed, a $>10$-fold increase in the numbers of cases between the weeks ending February 11 and 18, 2022, will require an unrealistically high increase in the reproduction number to explain.
+Indeed, a $>10$-fold increase in the numbers of cases between the weeks ending February 11 and 18, 2022, would require an unrealistically high increase in the contact reproduction number to explain.
 These simulations indicate that changes in distancing and testing policies and the increased transmissibility of the BA.2 subvariant alone are unlikely to be the direct causes of the outbreak.
@@
 
-Here, we present the analysis of SARS-CoV-2 outbreaks on the PU campus between fall 2020 and early 2022.
-Our analysis demonstrates strong spatiotemporal correlations between the patterns of spread of SARS-CoV-2 on campus and those from surrounding communities.
+Here, we analyze SARS-CoV-2 outbreaks on the PU campus between fall 2020 and early 2022.
+We demonstrate strong spatiotemporal correlations between the patterns of spread of SARS-CoV-2 on campus and those from surrounding communities.
 These correlations decreased with distance from Mercer County in fall 2021--2022, likely reflecting contact and commuting patterns as the university campus reopened.
-Mathematical modeling further suggests limited transmission between the university population during fall and spring semesters of the 2020--2021 academic year and an increased amount of community contacts during the fall semester of the 2021--2022 academic year compared previous semesters.
+Mathematical modeling further suggests limited transmission between the university population during fall and spring semesters of the 2020--2021 academic year and an increased frequency of infective community contacts during the fall semester of the 2021--2022 academic year, compared to previous semesters.
 An increase in the number of cases by the end of November 2021 is consistent with the increase in the levels of community cases and waning immunity.
-Finally, our analysis highlights the role of super-spreading events in driving the spread of the Omicron variant on PU campus.
+Finally, our analysis highlights the role of super-spreading events in driving the spread of the Omicron variant on the PU campus.
 
-Although previous outbreak reports from other universities primarily focused on transmission on university campuses \citep{wilson2020multiple,currie2021interventions}, a few studies identified off-campus infections as an important source of transmission \citep{fox2021response,hamer2021assessment}.
+Although previous outbreak reports from other universities primarily focused on within-campus transmission \citep{wilson2020multiple,currie2021interventions}, a few studies identified off-campus infections as an important source of transmission \citep{fox2021response,hamer2021assessment}.
 For example, extensive modeling efforts from Cornell University demonstrated an increase in the amount of transmission from outside the university campus during fall 2021 and found that community transmissions are the biggest risk for faculty and staff members \citep{frazier2022modeling}.
-Our study further extends these findings in demonstrating a strong spatiotemporal correlation in the spread of SARS-CoV-2 between university campuses and surrounding communities.
+Our study further extends these findings in demonstrating a strong spatiotemporal correlation in the spread of SARS-CoV-2 between university campuses and surrounding communities;
+however, when campus transmission is sustained, community coupling becomes less important.
 Our analysis suggest that comparing the ratios between the cases on university campuses and neighboring communities can also provide a useful measure for how well a university campus is controlling the epidemic;
@@
 For example, the ratios of cases can suddenly change during holidays when students are away from campus.
-Future studies may combine phylogenetic data to better understand spatial patterns of SARS-CoV-2 on campus.
+Future studies could combine viral phylogenetic data to better understand spatial patterns of SARS-CoV-2 on campus.
 
@@
 instead, we assume constant values for $\theta$ and $\Rc$ throughout a semester.
+% We note that the invasion of the Omicron variant was modeled separately for the spring semester of a 2021--2022 academic year by explicitly considering the immune evasion.
 We also do not explore parameter uncertainty, which can lead to underestimation of overall uncertainty \citep{elderd2006uncertainty}. 
@@
 Despite the simplicity of the analysis, our study provides important lessons for controlling SARS-CoV-2 outbreaks on university campuses in general.
-First, our analysis highlights the power of mass testing for epidemic measurement and planning.
+First, our analysis highlights the power of mass asymptomatic testing for epidemic measurement and planning.
 Combining other interventions measures, such as social distancing, mask wearing, and vaccination, can help provide a safe means of reopening university campuses---but the extent to which these interventions are implemented will necessarily depend on resource availability.
-Second, the safe reopening of a university campus must consider the spread of SARS-CoV-2 within the surrounding community as they can both drive transmission in each other.
+Second, the safe reopening of a university campus must consider the spread of SARS-CoV-2 within the surrounding community as they can both potentially drive transmission in each other---however, the degree to which infections spread from campus to community remains uncertain.
 Finally, intervention measures placed on campuses must continue to adapt and change to reflect changes in epidemiological conditions.
@@
 The emergence of new variants---in particular, their ability to evade prior immunity and transmit better---continues to add uncertainty to the future controllability of the ongoing SARS-CoV-2 pandemic.
-Nonetheless, as population-level immunity increases (either due to infection or vaccination), we are transitioning to an endemic phase, during which COVID-19 is expected to become less severe \citep{lavine2021immunological}.
-Many institutions have so far sought to minimize infections on their campuses, thereby implementing as many interventions as possible---but such measures can be difficult to maintain for prolonged periods both economically and societally.
-Instead, the expectations for reopening campuses (e.g., whether to minimize infections on campuses) need to be re-evaluated as we transition to next phases of the pandemic, accounting not only for changes in epidemic dynamics but also for our perception of the pathogen.
-The answers to these questions ultimately depend on the landscape of SARS-CoV-2 immunity \citep{saad2020immune,baker2021limits}.
+Nonetheless, as population-level immunity increases (either due to infection or vaccination), we are (hopefully) transitioning to an endemic phase, during which COVID-19 is expected to become less severe \citep{lavine2021immunological}.
+Many institutions have so far sought to minimize infections on their campuses early in the pandemic, thereby implementing as many interventions as possible---but such measures can be difficult to maintain for prolonged periods both economically and societally.
+As the Omicron variant began to spread, many campuses---including the PU campus---opted in for less intense interventions, reflecting difficulties in controlling the spread and a lack of severe cases among the majority of vaccinated students.
+As we continue to transition to future phases of the pandemic, the expectations for reopening campuses (e.g., whether to minimize infections on campuses) need to be re-evaluated, accounting not only for changes in epidemic dynamics but also for our perception of the pathogen.
+The answers to these questions ultimately depend on the landscape of SARS-CoV-2 immunity and its future evolutionary dynamics \citep{saad2020immune,baker2021limits}.
 

### 2022-07-12 doc/university-covid.tex
almost done

@@
 Sang Woo Park\textsuperscript{1}
-Irini Daskalaki\textsuperscript{?}
-Robin M. Izzo\textsuperscript{?}
+Irini Daskalaki\textsuperscript{2}
+Robin M. Izzo\textsuperscript{3}
+Irina Aranovich\textsuperscript{4}
+Aartjan J.W. te Velthuis\textsuperscript{4}
+Daniel A. Notterman\textsuperscript{4}
 C. Jessica E. Metcalf\textsuperscript{1}
-Daniel A. Notterman\textsuperscript{?}
-Bryan T.\ Grenfell\textsuperscript{1}
+Bryan T.\ Grenfell\textsuperscript{1,5}
 \\
@@
 \\
+\textbf{2} Princeton University Health Services, Princeton University, Princeton, NJ, USA
+\\
+\textbf{3} Environmental Health and Safety, Princeton University, Princeton, NJ, USA
+\\
+\textbf{4} Department of Molecular Biology, Princeton University, Princeton, NJ, USA
+\\
+\textbf{5} Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA
+\\
 \bigskip
 
-Corresponding author: ???
+Corresponding author: swp2@princeton.edu
 \bigskip
@@
 These findings are corroborated by our model predictions, which indicate that the amount of on-campus transmission was likely limited during much of the wider pandemic until the end of 2021.
-Finally, we find that a super-spreading event likely played a major role in driving the recent outbreak on the Princeton University campus during the spring semester of a 2021--2022 academic year.
-Despite large numbers of cases on campus, case levels in surrounding communities remained low, suggesting that there was little spillover transmission from campus to the local community.
+Finally, we find that a super-spreading event likely played a major role in driving the recent Omicron variant outbreak on the Princeton University campus during the spring semester of the 2021--2022 academic year.
+Despite large numbers of cases on campus in this period, case levels in surrounding communities remained low, suggesting that there was little spillover transmission from campus to the local community.
 
@@
 Social distancing measures have played major roles in preventing transmission, but can be difficult to maintain for a prolonged period \citep{galanti2021social}.
-The development of vaccines has provided a safe means of reopening society, but uncertainty remains in their long-term effectiveness in preventing infection and transmission, especially in the face of new viral variants.
+The development of vaccines has provided a safe means of reopening society, but uncertainty remains on their long-term effectiveness in preventing infection and transmission, especially in the face of new viral variants.
 
@@
 For example, Duke and Harvard Universities experienced moderate outbreaks at the beginning of the fall semester in 2021 when in-person classes were allowed, despite high vaccination rates and weekly asymptomatic testing protocols \citep{dukeoutbreak,harvardoutbreak}, whereas the number of cases remained low in Princeton University (PU) during the same time period with similar levels of testing and vaccination.
-Here, we focused on the dynamics of SARS-CoV-2 on the PU campus alone to eliminate heterogeneities inherent to such comparisons.
+Here, we focused on the dynamics of SARS-CoV-2 on the PU campus alone to eliminate heterogeneities inherent to such comparisons; we return to comparisons with other campuses later in the discussion.
 
-We analyzed the spread of SARS-CoV-2 on campus (\fref{princeton}).
-We begin with a descriptive analysis of the outbreak and then present modeling analyses of the various outbreaks during 2020--2022.
+We begin with a descriptive analysis of the PU outbreak (\fref{princeton}), and present modeling analyses of the individual epidemics during 2020--2022.
 PU is located in Mercer County, New Jersey, USA;
@@
 
-During the fall semester of the 2020--2021 academic year, roughly 1000 grad students and 2000 faculty and staff members were present on campus and participated in asymptomatic testing. 
+During the \textbf{fall semester of the 2020--2021 academic year}, roughly 1000 grad students and 2000 faculty and staff members were present on campus and participated in asymptomatic testing. 
 All classes were held virtually, and so only a few undergraduate students remained on campus ($<300$).  
@@
 
-In the beginning of the spring semester of the 2020--2021 academic year, the number of cases suddenly increased before classes started (\fref{princeton}B), reflecting $\approx 3000$ undergraduate students returning to campus.
+In the beginning of the \textbf{spring semester of the 2020--2021 academic year}, the number of cases suddenly increased before classes started (\fref{princeton}B), reflecting $\approx 3000$ undergraduate students returning to campus.
 Returning students were required to be tested and quarantine for 7 days regardless of their returning location.
@@
 
-For the fall semester of the 2021--2022 academic year, all students and faculty and staff members were required to be vaccinated, with very few medical and religious exemptions.
+For the \textbf{fall semester of the 2021--2022 academic year}, all students and faculty and staff members were required to be vaccinated, with very few medical and religious exemptions.
 By the beginning of the semesters, 97\% of undergraduate students, 96\% of graduate students, and 94\% of faculty and staff members were vaccinated.
@@
 
-For the spring semester of the 2021--2022 academic year, all eligible students and faculty and staff members were required to be boosted.
+For the \textbf{spring semester of the 2021--2022 academic year}, all eligible students and faculty and staff members were required to be boosted.
 By the beginning of the semesters, 65\% of undergraduate students, 71\% of graduate students, and 82\% of faculty and staff members were boosted.
@@
 
-\subsection{Comparison with community transmission}
+\subsection*{Comparisons of campus and community transmission}
 
@@
 The population size and testing frequencies (with twice weekly testing modeled as testing every 3 days) are set to reflect realistic campus settings.
-Although we account for heterogeneity in the number of individuals in each population group on campus (i.e., undergraduate students, graduate students, and faculty and staff members) and their respective testing patterns (e.g., twice a week for undergraduate and graduate students and once a week for faculty and staff members during fall and spring, 2020), we assume that all other parameters are equal across different groups for simplicity.
+Although we account for heterogeneity in the number of individuals in each population group on campus (i.e., undergraduate students, graduate students, and faculty and staff members) and their respective testing patterns (e.g., twice a week for undergraduate and graduate students and once a week for faculty and staff members during fall and spring, 2020), we assume, for simplicity, that all other parameters are equal across different groups
 We further assume that the population mixes homogeneously.
@@
 These interventions---as well as students returning back home as classes ended---likely would have reduced contact rates (and therefore transmission of the Delta variant).
-This reduction in transmission was likely counterbalanced by the introduction of the Omicron variant and its immune evasion, leading to similar and persistent growth in the number of cases.
+This reduction in transmission was likely counterbalanced by the introduction of the Omicron variant and its high transmissibility and immune evasion, leading to similar and persistent growth in the number of cases.
 
@@
 In reality, increases in cases were often associated with specific transmission clusters, suggesting heterogeneity in transmission patterns.
-Contact levels also likely differ between different groups;
-for example, faculty and staff members are more likely to interact with community members than undergraduate students and are therefore at a higher risk for community infections \citep{frazier2022modeling}.
+Contact levels also likely differ between different groups:
+for example, faculty and staff members are more likely to interact with community members than undergraduate students and would be at a higher risk for community infections \citep{frazier2022modeling};
+therefore, our homogeneous mixing assumption is conservatively pessimistic.
 We also do not account for changes in behavior; 

### 2022-07-13 doc/university-covid.tex
done

@@
 \\
-\textbf{2} Princeton University Health Services, Princeton University, Princeton, NJ, USA
+\textbf{2} University Health Services, Princeton University, Princeton, NJ, USA
 \\
@@
 \pagebreak
+
+\section*{Introduction}
 
@@
 \end{figure}
+
+\section*{Descriptive analysis}
 
@@
 Thus, combining some amount of immune waning and high campus and community contact rates likely best explains the epidemic growth near the end of the semester.
+We note that other factors, such as changes in behavior, could have also contributed to the increase in the numbers of cases.
 
@@
 however, when campus transmission is sustained, community coupling becomes less important.
-Our analysis suggest that comparing the ratios between the cases on university campuses and neighboring communities can also provide a useful measure for how well a university campus is controlling the epidemic;
-however, we note that this ratio needs to be interpreted with caution as it is sensitive to changes in testing patterns as well as the numbers of students on campus.
+The degree to which community coupling affects campus transmission also depends on the campus. 
+Although Princeton University is located in a small town (Mercer County) with a population of ~390,000 (\url{www.census.gov}), it is located near large cities, such as New York City and Philadelphia, which can drive infections in smaller cities nearby \citep{grenfell1997meta}.
+For example, contact tracing efforts from Boston University, which is located in a large metropolitan area, found that more than 50\% of infections among Boston University affiliates with known exposures could be attributed to sources outside of the university campus \citep{hamer2021assessment}.
+In contrast, other university campuses that are far from urban areas may experience weaker community coupling.
+The degree of coupling will also depend on intervention measures in surrounding communities and on campus.
+Understanding these heterogeneities is critical for preventing future campus outbreaks.
+
+Our analysis also suggests that comparing the ratios between the cases on university campuses and neighboring communities can also provide a useful measure for how well a university campus is controlling the epidemic;
+however, that this ratio needs to be interpreted with caution as it is sensitive to changes in testing patterns as well as the numbers of students on campus.
 For example, the ratios of cases can suddenly change during holidays when students are away from campus.
@@
 instead, we assume constant values for $\theta$ and $\Rc$ throughout a semester.
+While we cannot rule out the possibility that changes in behavior (and therefore transmission rates) could have contributed to various epidemics (e.g., the Omicron wave beginning in the fall semester of the 2021--2022 academic year), we were able to capture the majority of epidemic patterns without modeling them.
 % We note that the invasion of the Omicron variant was modeled separately for the spring semester of a 2021--2022 academic year by explicitly considering the immune evasion.
@@
 Combining other interventions measures, such as social distancing, mask wearing, and vaccination, can help provide a safe means of reopening university campuses---but the extent to which these interventions are implemented will necessarily depend on resource availability.
-Second, the safe reopening of a university campus must consider the spread of SARS-CoV-2 within the surrounding community as they can both potentially drive transmission in each other---however, the degree to which infections spread from campus to community remains uncertain.
+Second, we expect immune waning and superspreading to continue to play important roles in driving campus transmission---keeping vaccine statuses up-to-date within the campus community will be critical moving forward.
+In addition, preventing large gatherings can help prevent large superspreading events in the midst of a rising epidemic.
+Third, the safe reopening of a university campus must consider the spread of SARS-CoV-2 within the surrounding community as they can both potentially drive transmission in each other---however, the degree to which infections spread from campus to community remains uncertain.
 Finally, intervention measures placed on campuses must continue to adapt and change to reflect changes in epidemiological conditions.

### 2022-07-28 doc/university-covid.tex
updating

@@
 Irina Aranovich\textsuperscript{4}
-Aartjan J.W. te Velthuis\textsuperscript{4}
-Daniel A. Notterman\textsuperscript{4}
+Aartjan J.W. te Velthuis\textsuperscript{5}
+Daniel A. Notterman\textsuperscript{5}
 C. Jessica E. Metcalf\textsuperscript{1}
-Bryan T.\ Grenfell\textsuperscript{1,5}
+Bryan T.\ Grenfell\textsuperscript{1,6}
 \\
@@
 \\
-\textbf{4} Department of Molecular Biology, Princeton University, Princeton, NJ, USA
+\textbf{4} Princeton University Clinical Laboratory, Princeton University, Princeton, NJ, USA
 \\
-\textbf{5} Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA
+\textbf{5} Department of Molecular Biology, Princeton University, Princeton, NJ, USA
+\\
+\textbf{6} Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA
 \\
@@
 
+\section*{Data availability}
+
+All data and code are stored in a publicly available GitHub repository (\url{https://github.com/parksw3/university-covid}).
+
 \bibliography{university-covid}

### 2022-07-28 figure_diagram/diagram.tex
updating

\documentclass[tikz,border=2mm]{standalone}

\usetikzlibrary{arrows,positioning}

\begin{document}

\begin{tikzpicture}[node distance=2.7cm,auto,>=latex',every node/.append style={align=center},int/.style={draw, minimum size=1cm},inverter/.style={rectangle,draw,inner sep=2pt,minimum size=6mm},posnode/.style={shape=rectangle, draw=orange, line width=2},
  negnode/.style={shape=rectangle, draw=black, line width=2}]
    \node [int, ultra thick] (S)             {$S$};
    \node [int, right=of S, ultra thick] (E)             {$E$};
    \node [int, right=of E, ultra thick, orange] (Ip1)             {$I_{p,1}$};
    \node [int, right=of Ip1, ultra thick, orange] (Ip2)             {$I_{p,2}$};
    \node [int, above right=of Ip2, ultra thick, orange] (Is1)             {$I_{s,1}$};
    \node [int, right=of Is1, ultra thick, orange] (Is2)             {$I_{s,2}$};
    \node [int, below right=of Ip2, ultra thick, orange] (Ia1)             {$I_{a,1}$};
    \node [int, right=of Ia1, ultra thick, orange] (Ia2)             {$I_{a,2}$};
    \node [int, below right=of Is2, ultra thick] (R)             {$R$};
    \path[->, ultra thick] (S) edge (E);
    \path[->, ultra thick] (E) edge (Ip1);
    \path[->, ultra thick] (Ip1) edge (Ip2);
    \path[->, ultra thick] (Ip2) edge (Is1);
    \path[->, ultra thick] (Is1) edge (Is2);
    \path[->, ultra thick] (Ip2) edge (Ia1);
    \path[->, ultra thick] (Ia1) edge (Ia2);
    \path[->, ultra thick] (Is2) edge (R);
    \path[->, ultra thick] (Ia2) edge (R);
    \matrix [draw,below right] at (current bounding box.north west) {
        \node [negnode,label=right:PCR true negative] {}; \\
        \node [posnode,label=right:PCR true positive] {}; \\
    };
\end{tikzpicture}
\end{document}

### 2022-07-28 figure_princeton_new/figure_princeton_map_extended_new.R

@@ -1,0 +1,574 @@
+library(dplyr)
+library(tidyr)
+library(readxl)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(egg)
+library(lubridate)
+library(usmap)
+library(geosphere)
+
+data_asymp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=1) %>%
+  mutate(
+    type="asymp"
+  )
+
+data_symp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=2) %>%
+  mutate(
+    type="symp"
+  )
+
+data <- bind_rows(
+  data_asymp, data_symp
+)
+
+county <- read.csv("../data/us-counties-03-22-2022.csv")
+
+countylonlat <- read.csv("../data/us-counties-lonlat.csv")
+
+county2 <- county %>%
+  filter(state %in% c("New Jersey", "New York", "Pennsylvania")) %>%
+  group_by(state, county) %>%
+  arrange(state, county, date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+fall2020 <- data %>%
+  filter(`Week Ending` <= as.Date("2021-01-01")) %>%
+  select(`Week Ending`, `Undergrad Positive Tests`, `Grad Student Positive Tests`,
+         `Faculty Staff Positive Tests`) %>%
+  rename(
+    date=`Week Ending`,
+    Undergrad=`Undergrad Positive Tests`,
+    Grad=`Grad Student Positive Tests`,
+    `Faculty and Staff`=`Faculty Staff Positive Tests`
+  ) %>%
+  mutate(
+    date=as.Date(date)
+  )  %>%
+  mutate(
+    year=substr(date, 1, 4),
+    month=as.numeric(gsub("-.*", "", gsub("20..-", "", date))), 
+    day=as.numeric(gsub(".*-", "", gsub("20..-", "", date)))
+  ) %>%
+  arrange(date) %>%
+  mutate(
+    date=paste0(month.abb[month], " ", day, ", ", year),
+    date=factor(date, levels=unique(date))
+  ) %>%
+  gather(key, value, -date, -year, -month, -day) %>%
+  mutate(
+    date2=as.Date(paste0(year, "-", month, "-", day))
+  ) %>%
+  group_by(date, year, month, day, key, date2) %>%
+  summarize(
[…]

### 2022-07-28 figure_princeton_new/figure_princeton_map_new.R

@@ -1,0 +1,565 @@
+library(dplyr)
+library(tidyr)
+library(readxl)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(egg)
+library(lubridate)
+library(usmap)
+library(geosphere)
+
+data_asymp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=1) %>%
+  mutate(
+    type="asymp"
+  )
+
+data_symp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=2) %>%
+  mutate(
+    type="symp"
+  )
+
+data <- bind_rows(
+  data_asymp, data_symp
+)
+
+county <- read.csv("../data/us-counties-03-22-2022.csv")
+
+countylonlat <- read.csv("../data/us-counties-lonlat.csv")
+
+nj <- county %>%
+  filter(state=="New Jersey") %>%
+  group_by(county) %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+fall2020 <- data %>%
+  filter(`Week Ending` <= as.Date("2021-01-01")) %>%
+  select(`Week Ending`, `Undergrad Positive Tests`, `Grad Student Positive Tests`,
+         `Faculty Staff Positive Tests`) %>%
+  rename(
+    date=`Week Ending`,
+    Undergrad=`Undergrad Positive Tests`,
+    Grad=`Grad Student Positive Tests`,
+    `Faculty and Staff`=`Faculty Staff Positive Tests`
+  ) %>%
+  mutate(
+    date=as.Date(date)
+  )  %>%
+  mutate(
+    year=substr(date, 1, 4),
+    month=as.numeric(gsub("-.*", "", gsub("20..-", "", date))), 
+    day=as.numeric(gsub(".*-", "", gsub("20..-", "", date)))
+  ) %>%
+  arrange(date) %>%
+  mutate(
+    date=paste0(month.abb[month], " ", day, ", ", year),
+    date=factor(date, levels=unique(date))
+  ) %>%
+  gather(key, value, -date, -year, -month, -day) %>%
+  mutate(
+    date2=as.Date(paste0(year, "-", month, "-", day))
+  ) %>%
+  group_by(date, year, month, day, key, date2) %>%
+  summarize(
+    value=sum(value)
+  )
+
[…]

### 2022-07-28 figure_princeton_new/figure_princeton_new.R

@@ -1,0 +1,611 @@
+library(dplyr)
+library(tidyr)
+library(readxl)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(gridExtra)
+library(egg)
+library(lubridate)
+
+data_asymp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=1) %>%
+  mutate(
+    type="asymp"
+  )
+
+data_symp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=2) %>%
+  mutate(
+    type="symp"
+  )
+
+data <- bind_rows(
+  data_asymp, data_symp
+)
+
+county <- read.csv("../data/us-counties-03-22-2022.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+fall2020 <- data %>%
+  filter(`Week Ending` <= as.Date("2021-01-01")) %>%
+  select(`Week Ending`, `Undergrad Positive Tests`, `Grad Student Positive Tests`,
+         `Faculty Staff Positive Tests`) %>%
+  rename(
+    date=`Week Ending`,
+    Undergrad=`Undergrad Positive Tests`,
+    Grad=`Grad Student Positive Tests`,
+    `Faculty and Staff`=`Faculty Staff Positive Tests`
+  ) %>%
+  mutate(
+    date=as.Date(date)
+  )  %>%
+  mutate(
+    year=substr(date, 1, 4),
+    month=as.numeric(gsub("-.*", "", gsub("20..-", "", date))), 
+    day=as.numeric(gsub(".*-", "", gsub("20..-", "", date)))
+  ) %>%
+  arrange(date) %>%
+  mutate(
+    date=paste0(month.abb[month], " ", day, ", ", year),
+    date=factor(date, levels=unique(date))
+  ) %>%
+  gather(key, value, -date, -year, -month, -day) %>%
+  mutate(
+    date2=as.Date(paste0(year, "-", month, "-", day))
+  ) %>%
+  group_by(date, year, month, day, key, date2) %>%
+  summarize(
+    value=sum(value)
+  )
+
+spring2020 <- data %>%
+  filter(`Week Ending` <= as.Date("2021-05-14"),
+         `Week Ending` >= as.Date("2021-01-22")) %>%
[…]

### 2022-07-28 figure_princeton_new/figure_princeton_newyork_new.R

@@ -1,0 +1,461 @@
+library(dplyr)
+library(tidyr)
+library(readxl)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(gridExtra)
+library(egg)
+library(lubridate)
+
+data_asymp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=1) %>%
+  mutate(
+    type="asymp"
+  )
+
+data_symp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=2) %>%
+  mutate(
+    type="symp"
+  )
+
+data <- bind_rows(
+  data_asymp, data_symp
+)
+
+county <- read.csv("../data/us-counties-03-22-2022.csv")
+
+nyc <- county %>%
+  filter(county=="New York City", state=="New York") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+nyc$cases[389] <- (nyc$cases[388] + nyc$cases[390])/2
+
+fall2020 <- data %>%
+  filter(`Week Ending` <= as.Date("2021-01-01")) %>%
+  select(`Week Ending`, `Undergrad Positive Tests`, `Grad Student Positive Tests`,
+         `Faculty Staff Positive Tests`) %>%
+  rename(
+    date=`Week Ending`,
+    Undergrad=`Undergrad Positive Tests`,
+    Grad=`Grad Student Positive Tests`,
+    `Faculty and Staff`=`Faculty Staff Positive Tests`
+  ) %>%
+  mutate(
+    date=as.Date(date)
+  )  %>%
+  mutate(
+    year=substr(date, 1, 4),
+    month=as.numeric(gsub("-.*", "", gsub("20..-", "", date))), 
+    day=as.numeric(gsub(".*-", "", gsub("20..-", "", date)))
+  ) %>%
+  arrange(date) %>%
+  mutate(
+    date=paste0(month.abb[month], " ", day, ", ", year),
+    date=factor(date, levels=unique(date))
+  ) %>%
+  gather(key, value, -date, -year, -month, -day) %>%
+  mutate(
+    date2=as.Date(paste0(year, "-", month, "-", day))
+  ) %>%
+  group_by(date, year, month, day, key, date2) %>%
+  summarize(
+    value=sum(value)
+  )
+
+spring2020 <- data %>%
+  filter(`Week Ending` <= as.Date("2021-05-14"),
[…]

### 2022-07-28 figure_princeton_new/figure_princeton_philly_new.R

@@ -1,0 +1,459 @@
+library(dplyr)
+library(tidyr)
+library(readxl)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(gridExtra)
+library(egg)
+library(lubridate)
+
+data_asymp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=1) %>%
+  mutate(
+    type="asymp"
+  )
+
+data_symp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=2) %>%
+  mutate(
+    type="symp"
+  )
+
+data <- bind_rows(
+  data_asymp, data_symp
+)
+
+county <- read.csv("../data/us-counties-03-22-2022.csv")
+
+philly <- county %>%
+  filter(county=="Philadelphia", state=="Pennsylvania") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+fall2020 <- data %>%
+  filter(`Week Ending` <= as.Date("2021-01-01")) %>%
+  select(`Week Ending`, `Undergrad Positive Tests`, `Grad Student Positive Tests`,
+         `Faculty Staff Positive Tests`) %>%
+  rename(
+    date=`Week Ending`,
+    Undergrad=`Undergrad Positive Tests`,
+    Grad=`Grad Student Positive Tests`,
+    `Faculty and Staff`=`Faculty Staff Positive Tests`
+  ) %>%
+  mutate(
+    date=as.Date(date)
+  )  %>%
+  mutate(
+    year=substr(date, 1, 4),
+    month=as.numeric(gsub("-.*", "", gsub("20..-", "", date))), 
+    day=as.numeric(gsub(".*-", "", gsub("20..-", "", date)))
+  ) %>%
+  arrange(date) %>%
+  mutate(
+    date=paste0(month.abb[month], " ", day, ", ", year),
+    date=factor(date, levels=unique(date))
+  ) %>%
+  gather(key, value, -date, -year, -month, -day) %>%
+  mutate(
+    date2=as.Date(paste0(year, "-", month, "-", day))
+  ) %>%
+  group_by(date, year, month, day, key, date2) %>%
+  summarize(
+    value=sum(value)
+  )
+
+spring2020 <- data %>%
+  filter(`Week Ending` <= as.Date("2021-05-14"),
+         `Week Ending` >= as.Date("2021-01-22")) %>%
[…]

### 2022-07-28 figure_princeton_new/figure_princeton_simulation_new.R

@@ -1,0 +1,590 @@
+library(dplyr)
+library(tidyr)
+library(readxl)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(egg)
+library(lubridate)
+
+data_asymp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=1) %>%
+  mutate(
+    type="asymp"
+  )
+
+data_symp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=2) %>%
+  mutate(
+    type="symp"
+  )
+
+data <- bind_rows(
+  data_asymp, data_symp
+)
+
+fall2020 <- data %>%
+  filter(`Week Ending` <= as.Date("2021-01-01")) %>%
+  select(`Week Ending`, `Undergrad Positive Tests`, `Grad Student Positive Tests`,
+         `Faculty Staff Positive Tests`) %>%
+  rename(
+    date=`Week Ending`,
+    Undergrad=`Undergrad Positive Tests`,
+    Grad=`Grad Student Positive Tests`,
+    `Faculty and Staff`=`Faculty Staff Positive Tests`
+  ) %>%
+  mutate(
+    date=as.Date(date),
+    weekly=Undergrad+Grad+`Faculty and Staff`
+  ) %>%
+  group_by(date) %>%
+  summarize(
+    weekly=sum(weekly)
+  )
+
+spring2020 <- data %>%
+  filter(`Week Ending` <= as.Date("2021-05-14"),
+         `Week Ending` >= as.Date("2021-01-22"))  %>%
+  select(`Week Ending`, `Undergrad Positive Tests`, `Grad Student Positive Tests`,
+         `Faculty Staff Positive Tests`) %>%
+  rename(
+    date=`Week Ending`,
+    Undergrad=`Undergrad Positive Tests`,
+    Grad=`Grad Student Positive Tests`,
+    `Faculty and Staff`=`Faculty Staff Positive Tests`
+  ) %>%
+  mutate(
+    date=as.Date(date),
+    weekly=Undergrad+Grad+`Faculty and Staff`
+  ) %>%
+  group_by(date) %>%
+  summarize(
+    weekly=sum(weekly)
+  )
+
+fall2021 <- data %>%
+  filter(`Week Ending` >= as.Date("2021-08-20"),
+         `Week Ending` <= as.Date("2021-12-31")) %>%
[…]

### 2022-07-28 figure_princeton_new/figure_princeton_simulation_nowaning.R

@@ -1,0 +1,135 @@
+library(dplyr)
+library(tidyr)
+library(readxl)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(egg)
+library(lubridate)
+
+data_asymp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=1) %>%
+  mutate(
+    type="asymp"
+  )
+
+data_symp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=2) %>%
+  mutate(
+    type="symp"
+  )
+
+data <- bind_rows(
+  data_asymp, data_symp
+)
+
+fall2021 <- data %>%
+  filter(`Week Ending` >= as.Date("2021-08-20"),
+         `Week Ending` <= as.Date("2021-12-31")) %>%
+  select(`Week Ending`, `Undergrad Positive Tests`, `Grad Student Positive Tests`,
+         `Faculty Staff Positive Tests`) %>%
+  rename(
+    date=`Week Ending`,
+    Undergrad=`Undergrad Positive Tests`,
+    Grad=`Grad Student Positive Tests`,
+    `Faculty and Staff`=`Faculty Staff Positive Tests`
+  ) %>%
+  mutate(
+    date=as.Date(date),
+    weekly=Undergrad+Grad+`Faculty and Staff`
+  ) %>%
+  group_by(date) %>%
+  summarize(
+    weekly=sum(weekly)
+  )
+
+load("../simulation_princeton/simulation_fall_2021_nowaning.rda")
+
+simall_fall_2021 <- lapply(simulation_fall_2021_nowaning, function(x){
+  lapply(x, function(y) {
+    data.frame(
+      date=y$datevec,
+      time=1:length(y$incidencevec),
+      incidence=y$incidencevec,
+      cases=y$positivevec,
+      R0=y$R0,
+      scale=y$scalevec
+    )
+  }) %>%
+    bind_rows(.id="sim")
+})  %>%
+  bind_rows()
+
+simall_fall_2021_gof <- simall_fall_2021 %>%
+  mutate(
+    group=cut(as.numeric(as.Date(date)), breaks=unique(as.numeric(fall2021$date)))
+  ) %>%
+  group_by(R0, scale, sim, group) %>%
+  summarize(
+    cases=sum(cases),
+    date=max(date)
+  ) %>%
+  arrange(R0, scale, sim, date) %>%
+  merge(fall2021) %>%
[…]

### 2022-07-28 figure_princeton_new/figure_princeton_simulation_omicron_spring.R

@@ -1,0 +1,192 @@
+library(dplyr)
+library(tidyr)
+library(readxl)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(egg)
+library(lubridate)
+
+data_asymp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=1) %>%
+  mutate(
+    type="asymp"
+  )
+
+data_symp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=2) %>%
+  mutate(
+    type="symp"
+  )
+
+data <- bind_rows(
+  data_asymp, data_symp
+)
+
+spring2021 <- data %>%
+  filter(`Week Ending` > as.Date("2021-12-31"),
+         `Week Ending` <= as.Date("2022-03-18")) %>%
+  select(`Week Ending`, `Undergrad Positive Tests`) %>%
+  rename(
+    date=`Week Ending`,
+    Undergrad=`Undergrad Positive Tests`
+  ) %>%
+  mutate(
+    date=as.Date(date)
+  )  %>%
+  mutate(
+    year=substr(date, 1, 4),
+    month=as.numeric(gsub("-.*", "", gsub("20..-", "", date))), 
+    day=as.numeric(gsub(".*-", "", gsub("20..-", "", date)))
+  ) %>%
+  arrange(date) %>%
+  mutate(
+    date=paste0(month.abb[month], " ", day, ", ", year),
+    date=factor(date, levels=unique(date))
+  ) %>%
+  gather(key, value, -date, -year, -month, -day) %>%
+  mutate(
+    date2=as.Date(paste0(year, "-", month, "-", day))
+  ) %>%
+  group_by(date, year, month, day, key, date2) %>%
+  summarize(
+    value=sum(value)
+  )
+
+load("../simulation_omicron_spring/simulation_omicron_spring_noswitch.rda")
+load("../simulation_omicron_spring/simulation_omicron_spring.rda")
+load("../simulation_omicron_spring/simulation_omicron_spring_increase_20.rda")
+load("../simulation_omicron_spring/simulation_omicron_spring_increase_50.rda")
+load("../simulation_omicron_spring/simulation_omicron_spring_increase_100.rda")
+
+simall_spring_2021_noswitch <- lapply(simulation_omicron_spring_noswitch, function(x){
+  lapply(x, function(y) {
+    data.frame(
+      date=y$datevec,
+      time=1:length(y$incidencevec),
[…]

### 2022-07-28 figure_princeton_new/figure_princeton_simulation_omicron_spring_change.R

@@ -1,0 +1,99 @@
+library(dplyr)
+library(tidyr)
+library(readxl)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(egg)
+library(lubridate)
+
+data_asymp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=1) %>%
+  mutate(
+    type="asymp"
+  )
+
+data_symp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=2) %>%
+  mutate(
+    type="symp"
+  )
+
+data <- bind_rows(
+  data_asymp, data_symp
+)
+
+spring2021 <- data %>%
+  filter(`Week Ending` > as.Date("2021-12-31"),
+         `Week Ending` <= as.Date("2022-03-18")) %>%
+  select(`Week Ending`, `Undergrad Positive Tests`) %>%
+  rename(
+    date=`Week Ending`,
+    Undergrad=`Undergrad Positive Tests`
+  ) %>%
+  mutate(
+    date=as.Date(date)
+  )  %>%
+  mutate(
+    year=substr(date, 1, 4),
+    month=as.numeric(gsub("-.*", "", gsub("20..-", "", date))), 
+    day=as.numeric(gsub(".*-", "", gsub("20..-", "", date)))
+  ) %>%
+  arrange(date) %>%
+  mutate(
+    date=paste0(month.abb[month], " ", day, ", ", year),
+    date=factor(date, levels=unique(date))
+  ) %>%
+  gather(key, value, -date, -year, -month, -day) %>%
+  mutate(
+    date2=as.Date(paste0(year, "-", month, "-", day))
+  ) %>%
+  group_by(date, year, month, day, key, date2) %>%
+  summarize(
+    value=sum(value)
+  )
+
+load("../simulation_omicron_spring/simulation_omicron_spring_null.rda")
+
+simall_spring_2021_null <- lapply(simulation_omicron_spring_null, function(x){
+  lapply(x, function(y) {
+    data.frame(
+      date=y$datevec,
+      time=1:length(y$incidencevec),
+      incidence=y$incidencevec,
+      cases=y$positivevec,
+      R0=y$R0,
+      effSvec=y$effSvec/5000
+    )
+  }) %>%
+    bind_rows(.id="sim")
+})  %>%
+  bind_rows()
+
[…]

### 2022-07-28 figure_princeton_new/figure_princeton_simulation_omicron_spring_outbreak.R

@@ -1,0 +1,111 @@
+library(dplyr)
+library(tidyr)
+library(readxl)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(egg)
+library(lubridate)
+
+data_asymp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=1) %>%
+  mutate(
+    type="asymp"
+  )
+
+data_symp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=2) %>%
+  mutate(
+    type="symp"
+  )
+
+data <- bind_rows(
+  data_asymp, data_symp
+)
+
+spring2021 <- data %>%
+  filter(`Week Ending` > as.Date("2021-12-31"),
+         `Week Ending` <= as.Date("2022-03-18")) %>%
+  select(`Week Ending`, `Undergrad Positive Tests`) %>%
+  rename(
+    date=`Week Ending`,
+    Undergrad=`Undergrad Positive Tests`
+  ) %>%
+  mutate(
+    date=as.Date(date)
+  )  %>%
+  mutate(
+    year=substr(date, 1, 4),
+    month=as.numeric(gsub("-.*", "", gsub("20..-", "", date))), 
+    day=as.numeric(gsub(".*-", "", gsub("20..-", "", date)))
+  ) %>%
+  arrange(date) %>%
+  mutate(
+    date=paste0(month.abb[month], " ", day, ", ", year),
+    date=factor(date, levels=unique(date))
+  ) %>%
+  gather(key, value, -date, -year, -month, -day) %>%
+  mutate(
+    date2=as.Date(paste0(year, "-", month, "-", day))
+  ) %>%
+  group_by(date, year, month, day, key, date2) %>%
+  summarize(
+    value=sum(value)
+  )
+
+load("../simulation_omicron_spring/simulation_omicron_spring_outbreak.rda")
+
+simall_spring_2021_outbreak <- lapply(simulation_omicron_spring_outbreak, function(x){
+  lapply(x, function(y) {
+    data.frame(
+      date=y$datevec,
+      time=1:length(y$incidencevec),
+      incidence=y$incidencevec,
+      cases=y$positivevec,
+      R0=y$R0,
+      outbreaksize=y$outbreaksize
+    )
+  }) %>%
+    bind_rows(.id="sim")
+})  %>%
+  bind_rows()
+
[…]

### 2022-07-28 R/simulate.R
more

@@ -1,0 +1,158 @@
+drawpar <- function(D_e=2,
+                    D_p=3,
+                    D_s=3,
+                    p_a=0.4,
+                    R0=2.5,
+                    p_p=0.5, ## proportion of presymptomatic transmission for symptomatic individuals
+                    k=0.1,
+                    preport=1) {
+  
+  c(
+    D_e=D_e,
+    D_p=D_p,
+    D_s=D_s,
+    p_a=p_a,
+    R0=R0,
+    p_p=p_p,
+    k=k,
+    preport=preport
+  )
+}
+
+##' @param x parameter (from `drawpar()`)
+convertpar <- function(x) {
+  with(as.list(x), {
+    ## (1-p_p) * R0 = beta * D_s
+    ##
+    ##
+    
+    beta <- R0 * (1 - p_p)/D_s
+    Rs <- D_s * beta
+    
+    Rp <- p_p/(1-p_p) * Rs
+    
+    beta_p <- Rp/D_p
+    
+    list(
+      sigma=-log(1-1/D_e), 
+      gamma_a1=-log(1-2/D_s),
+      gamma_a2=-log(1-2/D_s),
+      gamma_p1=-log(1-2/D_p), 
+      gamma_p2=-log(1-2/D_p), 
+      gamma_s1=-log(1-2/D_s),
+      gamma_s2=-log(1-2/D_s),
+      p_a=p_a, 
+      beta=beta,
+      beta_p=beta_p,
+      k=k,
+      preport=preport
+    )
+  })
+}
+
+initialize <- function(param,
+                       N,
+                       I,
+                       R) {
+  with(as.list(param), {
+    .state <- c("S", "E", "Ip1", "Ip2", "Ia1", "Ia2", "Is1", "Is2", "R")
+    
+    distrb <- c(rmultinom(1, N, c(1-I-R, I, R)))
+    
+    prob <- c(D_e, D_p/2, D_p/2, D_s/2, D_s/2, 0, 0)
+    
+    prob <- prob/sum(prob)
+    
+    state <- rmultinom(1, distrb[2], prob)
+    
+    final <- c(distrb[1], state, distrb[3])
+    
+    sample(unname(unlist(mapply(function(x, y) rep(x, y), .state, final))))
+  })
+}
+
+simulate <- function(param=drawpar(),
+                     vacc,
+                     vprop=0.9,
+                     veff=0.9,
+                     return,
+                     returnfrac=1/7,
+                     N=8000,
[…]

### 2022-07-28 R/simulate_internal.R

@@ -1,0 +1,585 @@
+##' @param state (list) initial state of individuals (X_t^{(i)} in the Appendix)
+##' @param return (list) initial return state of individuals (X_t^{(i)} in the Appendix)
+##' @param sigma transition rate from the exposed stage
+##' @param p_a probability of asymptomatic infection
+##' @param gamma_a transition rate from the asymptomatic infection stage to the recovered stage
+##' @param gamma_p transition rate from the presymptomatic infection stage to the symptomatic infection stage
+##' @param gamma_s transition rate from the symptomatic infection stage to the recovered stage
+##' @param beta_a transmission rate during the asymptomatic infection stage
+##' @param beta_p transmission rate during the presymptomatic infection stage
+##' @param beta_s transmission rate during the symptomatic infection stage
+##' @param k negative binomial over-dispersion parameter
+##' @param epsilon within- and between-group mixing rates (n x n matrix representing contact from ith row -> jth column)
+##' @param theta infection from outside
+##' @param tmax final time
+##' @param internalstart internal test starting time
+##' @param teststart mass test starting time
+##' @param ntest number of internal tests that can be processed per day
+##' @param sensitivity sensitivity
+##' @param specificity specificity
+##' @param qperiod quarantine period
+##' @param qcontact (TRUE) infected individuals can "contact" quarantined individuals; (FALSE) infected individuals cannot "contact" quarantined individuals; analogous to densitiy and frequency dependence
+##' @param preport probability of self-reporting symptoms and getting tested (and isolate until the test result comes back)
+##' @param trace (logical) contact trace?
+##' @param traceeff tracing efficacy
+##' @param tracedelay days
+##' @param tracememory how far can we go back
+simulate_internal <- function(state,
+                              vacc,
+                              veff,
+                              return,
+                              returnfrac=1/7,
+                              sigma,
+                              p_a, gamma_a1, gamma_a2,
+                              gamma_p1, gamma_p2, gamma_s1, gamma_s2,
[…]

### 2022-07-28 R/simulate_internal_omicron.R

@@ -1,0 +1,631 @@
+##' @param state (list) initial state of individuals (X_t^{(i)} in the Appendix)
+##' @param return (list) initial return state of individuals (X_t^{(i)} in the Appendix)
+##' @param sigma transition rate from the exposed stage
+##' @param p_a probability of asymptomatic infection
+##' @param gamma_a transition rate from the asymptomatic infection stage to the recovered stage
+##' @param gamma_p transition rate from the presymptomatic infection stage to the symptomatic infection stage
+##' @param gamma_s transition rate from the symptomatic infection stage to the recovered stage
+##' @param beta_a transmission rate during the asymptomatic infection stage
+##' @param beta_p transmission rate during the presymptomatic infection stage
+##' @param beta_s transmission rate during the symptomatic infection stage
+##' @param k negative binomial over-dispersion parameter
+##' @param epsilon within- and between-group mixing rates (n x n matrix representing contact from ith row -> jth column)
+##' @param theta infection from outside
+##' @param tmax final time
+##' @param internalstart internal test starting time
+##' @param teststart mass test starting time
+##' @param ntest number of internal tests that can be processed per day
+##' @param sensitivity sensitivity
+##' @param specificity specificity
+##' @param qperiod quarantine period
+##' @param qcontact (TRUE) infected individuals can "contact" quarantined individuals; (FALSE) infected individuals cannot "contact" quarantined individuals; analogous to densitiy and frequency dependence
+##' @param preport probability of self-reporting symptoms and getting tested (and isolate until the test result comes back)
+##' @param trace (logical) contact trace?
+##' @param traceeff tracing efficacy
+##' @param tracedelay days
+##' @param tracememory how far can we go back
+simulate_internal_omicron <- function(state,
+                                      vacc,
+                                      boost,
+                                      veff_twodose,
+                                      veff_threedose,
+                                      veff_twodose_inf,
+                                      veff_threedose_inf,
+                                      vaccine_delay,
+                                      veff,
[…]

### 2022-07-28 R/simulate_omicron.R

@@ -1,0 +1,183 @@
+drawpar <- function(D_e=2,
+                    D_p=3,
+                    D_s=3,
+                    p_a=0.4,
+                    R0=2,
+                    p_p=0.5, ## proportion of presymptomatic transmission for symptomatic individuals
+                    k=0.1,
+                    preport=1,
+                    delta=0.6/140) {
+  
+  c(
+    D_e=D_e,
+    D_p=D_p,
+    D_s=D_s,
+    p_a=p_a,
+    R0=R0,
+    p_p=p_p,
+    k=k,
+    preport=preport,
+    delta=delta
+  )
+}
+
+##' @param x parameter (from `drawpar()`)
+convertpar <- function(x) {
+  with(as.list(x), {
+    ## (1-p_p) * R0 = beta * D_s
+    ##
+    ##
+    
+    beta <- R0 * (1 - p_p)/D_s
+    Rs <- D_s * beta
+    
+    Rp <- p_p/(1-p_p) * Rs
+    
+    beta_p <- Rp/D_p
+    
+    list(
+      sigma=-log(1-1/D_e), 
+      gamma_a1=-log(1-2/D_s),
+      gamma_a2=-log(1-2/D_s),
+      gamma_p1=-log(1-2/D_p), 
+      gamma_p2=-log(1-2/D_p), 
+      gamma_s1=-log(1-2/D_s),
+      gamma_s2=-log(1-2/D_s),
+      p_a=p_a, 
+      beta=beta,
+      beta_p=beta_p,
+      k=k,
+      preport=preport,
+      delta=delta
+    )
+  })
+}
+
+initialize <- function(param,
+                       N,
+                       I,
+                       R) {
+  with(as.list(param), {
+    .state <- c("S", "E", "Ip1", "Ip2", "Ia1", "Ia2", "Is1", "Is2", "R")
+    
+    distrb <- c(rmultinom(1, N, c(1-I-R, I, R)))
+    
+    prob <- c(D_e, D_p/2, D_p/2, D_s/2, D_s/2, 0, 0)
+    
+    prob <- prob/sum(prob)
+    
+    state <- rmultinom(1, distrb[2], prob)
+    
+    final <- c(distrb[1], state, distrb[3])
+    
+    sample(unname(unlist(mapply(function(x, y) rep(x, y), .state, final))))
+  })
+}
+
+simulate_omicron <- function(param=drawpar(),
+                     vprop=0.99,
+                     boostprop=0.65,
[…]

### 2022-07-28 simulation_omicron_spring/simulation_omicron_spring.R

@@ -1,0 +1,75 @@
+library(dplyr)
+source("../R/simulate_internal_omicron_switch.R")
+source("../R/simulate_omicron_switch.R")
+
+county <- read.csv("../data/us-counties-04-01-2022.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+mercer_spring2021 <- mercer %>%
+  filter(date >= as.Date("2022-01-08"), date <= as.Date("2022-03-25"))
+
+nsim <- 100
+
+simulation_omicron_spring <- vector('list')
+
+theta <- 0.15 * mercer_spring2021$cases * 5000/13000
+
+R0vec <- c(2, 4, 6, 8)
+
+paramvec <- expand.grid(R0vec)
+
+for (j in 1:nrow(paramvec)) {
+  tmplist <- vector('list')
+  
+  set.seed(101)
+  for (i in 1:nsim) {
+    ## 5000 undergrad
+    print(paste0(j, ", ", i))
+    ss <- simulate_omicron_switch(param=drawpar(preport=1, R0=paramvec[j, 1]),
+                           N=5000,
+                           I=0.14,
+                           R=100/5000,
+                           return=c(rep(TRUE, 700), rep(FALSE, 4300)),
+                           returnfrac=1/28,
+                           datestart=as.Date("2022-01-01"),
+                           datesemester=as.Date("2022-01-01"),
+                           dateinternal=as.Date("2022-01-01"),
+                           dateswitch=as.Date("2022-02-08"),
+                           dateend=as.Date("2022-03-18"),
+                           thetavec=theta,
+                           testfreq_unvacc = 3,
+                           testfreq_vacc = 3,
+                           testfreq_new=7,
+                           vprop=0.99,
+                           boostprop=0.6,
+                           boostdaily=70,
+                           veff_twodose=0.1,
+                           veff_threedose=0.7,
+                           veff_twodose_inf = 0.1 * 0.2/0.9,
+                           veff_threedose_inf = 0.7 * 0.2/0.9,
+                           vaccine_delay=7,
+                           iperiod=5,
+                           ntest = 8000,
+                           asymp_testing=TRUE,
+                           return_testing=FALSE,
[…]

### 2022-07-28 simulation_omicron_spring/simulation_omicron_spring_increase_10.R

@@ -1,0 +1,76 @@
+library(dplyr)
+source("../R/simulate_internal_omicron_switch2.R")
+source("../R/simulate_omicron_switch2.R")
+
+county <- read.csv("../data/us-counties-03-22-2022.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+mercer_spring2021 <- mercer %>%
+  filter(date > as.Date("2021-12-31"), date <= as.Date("2022-03-18"))
+
+nsim <- 100
+
+simulation_omicron_spring_increase_10 <- vector('list')
+
+theta <- 0.1 * mercer_spring2021$cases * 5000/13000
+
+R0vec <- c(2, 4, 6, 8)
+
+paramvec <- expand.grid(R0vec)
+
+for (j in 1:nrow(paramvec)) {
+  tmplist <- vector('list')
+  
+  set.seed(101)
+  for (i in 1:nsim) {
+    ## 5000 undergrad
+    print(paste0(j, ", ", i))
+    ss <- simulate_omicron_switch2(param=drawpar(preport=1, R0=paramvec[j, 1]),
+                           N=5000,
+                           I=0,
+                           R=100/5000,
+                           return=c(rep(TRUE, 700), rep(FALSE, 4300)),
+                           returnfrac=1/28,
+                           datestart=as.Date("2022-01-01"),
+                           datesemester=as.Date("2022-01-01"),
+                           dateinternal=as.Date("2022-01-01"),
+                           dateswitch=as.Date("2022-02-08"),
+                           dateend=as.Date("2022-03-18"),
+                           thetavec=theta,
+                           testfreq_unvacc = 3,
+                           testfreq_vacc = 3,
+                           testfreq_new=7,
+                           Rnewratio=1.1,
+                           vprop=0.99,
+                           boostprop=0.6,
+                           boostdaily=30,
+                           veff_twodose=0.1,
+                           veff_threedose=0.7,
+                           veff_twodose_inf = 0.1 * 0.2/0.9,
+                           veff_threedose_inf = 0.7 * 0.2/0.9,
+                           vaccine_delay=7,
+                           iperiod=5,
+                           ntest = 8000,
+                           asymp_testing=TRUE,
[…]

### 2022-07-28 simulation_omicron_spring/simulation_omicron_spring_increase_100.R

@@ -1,0 +1,76 @@
+library(dplyr)
+source("../R/simulate_internal_omicron_switch2.R")
+source("../R/simulate_omicron_switch2.R")
+
+county <- read.csv("../data/us-counties-04-01-2022.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+mercer_spring2021 <- mercer %>%
+  filter(date >= as.Date("2022-01-08"), date <= as.Date("2022-03-25"))
+
+nsim <- 100
+
+simulation_omicron_spring_increase_100 <- vector('list')
+
+theta <- 0.15 * mercer_spring2021$cases * 5000/13000
+
+R0vec <- c(2, 4, 6)
+
+paramvec <- expand.grid(R0vec)
+
+for (j in 1:nrow(paramvec)) {
+  tmplist <- vector('list')
+  
+  set.seed(101)
+  for (i in 1:nsim) {
+    ## 5000 undergrad
+    print(paste0(j, ", ", i))
+    ss <- simulate_omicron_switch2(param=drawpar(preport=1, R0=paramvec[j, 1]),
+                           N=5000,
+                           I=0.14,
+                           R=100/5000,
+                           return=c(rep(TRUE, 700), rep(FALSE, 4300)),
+                           returnfrac=1/28,
+                           datestart=as.Date("2022-01-01"),
+                           datesemester=as.Date("2022-01-01"),
+                           dateinternal=as.Date("2022-01-01"),
+                           dateswitch=as.Date("2022-02-08"),
+                           dateend=as.Date("2022-03-18"),
+                           thetavec=theta,
+                           testfreq_unvacc = 3,
+                           testfreq_vacc = 3,
+                           testfreq_new=7,
+                           Rnewratio=2,
+                           vprop=0.99,
+                           boostprop=0.6,
+                           boostdaily=70,
+                           veff_twodose=0.1,
+                           veff_threedose=0.7,
+                           veff_twodose_inf = 0.1 * 0.2/0.9,
+                           veff_threedose_inf = 0.7 * 0.2/0.9,
+                           vaccine_delay=7,
+                           iperiod=5,
+                           ntest = 8000,
+                           asymp_testing=TRUE,
[…]

### 2022-07-28 simulation_omicron_spring/simulation_omicron_spring_increase_20.R

@@ -1,0 +1,76 @@
+library(dplyr)
+source("../R/simulate_internal_omicron_switch2.R")
+source("../R/simulate_omicron_switch2.R")
+
+county <- read.csv("../data/us-counties-04-01-2022.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+mercer_spring2021 <- mercer %>%
+  filter(date >= as.Date("2022-01-08"), date <= as.Date("2022-03-25"))
+
+nsim <- 100
+
+simulation_omicron_spring_increase_20 <- vector('list')
+
+theta <- 0.15 * mercer_spring2021$cases * 5000/13000
+
+R0vec <- c(2, 4, 6, 8)
+
+paramvec <- expand.grid(R0vec)
+
+for (j in 1:nrow(paramvec)) {
+  tmplist <- vector('list')
+  
+  set.seed(101)
+  for (i in 1:nsim) {
+    ## 5000 undergrad
+    print(paste0(j, ", ", i))
+    ss <- simulate_omicron_switch2(param=drawpar(preport=1, R0=paramvec[j, 1]),
+                           N=5000,
+                           I=0.14,
+                           R=100/5000,
+                           return=c(rep(TRUE, 700), rep(FALSE, 4300)),
+                           returnfrac=1/28,
+                           datestart=as.Date("2022-01-01"),
+                           datesemester=as.Date("2022-01-01"),
+                           dateinternal=as.Date("2022-01-01"),
+                           dateswitch=as.Date("2022-02-08"),
+                           dateend=as.Date("2022-03-18"),
+                           thetavec=theta,
+                           testfreq_unvacc = 3,
+                           testfreq_vacc = 3,
+                           testfreq_new=7,
+                           Rnewratio=1.2,
+                           vprop=0.99,
+                           boostprop=0.6,
+                           boostdaily=70,
+                           veff_twodose=0.1,
+                           veff_threedose=0.7,
+                           veff_twodose_inf = 0.1 * 0.2/0.9,
+                           veff_threedose_inf = 0.7 * 0.2/0.9,
+                           vaccine_delay=7,
+                           iperiod=5,
+                           ntest = 8000,
+                           asymp_testing=TRUE,
[…]

### 2022-07-28 simulation_omicron_spring/simulation_omicron_spring_increase_50.R

@@ -1,0 +1,76 @@
+library(dplyr)
+source("../R/simulate_internal_omicron_switch2.R")
+source("../R/simulate_omicron_switch2.R")
+
+county <- read.csv("../data/us-counties-04-01-2022.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+mercer_spring2021 <- mercer %>%
+  filter(date >= as.Date("2022-01-08"), date <= as.Date("2022-03-25"))
+
+nsim <- 100
+
+simulation_omicron_spring_increase_50 <- vector('list')
+
+theta <- 0.15 * mercer_spring2021$cases * 5000/13000
+
+R0vec <- c(2, 4, 6, 8)
+
+paramvec <- expand.grid(R0vec)
+
+for (j in 1:nrow(paramvec)) {
+  tmplist <- vector('list')
+  
+  set.seed(101)
+  for (i in 1:nsim) {
+    ## 5000 undergrad
+    print(paste0(j, ", ", i))
+    ss <- simulate_omicron_switch2(param=drawpar(preport=1, R0=paramvec[j, 1]),
+                           N=5000,
+                           I=0.14,
+                           R=100/5000,
+                           return=c(rep(TRUE, 700), rep(FALSE, 4300)),
+                           returnfrac=1/28,
+                           datestart=as.Date("2022-01-01"),
+                           datesemester=as.Date("2022-01-01"),
+                           dateinternal=as.Date("2022-01-01"),
+                           dateswitch=as.Date("2022-02-08"),
+                           dateend=as.Date("2022-03-18"),
+                           thetavec=theta,
+                           testfreq_unvacc = 3,
+                           testfreq_vacc = 3,
+                           testfreq_new=7,
+                           Rnewratio=1.5,
+                           vprop=0.99,
+                           boostprop=0.6,
+                           boostdaily=70,
+                           veff_twodose=0.1,
+                           veff_threedose=0.7,
+                           veff_twodose_inf = 0.1 * 0.2/0.9,
+                           veff_threedose_inf = 0.7 * 0.2/0.9,
+                           vaccine_delay=7,
+                           iperiod=5,
+                           ntest = 8000,
+                           asymp_testing=TRUE,
[…]

### 2022-07-28 simulation_omicron_spring/simulation_omicron_spring_noswitch.R

@@ -1,0 +1,75 @@
+library(dplyr)
+source("../R/simulate_internal_omicron_switch.R")
+source("../R/simulate_omicron_switch.R")
+
+county <- read.csv("../data/us-counties-04-01-2022.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+mercer_spring2021 <- mercer %>%
+  filter(date >= as.Date("2022-01-08"), date <= as.Date("2022-03-25"))
+
+nsim <- 100
+
+simulation_omicron_spring_noswitch <- vector('list')
+
+theta <- 0.15 * mercer_spring2021$cases * 5000/13000
+
+R0vec <- c(2, 4, 6, 8)
+
+paramvec <- expand.grid(R0vec)
+
+for (j in 1:nrow(paramvec)) {
+  tmplist <- vector('list')
+  
+  set.seed(101)
+  for (i in 1:nsim) {
+    ## 5000 undergrad
+    print(paste0(j, ", ", i))
+    ss <- simulate_omicron_switch(param=drawpar(preport=1, R0=paramvec[j, 1]),
+                           N=5000,
+                           I=0.14,
+                           R=100/5000,
+                           return=c(rep(TRUE, 700), rep(FALSE, 4300)),
+                           returnfrac=1/28,
+                           datestart=as.Date("2022-01-01"),
+                           datesemester=as.Date("2022-01-01"),
+                           dateinternal=as.Date("2022-01-01"),
+                           dateswitch=as.Date("2022-02-08"),
+                           dateend=as.Date("2022-03-18"),
+                           thetavec=theta,
+                           testfreq_unvacc = 3,
+                           testfreq_vacc = 3,
+                           testfreq_new=3,
+                           vprop=0.99,
+                           boostprop=0.6,
+                           boostdaily=70,
+                           veff_twodose=0.1,
+                           veff_threedose=0.7,
+                           veff_twodose_inf = 0.1 * 0.2/0.9,
+                           veff_threedose_inf = 0.7 * 0.2/0.9,
+                           vaccine_delay=7,
+                           iperiod=5,
+                           ntest = 8000,
+                           asymp_testing=TRUE,
+                           return_testing=FALSE,
[…]

### 2022-07-28 simulation_omicron_spring/simulation_omicron_spring_null.R

@@ -1,0 +1,75 @@
+library(dplyr)
+source("../R/simulate_internal_omicron_switch.R")
+source("../R/simulate_omicron_switch.R")
+
+county <- read.csv("../data/us-counties-03-22-2022.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+mercer_spring2021 <- mercer %>%
+  filter(date > as.Date("2021-12-31"), date <= as.Date("2022-03-18"))
+
+nsim <- 100
+
+simulation_omicron_spring_null <- vector('list')
+
+theta <- 0.1 * mercer_spring2021$cases * 5000/13000
+
+R0vec <- c(0)
+
+paramvec <- expand.grid(R0vec)
+
+for (j in 1:nrow(paramvec)) {
+  tmplist <- vector('list')
+  
+  set.seed(101)
+  for (i in 1:nsim) {
+    ## 5000 undergrad
+    print(paste0(j, ", ", i))
+    ss <- simulate_omicron_switch(param=drawpar(preport=1, R0=paramvec[j, 1]),
+                           N=5000,
+                           I=0,
+                           R=100/5000,
+                           return=c(rep(TRUE, 700), rep(FALSE, 4300)),
+                           returnfrac=1/28,
+                           datestart=as.Date("2022-01-01"),
+                           datesemester=as.Date("2022-01-01"),
+                           dateinternal=as.Date("2022-01-01"),
+                           dateswitch=as.Date("2022-02-08"),
+                           dateend=as.Date("2022-03-18"),
+                           thetavec=theta,
+                           testfreq_unvacc = 3,
+                           testfreq_vacc = 3,
+                           testfreq_new=7,
+                           vprop=0.99,
+                           boostprop=0.6,
+                           boostdaily=30,
+                           veff_twodose=0.1,
+                           veff_threedose=0.7,
+                           veff_twodose_inf = 0.1 * 0.2/0.9,
+                           veff_threedose_inf = 0.7 * 0.2/0.9,
+                           vaccine_delay=7,
+                           iperiod=5,
+                           ntest = 8000,
+                           asymp_testing=TRUE,
+                           return_testing=FALSE,
+                           trace=FALSE,
[…]

### 2022-07-28 simulation_omicron_spring/simulation_omicron_spring_outbreak.R

@@ -1,0 +1,79 @@
+library(dplyr)
+source("../R/simulate_internal_omicron_switch_outbreak.R")
+source("../R/simulate_omicron_switch_outbreak.R")
+
+county <- read.csv("../data/us-counties-04-01-2022.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+mercer_spring2021 <- mercer %>%
+  filter(date >= as.Date("2022-01-08"), date <= as.Date("2022-03-25"))
+
+nsim <- 100
+
+simulation_omicron_spring_outbreak <- vector('list')
+
+theta <- 0.15 * mercer_spring2021$cases * 5000/13000
+
+R0vec <- c(2, 4, 6)
+outbreaksize <- c(100, 150, 200, 250, 300)
+
+paramvec <- expand.grid(R0vec, outbreaksize)
+
+for (j in 1:nrow(paramvec)) {
+  tmplist <- vector('list')
+  
+  set.seed(101)
+  for (i in 1:nsim) {
+    ## 5000 undergrad
+    print(paste0(j, ", ", i))
+    ss <- simulate_omicron_switch_outbreak(param=drawpar(preport=1, R0=paramvec[j, 1]),
+                           N=5000,
+                           I=0.14,
+                           R=100/5000,
+                           return=c(rep(TRUE, 700), rep(FALSE, 4300)),
+                           returnfrac=1/28,
+                           datestart=as.Date("2022-01-01"),
+                           datesemester=as.Date("2022-01-01"),
+                           dateinternal=as.Date("2022-01-01"),
+                           dateswitch=as.Date("2022-02-08"),
+                           dateend=as.Date("2022-03-18"),
+                           dateoutbreak = as.Date("2022-02-12"),
+                           outbreaksize=paramvec[j,2],
+                           thetavec=theta,
+                           testfreq_unvacc = 3,
+                           testfreq_vacc = 3,
+                           testfreq_new=7,
+                           vprop=0.99,
+                           boostprop=0.6,
+                           boostdaily=70,
+                           veff_twodose=0.1,
+                           veff_threedose=0.7,
+                           veff_twodose_inf = 0.1 * 0.2/0.9,
+                           veff_threedose_inf = 0.7 * 0.2/0.9,
[…]

### 2022-07-28 simulation_princeton/simulation_fall_2020.R

@@ -1,0 +1,73 @@
+library(dplyr)
+library(lubridate)
+source("../R/simulate_internal.R")
+source("../R/simulate.R")
+
+county <- read.csv("../data/us-counties-12-30-2021.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+mercer_fall2020 <- mercer %>%
+  filter(date >= "2020-08-30", date<= "2021-01-08")
+
+mercer_fall2020$cases[128] <- (mercer_fall2020$cases[127]+mercer_fall2020$cases[129])/2
+
+nsim <- 100
+
+simulation_fall_2020 <- vector('list')
+
+R0vec <- c(0.5, 1, 2)
+scalevec <- c(0.01, 0.015, 0.02, 0.025, 0.03)
+
+paramvec <- expand.grid(R0vec, scalevec)
+
+for (j in 1:nrow(paramvec)) {
+  tmplist <- vector('list')
+  
+  set.seed(101)
+  for (i in 1:nsim) {
+    ## 1000 grad and 
+    ## 2000 staff
+    ## and a small number of undergrads
+    ## assuming 3000 for simplicity
+    ## trick to change testing... 
+    ## assume 1/3 are vaccinated with no efficacy but assume vaccinated are tested 
+    ## at different rates
+    print(paste0(j, ", ", i))
+    ss <- simulate(param=drawpar(preport=1, R0=paramvec[j, 1]),
+                   N=3000,
+                   I=0,
+                   return=rep(TRUE, 3000),
+                   returnfrac=1,
+                   thetavec=paramvec[j,2] * mercer_fall2020$cases,
+                   datestart=as.Date("2020-08-23"),
+                   datesemester=as.Date("2020-08-31"),
+                   dateinternal=as.Date("2020-08-31"),
+                   dateend=as.Date("2021-01-01"),
+                   testfreq_unvacc = 7,
+                   testfreq_vacc = 3,
+                   ntest = 8000,
+                   vprop=0.33, veff=0,
+                   asymp_testing=TRUE,
+                   return_testing=FALSE,
+                   trace=FALSE,
+                   returnq=0)
+    
+    ss2 <- ss$res
+    
[…]

### 2022-07-28 simulation_princeton/simulation_fall_2021.R

@@ -1,0 +1,75 @@
+library(dplyr)
+library(lubridate)
+source("../R/simulate_internal_omicron.R")
+source("../R/simulate_omicron.R")
+
+county <- read.csv("../data/us-counties-01-18-2022.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+mercer_fall2021 <- mercer %>%
+  filter(date >= "2021-08-21", date<= "2022-01-14")
+
+nsim <- 100
+
+simulation_fall_2021 <- vector('list')
+
+R0vec <- c(0.5, 1, 2, 4, 8)
+scalevec <- c(0.005, 0.01, 0.015, 0.02, 0.025) * 10
+
+paramvec <- expand.grid(R0vec, scalevec)
+
+for (j in 1:nrow(paramvec)) {
+  tmplist <- vector('list')
+  
+  set.seed(101)
+  for (i in 1:nsim) {
+    ## 5000 undergrad
+    ## 2000 grad
+    ## 6000 staff and faculty
+    ## turn off omicron parameters
+    print(paste0(j, ", ", i))
+    ss <- simulate_omicron(param=drawpar(preport=1, R0=paramvec[j, 1]),
+                   N=13000,
+                   I=0.005,
+                   vprop=0.98,
+                   boostprop=0,
+                   veff_twodose=0.9,
+                   veff_threedose=0,
+                   veff_twodose_inf=0.2,
+                   veff_threedose_inf=0,
+                   vaccine_delay=14,
+                   boostdaily=0,
+                   return=c(rep(FALSE, 5000), rep(TRUE, 8000)),
+                   returnfrac=1/16,
+                   thetavec=paramvec[j,2] * mercer_fall2021$cases,
+                   datestart=as.Date("2021-08-14"),
+                   datesemester=as.Date("2021-09-01"),
+                   dateinternal=as.Date("2021-09-01"),
+                   dateend=as.Date("2022-01-07"),
+                   testfreq_unvacc = 3,
+                   testfreq_vacc = 7,
+                   ntest = 8000,
+                   asymp_testing=TRUE,
+                   return_testing=TRUE,
+                   trace=FALSE,
+                   returnq=0)
[…]

### 2022-07-28 simulation_princeton/simulation_fall_2021_nowaning.R

@@ -1,0 +1,75 @@
+library(dplyr)
+library(lubridate)
+source("../R/simulate_internal_omicron.R")
+source("../R/simulate_omicron.R")
+
+county <- read.csv("../data/us-counties-01-18-2022.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+mercer_fall2021 <- mercer %>%
+  filter(date >= "2021-08-21", date<= "2022-01-14")
+
+nsim <- 100
+
+simulation_fall_2021_nowaning <- vector('list')
+
+R0vec <- c(0.5, 1, 2, 4, 8)
+scalevec <- c(0.005, 0.01, 0.015, 0.02, 0.025) * 10
+
+paramvec <- expand.grid(R0vec, scalevec)
+
+for (j in 1:nrow(paramvec)) {
+  tmplist <- vector('list')
+  
+  set.seed(101)
+  for (i in 1:nsim) {
+    ## 5000 undergrad
+    ## 2000 grad
+    ## 6000 staff and faculty
+    ## turn off omicron parameters
+    print(paste0(j, ", ", i))
+    ss <- simulate_omicron(param=drawpar(preport=1, R0=paramvec[j, 1], delta=0),
+                   N=13000,
+                   I=0.005,
+                   vprop=0.98,
+                   boostprop=0,
+                   veff_twodose=0.9,
+                   veff_threedose=0,
+                   veff_twodose_inf=0.2,
+                   veff_threedose_inf=0,
+                   vaccine_delay=14,
+                   boostdaily=0,
+                   return=c(rep(FALSE, 5000), rep(TRUE, 8000)),
+                   returnfrac=1/16,
+                   thetavec=paramvec[j,2] * mercer_fall2021$cases,
+                   datestart=as.Date("2021-08-14"),
+                   datesemester=as.Date("2021-09-01"),
+                   dateinternal=as.Date("2021-09-01"),
+                   dateend=as.Date("2022-01-07"),
+                   testfreq_unvacc = 3,
+                   testfreq_vacc = 7,
+                   ntest = 8000,
+                   asymp_testing=TRUE,
+                   return_testing=TRUE,
+                   trace=FALSE,
[…]

### 2022-07-28 simulation_princeton/simulation_spring_2020.R

@@ -1,0 +1,72 @@
+library(dplyr)
+library(lubridate)
+source("../R/simulate_internal.R")
+source("../R/simulate.R")
+
+county <- read.csv("../data/us-counties-12-30-2021.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+mercer_spring2020 <- mercer %>%
+  filter(date >= "2021-01-23", date<= "2021-05-21")
+
+nsim <- 100
+
+simulation_spring_2020 <- vector('list')
+
+R0vec <- c(0.5, 1, 2)
+scalevec <- c(0.01, 0.015, 0.02, 0.025, 0.03)
+
+paramvec <- expand.grid(R0vec, scalevec)
+
+for (j in 1:nrow(paramvec)) {
+  tmplist <- vector('list')
+  
+  set.seed(101)
+  for (i in 1:nsim) {
+    ## 3000 undergrad
+    ## 2000 grad
+    ## 3000 staff
+    ## assuming 8000 for simplicity
+    ## trick to change testing... 
+    ## assume 5/8 are vaccinated with no efficacy but assume vaccinated are tested 
+    ## at different rates
+    print(paste0(j, ", ", i))
+    ss <- simulate(param=drawpar(preport=1, R0=paramvec[j,1]),
+                   N=8000,
+                   I=0.01,
+                   return=c(rep(FALSE, 3000), rep(FALSE, 1000), rep(TRUE, 1000), rep(TRUE, 3000)),
+                   returnfrac=1/14,
+                   thetavec=paramvec[j,2] * mercer_spring2020$cases,
+                   datestart=as.Date("2021-01-16"),
+                   datesemester=as.Date("2021-02-01"),
+                   dateinternal=as.Date("2021-02-01"),
+                   dateend=as.Date("2021-05-14"),
+                   testfreq_unvacc = 7,
+                   testfreq_vacc = 3,
+                   ntest = 8000,
+                   vacc=c(rep(TRUE, 5000), rep(FALSE, 3000)),
+                   vprop=5/8, veff=0,
+                   asymp_testing=TRUE,
+                   return_testing=TRUE,
+                   trace=FALSE,
+                   returnq=1)
+    
[…]

### 2022-07-29 R/simulate_internal_omicron_switch.R
more files

@@ -1,0 +1,642 @@
+##' @param state (list) initial state of individuals (X_t^{(i)} in the Appendix)
+##' @param return (list) initial return state of individuals (X_t^{(i)} in the Appendix)
+##' @param sigma transition rate from the exposed stage
+##' @param p_a probability of asymptomatic infection
+##' @param gamma_a transition rate from the asymptomatic infection stage to the recovered stage
+##' @param gamma_p transition rate from the presymptomatic infection stage to the symptomatic infection stage
+##' @param gamma_s transition rate from the symptomatic infection stage to the recovered stage
+##' @param beta_a transmission rate during the asymptomatic infection stage
+##' @param beta_p transmission rate during the presymptomatic infection stage
+##' @param beta_s transmission rate during the symptomatic infection stage
+##' @param k negative binomial over-dispersion parameter
+##' @param epsilon within- and between-group mixing rates (n x n matrix representing contact from ith row -> jth column)
+##' @param theta infection from outside
+##' @param tmax final time
+##' @param internalstart internal test starting time
+##' @param teststart mass test starting time
+##' @param ntest number of internal tests that can be processed per day
+##' @param sensitivity sensitivity
+##' @param specificity specificity
+##' @param qperiod quarantine period
+##' @param qcontact (TRUE) infected individuals can "contact" quarantined individuals; (FALSE) infected individuals cannot "contact" quarantined individuals; analogous to densitiy and frequency dependence
+##' @param preport probability of self-reporting symptoms and getting tested (and isolate until the test result comes back)
+##' @param trace (logical) contact trace?
+##' @param traceeff tracing efficacy
+##' @param tracedelay days
+##' @param tracememory how far can we go back
+simulate_internal_omicron_switch <- function(state,
+                                      vacc,
+                                      boost,
+                                      veff_twodose,
+                                      veff_threedose,
+                                      veff_twodose_inf,
+                                      veff_threedose_inf,
+                                      vaccine_delay,
+                                      veff,
+                                      veff_inf,
[…]

### 2022-07-29 R/simulate_internal_omicron_switch_outbreak.R

@@ -1,0 +1,655 @@
+##' @param state (list) initial state of individuals (X_t^{(i)} in the Appendix)
+##' @param return (list) initial return state of individuals (X_t^{(i)} in the Appendix)
+##' @param sigma transition rate from the exposed stage
+##' @param p_a probability of asymptomatic infection
+##' @param gamma_a transition rate from the asymptomatic infection stage to the recovered stage
+##' @param gamma_p transition rate from the presymptomatic infection stage to the symptomatic infection stage
+##' @param gamma_s transition rate from the symptomatic infection stage to the recovered stage
+##' @param beta_a transmission rate during the asymptomatic infection stage
+##' @param beta_p transmission rate during the presymptomatic infection stage
+##' @param beta_s transmission rate during the symptomatic infection stage
+##' @param k negative binomial over-dispersion parameter
+##' @param epsilon within- and between-group mixing rates (n x n matrix representing contact from ith row -> jth column)
+##' @param theta infection from outside
+##' @param tmax final time
+##' @param internalstart internal test starting time
+##' @param teststart mass test starting time
+##' @param ntest number of internal tests that can be processed per day
+##' @param sensitivity sensitivity
+##' @param specificity specificity
+##' @param qperiod quarantine period
+##' @param qcontact (TRUE) infected individuals can "contact" quarantined individuals; (FALSE) infected individuals cannot "contact" quarantined individuals; analogous to densitiy and frequency dependence
+##' @param preport probability of self-reporting symptoms and getting tested (and isolate until the test result comes back)
+##' @param trace (logical) contact trace?
+##' @param traceeff tracing efficacy
+##' @param tracedelay days
+##' @param tracememory how far can we go back
+simulate_internal_omicron_switch_outbreak <- function(state,
+                                      vacc,
+                                      boost,
+                                      veff_twodose,
+                                      veff_threedose,
+                                      veff_twodose_inf,
+                                      veff_threedose_inf,
+                                      vaccine_delay,
+                                      veff,
[…]

### 2022-07-29 R/simulate_omicron_switch.R

@@ -1,0 +1,187 @@
+drawpar <- function(D_e=2,
+                    D_p=3,
+                    D_s=3,
+                    p_a=0.4,
+                    R0=2,
+                    p_p=0.5, ## proportion of presymptomatic transmission for symptomatic individuals
+                    k=0.1,
+                    preport=1,
+                    delta=0.6/140) {
+  
+  c(
+    D_e=D_e,
+    D_p=D_p,
+    D_s=D_s,
+    p_a=p_a,
+    R0=R0,
+    p_p=p_p,
+    k=k,
+    preport=preport,
+    delta=delta
+  )
+}
+
+##' @param x parameter (from `drawpar()`)
+convertpar <- function(x) {
+  with(as.list(x), {
+    ## (1-p_p) * R0 = beta * D_s
+    ##
+    ##
+    
+    beta <- R0 * (1 - p_p)/D_s
+    Rs <- D_s * beta
+    
+    Rp <- p_p/(1-p_p) * Rs
+    
+    beta_p <- Rp/D_p
+    
+    list(
+      sigma=-log(1-1/D_e), 
+      gamma_a1=-log(1-2/D_s),
+      gamma_a2=-log(1-2/D_s),
+      gamma_p1=-log(1-2/D_p), 
+      gamma_p2=-log(1-2/D_p), 
+      gamma_s1=-log(1-2/D_s),
+      gamma_s2=-log(1-2/D_s),
+      p_a=p_a, 
+      beta=beta,
+      beta_p=beta_p,
+      k=k,
+      preport=preport,
+      delta=delta
+    )
+  })
+}
+
+initialize <- function(param,
+                       N,
+                       I,
+                       R) {
+  with(as.list(param), {
+    .state <- c("S", "E", "Ip1", "Ip2", "Ia1", "Ia2", "Is1", "Is2", "R")
+    
+    distrb <- c(rmultinom(1, N, c(1-I-R, I, R)))
+    
+    prob <- c(D_e, D_p/2, D_p/2, D_s/2, D_s/2, 0, 0)
+    
+    prob <- prob/sum(prob)
+    
+    state <- rmultinom(1, distrb[2], prob)
+    
+    final <- c(distrb[1], state, distrb[3])
+    
+    sample(unname(unlist(mapply(function(x, y) rep(x, y), .state, final))))
+  })
+}
+
+simulate_omicron_switch <- function(param=drawpar(),
+                     vprop=0.99,
+                     boostprop=0.65,
[…]

### 2022-07-29 R/simulate_omicron_switch_outbreak.R

@@ -1,0 +1,193 @@
+drawpar <- function(D_e=2,
+                    D_p=3,
+                    D_s=3,
+                    p_a=0.4,
+                    R0=2,
+                    p_p=0.5, ## proportion of presymptomatic transmission for symptomatic individuals
+                    k=0.1,
+                    preport=1,
+                    delta=0.6/140) {
+  
+  c(
+    D_e=D_e,
+    D_p=D_p,
+    D_s=D_s,
+    p_a=p_a,
+    R0=R0,
+    p_p=p_p,
+    k=k,
+    preport=preport,
+    delta=delta
+  )
+}
+
+##' @param x parameter (from `drawpar()`)
+convertpar <- function(x) {
+  with(as.list(x), {
+    ## (1-p_p) * R0 = beta * D_s
+    ##
+    ##
+    
+    beta <- R0 * (1 - p_p)/D_s
+    Rs <- D_s * beta
+    
+    Rp <- p_p/(1-p_p) * Rs
+    
+    beta_p <- Rp/D_p
+    
+    list(
+      sigma=-log(1-1/D_e), 
+      gamma_a1=-log(1-2/D_s),
+      gamma_a2=-log(1-2/D_s),
+      gamma_p1=-log(1-2/D_p), 
+      gamma_p2=-log(1-2/D_p), 
+      gamma_s1=-log(1-2/D_s),
+      gamma_s2=-log(1-2/D_s),
+      p_a=p_a, 
+      beta=beta,
+      beta_p=beta_p,
+      k=k,
+      preport=preport,
+      delta=delta
+    )
+  })
+}
+
+initialize <- function(param,
+                       N,
+                       I,
+                       R) {
+  with(as.list(param), {
+    .state <- c("S", "E", "Ip1", "Ip2", "Ia1", "Ia2", "Is1", "Is2", "R")
+    
+    distrb <- c(rmultinom(1, N, c(1-I-R, I, R)))
+    
+    prob <- c(D_e, D_p/2, D_p/2, D_s/2, D_s/2, 0, 0)
+    
+    prob <- prob/sum(prob)
+    
+    state <- rmultinom(1, distrb[2], prob)
+    
+    final <- c(distrb[1], state, distrb[3])
+    
+    sample(unname(unlist(mapply(function(x, y) rep(x, y), .state, final))))
+  })
+}
+
+simulate_omicron_switch_outbreak <- function(param=drawpar(),
+                     vprop=0.99,
+                     boostprop=0.65,
[…]

### 2022-07-29 doc/university-covid.tex
sig

@@
 
+\usepackage{lineno}
+\linenumbers
+
 \begin{document}
@@
 Despite large numbers of cases on campus in this period, case levels in surrounding communities remained low, suggesting that there was little spillover transmission from campus to the local community.
+
+\section*{Significance}
+
+University campuses present challenges to preventing SARS-CoV-2 transmission due to a high proportion of asymptomatic infections and high contact rates.
+SARS-CoV-2 outbreaks on the Princeton University campus offer an unusually well-documented perspective, rooted in mass asymptomatic testing, which was further informed by mathematical modeling approaches.
+Here, we test whether this model, which was previously used for guiding policy decisions, can capture the observed outbreak patterns on campus.
+Our analysis reveals strong coupling in epidemic dynamics on campus and in local communities.
+Super-spreading events can drive transmission on campus, thereby weakening the dynamical coupling in campus and community outbreaks.
 

### 2023-01-16 figure_princeton_new/figure_princeton_new_per_1000.R
trying

@@ -1,0 +1,636 @@
+library(dplyr)
+library(tidyr)
+library(readxl)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(gridExtra)
+library(egg)
+library(lubridate)
+
+data_asymp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=1) %>%
+  mutate(
+    type="asymp"
+  )
+
+data_symp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=2) %>%
+  mutate(
+    type="symp"
+  )
+
+data <- bind_rows(
+  data_asymp, data_symp
+)
+
+county <- read.csv("../data/us-counties-03-22-2022.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+fall2020 <- data %>%
+  filter(`Week Ending` <= as.Date("2021-01-01")) %>%
+  select(`Week Ending`, `Undergrad Positive Tests`, `Grad Student Positive Tests`,
+         `Faculty Staff Positive Tests`) %>%
+  rename(
+    date=`Week Ending`,
+    Undergrad=`Undergrad Positive Tests`,
+    Grad=`Grad Student Positive Tests`,
+    `Faculty and Staff`=`Faculty Staff Positive Tests`
+  ) %>%
+  mutate(
+    date=as.Date(date)
+  )  %>%
+  mutate(
+    year=substr(date, 1, 4),
+    month=as.numeric(gsub("-.*", "", gsub("20..-", "", date))), 
+    day=as.numeric(gsub(".*-", "", gsub("20..-", "", date)))
+  ) %>%
+  arrange(date) %>%
+  mutate(
+    date=paste0(month.abb[month], " ", day, ", ", year),
+    date=factor(date, levels=unique(date))
+  ) %>%
+  gather(key, value, -date, -year, -month, -day) %>%
+  mutate(
+    date2=as.Date(paste0(year, "-", month, "-", day))
+  ) %>%
+  group_by(date, year, month, day, key, date2) %>%
+  summarize(
+    value=sum(value)
+  )
+
+spring2020 <- data %>%
+  filter(`Week Ending` <= as.Date("2021-05-14"),
+         `Week Ending` >= as.Date("2021-01-22")) %>%
[…]

### 2023-01-16 figure_princeton_new/figure_princeton_new_per_1000_bypop.R

@@ -20,18 +20,6 @@ data <- bind_rows(
   data_asymp, data_symp
 )
 
-county <- read.csv("../data/us-counties-03-22-2022.csv")
-
-nyc <- county %>%
-  filter(county=="New York City", state=="New York") %>%
-  arrange(date) %>%
-  mutate(
-    cases=c(0, diff(cases)),
-    cases=ifelse(cases < 0, 0, cases)
-  )
-
-nyc$cases[389] <- (nyc$cases[388] + nyc$cases[390])/2
-
 fall2020 <- data %>%
   filter(`Week Ending` <= as.Date("2021-01-01")) %>%
   select(`Week Ending`, `Undergrad Positive Tests`, `Grad Student Positive Tests`,
@@ -43,6 +31,8 @@ fall2020 <- data %>%
     `Faculty and Staff`=`Faculty Staff Positive Tests`
   ) %>%
   mutate(
+    Grad=Grad/1000,
+    `Faculty and Staff`=`Faculty and Staff`/2000,
     date=as.Date(date)
   )  %>%
   mutate(
@@ -76,6 +66,9 @@ spring2020 <- data %>%
     `Faculty and Staff`=`Faculty Staff Positive Tests`
   )  %>%
   mutate(
+    Undergrad=Undergrad/3000,
+    Grad=Grad/2000,
+    `Faculty and Staff`=`Faculty and Staff`/3000,
     date=as.Date(date)
   )  %>%
   mutate(
@@ -109,6 +102,9 @@ fall2021 <- data %>%
     `Faculty and Staff`=`Faculty Staff Positive Tests`
   )  %>%
   mutate(
+    Undergrad=Undergrad/5000,
+    Grad=Grad/2000,
+    `Faculty and Staff`=`Faculty and Staff`/6000,
     date=as.Date(date)
   )  %>%
   mutate(
@@ -142,6 +138,9 @@ spring2021 <- data %>%
     `Faculty and Staff`=`Faculty Staff Positive Tests`
   )  %>%
   mutate(
+    Undergrad=Undergrad/3000,
+    Grad=Grad/2000,
+    `Faculty and Staff`=`Faculty and Staff`/3000,
     date=as.Date(date)
   )  %>%
   mutate(
@@ -163,94 +162,15 @@ spring2021 <- data %>%
     value=sum(value)
   )
 
-nyc_fall2020 <- nyc %>%
[…]

### 2023-01-16 figure_princeton_new/figure_princeton_philly_new.R

@@ -218,12 +218,12 @@ philly_spring2021 <- philly %>%
   )
 
 g1 <- ggplot(fall2020) +
-  geom_bar(aes(date, value, fill=key), stat="identity") +
-  geom_line(data=philly_fall2020, aes(date, cases/400, group=1), col="red") +
-  geom_point(data=philly_fall2020, aes(date, cases/400), col="red") +
- scale_x_discrete("Date") +
-  scale_y_continuous("Princeton cases", expand=c(0, 0), limits=c(0, 50),
-                     sec.axis = sec_axis(trans=~.*400, "New York City cases")) +
+  geom_bar(aes(date, value/3, fill=key), stat="identity") +
+  geom_line(data=philly_fall2020, aes(date, cases/1.6e6*1000, group=1), col="red") +
+  geom_point(data=philly_fall2020, aes(date, cases/1.6e6*1000), col="red") +
+  scale_x_discrete("Date") +
+  scale_y_continuous("Princeton cases per 1000", expand=c(0, 0), limits=c(0, 20),
+                     sec.axis = sec_axis(trans=~., "Philadelphia cases per 1000")) +
   scale_fill_viridis_d() +
   ggtitle("A. Fall 2020-2021") +
   theme(
@@ -241,12 +241,12 @@ g1 <- ggplot(fall2020) +
   )
 
 g2 <- ggplot(spring2020) +
-  geom_bar(aes(as.factor(date), value, fill=key), stat="identity") +
-  geom_line(data=philly_spring2020, aes(date, cases/400, group=1), col="red") +
-  geom_point(data=philly_spring2020, aes(date, cases/400), col="red") +
+  geom_bar(aes(date, value/8, fill=key), stat="identity") +
+  geom_line(data=philly_spring2020, aes(date, cases/1.6e6*1000, group=1), col="red") +
+  geom_point(data=philly_spring2020, aes(date, cases/1.6e6*1000), col="red") +
   scale_x_discrete("Date") +
-  scale_y_continuous("Princeton cases", expand=c(0, 0), limits=c(0, 60),
-                     sec.axis = sec_axis(trans=~.*400, "New York City cases")) +
+  scale_y_continuous("Princeton cases per 1000", expand=c(0, 0), limits=c(0, 10),
+                     sec.axis = sec_axis(trans=~., "Philadelphia cases per 1000")) +
[…]

### 2023-01-16 doc/university-covid-firstsub.tex
more

--- 2022-07-28 doc/university-covid.tex
@@
 Despite large numbers of cases on campus in this period, case levels in surrounding communities remained low, suggesting that there was little spillover transmission from campus to the local community.
+
+\section*{Significance}
+
+University campuses present challenges to preventing SARS-CoV-2 transmission, due to a high proportion of asymptomatic infections and high contact rates.
+SARS-CoV-2 outbreaks on the Princeton University campus offer an unusually well-documented perspective, rooted in mass asymptomatic testing, further informed by mathematical modeling aimed at guiding policy decisions.
+Here, we show that this model can parsimoniously capture observed outbreak patterns on campus during different eras of control.
+Our analysis reveals that strong coupling between epidemic dynamics on campus and in local communities drives the early epidemic.
+Subsequently, especially in the omicron era, superspreading events came to dominate transmission on campus, thereby weakening the dynamical coupling of campus and community outbreaks.
 

### 2023-01-17 doc/university-covid.tex
more changes

--- 2022-07-28 doc/university-covid.tex
@@
 
+\section*{Significance}
+
+University campuses present challenges to preventing SARS-CoV-2 transmission, due to a high proportion of asymptomatic infections and high contact rates.
+SARS-CoV-2 outbreaks on the Princeton University campus offer an unusually well-documented perspective, rooted in mass asymptomatic testing, further informed by mathematical modeling aimed at guiding policy decisions.
+Here, we show that this model can parsimoniously capture observed outbreak patterns on campus during different eras of control.
+Our analysis reveals that strong coupling between epidemic dynamics on campus and in local communities drives the early epidemic.
+Subsequently, especially in the Omicron era, superspreading events came to dominate transmission on campus, thereby weakening the dynamical coupling of campus and community outbreaks.
+
 \pagebreak
@@
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_new.pdf}
+\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_new_per_1000.pdf}
 \caption{
@@
 (A--D) Epidemic trajectories across four semesters: Fall 2020--2021 (A), Spring 2020--2021 (B), Fall 2021--2022 (C), and Spring 2021--2022 (D).
-Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in PU.
-Red lines represent the weekly number of cases in Mercer County.
+Colored bar plots represent the weekly number of cases per 1000 from both asymptomatic and symptomatic testing in PU.
+Red lines represent the weekly number of cases per 1000 in Mercer County.
 Number of cases in Mercer County is obtained from \url{https://github.com/nytimes/covid-19-data}.
-(E--H) Correlations between the weekly number of cases in PU and in Mercer County.
+The weekly number of cases per 1000 in Princeton is normalized by the approximate total size of the PU population present on campus for each semester: 3000 for Fall 2020--2021, 8000 for Spring 2020--2021, 13000 for Fall 2021--2022, and 15000 for Spring 2021--2022.
+The weekly number of cases per 1000 in Princeton in Mercer County is calculated based on the total population size as of 2020: 387340 (\url{www.census.gov}).
+(E--H) Correlations between the weekly number of cases per 1000 in PU and in Mercer County.
 Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
-(I--L) Ratios between weekly numbers of cases reported from PU and Mercer County.
-The shaded orange area covers the region between 1:100 and 1:25 ratios.
+(I--L) Ratios between weekly number of cases per 1000 rin PU and Mercer County.
+The dashed lines represent the 1:1 ratio.
 \label{fig:princeton}
@@
 The number of cases remained relatively low throughout the semester with a peak occurring in early December, coinciding with the epidemic trajectory in Mercer County (\fref{princeton}A).  
-A sudden decrease in the number of cases around Thanksgiving partly reflects the reduced number of tests (3852 and 2972 asymptomatic tests performed on the week ending November 20th and 27th, respectively).
+A sudden decrease in the number of cases around Thanksgiving---a national holiday in the US during which many students---partly reflects the reduced number of tests (3852 and 2972 asymptomatic tests performed on the week ending November 20th and 27th, respectively).
 The highest number of cases was reported among faculty and staff members ($=169$), followed by graduate students ($=41$), and undergraduate students ($=4$).
+Even when we control for the differing population sizes among these groups, we find that a considerably larger amount of cases were reported among faculty and staff members (84.5 cases per 1000) than from graduate students (41 cases per 1000) (Supplementary Figure S1A)---we exclude undergraduate student population from this comparison due to a very low number of undergraduate students present on campus during this period.
 
@@
 The highest number of cases was reported among faculty and staff members ($=111$), followed by undergraduate students ($=101$), and graduate students ($=29$).
+This ordering is robust to differences in population sizes: 37 cases per 1000 among faculty and staff members, 34.3 cases per 1000 among undergraduate students, and 14.5 cases per 1000 among graduate students (Supplementary Figure S1B).
 
@@
 The number of cases decreased slightly as classes ended but soon increased again as the Omicron (BA.1) variant began to spread on campus and in Mercer county.
+The total number of reported cases per 1000 remained high for all three population groups: 74 cases per 1000 among faculty and staff members, 63.5 cases per 1000 among graduate students, and 52.4 cases per 1000 among undergraduate students (Supplementary Figure S1C).
 
@@
 On March 7, 2022, mask mandates were lifted and testing frequency was reduced to once a month.
+Cases were largely concentrated among undergraduate students during this semester: 309 cases per 1000 among undergraduate students, 106 cases per 1000 among faculty and staff members, and 65.3 cases per 1000 among graduate students.
 
@@
 
-Across the first three semesters, we find strong and significant correlations between the weekly logged numbers of cases from PU and those from Mercer county:
-fall 2020--2021 ($\rho = 0.81$ (95\% CI: 0.55--0.92; $p < 0.001$), \fref{princeton}E); spring 2020--2021 ($\rho = 0.80$ (95\% CI: 0.51--0.92; $p < 0.001$), \fref{princeton}F); and fall 2021--2022 ($\rho = 0.86$ (95\% CI: 0.67--0.94; $p < 0.001$), \fref{princeton}G).
-These correlations are robust even when we stratify cases by the population, except for undergraduate students during Fall 2020, when most were not physically present on campus (Supplementary Figure S1).
-However, the case patterns in PU were decoupled from those in Mercer county for the spring semester of the 2021--2022 academic year with unclear correlations (\fref{princeton}H): $\rho = 0.32$ (95\% CI: $-0.34$--0.77; $p = 0.33$).
-Stratifying cases by subpopulation shows that case patterns in graduate students and faculty and staff members were still strongly correlated with case patterns in Mercer county, meaning that the campus transmission was limited to undergraduate students (Supplementary Figure S1).
-We also find strong correlations between the weekly logged numbers of cases from PU and those from other counties in New Jersey (Supplementary Figure S2)---these correlations significantly decreased with distance from Mercer county in both spring 2020--2021 ($\rho=-0.48$ (95\% CI: $-0.75$--$-0.06$; $p = 0.03$) for all cases and $\rho=-0.51$ (95\% CI: $-0.77$--$-0.10$; $p = 0.02$) for faculty and staff cases) and fall 2021--2022 ($\rho=-0.68$ (95\% CI: $-0.86$--$-0.35$, $p < 0.001$) for all cases and $\rho=-0.74$ (95\% CI: $-0.89$--$-0.46$, $p < 0.001$) for faculty and staff cases).
+Across the first three semesters, we find strong and significant correlations between the weekly numbers of cases from PU and those from Mercer county:
+fall 2020--2021 ($\rho = 0.79$ (95\% CI: 0.52--0.91; $p < 0.001$), \fref{princeton}E); spring 2020--2021 ($\rho = 0.84$ (95\% CI: 0.60--0.94; $p < 0.001$), \fref{princeton}F); and fall 2021--2022 ($\rho = 0.93$ (95\% CI: 0.84--0.97; $p < 0.001$), \fref{princeton}G).
+These correlations are robust even when we stratify cases by the population, except for undergraduate students during Fall 2020, when most were not physically present on campus (Supplementary Figure S2).
+However, the case patterns in PU were decoupled from those in Mercer county for the spring semester of the 2021--2022 academic year with unclear correlations (\fref{princeton}H): $\rho = 0.47$ (95\% CI: $-0.18$--0.83; $p = 0.15$).
+Stratifying cases by subpopulation shows that case patterns in graduate students and faculty and staff members were still strongly correlated with case patterns in Mercer county, meaning that the campus transmission was limited to undergraduate students (Supplementary Figure S2).
+We also find strong correlations between the weekly logged numbers of cases from PU and those from other counties in New Jersey (Supplementary Figure S3)---these correlations significantly decreased with distance from Mercer county in both spring 2020--2021 ($\rho=-0.48$ (95\% CI: $-0.75$--$-0.06$; $p = 0.03$) for all cases and $\rho=-0.51$ (95\% CI: $-0.77$--$-0.10$; $p = 0.02$) for faculty and staff cases) and fall 2021--2022 ($\rho=-0.68$ (95\% CI: $-0.86$--$-0.35$, $p < 0.001$) for all cases and $\rho=-0.74$ (95\% CI: $-0.89$--$-0.46$, $p < 0.001$) for faculty and staff cases).
 Across the first three semesters, both the total cases and faculty and staff cases showed similar levels of correlations with local cases.
-For the spring semester of the 2021--2022 academic year, we still find high correlations between faculty and staff cases and local cases throughout other countries ($\rho > 0.8$ across all counties in New Jersey); however, the total cases exhibit considerably weaker correlations due to student-to-student transmission on campus (Supplementary Figure S2).
+For the spring semester of the 2021--2022 academic year, we still find high correlations between faculty and staff cases and local cases throughout other counties ($\rho > 0.8$ across all counties in New Jersey); however, the total cases exhibit considerably weaker correlations due to student-to-student transmission on campus (Supplementary Figure S3).
 
 These correlations likely reflect commuting and contact patterns, and therefore we expect SARS-CoV-2 dynamics on campus to be correlated with those from nearby large cities as well. 
-We find similarly strong correlations with New York City for the first three semesters: fall 2020--2021 ($\rho = 0.74$ (95\% CI: 0.44--0.90; $p < 0.001$), Supplementary Figure S3A); spring 2020--2021 ($\rho = 0.84$ (95\% CI: 0.61--0.94; $p < 0.001$), Supplementary Figure S3B); fall 2021--2022 ($\rho = 0.78$ (95\% CI: 0.51--0.91; $p < 0.001$), Supplementary Figure S3C); and spring 2021--2022 ($\rho = 0.33$ (95\% CI: $-0.34$--0.78; $p=0.3$), Supplementary Figure S3D).
+We find similarly strong correlations with New York City for the first three semesters: fall 2020--2021 ($\rho = 0.64$ (95\% CI: 0.26--0.85; $p = 0.003$), Supplementary Figure S4A); spring 2020--2021 ($\rho = 0.80$ (95\% CI: 0.53--0.93; $p < 0.001$), Supplementary Figure S4B); fall 2021--2022 ($\rho = 0.89 (95\% CI: 0.73--0.96; $p < 0.001$), Supplementary Figure S4C); and spring 2021--2022 ($\rho = 0.50$ (95\% CI: $-0.14$--0.85; $p=0.1$), Supplementary Figure S4D).
 
-The same picture emerges for Philadelphia except for spring 2020: fall 2020--2021 ($\rho = 0.83$ (95\% CI: 0.59--0.93; $p < 0.001$), Supplementary Figure S4A); spring 2020--2021 ($\rho = 0.29$ (95\% CI: $-0.22$--0.68; $p = 0.25$), Supplementary Figure S4B); fall 2021--2022 ($\rho = 0.78$ (95\% CI: 0.51--0.91; $p < 0.001$), Supplementary Figure S4C);
-and spring 2021--2022 ($\rho=0.37$ (95\% CI: $-0.30$--0.80; $p=0.27$), Supplementary Figure S4D).
-Including counties from New York and Pennsylvania states into the spatial correlation analysis yields additional insights (Supplementary Figure S5):
+Similar picture emerges for Philadelphia except for spring 2020: fall 2020--2021 ($\rho = 0.87$ (95\% CI: 0.68--0.95; $p < 0.001$), Supplementary Figure S5A); spring 2020--2021 ($\rho = 0.27$ (95\% CI: $-0.24$--0.66; $p = 0.30$), Supplementary Figure S5B); fall 2021--2022 ($\rho = 0.89$ (95\% CI: 0.74--0.96; $p < 0.001$), Supplementary Figure S5C);
+and spring 2021--2022 ($\rho=0.46$ (95\% CI: $-0.19$--0.83; $p=0.15$), Supplementary Figure S5D).
+Including counties from New York and Pennsylvania states into the spatial correlation analysis yields additional insights (Supplementary Figure S6):
 epidemic dynamics were highly synchronized across all counties in fall 2020--2021 and became less synchronized over time. 
@@
 
-Infection processes are modeled based on standard compartmental structures (Supplementary Figure S6).
+Infection processes are modeled based on standard compartmental structures (Supplementary Figure S7).
 Once infected, susceptible individuals remain in the exposed stage for $D_e = 2$ days on average, during which they cannot transmit or test positive. 
@@
 The median predictions are positively correlated with the observed dynamics ($\rho = 0.83$; 95\% CI: 0.61--0.93; \fref{matching}C).
-Although a wide range of assumptions about the levels of community transmission $\theta$ are consistent with the observed dynamics, our simulations preclude high levels of contact, $\Rc > 2$ (Supplementary Figure S7).
+Although a wide range of assumptions about the levels of community transmission $\theta$ are consistent with the observed dynamics, our simulations preclude high levels of contact, $\Rc > 2$ (Supplementary Figure S8).
 Distancing measures on campus and contact tracing efforts likely contributed to lowering contact levels $\Rc$.
@@
 A similar set of parameters can capture the observed dynamics in spring 2020--2021.
-The best matching parameter predicts a slightly higher levels of community transmission $\theta=0.02$ (\fref{matching}D), but a wide range of parameters are consistent with the observed dynamics as before (Supplementary Figure S8).
+The best matching parameter predicts a slightly higher levels of community transmission $\theta=0.02$ (\fref{matching}D), but a wide range of parameters are consistent with the observed dynamics as before (Supplementary Figure S9).
 Simulations also preclude high $\Rc > 2$ again, suggesting that transmission between students were likely limited even though they had returned to campus---the absence of in-person teaching is likely to have contributed to lowering $\Rc$.
@@
 
-While $\theta = 0.15$ and $\Rc = 0.5$ gives the best matching parameter set with a median logged sum of squared errors of 8.88 (95\% quantile: 6.55--12.6), other parameter sets also give nearly identical fits (\fref{matching}H; Supplementary Figure S9):
+While $\theta = 0.15$ and $\Rc = 0.5$ gives the best matching parameter set with a median logged sum of squared errors of 8.88 (95\% quantile: 6.55--12.6), other parameter sets also give nearly identical fits (\fref{matching}H; Supplementary Figure S10):
 for example, $\theta = 0.1$ and $\Rc = 1$ gives a median logged sum of squared errors of 8.9 (95\% quantile: 5.79--13.6).
@@
 These simulations suggest that an increase in the number of cases in November can be explained by a combination of waning immunity alone without requiring additional changes in transmission dynamics (note we do not allow $\theta$ or $\Rc$ to vary over time).
-When we exclude immune waning from the model, predicted epidemic dynamics exhibit slower growth and require even higher values of campus and community contact rates ($\Rc$ and $\theta$) to qualitatively match the observed dynamics (Supplementary Figure S10)---even so, the logged sum of squared differences are generally higher (with median logged sum of squared differences ranging from 6.9 to 41.3 for the same parameter regime).
+When we exclude immune waning from the model, predicted epidemic dynamics exhibit slower growth and require even higher values of campus and community contact rates ($\Rc$ and $\theta$) to qualitatively match the observed dynamics (Supplementary Figure S11)---even so, the logged sum of squared differences are generally higher (with median logged sum of squared differences ranging from 6.9 to 41.3 for the same parameter regime).
 Thus, combining some amount of immune waning and high campus and community contact rates likely best explains the epidemic growth near the end of the semester.
@@
 \begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_correlation2.pdf}
+\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_new_per_1000_bypop.pdf}
+\caption{
+\textbf{The weekly number of cases per 1000 stratified by subpopulations.}
+The case trajectory for undergraduate students was omitted for the fall semester of 2020--2021 academic year due to a very low number of undergraduate students present on campus.
+}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!htp]
+\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_correlation_per_1000.pdf}
 \caption{

### 2023-01-17 figure_princeton_new/figure_princeton_new_per_1000.R

@@ -344,7 +344,7 @@ g3 <- g3_base +
 g4 <- ggplot(spring2021) +
   geom_segment(x=3, y=43/13, xend=3, yend=355/13, lty=3) +
   geom_segment(x=10, y=43/13, xend=10, yend=355/13, lty=3) +
-  geom_bar(aes(as.factor(date), value/13, fill=key), stat="identity") +
+  geom_bar(aes(as.factor(date), value/15, fill=key), stat="identity") +
   geom_line(data=mercer_spring2021, aes(date, cases/387340*1000, group=1), col="red") +
   geom_point(data=mercer_spring2021, aes(date, cases/387340*1000), col="red") +
   annotate("text", x=3, y=390/13, label="Classes\nbegin", family="Times") +
@@ -441,11 +441,11 @@ g7 <- ggplot(fall2021a) +
     axis.line = element_line()
   )
 
-cor.test(log(spring2021a$value+1), log(spring2021a$cases+1))
+cor.test(spring2021a$value, spring2021a$cases)
 
 g8 <- ggplot(spring2021a) +
-  geom_point(aes(cases/387340*1000, value/13)) +
-  geom_smooth(aes(cases/387340*1000, value/13), method="lm", col="black") +
+  geom_point(aes(cases/387340*1000, value/15)) +
+  geom_smooth(aes(cases/387340*1000, value/15), method="lm", col="black") +
   geom_abline(intercept=0, slope=1, lty=2) +
   scale_x_continuous("Mercer county cases per 1000") +
   scale_y_continuous("Princeton cases per 1000") +
@@ -502,8 +502,8 @@ g11 <- ggplot(fall2021a) +
   )
 
 g12 <- ggplot(spring2021a) +
-  geom_point(aes(date, value/13/(cases/387340*1000)))  +
-  geom_line(aes(date, value/13/(cases/387340*1000), group=1)) +
+  geom_point(aes(date, value/15/(cases/387340*1000)))  +
+  geom_line(aes(date, value/15/(cases/387340*1000), group=1)) +
   geom_hline(yintercept=1, lty=2) +
   scale_x_discrete("Date") +
   scale_y_continuous("Princeton to Mercer case ratio") +
@@ -552,7 +552,7 @@ fall2021b <- fall2021 %>%
 
 spring2021_pop <- data.frame(
   key=c("Faculty and Staff", "Grad", "Undergrad"),
[…]

### 2023-01-17 figure_princeton_new/figure_princeton_new_per_1000_bypop.R

@@ -138,9 +138,9 @@ spring2021 <- data %>%
     `Faculty and Staff`=`Faculty Staff Positive Tests`
   )  %>%
   mutate(
-    Undergrad=Undergrad/3000,
-    Grad=Grad/2000,
-    `Faculty and Staff`=`Faculty and Staff`/3000,
+    Undergrad=Undergrad/5000,
+    Grad=Grad/3000,
+    `Faculty and Staff`=`Faculty and Staff`/7000,
     date=as.Date(date)
   )  %>%
   mutate(

### 2023-01-17 figure_princeton_new/figure_princeton_newyork_new.R

@@ -289,7 +289,7 @@ g3 <- ggplot(fall2021) +
   )
 
 g4 <- ggplot(spring2021) +
-  geom_bar(aes(as.factor(date), value/13, fill=key), stat="identity") +
+  geom_bar(aes(as.factor(date), value/15, fill=key), stat="identity") +
   geom_line(data=nyc_spring2021, aes(date, cases/8.8e6*1000, group=1), col="red") +
   geom_point(data=nyc_spring2021, aes(date, cases/8.8e6*1000), col="red") +
   scale_x_discrete("Date") +
@@ -384,11 +384,11 @@ g7 <- ggplot(fall2021a) +
     axis.line = element_line()
   )
 
-cor.test(log(spring2021a$value+1), log(spring2021a$cases+1))
+cor.test(spring2021a$value, spring2021a$cases)
 
 g8 <- ggplot(spring2021a) +
-  geom_point(aes(cases/8.8e6*1000, value/13)) +
-  geom_smooth(aes(cases/8.8e6*1000, value/13), method="lm", col="black") +
+  geom_point(aes(cases/8.8e6*1000, value/15)) +
+  geom_smooth(aes(cases/8.8e6*1000, value/15), method="lm", col="black") +
   geom_abline(intercept=0, slope=1, lty=2) +
   scale_x_continuous("New York City cases per 1000") +
   scale_y_continuous("Princeton cases per 1000") +

### 2023-01-17 figure_princeton_new/figure_princeton_philly_new.R

@@ -287,7 +287,7 @@ g3 <- ggplot(fall2021) +
   )
 
 g4 <- ggplot(spring2021) +
-  geom_bar(aes(as.factor(date), value/13, fill=key), stat="identity") +
+  geom_bar(aes(as.factor(date), value/15, fill=key), stat="identity") +
   geom_line(data=philly_spring2021, aes(date, cases/1.6e6*1000, group=1), col="red") +
   geom_point(data=philly_spring2021, aes(date, cases/1.6e6*1000), col="red") +
   scale_x_discrete("Date") +
@@ -401,63 +401,3 @@ gcomb <- ggarrange(g1, g2, g3, g4, g5, g6, g7, g8, nrow=4, byrow=FALSE,
                     widths=c(2, 1), draw=FALSE)
 
 ggsave("figure_princeton_philly_new.pdf", gcomb, width=12*1.2, height=10*1.2)
-
-fall2020b <- fall2020 %>%
-  merge(philly_fall2020)
-
-spring2020b <- spring2020 %>%
-  merge(philly_spring2020)
-
-fall2021b <- fall2021 %>%
-  merge(philly_fall2021)
-
-spring2021b <- spring2021 %>%
-  merge(philly_spring2021)
-
-g9 <- ggplot(fall2020b) +
-  geom_point(aes(cases+1, value+1)) +
-  geom_smooth(aes(cases+1, value+1), method="lm", col="black") +
-  scale_x_log10("Philadelphia cases") +
-  scale_y_log10("Princeton cases") +
-  ggtitle("A. Fall 2020-2021") +
-  facet_wrap(~key, scale="free") +
-  theme(
-    panel.grid = element_blank()
-  )
-
-g10 <- ggplot(spring2020b) +
-  geom_point(aes(cases+1, value+1)) +
-  geom_smooth(aes(cases+1, value+1), method="lm", col="black") +
-  scale_x_log10("Philadelphia cases") +
-  scale_y_log10("Princeton cases") +
-  ggtitle("B. Spring 2020-2021") +
-  facet_wrap(~key, scale="free") +
-  theme(
-    panel.grid = element_blank()
-  )
-
-g11 <- ggplot(fall2021b) +
-  geom_point(aes(cases+1, value+1)) +
-  geom_smooth(aes(cases+1, value+1), method="lm", col="black") +
-  scale_x_log10("Philadelphia cases") +
-  scale_y_log10("Princeton cases") +
[…]

### 2023-01-19 doc/university-covid.tex
getting closer to finishing

@@
 These correlations likely reflect commuting and contact patterns, and therefore we expect SARS-CoV-2 dynamics on campus to be correlated with those from nearby large cities as well. 
-We find similarly strong correlations with New York City for the first three semesters: fall 2020--2021 ($\rho = 0.64$ (95\% CI: 0.26--0.85; $p = 0.003$), Supplementary Figure S4A); spring 2020--2021 ($\rho = 0.80$ (95\% CI: 0.53--0.93; $p < 0.001$), Supplementary Figure S4B); fall 2021--2022 ($\rho = 0.89 (95\% CI: 0.73--0.96; $p < 0.001$), Supplementary Figure S4C); and spring 2021--2022 ($\rho = 0.50$ (95\% CI: $-0.14$--0.85; $p=0.1$), Supplementary Figure S4D).
+We find similarly strong correlations with New York City for the first three semesters: fall 2020--2021 ($\rho = 0.64$ (95\% CI: 0.26--0.85; $p = 0.003$), Supplementary Figure S4A); spring 2020--2021 ($\rho = 0.80$ (95\% CI: 0.53--0.93; $p < 0.001$), Supplementary Figure S4B); fall 2021--2022 ($\rho = 0.89$ (95\% CI: 0.73--0.96; $p < 0.001$), Supplementary Figure S4C); and spring 2021--2022 ($\rho = 0.50$ (95\% CI: $-0.14$--0.85; $p=0.1$), Supplementary Figure S4D).
 
@@
 
-Finally, mass testing allows us to infer the ratio between the weekly numbers of cases from Princeton and those from Mercer county---
-we expect this ratio to remain constant over time if the majority of infections on campus is caused by community transmission, provided that testing patterns remain roughly constant in both places.
-We find that the ratio between the weekly numbers of cases from Princeton and those from Mercer county stayed between 1:100 and 1:25 until the end of the fall semester of the 2021--2022 academic year (\fref{princeton}I--K).
+Finally, mass testing allows us to infer the ratio between the weekly numbers of cases per 1000 from Princeton and those from Mercer county---
+we expect this ratio to remain constant around 1 over time when (1) there is random, homogeneous mixing between the campus and community and (2) testing patterns remains constant in both places within each semester.
+In this case, the majority of infections on PU campus would be caused by community transmission owing to its small population size.
+Instead, we find that the ratio generally hovers above 1 during the fall semester of the 2020--2021 academic year even though there was little to no documented transmission on campus (\fref{princeton}I).
+This pattern likely reflects higher testing rate on campus, thereby resulting in a higher case ascertainment rate.
+For the most part of the spring semester of the 2020--2021 academic year and the fall semester of the 2021--2022 academic year, the case ratios hover around 1 (\fref{princeton}J--K).
+Deviations from the one-to-one ratio were often associated with large campus events, such as school holidays and the beginning and end of semesters.
 An increase in this ratio at the end of November 2021 was associated with the campus outbreak before Thanksgiving followed by an introduction of the Omicron variant in December---this deviation indicates an increase in the amount of transmission on campus.
-During the spring semester of the 2021--2022 academic year, the ratio between PU cases and Mercer county cases increased above one, meaning that more cases were reported in PU than elsewhere in Mercer county (\fref{princeton}L);
-notably, we did not see an increase in Mercer county cases, meaning that there was little-to-no transmission from campus to local community.
-
-% We do not find significant changes in the ratio between the weekly numbers of cases from Princeton and those from Mercer county (\fref{princeton}G).
-% Instead, the mean ratios remained similar across three semesters despite changes in campus populations (e.g., undergraduate student returning in spring 2020) and vaccination levels: 0.034 (95\% CI: 0.021--0.047) for fall 2020, 0.019 (95\% CI: 0.015--0.023) for spring 2020, and 0.046 (95\% CI: 0.035--0.057) for fall 2021.
-% However, this does not necessarily imply that the contact or transmission levels stayed constant over time: given extremely high vaccination coverage in PU during fall 2021, the university population would been subject to a much higher forces of infection (e.g., a greater amount of mixing with the community population) to obtain similar levels of cases.
-% We also find that some variations in the ratio around the mean can be explained by holiday effects.
-% For example, sudden decreases in the number of cases around Thanksgiving of 2020, fall recess of 2021, and end of classes of 2021 correspond to a reduction in the ratio between university and community cases.
-% Similarly, a sudden increase in the number of cases during Thanksgiving of 2021 caused the ratio to increase, reflecting decoupling of epidemic dynamics due to on-campus transmission.
+During the spring semester of the 2021--2022 academic year, the ratio between PU cases and Mercer county cases increased above one due to a large outbreak on campus (\fref{princeton}L);
+notably, we did not see an increase in Mercer county cases (\fref{princeton}D), meaning that there was little-to-no transmission from campus to local community.
 
@@
 here, infectious contacts refer to contacts that would result in infection when the contacted individual is susceptible to infection.
+We note that the definition of the contact reproduction number $\Rc$ is similar to the standard definition of basic reproduction number $\RR_0$.
+The main difference is that the contact reproduction number models the number of total contacts, rather than infections. 
+Since infected individuals make their contacts at random with replacement, the same susceptible person could be contacted multiple times by the same or different infected individual during a time step---all these overlapping contacts will result in one infection.
+Therefore, the number of actual infections may be smaller than the number of contacts, especially since contacts can also land on non-susceptible individuals.
+
 The contact reproduction number implicitly accounts for all intervention measures that we do not model explicitly, such as social distancing and contact tracing---in other words, this contact reproduction number does not account for asymptomatic testing or vaccination, which are modeled separately.
@@
 On each day, all infected individuals who have not yet been isolated then make infectious contacts at random to anyone on campus; the number of infectious contacts are drawn from a negative binomial distribution with a mean of either $\beta_p$ or $\beta_s$ and an overdispersion parameter of $k=0.1$ to account for the possibility of super-spreading events \citep{10.12688/wellcomeopenres.15842.3}.
+
 We also rely on cases from Mercer County to crudely capture community dynamics.
@@
 vaccine efficacy against transmissibility is also allowed to wane at the same rate (i.e., from 20\% to 11\% in 20 weeks).
+We note that these assumptions are specific to the Delta variant---we discuss vaccine effectiveness against the Omicron variant later on.
 
@@
 For \textbf{spring 2020--2021}, we simulate the model assuming 8000 individuals (3000 undergraduate students, 2000 graduate students, and 3000 faculty and staff members) on campus with 5000 of them participating in asymptomatic testing twice a week.
-We further assume that 4000 individuals (3000 undergraduate students and 1000 graduate students) returned to campus over 14 days (January 16, 2021--January 29, 2021);
-all returning individuals are assumed to be quarantined for 14 days and tested upon returning. 
+We further assume that 4000 individuals (3000 undergraduate students and 1000 graduate students) returned to campus over 14 days (January 16, 2021--January 29, 2021).
+In the beginning of the semester, all returning students were required to quarantine in their rooms for 14 and tested upon returning by the university---in our model, this was implemented by preventing returning students from getting infected or infecting other individuals.
 Finally, to match the initial influx of cases, we assume that 1\% of both returning and on-campus populations are infected at the beginning of simulation (January 16, 2021).
@@
 therefore, our homogeneous mixing assumption is conservatively pessimistic.
-We also do not account for changes in behavior; 
-instead, we assume constant values for $\theta$ and $\Rc$ throughout a semester.
-While we cannot rule out the possibility that changes in behavior (and therefore transmission rates) could have contributed to various epidemics (e.g., the Omicron wave beginning in the fall semester of the 2021--2022 academic year), we were able to capture the majority of epidemic patterns without modeling them.
-% We note that the invasion of the Omicron variant was modeled separately for the spring semester of a 2021--2022 academic year by explicitly considering the immune evasion.
+We also do not account for explicit changes in behavior on campus and assume constant $\Rc$ throughout each semester.
+Instead, we implicitly account for behavioral changes in the community by modeling community transmission to campus as a function of community case numbers.
+While we cannot rule out the possibility that behavioral changes on campus could have contributed to various epidemics (e.g., the Omicron wave beginning in the fall semester of the 2021--2022 academic year), we were able to capture the majority of epidemic patterns without modeling them---when the majority of transmission is caused by imported cases from the community, we expect behavioral changes on campus to have relatively weaker effects on overall transmission dynamics.
 We also do not explore parameter uncertainty, which can lead to underestimation of overall uncertainty \citep{elderd2006uncertainty}. 

### 2023-01-19 figure_princeton_new/figure_princeton_philly_new.R

@@ -385,8 +385,8 @@ g7 <- ggplot(fall2021a) +
 cor.test(spring2021a$value, spring2021a$cases)
 
 g8 <- ggplot(spring2021a) +
-  geom_point(aes(cases/1.6e6*1000, value/13)) +
-  geom_smooth(aes(cases/1.6e6*1000, value/13), method="lm", col="black") +
+  geom_point(aes(cases/1.6e6*1000, value/15)) +
+  geom_smooth(aes(cases/1.6e6*1000, value/15), method="lm", col="black") +
   geom_abline(intercept=0, slope=1, lty=2) +
   scale_x_continuous("Philadelphia cases per 1000") +
   scale_y_continuous("Princeton cases per 1000") +

### 2023-01-19 figure_princeton_new/figure_princeton_simulation_new.R

@@ -121,13 +121,19 @@ simall_fall_2020_gof2 <- simall_fall_2020_gof %>%
 
 g1 <- ggplot(simall_fall_2020_gof2) +
   geom_tile(aes(R0, scale, fill=log(gof))) +
-  scale_x_log10(expression(R[contact]), expand=c(0, 0)) +
-  scale_y_continuous(expression(theta), expand=c(0, 0)) +
+  geom_point(aes(x=0.25, y=0.0000075), col="white", size=5, shape=1, stroke=2) +
+  scale_x_log10(expression(R[contact]), expand=c(0, 0), breaks=c(0.25, 0.5, 1, 2)) +
+  scale_y_continuous(expression(theta), expand=c(0, 0),
+                     breaks=c(5e-6, 7.5e-6, 10e-6, 12.5e-6, 15e-6)) +
   scale_fill_viridis_c("log SSQ ") +
   ggtitle("A. Fall 2020-2021")
 
+simall_fall_2020_gof2 %>%
+  ungroup %>%
+  filter(gof==min(gof))
+
 simall_fall_2020_best <- simall_fall_2020 %>%
-  filter(R0==0.5, scale==0.02) %>%
+  filter(R0==0.25, scale==0.0000075) %>%
   mutate(
     group=cut(as.numeric(as.Date(date)), breaks=unique(as.numeric(fall2020$date)))
   ) %>%
@@ -153,9 +159,35 @@ simall_fall_2020_best <- simall_fall_2020 %>%
     date=factor(date, levels=date)
   )
 
+simall_fall_2020_gof %>%
+  filter(R0==0.25, scale==0.0000075) %>%
+  filter(gof==min(gof)) ## sim 90
+
+simall_fall_2020_best2 <- simall_fall_2020 %>%
+  filter(R0==0.25, scale==0.0000075, sim==90)  %>%
+  mutate(
+    group=cut(as.numeric(as.Date(date)), breaks=unique(as.numeric(fall2020$date)))
+  ) %>%
+  group_by(sim, group) %>%
+  summarize(
+    cases=sum(cases),
+    date=max(date)
+  ) %>%
+  merge(fall2020) %>%
+  mutate(
+    year=substr(date, 1, 4),
+    month=as.numeric(gsub("-.*", "", gsub("20..-", "", date))), 
+    day=as.numeric(gsub(".*-", "", gsub("20..-", "", date)))
+  ) %>%
+  mutate(
+    date=paste0(month.abb[month], " ", day, ", ", year),
[…]

### 2023-01-19 simulation_princeton/simulation_fall_2020.R

@@ -22,8 +22,8 @@ nsim <- 100
 
 simulation_fall_2020 <- vector('list')
 
-R0vec <- c(0.5, 1, 2)
-scalevec <- c(0.01, 0.015, 0.02, 0.025, 0.03)
+R0vec <- c(0.25, 0.5, 1, 2)
+scalevec <- c(5e-6, 7.5e-6, 10e-6, 12.5e-6, 15e-6)
 
 paramvec <- expand.grid(R0vec, scalevec)
 
@@ -45,7 +45,7 @@ for (j in 1:nrow(paramvec)) {
                    I=0,
                    return=rep(TRUE, 3000),
                    returnfrac=1,
-                   thetavec=paramvec[j,2] * mercer_fall2020$cases,
+                   thetavec=paramvec[j,2] * mercer_fall2020$cases * 3000,
                    datestart=as.Date("2020-08-23"),
                    datesemester=as.Date("2020-08-31"),
                    dateinternal=as.Date("2020-08-31"),

### 2023-01-19 simulation_princeton/simulation_fall_2021.R

@@ -20,8 +20,8 @@ nsim <- 100
 
 simulation_fall_2021 <- vector('list')
 
-R0vec <- c(0.5, 1, 2, 4, 8)
-scalevec <- c(0.005, 0.01, 0.015, 0.02, 0.025) * 10
+R0vec <- c(0.25, 0.5, 1, 2, 4, 8)
+scalevec <- c(5e-6, 7.5e-6, 10e-6, 12.5e-6, 15e-6)
 
 paramvec <- expand.grid(R0vec, scalevec)
 
@@ -48,7 +48,7 @@ for (j in 1:nrow(paramvec)) {
                    boostdaily=0,
                    return=c(rep(FALSE, 5000), rep(TRUE, 8000)),
                    returnfrac=1/16,
-                   thetavec=paramvec[j,2] * mercer_fall2021$cases,
+                   thetavec=paramvec[j,2] * mercer_fall2021$cases * 13000,
                    datestart=as.Date("2021-08-14"),
                    datesemester=as.Date("2021-09-01"),
                    dateinternal=as.Date("2021-09-01"),

### 2023-01-19 simulation_princeton/simulation_spring_2020.R

@@ -20,8 +20,8 @@ nsim <- 100
 
 simulation_spring_2020 <- vector('list')
 
-R0vec <- c(0.5, 1, 2)
-scalevec <- c(0.01, 0.015, 0.02, 0.025, 0.03)
+R0vec <- c(0.25, 0.5, 1, 2)
+scalevec <- c(1e-6, 2e-6, 3e-6, 4e-6, 5e-6)
 
 paramvec <- expand.grid(R0vec, scalevec)
 
@@ -43,7 +43,7 @@ for (j in 1:nrow(paramvec)) {
                    I=0.01,
                    return=c(rep(FALSE, 3000), rep(FALSE, 1000), rep(TRUE, 1000), rep(TRUE, 3000)),
                    returnfrac=1/14,
-                   thetavec=paramvec[j,2] * mercer_spring2020$cases,
+                   thetavec=paramvec[j,2] * mercer_spring2020$cases * 8000,
                    datestart=as.Date("2021-01-16"),
                    datesemester=as.Date("2021-02-01"),
                    dateinternal=as.Date("2021-02-01"),

### 2023-01-22 doc/university-covid.tex
a bunch of new things

@@
 These kinds of contacts are inherently difficult to keep track of, making contact tracing less effective.
-The impact of intervention measures is expected to vary across different university campuses, reflecting heterogeneity in campus settings such as compliance, resources, community prevalence, as well as effects of other interventions present on campuses.
+The impact of intervention measures is expected to vary across different university campuses, reflecting heterogeneity in campus settings such as compliance, resources, community prevalence, ventilation infrastructure, as well as effects of other interventions present on campuses.
 For example, Duke and Harvard Universities experienced moderate outbreaks at the beginning of the fall semester in 2021 when in-person classes were allowed, despite high vaccination rates and weekly asymptomatic testing protocols \citep{dukeoutbreak,harvardoutbreak}, whereas the number of cases remained low in Princeton University (PU) during the same time period with similar levels of testing and vaccination.
@@
 This in turn likely reduced the accuracy of surveillance; therefore, we chose to focus on epidemic patterns before this change was implemented.
+In Supplementary Figure S1, we show the testing volumes over time.
 Throughout the study period, contact tracing was also performed for positive cases to alert their close contacts to either quarantine or test more frequently according to the close contacts' vaccination status, and to gather data that could help uncover clusters of transmission or superspreader events.
@@
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_new_per_1000.pdf}
+\includegraphics[width=0.95\textwidth]{../figure_princeton_new/figure_princeton_new_per_1000.pdf}
 \caption{
@@
 Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
-(I--L) Ratios between weekly number of cases per 1000 rin PU and Mercer County.
+(I--L) Ratios between weekly number of cases per 1000 in PU and Mercer County.
 The dashed lines represent the 1:1 ratio.
@@
 The highest number of cases was reported among faculty and staff members ($=169$), followed by graduate students ($=41$), and undergraduate students ($=4$).
-Even when we control for the differing population sizes among these groups, we find that a considerably larger amount of cases were reported among faculty and staff members (84.5 cases per 1000) than from graduate students (41 cases per 1000) (Supplementary Figure S1A)---we exclude undergraduate student population from this comparison due to a very low number of undergraduate students present on campus during this period.
+Even when we control for the differing population sizes among these groups, we find that a considerably larger amount of cases were reported among faculty and staff members (84.5 cases per 1000) than from graduate students (41 cases per 1000) (Supplementary Figure S2A)---we exclude undergraduate student population from this comparison due to a very low number of undergraduate students present on campus during this period.
 
@@
 The highest number of cases was reported among faculty and staff members ($=111$), followed by undergraduate students ($=101$), and graduate students ($=29$).
-This ordering is robust to differences in population sizes: 37 cases per 1000 among faculty and staff members, 34.3 cases per 1000 among undergraduate students, and 14.5 cases per 1000 among graduate students (Supplementary Figure S1B).
+This ordering is robust to differences in population sizes: 37 cases per 1000 among faculty and staff members, 34.3 cases per 1000 among undergraduate students, and 14.5 cases per 1000 among graduate students (Supplementary Figure S2B).
 
@@
 The number of cases decreased slightly as classes ended but soon increased again as the Omicron (BA.1) variant began to spread on campus and in Mercer county.
-The total number of reported cases per 1000 remained high for all three population groups: 74 cases per 1000 among faculty and staff members, 63.5 cases per 1000 among graduate students, and 52.4 cases per 1000 among undergraduate students (Supplementary Figure S1C).
+The total number of reported cases per 1000 remained high for all three population groups: 74 cases per 1000 among faculty and staff members, 63.5 cases per 1000 among graduate students, and 52.4 cases per 1000 among undergraduate students (Supplementary Figure S2C).
 
@@
 The number of cases remained high before classes began but decreased over time, following epidemic patterns in Mercer County (\fref{princeton}D).
-Coinciding with the decrease in the campus and local cases numbers, the gathering policy was updated to allow food in events were no longer limited to 20 people;
+Coinciding with the decrease in the campus and local cases numbers, the gathering policy was updated on February 8, 2022 to allow food in events were no longer limited to 20 people;
 in addition, the testing frequency was reduced to once a week.
-Following the policy change, a large outbreak occurred on campus with high case numbers persisting until Spring Break (March 5th, 2022).
+Following the policy change, a large gathering event was held on campus, which resulted in an outbreak with high case numbers persisting until Spring Break (March 5th, 2022).
 The timing of this outbreak also coincided with a rapid turnover of the Omicron subvariant BA.2---the proportion of BA.2 subvariant reached 93.5\% (372/398) compared to 26.9\% (14/52) from the previous week.
@@
 fall 2020--2021 ($\rho = 0.79$ (95\% CI: 0.52--0.91; $p < 0.001$), \fref{princeton}E); spring 2020--2021 ($\rho = 0.84$ (95\% CI: 0.60--0.94; $p < 0.001$), \fref{princeton}F); and fall 2021--2022 ($\rho = 0.93$ (95\% CI: 0.84--0.97; $p < 0.001$), \fref{princeton}G).
-These correlations are robust even when we stratify cases by the population, except for undergraduate students during Fall 2020, when most were not physically present on campus (Supplementary Figure S2).
+These correlations are robust even when we stratify cases by the population, except for undergraduate students during Fall 2020, when most were not physically present on campus (Supplementary Figure S3).
 However, the case patterns in PU were decoupled from those in Mercer county for the spring semester of the 2021--2022 academic year with unclear correlations (\fref{princeton}H): $\rho = 0.47$ (95\% CI: $-0.18$--0.83; $p = 0.15$).
-Stratifying cases by subpopulation shows that case patterns in graduate students and faculty and staff members were still strongly correlated with case patterns in Mercer county, meaning that the campus transmission was limited to undergraduate students (Supplementary Figure S2).
-We also find strong correlations between the weekly logged numbers of cases from PU and those from other counties in New Jersey (Supplementary Figure S3)---these correlations significantly decreased with distance from Mercer county in both spring 2020--2021 ($\rho=-0.48$ (95\% CI: $-0.75$--$-0.06$; $p = 0.03$) for all cases and $\rho=-0.51$ (95\% CI: $-0.77$--$-0.10$; $p = 0.02$) for faculty and staff cases) and fall 2021--2022 ($\rho=-0.68$ (95\% CI: $-0.86$--$-0.35$, $p < 0.001$) for all cases and $\rho=-0.74$ (95\% CI: $-0.89$--$-0.46$, $p < 0.001$) for faculty and staff cases).
+Stratifying cases by subpopulation shows that case patterns in graduate students and faculty and staff members were still strongly correlated with case patterns in Mercer county, meaning that the campus transmission was limited to undergraduate students (Supplementary Figure S3).
+We also find strong correlations between the weekly logged numbers of cases from PU and those from other counties in New Jersey (Supplementary Figure S4)---these correlations significantly decreased with distance from Mercer county in both spring 2020--2021 ($\rho=-0.48$ (95\% CI: $-0.75$--$-0.06$; $p = 0.03$) for all cases and $\rho=-0.51$ (95\% CI: $-0.77$--$-0.10$; $p = 0.02$) for faculty and staff cases) and fall 2021--2022 ($\rho=-0.68$ (95\% CI: $-0.86$--$-0.35$, $p < 0.001$) for all cases and $\rho=-0.74$ (95\% CI: $-0.89$--$-0.46$, $p < 0.001$) for faculty and staff cases).
 Across the first three semesters, both the total cases and faculty and staff cases showed similar levels of correlations with local cases.
-For the spring semester of the 2021--2022 academic year, we still find high correlations between faculty and staff cases and local cases throughout other counties ($\rho > 0.8$ across all counties in New Jersey); however, the total cases exhibit considerably weaker correlations due to student-to-student transmission on campus (Supplementary Figure S3).
+For the spring semester of the 2021--2022 academic year, we still find high correlations between faculty and staff cases and local cases throughout other counties ($\rho > 0.8$ across all counties in New Jersey); however, the total cases exhibit considerably weaker correlations due to student-to-student transmission on campus (Supplementary Figure S4).
 
 These correlations likely reflect commuting and contact patterns, and therefore we expect SARS-CoV-2 dynamics on campus to be correlated with those from nearby large cities as well. 
-We find similarly strong correlations with New York City for the first three semesters: fall 2020--2021 ($\rho = 0.64$ (95\% CI: 0.26--0.85; $p = 0.003$), Supplementary Figure S4A); spring 2020--2021 ($\rho = 0.80$ (95\% CI: 0.53--0.93; $p < 0.001$), Supplementary Figure S4B); fall 2021--2022 ($\rho = 0.89$ (95\% CI: 0.73--0.96; $p < 0.001$), Supplementary Figure S4C); and spring 2021--2022 ($\rho = 0.50$ (95\% CI: $-0.14$--0.85; $p=0.1$), Supplementary Figure S4D).
+We find similarly strong correlations with New York City for the first three semesters: fall 2020--2021 ($\rho = 0.64$ (95\% CI: 0.26--0.85; $p = 0.003$), Supplementary Figure S5A); spring 2020--2021 ($\rho = 0.80$ (95\% CI: 0.53--0.93; $p < 0.001$), Supplementary Figure S5B); fall 2021--2022 ($\rho = 0.89$ (95\% CI: 0.73--0.96; $p < 0.001$), Supplementary Figure S5C); and spring 2021--2022 ($\rho = 0.50$ (95\% CI: $-0.14$--0.85; $p=0.1$), Supplementary Figure S5D).
 
-Similar picture emerges for Philadelphia except for spring 2020: fall 2020--2021 ($\rho = 0.87$ (95\% CI: 0.68--0.95; $p < 0.001$), Supplementary Figure S5A); spring 2020--2021 ($\rho = 0.27$ (95\% CI: $-0.24$--0.66; $p = 0.30$), Supplementary Figure S5B); fall 2021--2022 ($\rho = 0.89$ (95\% CI: 0.74--0.96; $p < 0.001$), Supplementary Figure S5C);
-and spring 2021--2022 ($\rho=0.46$ (95\% CI: $-0.19$--0.83; $p=0.15$), Supplementary Figure S5D).
-Including counties from New York and Pennsylvania states into the spatial correlation analysis yields additional insights (Supplementary Figure S6):
+Similar picture emerges for Philadelphia except for spring 2020: fall 2020--2021 ($\rho = 0.87$ (95\% CI: 0.68--0.95; $p < 0.001$), Supplementary Figure S6A); spring 2020--2021 ($\rho = 0.27$ (95\% CI: $-0.24$--0.66; $p = 0.30$), Supplementary Figure S6B); fall 2021--2022 ($\rho = 0.89$ (95\% CI: 0.74--0.96; $p < 0.001$), Supplementary Figure S6C);
+and spring 2021--2022 ($\rho=0.46$ (95\% CI: $-0.19$--0.83; $p=0.15$), Supplementary Figure S6D).
+Including counties from New York and Pennsylvania states into the spatial correlation analysis yields additional insights (Supplementary Figure S7):
 epidemic dynamics were highly synchronized across all counties in fall 2020--2021 and became less synchronized over time. 
@@
 
-Infection processes are modeled based on standard compartmental structures (Supplementary Figure S7).
+Infection processes are modeled based on standard compartmental structures (Supplementary Figure S8).
 Once infected, susceptible individuals remain in the exposed stage for $D_e = 2$ days on average, during which they cannot transmit or test positive. 
@@
 Therefore, the number of actual infections may be smaller than the number of contacts, especially since contacts can also land on non-susceptible individuals.
-
-The contact reproduction number implicitly accounts for all intervention measures that we do not model explicitly, such as social distancing and contact tracing---in other words, this contact reproduction number does not account for asymptomatic testing or vaccination, which are modeled separately.
+We also note that the contact reproduction number implicitly accounts for all intervention measures that we do not model explicitly, such as social distancing and contact tracing---this is different from $\RR_0$, which typically does not include the effects of intervention measures.
+However, the contact reproduction number does not include asymptomatic testing or vaccination, which are modeled separately.
 We further decompose $\Rc$ into pre-symptomatic $\mathcal R_p = \beta_p D_p$ and (a)symptomatic $\mathcal R_s = \beta_s D_s$ reproduction numbers, where $\beta_p$ and $\beta_s$ represent the corresponding infectious contact rates during pre-symptomatic and (a)symptomatic stages, respectively.
@@
 
+In reality, the contact structure among the campus population is likely more structured and exhibit strong assortativity.
+For example, undergraduate students are more likely to mix with other undergraduate students, rather than graduate students or faculty and staff members.
+Even among undergraduate students, students are more likely to mix with their close friend group than with other students.
+On one hand, assortative mixing may lead to faster epidemic growth within certain population groups.
+On the other hand, it can also make the disease more difficult to spread among other groups that have lower contact rates.
+Therefore, predicting the impact of structured contact network requires a more detailed information about whether the majority of cases were infected at random or from certain groups of the campus population.
+For simplicity, we assume random mixing throughout the paper---nonetheless, allowing for overdispersion in transmission is expected to emulate variability in epidemic growth rates driven by complex contact structures \citep{lloyd2005superspreading}.
+
 We also rely on cases from Mercer County to crudely capture community dynamics.
-In particular, we assume that infectious contacts from local or regional community can be made at random to anyone on campus; these contacts are modeled using a Poisson distribution with a time-varying mean, which is calculated by scaling the daily number of cases by $\theta$ and shifting it by 1 week to account for reporting delays.
+In particular, we assume that infectious contacts from local or regional community can be made at random to anyone on campus.
+These contacts are modeled using a Poisson distribution with a time-varying mean, which is calculated by multiplying the daily number of cases by the community contact rate $\theta$ and the population size on campus $N$.
+More precisely, $\theta$ is the probability that an infected individual from the community makes an infectious contact with an individual on campus per capita campus population. 
+By further multiplying this probability with the population size $N$, we are essentially assuming a density dependent contact, where a higher population size on campus leads to more infections from the community.
+We further shift community contacts by 1 week to account for reporting delays.
 Infectious contacts, whether made by individuals on campus or from outside, result in infection only when the contacted individuals are susceptible; when the contacted individuals are vaccinated, and therefore partially susceptible to infection, they have a reduced probability of infection corresponding to their susceptibility (discussed later).
@@
 We further assume that symptomatic individuals are isolated immediately when they submit their samples until they receive negative results.
-All individuals who test positive are required to isolate (following the same isolation rule as described earlier) and are exempt from asymptomatic testing for 90 days.
+All individuals who test positive are required to isolate (following the same isolation rule as described earlier) and are exempt from asymptomatic testing for 90 days---while this assumption reflects the isolation policy in Princeton University during the investigation period, it may be inapplicable in studying other institutional outbreaks.
 Isolated individuals are assumed to no longer transmit infections.
 We assume that PCR tests can detect infections from individuals who are in pre-symptomatic, symptomatic, and asymptomatic stages with 95\% sensitivity and 100\% specificity.
+
+The 95\% sensitivity assumption may seem too high. 
+For example, \cite{hellewell2021estimating} estimated that the probability of detecting an infection from a PCR peaks at 77\% (54–88\%) 4 days after infection, and decreases to 50\% (38–65\%) by 10 days after infection.
+These estimates are considerably lower than our assumption because their estimates implicitly account for the latent period.
+At the individual level, we assume that an infected individual has no detectable infection (0\% sensitivity) during their latent period and 95\% sensitivity during their various stages of infectious periods.
+At the population level, this assumption translates to a peak sensitivity of 92\% by 4 days after infection, decreasing to 20\% sensitivity by 10 days after infection.
+Our assumption leads to a much lower PCR sensitivity 10 days after infection because we only model the PCR sensitivity during the infectious period.
+In reality, PCR can detect infections even after a person stops shedding infectious virus---we did not include this component in our model because it would not affect the effectiveness of isolation strategy in reducing transmission.
 
@@
 In this study, we use this model to retrospectively analyze past outbreaks.
-First, we try to match our model to epidemic patterns seen on campus for the first three semesters, during which there was limited campus transmission, by varying the contact reproduction number $\Rc$ and the amount of community transmission $\theta$ and holding all other parameters constant.
+First, we try to match our model to epidemic patterns seen on campus for the first three semesters, during which there was limited campus transmission, by varying the contact reproduction number $\Rc$ and the amount of community contact $\theta$ and holding all other parameters constant.
 For each parameter combination, we simulate 100 epidemic trajectories and calculate the sum of squared differences between the weekly numbers of the observed and predicted positive cases.
@@
 \textbf{Retrospective analysis of past SARS-CoV-2 outbreaks on Princeton University (PU) campus.}
-(A, D, G) Time series comparisons of model predictions with observed data across ranges of contact reproduction number $\Rc$ and scaling parameter for community transmission $\theta$.
+(A, D, G) Time series comparisons of model predictions with observed data across ranges of contact reproduction number $\Rc$ and scaling parameter for community contact $\theta$.
 For each parameter combination, we simulate the model 100 times and calculate the sum of squared differences (SSQ) between the reported number of positive cases and the model-predicted number of positive cases.
 Heat maps represent medians of the logged sum of squared differences.
+Circles represent the best fitting parameter set.
 (B, E, H) Model predictions. 
 Solid lines represent median predictions.
+Dashed lines represent a realization with the least sum of squared errors.
 Shaded areas represent 90\% quantiles for the best matching parameter set.
@@
 For \textbf{fall 2020--2021}, we simulate the model assuming 3000 individuals (1000 graduate students and 2000 faculty and staff members) on campus with 1000 of them participating in asymptomatic testing twice a week.
-We find that a low level of contacts $\Rc=0.5$ and a small amount of community transmission $\theta=0.015$ is most consistent with the observed epidemic dynamics in fall 2020 (\fref{matching}A).
+We find that a low level of contacts $\Rc=0.25$ and a small amount of community contact $\theta=7.5\times 10^{-6}$ is most consistent with the observed epidemic dynamics in fall 2020 (\fref{matching}A).
 With these parameters, the model is able to capture the rise and fall in the number of cases with the exception of a sudden decrease in the number of cases around Thanksgiving, which we do not model explicitly (\fref{matching}B).
 The median predictions are positively correlated with the observed dynamics ($\rho = 0.83$; 95\% CI: 0.61--0.93; \fref{matching}C).
-Although a wide range of assumptions about the levels of community transmission $\theta$ are consistent with the observed dynamics, our simulations preclude high levels of contact, $\Rc > 2$ (Supplementary Figure S8).
+Although a wide range of assumptions about the levels of community contact $\theta$ are consistent with the observed dynamics, our simulations preclude high levels of contact, $\Rc > 2$ (Supplementary Figure S9).
 Distancing measures on campus and contact tracing efforts likely contributed to lowering contact levels $\Rc$.
@@
 A similar set of parameters can capture the observed dynamics in spring 2020--2021.
-The best matching parameter predicts a slightly higher levels of community transmission $\theta=0.02$ (\fref{matching}D), but a wide range of parameters are consistent with the observed dynamics as before (Supplementary Figure S9).
+The best matching parameter predicts considerably lower levels of community contact $\theta=3\times 10^{-6}$ (\fref{matching}D), but a wide range of parameters are consistent with the observed dynamics as before (Supplementary Figure S10).
 Simulations also preclude high $\Rc > 2$ again, suggesting that transmission between students were likely limited even though they had returned to campus---the absence of in-person teaching is likely to have contributed to lowering $\Rc$.
@@
 
-Even though the numbers of cases during fall 2021 (before a large outbreak) were similar to those during previous semesters, we find that considerably higher levels of community contact $\theta$ ($\approx 10$ fold higher) are required to explain the observed dynamics due to a decreased susceptibility derived from vaccination (\fref{matching}G).
+The per capita numbers of cases during fall 2021 (before a large outbreak) were considerably lower than to those during previous semesters.
+Nonetheless, we find that higher levels of community contact $\theta=1 \times 10^{-5}$ are required to explain the observed dynamics due to a decreased susceptibility from vaccination (\fref{matching}G).
 We note that the parameter $\theta$ necessarily depends on our assumed vaccine efficacy against susceptibility, and $\theta$ would decrease if we assume a lower vaccine efficacy.
@@
 
-While $\theta = 0.15$ and $\Rc = 0.5$ gives the best matching parameter set with a median logged sum of squared errors of 8.88 (95\% quantile: 6.55--12.6), other parameter sets also give nearly identical fits (\fref{matching}H; Supplementary Figure S10):
-for example, $\theta = 0.1$ and $\Rc = 1$ gives a median logged sum of squared errors of 8.9 (95\% quantile: 5.79--13.6).
-Comparing simulations across a wide range of $\Rc$ (0.5--8) with $\theta = 0.1$ further illustrates that the predicted dynamics are largely insensitive to $\Rc$ until November 26th (\fref{matching}H).
+While $\theta = 1 \times 10^{-5}$ and $\Rc = 0.25$ gives the best matching parameter set with a median logged sum of squared errors of 8.61 (95\% quantile: 5.80--11.6), other parameter sets also give nearly identical fits (\fref{matching}H; Supplementary Figure S11).
+Comparing simulations across a wide range of $\Rc$ (0.25--8) with $\theta = 1 \times 10^{-5}$ further illustrates that the predicted dynamics are largely insensitive to $\Rc$ until November 26th (\fref{matching}H).
 All simulations shown in \fref{matching}H, except for the $\Rc=8$ scenario, are similarly correlated with the observed numbers of cases (\fref{matching}G). 
@@
 
-These simulations suggest that an increase in the number of cases in November can be explained by a combination of waning immunity alone without requiring additional changes in transmission dynamics (note we do not allow $\theta$ or $\Rc$ to vary over time).
-When we exclude immune waning from the model, predicted epidemic dynamics exhibit slower growth and require even higher values of campus and community contact rates ($\Rc$ and $\theta$) to qualitatively match the observed dynamics (Supplementary Figure S11)---even so, the logged sum of squared differences are generally higher (with median logged sum of squared differences ranging from 6.9 to 41.3 for the same parameter regime).
+These simulations suggest that an increase in the number of cases in November can be explained by a combination of waning immunity alone without requiring additional changes in transmission dynamics (note we do not allow $\theta$ or $\Rc$ to vary over time)---we see that extending the simulation beyond November 26th still captures the increase in cases.
+When we exclude immune waning from the model, predicted epidemic dynamics exhibit slower growth and require even higher values of campus and community contact rates ($\Rc$ and $\theta$) to qualitatively match the observed dynamics (Supplementary Figure S12)---even so, the logged sum of squared differences are generally higher (with median logged sum of squared differences ranging from 7.2 to 33.9 for the same parameter regime).
 Thus, combining some amount of immune waning and high campus and community contact rates likely best explains the epidemic growth near the end of the semester.
@@
 To match the realistic campus setting, we assume that 700 students are present on campus as of January 1, 2022, and the remaining 4300 students come back to campus across 28 days.
-We assume that 99\% of students are vaccinated with 60\% of them being boosted---we further allow 70 booster shots on each day such that most students will be boosted by the time everyone is back on campus.
+Based on known vaccination statuses, we assume that 99\% of students are vaccinated with 60\% of them being boosted as of January 1, 2022.
+Since all students were required to receive booster shots before returning to campus, we assumed that 70 booster shots were given on each day---this assumption allows all students to be boosted in 28 days.
 To match the high numbers of cases on the week ending January 7, 2022, we assume 14\% of the students present on campus are infected as of January 7, 2022 (roughly 100/700).
 To account for students who were infected with the Omicron variant during the fall semester, we assume that 100 students are already immune to Omicron infection at the beginning of the spring semester---this roughly corresponds to the number of PU cases that were reported in December.
-Finally, we take the best matching $\theta$ value for the previous semester and scale it by the number of undergraduate students relative to the entire population (therefore using $\theta = 0.15 \times 5000/13000$ throughout).
 
@@
 Vertical dashed lines represent the week including February 8, 2022, when distancing and testing policies were updated on PU campus.
+For each row, we assume a fixed value of baseline contact reproduction number $\Rc$ ranging from 2 to 6 across rows.
+Then, we simulate increase in $\Rc$ at the time of policy change (indicated by column labels).
 }
@@
 \end{figure}
+
+We considered the possibility that the Omicron variant can have shorter latent and infectious periods by decreasing the mean duration of latent, pre-symptomatic, and (a)symptomatic stages of infection by 0.5 days (therefore a total of 1.5 reduction in the duration of infection).
+In this case, a shorter generation interval can lead to faster growth rate given the same values of $\Rc$ \citep{wallinga2007generation}.
+However, we find that the effects of shorter infection has small effects on the overall dynamics (Supplementary Figure S13).
+
+We also considered the possibility that the vaccine effectiveness against the Omicron variant might be lower and run the same analysis with 30\% effectiveness against infection \citep{tan2023vaccine}.
+When the baseline $\Rc$ is low ($\Rc=2$), increasing $\Rc$ still does not increase the number of cases sufficiently.
+When we assume an intermediate value of $\Rc=4$, the model does a better job at capturing the dynamics but it does so by overestimating the trough before the policy change and underestimating the peak after the policy change.
+When we assume a high value of $\Rc=6$, the model overestimates both the trough and the peak (Supplementary Figure S14).
 
@@
 Overall, the observed patterns in cases are more consistent with the epidemic dynamics driven by super-spreading events.
+This is also consistent with the observation that this outbreak was associated with a large gathering event on campus.
 
@@
 For example, we assume conservatively that the entire university populations mix homogeneously and have identical campus and community contact rates (captured by $\Rc$ and $\theta$, respectively).
+This assumption can lead to the fastest epidemic growth rates because transmission is not limited by the size of the contact network---in other words, our estimates of the reproduction will be necessarily low, making the epidemic easier to control.
 In reality, increases in cases were often associated with specific transmission clusters, suggesting heterogeneity in transmission patterns.
 Contact levels also likely differ between different groups:
-for example, faculty and staff members are more likely to interact with community members than undergraduate students and would be at a higher risk for community infections \citep{frazier2022modeling};
-therefore, our homogeneous mixing assumption is conservatively pessimistic.
+for example, faculty and staff members are more likely to interact with community members than undergraduate students and would be at a higher risk for community infections \citep{frazier2022modeling}.
 We also do not account for explicit changes in behavior on campus and assume constant $\Rc$ throughout each semester.
@@
 Despite the simplicity of the analysis, our study provides important lessons for controlling SARS-CoV-2 outbreaks on university campuses in general.
-First, our analysis highlights the power of mass asymptomatic testing for epidemic measurement and planning.
+First, our analysis highlights the power of mass asymptomatic testing for epidemic measurement and planning---even if PCR testing may have lower sensitivity than what we assumed here \citep{hellewell2021estimating}, mass asymptomatic testing can still help track ongoing epidemic dynamics in real time.
 Combining other interventions measures, such as social distancing, mask wearing, and vaccination, can help provide a safe means of reopening university campuses---but the extent to which these interventions are implemented will necessarily depend on resource availability.
 Second, we expect immune waning and superspreading to continue to play important roles in driving campus transmission---keeping vaccine statuses up-to-date within the campus community will be critical moving forward.
-In addition, preventing large gatherings can help prevent large superspreading events in the midst of a rising epidemic.
 Third, the safe reopening of a university campus must consider the spread of SARS-CoV-2 within the surrounding community as they can both potentially drive transmission in each other---however, the degree to which infections spread from campus to community remains uncertain.
@@
 \setcounter{figure}{0}
+
+\begin{figure}[!htp]
+\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_testing.pdf}
+\caption{
+\textbf{The weekly number of PCR tests stratified by subpopulations.}
+}
+\end{figure}
+
+\pagebreak
 
@@
 
+\pagebreak
+
+\begin{figure}[!htp]
+\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_omicron_spring_shortI.pdf}
+\caption{
+\textbf{The impact of changes in testing frequency and an increased reproduction number on the spread of the Omicron variant, assuming a shorter latent and infectious periods.}
+Solid lines represent median predictions.
+Shaded areas represent 90\% quantiles across 100 simulations.
+Points represent the observed data.
+Vertical dashed lines represent the week including February 8, 2022, when distancing and testing policies were updated on PU campus.
+For each row, we assume a fixed value of baseline contact reproduction number $\Rc$ ranging from 2 to 6 across rows.
+Then, we simulate increase in $\Rc$ at the time of policy change (indicated by column labels).
+Here, the assumed mean latent, presymptomatic, and asymptomatic stages of infection are 0.5 days shorter than the assumed values for simulations presented in Figure 3 in the main text.
+}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[!htp]
+\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_omicron_spring_shortI.pdf}
+\caption{
+\textbf{The impact of changes in testing frequency and an increased reproduction number on the spread of the Omicron variant, assuming a lower vaccine effectiveness.}
+Solid lines represent median predictions.
+Shaded areas represent 90\% quantiles across 100 simulations.
+Points represent the observed data.
+Vertical dashed lines represent the week including February 8, 2022, when distancing and testing policies were updated on PU campus.
+For each row, we assume a fixed value of baseline contact reproduction number $\Rc$ ranging from 2 to 6 across rows.
+Then, we simulate increase in $\Rc$ at the time of policy change (indicated by column labels).
+Here, we assume 30\% vaccine effectiveness against infection.
+}
+\end{figure}
+
 \end{document}

### 2023-01-22 R/simulate_internal_omicron_switch2.R

@@ -1,0 +1,647 @@
+##' @param state (list) initial state of individuals (X_t^{(i)} in the Appendix)
+##' @param return (list) initial return state of individuals (X_t^{(i)} in the Appendix)
+##' @param sigma transition rate from the exposed stage
+##' @param p_a probability of asymptomatic infection
+##' @param gamma_a transition rate from the asymptomatic infection stage to the recovered stage
+##' @param gamma_p transition rate from the presymptomatic infection stage to the symptomatic infection stage
+##' @param gamma_s transition rate from the symptomatic infection stage to the recovered stage
+##' @param beta_a transmission rate during the asymptomatic infection stage
+##' @param beta_p transmission rate during the presymptomatic infection stage
+##' @param beta_s transmission rate during the symptomatic infection stage
+##' @param k negative binomial over-dispersion parameter
+##' @param epsilon within- and between-group mixing rates (n x n matrix representing contact from ith row -> jth column)
+##' @param theta infection from outside
+##' @param tmax final time
+##' @param internalstart internal test starting time
+##' @param teststart mass test starting time
+##' @param ntest number of internal tests that can be processed per day
+##' @param sensitivity sensitivity
+##' @param specificity specificity
+##' @param qperiod quarantine period
+##' @param qcontact (TRUE) infected individuals can "contact" quarantined individuals; (FALSE) infected individuals cannot "contact" quarantined individuals; analogous to densitiy and frequency dependence
+##' @param preport probability of self-reporting symptoms and getting tested (and isolate until the test result comes back)
+##' @param trace (logical) contact trace?
+##' @param traceeff tracing efficacy
+##' @param tracedelay days
+##' @param tracememory how far can we go back
+simulate_internal_omicron_switch2 <- function(state,
+                                      vacc,
+                                      boost,
+                                      veff_twodose,
+                                      veff_threedose,
+                                      veff_twodose_inf,
+                                      veff_threedose_inf,
+                                      vaccine_delay,
+                                      veff,
+                                      veff_inf,
[…]

### 2023-01-22 R/simulate_omicron_switch2.R

@@ -1,0 +1,189 @@
+drawpar <- function(D_e=2,
+                    D_p=3,
+                    D_s=3,
+                    p_a=0.4,
+                    R0=2,
+                    p_p=0.5, ## proportion of presymptomatic transmission for symptomatic individuals
+                    k=0.1,
+                    preport=1,
+                    delta=0.6/140) {
+  
+  c(
+    D_e=D_e,
+    D_p=D_p,
+    D_s=D_s,
+    p_a=p_a,
+    R0=R0,
+    p_p=p_p,
+    k=k,
+    preport=preport,
+    delta=delta
+  )
+}
+
+##' @param x parameter (from `drawpar()`)
+convertpar <- function(x) {
+  with(as.list(x), {
+    ## (1-p_p) * R0 = beta * D_s
+    ##
+    ##
+    
+    beta <- R0 * (1 - p_p)/D_s
+    Rs <- D_s * beta
+    
+    Rp <- p_p/(1-p_p) * Rs
+    
+    beta_p <- Rp/D_p
+    
+    list(
+      sigma=-log(1-1/D_e), 
+      gamma_a1=-log(1-2/D_s),
+      gamma_a2=-log(1-2/D_s),
+      gamma_p1=-log(1-2/D_p), 
+      gamma_p2=-log(1-2/D_p), 
+      gamma_s1=-log(1-2/D_s),
+      gamma_s2=-log(1-2/D_s),
+      p_a=p_a, 
+      beta=beta,
+      beta_p=beta_p,
+      k=k,
+      preport=preport,
+      delta=delta
+    )
+  })
+}
+
+initialize <- function(param,
+                       N,
+                       I,
+                       R) {
+  with(as.list(param), {
+    .state <- c("S", "E", "Ip1", "Ip2", "Ia1", "Ia2", "Is1", "Is2", "R")
+    
+    distrb <- c(rmultinom(1, N, c(1-I-R, I, R)))
+    
+    prob <- c(D_e, D_p/2, D_p/2, D_s/2, D_s/2, 0, 0)
+    
+    prob <- prob/sum(prob)
+    
+    state <- rmultinom(1, distrb[2], prob)
+    
+    final <- c(distrb[1], state, distrb[3])
+    
+    sample(unname(unlist(mapply(function(x, y) rep(x, y), .state, final))))
+  })
+}
+
+simulate_omicron_switch2 <- function(param=drawpar(),
+                     vprop=0.99,
+                     boostprop=0.65,
[…]

### 2023-01-22 figure_princeton_new/figure_princeton_simulation_nowaning.R

@@ -106,6 +106,11 @@ simall_fall_2021_best <- simall_fall_2021 %>%
     date=paste0(month.abb[month], " ", day, ", ", year),
     date=factor(date, levels=unique(date)),
     scale=paste0("theta==", scale),
+    scale=factor(scale, levels=c("theta==5e-06",
+                                 "theta==7.5e-06",
+                                 "theta==1e-05",
+                                 "theta==1.25e-05",
+                                 "theta==1.5e-05")),
     R0=paste0("R[contact]==",R0)
   )
 

### 2023-01-22 figure_princeton_new/figure_princeton_simulation_omicron_spring_lowerV.R

@@ -1,0 +1,143 @@
+library(dplyr)
+library(tidyr)
+library(readxl)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(egg)
+library(lubridate)
+
+data_asymp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=1) %>%
+  mutate(
+    type="asymp"
+  )
+
+data_symp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=2) %>%
+  mutate(
+    type="symp"
+  )
+
+data <- bind_rows(
+  data_asymp, data_symp
+)
+
+spring2021 <- data %>%
+  filter(`Week Ending` > as.Date("2021-12-31"),
+         `Week Ending` <= as.Date("2022-03-18")) %>%
+  select(`Week Ending`, `Undergrad Positive Tests`) %>%
+  rename(
+    date=`Week Ending`,
+    Undergrad=`Undergrad Positive Tests`
+  ) %>%
+  mutate(
+    date=as.Date(date)
+  )  %>%
+  mutate(
+    year=substr(date, 1, 4),
+    month=as.numeric(gsub("-.*", "", gsub("20..-", "", date))), 
+    day=as.numeric(gsub(".*-", "", gsub("20..-", "", date)))
+  ) %>%
+  arrange(date) %>%
+  mutate(
+    date=paste0(month.abb[month], " ", day, ", ", year),
+    date=factor(date, levels=unique(date))
+  ) %>%
+  gather(key, value, -date, -year, -month, -day) %>%
+  mutate(
+    date2=as.Date(paste0(year, "-", month, "-", day))
+  ) %>%
+  group_by(date, year, month, day, key, date2) %>%
+  summarize(
+    value=sum(value)
+  )
+
+load("../simulation_omicron_spring/simulation_omicron_spring_increase_50_lowerV.rda")
+load("../simulation_omicron_spring/simulation_omicron_spring_increase_100_lowerV.rda")
+
+simall_spring_2021_increase_50 <- lapply(simulation_omicron_spring_increase_50_lowerV, function(x){
+  lapply(x, function(y) {
+    data.frame(
+      date=y$datevec,
+      time=1:length(y$incidencevec),
+      incidence=y$incidencevec,
+      cases=y$positivevec,
+      R0=y$R0
+    )
+  }) %>%
+    bind_rows(.id="sim")
+})  %>%
+  bind_rows() 
+
[…]

### 2023-01-22 figure_princeton_new/figure_princeton_simulation_omicron_spring_shortI.R

@@ -1,0 +1,143 @@
+library(dplyr)
+library(tidyr)
+library(readxl)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(egg)
+library(lubridate)
+
+data_asymp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=1) %>%
+  mutate(
+    type="asymp"
+  )
+
+data_symp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=2) %>%
+  mutate(
+    type="symp"
+  )
+
+data <- bind_rows(
+  data_asymp, data_symp
+)
+
+spring2021 <- data %>%
+  filter(`Week Ending` > as.Date("2021-12-31"),
+         `Week Ending` <= as.Date("2022-03-18")) %>%
+  select(`Week Ending`, `Undergrad Positive Tests`) %>%
+  rename(
+    date=`Week Ending`,
+    Undergrad=`Undergrad Positive Tests`
+  ) %>%
+  mutate(
+    date=as.Date(date)
+  )  %>%
+  mutate(
+    year=substr(date, 1, 4),
+    month=as.numeric(gsub("-.*", "", gsub("20..-", "", date))), 
+    day=as.numeric(gsub(".*-", "", gsub("20..-", "", date)))
+  ) %>%
+  arrange(date) %>%
+  mutate(
+    date=paste0(month.abb[month], " ", day, ", ", year),
+    date=factor(date, levels=unique(date))
+  ) %>%
+  gather(key, value, -date, -year, -month, -day) %>%
+  mutate(
+    date2=as.Date(paste0(year, "-", month, "-", day))
+  ) %>%
+  group_by(date, year, month, day, key, date2) %>%
+  summarize(
+    value=sum(value)
+  )
+
+load("../simulation_omicron_spring/simulation_omicron_spring_increase_50_shortI.rda")
+load("../simulation_omicron_spring/simulation_omicron_spring_increase_100_shortI.rda")
+
+simall_spring_2021_increase_50 <- lapply(simulation_omicron_spring_increase_50_shortI, function(x){
+  lapply(x, function(y) {
+    data.frame(
+      date=y$datevec,
+      time=1:length(y$incidencevec),
+      incidence=y$incidencevec,
+      cases=y$positivevec,
+      R0=y$R0
+    )
+  }) %>%
+    bind_rows(.id="sim")
+})  %>%
+  bind_rows() 
+
[…]

### 2023-01-22 figure_princeton_new/figure_princeton_testing_new.R

@@ -1,0 +1,247 @@
+library(dplyr)
+library(tidyr)
+library(readxl)
+library(ggplot2); theme_set(theme_bw(base_family = "Times"))
+library(gridExtra)
+library(egg)
+library(lubridate)
+
+data_asymp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=1) %>%
+  mutate(
+    type="asymp"
+  )
+
+data_symp <- read_xlsx("../data/princeton-weekly-positive-03-22-2022.xlsx", sheet=2) %>%
+  mutate(
+    type="symp"
+  )
+
+data <- bind_rows(
+  data_asymp, data_symp
+)
+
+county <- read.csv("../data/us-counties-03-22-2022.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+fall2020 <- data %>%
+  filter(`Week Ending` <= as.Date("2021-01-01")) %>%
+  select(`Week Ending`, `Undergrad Tests`, `Grad Student Tests`,
+         `Faculty Staff Tests`) %>%
+  rename(
+    date=`Week Ending`,
+    Undergrad=`Undergrad Tests`,
+    Grad=`Grad Student Tests`,
+    `Faculty and Staff`=`Faculty Staff Tests`
+  ) %>%
+  mutate(
+    date=as.Date(date)
+  )  %>%
+  mutate(
+    year=substr(date, 1, 4),
+    month=as.numeric(gsub("-.*", "", gsub("20..-", "", date))), 
+    day=as.numeric(gsub(".*-", "", gsub("20..-", "", date)))
+  ) %>%
+  arrange(date) %>%
+  mutate(
+    date=paste0(month.abb[month], " ", day, ", ", year),
+    date=factor(date, levels=unique(date))
+  ) %>%
+  gather(key, value, -date, -year, -month, -day) %>%
+  mutate(
+    date2=as.Date(paste0(year, "-", month, "-", day))
+  ) %>%
+  group_by(date, year, month, day, key, date2) %>%
+  summarize(
+    value=sum(value)
+  )
+
+spring2020 <- data %>%
+  filter(`Week Ending` <= as.Date("2021-05-14"),
[…]

### 2023-01-22 simulation_omicron_spring/simulation_omicron_spring.R

@@ -19,9 +19,9 @@ nsim <- 100
 
 simulation_omicron_spring <- vector('list')
 
-theta <- 0.15 * mercer_spring2021$cases * 5000/13000
+theta <- 1e-5 * mercer_spring2021$cases * 5000
 
-R0vec <- c(2, 4, 6, 8)
+R0vec <- c(2, 4, 6)
 
 paramvec <- expand.grid(R0vec)
 

### 2023-01-22 simulation_omicron_spring/simulation_omicron_spring_increase_10.R

@@ -1,76 +1,0 @@
-library(dplyr)
-source("../R/simulate_internal_omicron_switch2.R")
-source("../R/simulate_omicron_switch2.R")
-
-county <- read.csv("../data/us-counties-03-22-2022.csv")
-
-mercer <- county %>%
-  filter(county=="Mercer", state=="New Jersey") %>%
-  arrange(date) %>%
-  mutate(
-    cases=c(0, diff(cases)),
-    cases=ifelse(cases < 0, 0, cases)
-  )
-
-mercer_spring2021 <- mercer %>%
-  filter(date > as.Date("2021-12-31"), date <= as.Date("2022-03-18"))
-
-nsim <- 100
-
-simulation_omicron_spring_increase_10 <- vector('list')
-
-theta <- 0.1 * mercer_spring2021$cases * 5000/13000
-
-R0vec <- c(2, 4, 6, 8)
-
-paramvec <- expand.grid(R0vec)
-
-for (j in 1:nrow(paramvec)) {
-  tmplist <- vector('list')
-  
-  set.seed(101)
-  for (i in 1:nsim) {
-    ## 5000 undergrad
-    print(paste0(j, ", ", i))
-    ss <- simulate_omicron_switch2(param=drawpar(preport=1, R0=paramvec[j, 1]),
-                           N=5000,
-                           I=0,
-                           R=100/5000,
-                           return=c(rep(TRUE, 700), rep(FALSE, 4300)),
-                           returnfrac=1/28,
-                           datestart=as.Date("2022-01-01"),
-                           datesemester=as.Date("2022-01-01"),
-                           dateinternal=as.Date("2022-01-01"),
-                           dateswitch=as.Date("2022-02-08"),
-                           dateend=as.Date("2022-03-18"),
-                           thetavec=theta,
-                           testfreq_unvacc = 3,
-                           testfreq_vacc = 3,
-                           testfreq_new=7,
-                           Rnewratio=1.1,
-                           vprop=0.99,
-                           boostprop=0.6,
-                           boostdaily=30,
-                           veff_twodose=0.1,
-                           veff_threedose=0.7,
-                           veff_twodose_inf = 0.1 * 0.2/0.9,
-                           veff_threedose_inf = 0.7 * 0.2/0.9,
-                           vaccine_delay=7,
-                           iperiod=5,
-                           ntest = 8000,
-                           asymp_testing=TRUE,
[…]

### 2023-01-22 simulation_omicron_spring/simulation_omicron_spring_increase_100.R

@@ -19,7 +19,7 @@ nsim <- 100
 
 simulation_omicron_spring_increase_100 <- vector('list')
 
-theta <- 0.15 * mercer_spring2021$cases * 5000/13000
+theta <- 1e-5 * mercer_spring2021$cases * 5000
 
 R0vec <- c(2, 4, 6)
 

### 2023-01-22 simulation_omicron_spring/simulation_omicron_spring_increase_100_lowerV.R

@@ -1,0 +1,76 @@
+library(dplyr)
+source("../R/simulate_internal_omicron_switch2.R")
+source("../R/simulate_omicron_switch2.R")
+
+county <- read.csv("../data/us-counties-04-01-2022.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+mercer_spring2021 <- mercer %>%
+  filter(date >= as.Date("2022-01-08"), date <= as.Date("2022-03-25"))
+
+nsim <- 100
+
+simulation_omicron_spring_increase_100_lowerV <- vector('list')
+
+theta <- 1e-5 * mercer_spring2021$cases * 5000
+
+R0vec <- c(2, 4, 6)
+
+paramvec <- expand.grid(R0vec)
+
+for (j in 1:nrow(paramvec)) {
+  tmplist <- vector('list')
+  
+  set.seed(101)
+  for (i in 1:nsim) {
+    ## 5000 undergrad
+    print(paste0(j, ", ", i))
+    ss <- simulate_omicron_switch2(param=drawpar(preport=1, R0=paramvec[j, 1]),
+                           N=5000,
+                           I=0.14,
+                           R=100/5000,
+                           return=c(rep(TRUE, 700), rep(FALSE, 4300)),
+                           returnfrac=1/28,
+                           datestart=as.Date("2022-01-01"),
+                           datesemester=as.Date("2022-01-01"),
+                           dateinternal=as.Date("2022-01-01"),
+                           dateswitch=as.Date("2022-02-08"),
+                           dateend=as.Date("2022-03-18"),
+                           thetavec=theta,
+                           testfreq_unvacc = 3,
+                           testfreq_vacc = 3,
+                           testfreq_new=7,
+                           Rnewratio=2,
+                           vprop=0.99,
+                           boostprop=0.6,
+                           boostdaily=70,
+                           veff_twodose=0.1,
+                           veff_threedose=0.3,
+                           veff_twodose_inf = 0.1 * 0.2/0.9,
+                           veff_threedose_inf = 0.3 * 0.2/0.9,
+                           vaccine_delay=7,
+                           iperiod=5,
+                           ntest = 8000,
+                           asymp_testing=TRUE,
[…]

### 2023-01-22 simulation_omicron_spring/simulation_omicron_spring_increase_100_shortI.R

@@ -1,0 +1,76 @@
+library(dplyr)
+source("../R/simulate_internal_omicron_switch2.R")
+source("../R/simulate_omicron_switch2.R")
+
+county <- read.csv("../data/us-counties-04-01-2022.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+mercer_spring2021 <- mercer %>%
+  filter(date >= as.Date("2022-01-08"), date <= as.Date("2022-03-25"))
+
+nsim <- 100
+
+simulation_omicron_spring_increase_100_shortI <- vector('list')
+
+theta <- 1e-5 * mercer_spring2021$cases * 5000
+
+R0vec <- c(2, 4, 6)
+
+paramvec <- expand.grid(R0vec)
+
+for (j in 1:nrow(paramvec)) {
+  tmplist <- vector('list')
+  
+  set.seed(101)
+  for (i in 1:nsim) {
+    ## 5000 undergrad
+    print(paste0(j, ", ", i))
+    ss <- simulate_omicron_switch2(param=drawpar(preport=1, R0=paramvec[j, 1], D_e=1.5, D_p=2.5, D_s=2.5),
+                           N=5000,
+                           I=0.14,
+                           R=100/5000,
+                           return=c(rep(TRUE, 700), rep(FALSE, 4300)),
+                           returnfrac=1/28,
+                           datestart=as.Date("2022-01-01"),
+                           datesemester=as.Date("2022-01-01"),
+                           dateinternal=as.Date("2022-01-01"),
+                           dateswitch=as.Date("2022-02-08"),
+                           dateend=as.Date("2022-03-18"),
+                           thetavec=theta,
+                           testfreq_unvacc = 3,
+                           testfreq_vacc = 3,
+                           testfreq_new=7,
+                           Rnewratio=2,
+                           vprop=0.99,
+                           boostprop=0.6,
+                           boostdaily=70,
+                           veff_twodose=0.1,
+                           veff_threedose=0.7,
+                           veff_twodose_inf = 0.1 * 0.2/0.9,
+                           veff_threedose_inf = 0.7 * 0.2/0.9,
+                           vaccine_delay=7,
+                           iperiod=5,
+                           ntest = 8000,
[…]

### 2023-01-22 simulation_omicron_spring/simulation_omicron_spring_increase_20.R

@@ -19,9 +19,9 @@ nsim <- 100
 
 simulation_omicron_spring_increase_20 <- vector('list')
 
-theta <- 0.15 * mercer_spring2021$cases * 5000/13000
+theta <- 1e-5 * mercer_spring2021$cases * 5000
 
-R0vec <- c(2, 4, 6, 8)
+R0vec <- c(2, 4, 6)
 
 paramvec <- expand.grid(R0vec)
 

### 2023-01-22 simulation_omicron_spring/simulation_omicron_spring_increase_50.R

@@ -19,9 +19,9 @@ nsim <- 100
 
 simulation_omicron_spring_increase_50 <- vector('list')
 
-theta <- 0.15 * mercer_spring2021$cases * 5000/13000
+theta <- 1e-5 * mercer_spring2021$cases * 5000
 
-R0vec <- c(2, 4, 6, 8)
+R0vec <- c(2, 4, 6)
 
 paramvec <- expand.grid(R0vec)
 

### 2023-01-22 simulation_omicron_spring/simulation_omicron_spring_increase_50_lowerV.R

@@ -1,0 +1,76 @@
+library(dplyr)
+source("../R/simulate_internal_omicron_switch2.R")
+source("../R/simulate_omicron_switch2.R")
+
+county <- read.csv("../data/us-counties-04-01-2022.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+mercer_spring2021 <- mercer %>%
+  filter(date >= as.Date("2022-01-08"), date <= as.Date("2022-03-25"))
+
+nsim <- 100
+
+simulation_omicron_spring_increase_50_lowerV <- vector('list')
+
+theta <- 1e-5 * mercer_spring2021$cases * 5000
+
+R0vec <- c(2, 4, 6)
+
+paramvec <- expand.grid(R0vec)
+
+for (j in 1:nrow(paramvec)) {
+  tmplist <- vector('list')
+  
+  set.seed(101)
+  for (i in 1:nsim) {
+    ## 5000 undergrad
+    print(paste0(j, ", ", i))
+    ss <- simulate_omicron_switch2(param=drawpar(preport=1, R0=paramvec[j, 1]),
+                           N=5000,
+                           I=0.14,
+                           R=100/5000,
+                           return=c(rep(TRUE, 700), rep(FALSE, 4300)),
+                           returnfrac=1/28,
+                           datestart=as.Date("2022-01-01"),
+                           datesemester=as.Date("2022-01-01"),
+                           dateinternal=as.Date("2022-01-01"),
+                           dateswitch=as.Date("2022-02-08"),
+                           dateend=as.Date("2022-03-18"),
+                           thetavec=theta,
+                           testfreq_unvacc = 3,
+                           testfreq_vacc = 3,
+                           testfreq_new=7,
+                           Rnewratio=1.5,
+                           vprop=0.99,
+                           boostprop=0.6,
+                           boostdaily=70,
+                           veff_twodose=0.1,
+                           veff_threedose=0.3,
+                           veff_twodose_inf = 0.1 * 0.2/0.9,
+                           veff_threedose_inf = 0.3 * 0.2/0.9,
+                           vaccine_delay=7,
+                           iperiod=5,
+                           ntest = 8000,
+                           asymp_testing=TRUE,
[…]

### 2023-01-22 simulation_omicron_spring/simulation_omicron_spring_increase_50_shortI.R

@@ -1,0 +1,76 @@
+library(dplyr)
+source("../R/simulate_internal_omicron_switch2.R")
+source("../R/simulate_omicron_switch2.R")
+
+county <- read.csv("../data/us-counties-04-01-2022.csv")
+
+mercer <- county %>%
+  filter(county=="Mercer", state=="New Jersey") %>%
+  arrange(date) %>%
+  mutate(
+    cases=c(0, diff(cases)),
+    cases=ifelse(cases < 0, 0, cases)
+  )
+
+mercer_spring2021 <- mercer %>%
+  filter(date >= as.Date("2022-01-08"), date <= as.Date("2022-03-25"))
+
+nsim <- 100
+
+simulation_omicron_spring_increase_50_shortI <- vector('list')
+
+theta <- 1e-5 * mercer_spring2021$cases * 5000
+
+R0vec <- c(2, 4, 6)
+
+paramvec <- expand.grid(R0vec)
+
+for (j in 1:nrow(paramvec)) {
+  tmplist <- vector('list')
+  
+  set.seed(101)
+  for (i in 1:nsim) {
+    ## 5000 undergrad
+    print(paste0(j, ", ", i))
+    ss <- simulate_omicron_switch2(param=drawpar(preport=1, R0=paramvec[j, 1], D_e=1.5, D_p=2.5, D_s=2.5),
+                           N=5000,
+                           I=0.14,
+                           R=100/5000,
+                           return=c(rep(TRUE, 700), rep(FALSE, 4300)),
+                           returnfrac=1/28,
+                           datestart=as.Date("2022-01-01"),
+                           datesemester=as.Date("2022-01-01"),
+                           dateinternal=as.Date("2022-01-01"),
+                           dateswitch=as.Date("2022-02-08"),
+                           dateend=as.Date("2022-03-18"),
+                           thetavec=theta,
+                           testfreq_unvacc = 3,
+                           testfreq_vacc = 3,
+                           testfreq_new=7,
+                           Rnewratio=1.5,
+                           vprop=0.99,
+                           boostprop=0.6,
+                           boostdaily=70,
+                           veff_twodose=0.1,
+                           veff_threedose=0.7,
+                           veff_twodose_inf = 0.1 * 0.2/0.9,
+                           veff_threedose_inf = 0.7 * 0.2/0.9,
+                           vaccine_delay=7,
+                           iperiod=5,
+                           ntest = 8000,
[…]

### 2023-01-22 simulation_omicron_spring/simulation_omicron_spring_noswitch.R

@@ -19,9 +19,9 @@ nsim <- 100
 
 simulation_omicron_spring_noswitch <- vector('list')
 
-theta <- 0.15 * mercer_spring2021$cases * 5000/13000
+theta <- 1e-5 * mercer_spring2021$cases * 5000
 
-R0vec <- c(2, 4, 6, 8)
+R0vec <- c(2, 4, 6)
 
 paramvec <- expand.grid(R0vec)
 

### 2023-01-22 simulation_omicron_spring/simulation_omicron_spring_null.R

@@ -19,7 +19,7 @@ nsim <- 100
 
 simulation_omicron_spring_null <- vector('list')
 
-theta <- 0.1 * mercer_spring2021$cases * 5000/13000
+theta <- 1e-5 * mercer_spring2021$cases * 5000
 
 R0vec <- c(0)
 

### 2023-01-22 simulation_princeton/simulation_fall_2021_nowaning.R

@@ -20,8 +20,8 @@ nsim <- 100
 
 simulation_fall_2021_nowaning <- vector('list')
 
-R0vec <- c(0.5, 1, 2, 4, 8)
-scalevec <- c(0.005, 0.01, 0.015, 0.02, 0.025) * 10
+R0vec <- c(0.25, 0.5, 1, 2, 4, 8)
+scalevec <- c(5e-6, 7.5e-6, 10e-6, 12.5e-6, 15e-6)
 
 paramvec <- expand.grid(R0vec, scalevec)
 
@@ -48,7 +48,7 @@ for (j in 1:nrow(paramvec)) {
                    boostdaily=0,
                    return=c(rep(FALSE, 5000), rep(TRUE, 8000)),
                    returnfrac=1/16,
-                   thetavec=paramvec[j,2] * mercer_fall2021$cases,
+                   thetavec=paramvec[j,2] * mercer_fall2021$cases * 13000,
                    datestart=as.Date("2021-08-14"),
                    datesemester=as.Date("2021-09-01"),
                    dateinternal=as.Date("2021-09-01"),

### 2023-01-22 doc/response.tex
final edits

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage{color}

\usepackage{xspace}

\usepackage{lmodern}
\usepackage{amssymb,amsmath}

\usepackage[pdfencoding=auto, psdextra]{hyperref}

\usepackage{natbib}
\bibliographystyle{chicago}

\newcommand{\eref}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}} 
\newcommand{\Ro}{\Rx{0}}
\newcommand{\Rc}{\Rx{\textrm{\tiny{contact}}}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}}
\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}

\newcommand{\dd}[1]{\ensuremath{\, \mathrm{d}#1}}
\newcommand{\dtau}{\dd{\tau}}
\newcommand{\dx}{\dd{x}}
\newcommand{\dsigma}{\dd{\sigma}}

\newcommand{\rev}{\subsection*}
\newcommand{\revtext}{\textsf}
\setlength{\parskip}{\baselineskip}
\setlength{\parindent}{0em}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
\newcommand{\jsw}[1]{\comment{green}{JSW}{#1}}
\newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}

\newcommand{\psymp}{\ensuremath{p}} %% primary symptom time
\newcommand{\ssymp}{\ensuremath{s}} %% secondary symptom time
\newcommand{\pinf}{\ensuremath{\alpha_1}} %% primary infection time
\newcommand{\sinf}{\ensuremath{\alpha_2}} %% secondary infection time

\newcommand{\psize}{{\mathcal P}} %% primary cohort size
\newcommand{\ssize}{{\mathcal S}} %% secondary cohort size

\newcommand{\gtime}{\tau_{\rm g}} %% generation interval
\newcommand{\gdist}{g} %% generation-interval distribution
\newcommand{\idist}{\ell} %% incubation period distribution

\newcommand{\total}{{\mathcal T}} %% total number of serial intervals

\begin{document}

\noindent Dear Editor:

Thank you for the chance to revise and resubmit our manuscript.
We have made major revisions to our manuscript to address reviewers' concerns about the methods and assumptions of our work.
We now compare normalized cases divided by the population size---this allows us to directly compare case patterns on campus and those around the surrounding community.
We have reparameterized the community contact rate in our model to account for variation in population sizes across semesters and refitted the model.
We have also added sensitivity analyses and additional figures to show the robustness of our conclusion.
Below please find our detailed responses to reviewers.

\rev{Editor}

\revtext{Associate Editor Comments: }

\revtext{This paper has turned out to be extremely difficult to find sufficient expert reviewers who were willing to review. We have now received two reviews. The first reviewer \#1 is more supportive but the second reviewer \#2 has major concerns that need to be addressed. Please revise the paper to address the concerns and comments of the two reviewers. The revised paper will require further review.}

\revtext{Reviewer \#1 Comments for the Author:}

\revtext{The authors present a retrospective analysis of the COVID-19 epidemic at Princeton University and provide an extensive analysis that generally supports the hypothesis that a superspreading event led to a large uptick in case numbers after the introduction of the Omicron variant. Overall I like the paper. It's well written, and provide a detailed description of the application of a detailed computational model for policy support at an institutional level. It is therefore potentially a valuable contribution and I would support publication after some minor revisions.}

Thank you.

\revtext{Chief among these is the sensitivity of model results (i.e., calibration, and ultimate policy commentary) on the assumptions regarding test sensitivity. The authors claim "Our analysis highlights the power of mass asymptomatic testing" but the assumption is that a PCR test is 95\% sensitive after latency, and that case detection leads to full case isolation. These are both extremely strong assumptions, that are not conservative with respect to the claim. I recommend examining the results of Hellewell et al ("Estimating the effectiveness of routine asymptomatic PCR testing...") for strong evidence that sensitivity is much lower than claimed here. Also see Zachreson et al ("COVID-19 in low-tolerance border quarantine systems...") for a detailed individual-based implementation of dynamic test sensitivity.}

We agree that 95\% sensitivity seem too strong.
Hellewell et al estimated that the probability of detecting an infection from a PCR peaks at 77\% (54–88\%) 4 days after infection, and decreases to 50\% (38–65\%) by 10 days after infection.
But this estimate implicitly accounts for latency, whereas we assume 95\% sensitivity after sensitivity.
Therefore, while our sensitivity peaks at 92\% 5 days after infection, it decreases to 20\% by 10 days after infection (due to recovery from infections), which is considerably lower than what was estimated by Hellewell et al.
Therefore, while the individual-level sensitivity is not dynamic (a step function changing from 0 to 0.95 and back down to 0), the population-level sensitivity is dynamic.
We have added the following paragraph to our main text:

``The 95\% sensitivity assumption may seem too high. 
For example, \cite{hellewell2021estimating} estimated that the probability of detecting an infection from a PCR peaks at 77\% (54–88\%) 4 days after infection, and decreases to 50\% (38–65\%) by 10 days after infection.
These estimates are considerably lower than our assumption because their estimates implicitly account for the latent period.
At the individual level, we assume that an infected individual has no detectable infection (0\% sensitivity) during their latent period and 95\% sensitivity during their various stages of infectious periods.
At the population level, this assumption translates to a peak sensitivity of 92\% by 4 days after infection, decreasing to 20\% sensitivity by 10 days after infection.
Our assumption leads to a much lower PCR sensitivity 10 days after infection because we only model the PCR sensitivity during the infectious period.
In reality, PCR can detect infections even after a person stops shedding infectious virus---we did not include this component in our model because it would not affect the effectiveness of isolation strategy in reducing transmission.''

We further note that the full case isolation was what was happening at Princeton University during the investigation of the study period.
However, we agree that this might be feasible in some institutions. We have added the following phrase:

``while this assumption reflects the isolation policy in Princeton University during the investigation period, it may be inapplicable in studying institutional outbreaks in general.''

\revtext{There are a number of other issues that are less important to my overall recommendation that I list below by section:}

\revtext{Results:}

\revtext{1) Figure 2 (B, E, H) - I would personally prefer to see a representative trajectory (i.e., one that is central in the ensemble based on summary statistics) than a median, since it's being compared to a single real-world trajectory, the 90\% quantiles can stay though. This would give a better illustration of the model dynamics (w.r.t. temporal fluctuations). 
This figure could also benefit from a clearer illustration of which parameters from the grids in (A, D, G) correspond to the ensembles in B, E, H.}

We now show a simulation with the best goodness of fit (least sum of squared errors) alongside a media. 
We agree that showing an actual simulation is important for capturing the variability in the case trajectory.
We also feel that the median is useful for understanding the general trend and decided to keep it in place.
We also added a circle to panels A, D, and G to show the best fitting parameter.

\revtext{2) There is a minor contradiction between Fig 2. A and the statement: "We find that a low level of contacts R contact = 0.5 and a small amount of community transmission $\theta = 0.015$ is most consistent with the observed epidemic dynamics in fall 2020 (Fig. 2A)." in Fig 2A, the minimum sits at Theta = 0.02, and is not definitive over the range > 0.01.}

This was a typo on our side. We have now fixed this.

\revtext{3) it looks like lower values of R could have been explored - what is the reason for the lower bound of 0.5?}

We did this for computationally efficiency. We now explore R as low as 0.25.
We could explore even lower values of R but we feel that lower values may be unrealistic.

\revtext{4) Please explain "Thanksgiving" to non-Americans. This should be indicated visually in Fig. 2B}

We now explain Thanksgiving the first time it is mentioned:

``Thanksgiving---a national holiday in the US during which many students''

\revtext{5) "all returning individuals are assumed to be quarantined for 14 days and tested upon returning." please elaborate (how is quarantine implemented?)}

We have revised this sentence:

``In the beginning of the semester, all returning students were required to quarantine in their rooms for 14 and tested upon returning by the university---in our model, this was implemented by preventing returning students from getting infected or infecting other individuals.''

\revtext{6) Regarding Theta: The model should be refactored so that theta is per-capita on-campus population, the decision not to do so leads to re-scaling of the parameter to campus population fractions in the final section of the results and is confusing. This would probably also help explain why the values of theta are so much higher for the Delta period after vaccination (more people on campus). Finally, the parameter is variously referred to as 'contact rate' and 'transmission rate', this should be clarified in the methods description (they seem to mean the same thing for unvaccinated populations, but the definition gets fuzzier once the transmission event itself is subject to a Bernoulli trial given "infectious contact").}

Thank you for pointing this out. 
We have reparameterized the parameter theta and reran our simulations.
We have also added the following description .

``In particular, we assume that infectious contacts from local or regional community can be made at random to anyone on campus.
These contacts are modeled using a Poisson distribution with a time-varying mean, which is calculated by multiplying the daily number of cases by the community contact rate $\theta$ and the population size on campus $N$.
More precisely, $\theta$ is the probability that an infected individual from the community makes an infectious contact with an individual on campus per capita campus population. 
By further multiplying this probability with the population size $N$, we are essentially assuming a density dependent contact, where a higher population size on campus leads to more infections from the community.''

\revtext{7) The Omicron wave part has a lot of ad-hoc assumptions about policies (i.e., 70 boosters per day), it would be good to know which of these are guessed, and which are supported through consultation etc.}

We have revised this section to make our assumptions clearer:

``Based on known vaccination statuses, we assume that 99\% of students are vaccinated with 60\% of them being boosted as of January 1, 2022.
Since all students were required to receive booster shots before returning to campus, we assumed that 70 booster shots were given on each day---this assumption allows all students to be boosted in 28 days.
To match the high numbers of cases on the week ending January 7, 2022, we assume 14\% of the students present on campus are infected as of January 7, 2022 (roughly 100/700).
To account for students who were infected with the Omicron variant during the fall semester, we assume that 100 students are already immune to Omicron infection at the beginning of the spring semester---this roughly corresponds to the number of PU cases that were reported in December.''

\revtext{8) Figure 3: do the 9 plots on the right have different reproductive ratios? This should be clearly stated (Rcontact is given in the row label, but the column label gives a reproductive ratio multiplier, so it's unclear how assumed transmissibility varies between panels).}

Thank you for pointing this out. The baseline Rcontact does not change within each row; only the reproductive ratio multiplier changes. 
We have added the following sentences to clarify this:

``
For each row, we assume a fixed value of baseline contact reproduction number $\Rc$ ranging from 2 to 6 across rows.
Then, we simulate increase in $\Rc$ at the time of policy change (indicated by column labels).
''

\revtext{9) Please be more clear about the policy change (dashed line in Fig 3). Was this strictly a change to gathering restrictions? Are you implying causation with the super-spreading event? I think the messaging around the relationship between the observed dynamics and the enacted policies needs to be stated more clearly and explicitly.}

There was a change to gathering restriction and testing frequency. This is explained in the Descriptive analysis section:

``Coinciding with the decrease in the campus and local cases numbers, the gathering policy was updated on February 8, 2022 to allow food in events were no longer limited to 20 people;
in addition, the testing frequency was reduced to once a week.''

We have also rephrased the following sentence to make the causality clearer:

``Following the policy change, a large gathering event was held on campus, which resulted in an outbreak with high case numbers persisting until Spring Break (March 5th, 2022).''

\revtext{Discussion:}

\revtext{10) Please explain the phrase "our homogeneous mixing assumption is conservatively pessimistic." One could argue that assuming homogeneity is not 'conservatively pessimistic', rather, it assumes the highest rate of case growth possible for a given natural transmissibility and secondary case dispersion (because transmission is unconstrained by group size beyond the size of the total population), this means the calibrated natural transmission rates (i.e., Rcontact) will be biased down, for the same case incidence. Lower estimates of natural transmission rates (vis. lower viral load of an index case) could lead to liberal conclusions regarding the required stringency of interventions, so it needs to be made clear in what policy decision context the statement 'conservatively pessimistic' applies.}

We meant conservatively pessimistic from a dynamical perspective, allowing for rapid growth.
We agree that this terminology is confusing in a policy decision context. 
We have tried to clarify the consequences our assumption and decided to remove the term ``pessimistic'' to avoid confusion.

``For example, we assume conservatively that the entire university populations mix homogeneously and have identical campus and community contact rates (captured by $\Rc$ and $\theta$, respectively).
This assumption can lead to the fastest epidemic growth rates because transmission is not limited by the size of the contact network---in other words, our estimates of the reproduction will be necessarily low, making the epidemic easier to control.
In reality, increases in cases were often associated with specific transmission clusters, suggesting heterogeneity in transmission patterns.
Contact levels also likely differ between different groups:
for example, faculty and staff members are more likely to interact with community members than undergraduate students and would be at a higher risk for community infections \citep{frazier2022modeling}.''

\revtext{11) The paper claims not to model behaviour. While this is strictly true, because the model does not incorporate behaviour explicitly, it is a bit misleading because the model does implicitly model behaviour by attaching case importation rates to the community case numbers (which are a function of behaviour). A better way of explaining the model would be (in my opinion) that it implicitly accounts for changes in community behaviour but does not explicitly simulate behavioural change in the campus environment.}

Thank you for pointing this out. We have reworded this sentence as the following:

``We also do not account for explicit changes in behavior on campus and assume constant $\Rc$ throughout each semester.
Instead, we implicitly account for behavioral changes in the community by modeling community transmission to campus as a function of community case numbers.
While we cannot rule out the possibility that behavioral changes on campus could have contributed to various epidemics (e.g., the Omicron wave beginning in the fall semester of the 2021--2022 academic year), we were able to capture the majority of epidemic patterns without modeling them---when the majority of transmission is caused by imported cases from the community, we expect behavioral changes on campus to have relatively weaker effects on overall transmission dynamics.''

\revtext{12) "Our analysis highlights the power of mass asymptomatic testing" (see above main comment)}

We have rephrased this sentence:

``First, our analysis highlights the power of mass asymptomatic testing for epidemic measurement and planning---even if PCR testing may have lower sensitivity than what we assumed here \citep{hellewell2021estimating}, mass asymptomatic testing can still help track ongoing epidemic dynamics in real time''

\revtext{13) "preventing large gatherings can help prevent large superspreading events in the midst of a rising epidemic" The model does not simulate gatherings, so this claim should be connected to the results explicitly or it should be removed.}

This was a comment from our Princeton experience rather than simulation. We agree that our model does not simulate gathering and so we have removed this comment.

\revtext{14) "intervention measures placed on campuses must continue to adapt and change" is there a clear statement in this work about the policy changes that led up to the Omicron outbreak? The paper implies such claims, but I do not see them concretely stated (which is frustrating - this is related to point (9)).}

Please see our response to point 9.

\revtext{Methods:}

\revtext{15) Regarding Rcontact, is there a difference between this and the generally-used R0? The definition seems the same... if not, it would be good to know the precise difference.}

Rcontact and R0 are largely similar. But we wanted to use a different terminology to emphasize their subtle differences because Rcontact is a parameter in our individual-based model, which makes a specific set of assumptions that are different from continuous time SIR models using ordinary differential equations. We have tried to clarify the definition of Rcontact.

``We note that the definition of the contact reproduction number $\Rc$ is similar to the standard definition of basic reproduction number $\RR_0$.
The main difference is that the contact reproduction number models the number of total contacts, rather than infections. 
Since infected individuals make their contacts at random with replacement, the same susceptible person could be contacted multiple times by the same or different infected individual during a time step---all these overlapping contacts will result in one infection.
Therefore, the number of actual infections may be smaller than the number of contacts, especially since contacts can also land on non-susceptible individuals.
We also note that the contact reproduction number implicitly accounts for all intervention measures that we do not model explicitly, such as social distancing and contact tracing---therefore, $\Rc$ is similar to the effective reproduction number, which typically accounts for intervention efforts.
However, our contact reproduction number does not account for the effects of asymptomatic testing or vaccination, which are modeled separately.''

\revtext{Overall, I like the paper and appreciate having the opportunity to review it. Understand that my critiques are intended to help improve the quality of the work prior to publication, which I generally support.}

Thank you for your review.

\revtext{Reviewer \#2 Comments for the Author:}

\revtext{In this manuscript, Park et al. retrospectively examine the transmission patterns of COVID-19 at Princeton University and across the wider community in Mercer County, NJ. The authors explore correlations between university and community case counts and also employ a mechanistic modeling framework to see if patterns in transmission can be explained by changes in R-effective and transmission from the broader community. Retrospective analyses like this are important, and valuable evidence generation for decision makers.}

Thank you.

\revtext{I had two main concerns about the conclusions as presented that I hope the authors can address. The first is that while Princeton is an ideal ecosystem because of its asymptomatic screening protocols, the surrounding community is not necessarily one—this raises concerns about how differences in testing practices could be controlled for/accounted for in the correlation analysis.}

\revtext{Most estimates seem to a huge reduction in case ascertainment during the Omicron surge, which makes comparisons between the broader community and Princeton tricky for that time point. At the very least, I would encourage the authors to control for differences in denominator (e.g., converting to per 100K/day) and/or exploring test positivity as an additional metric to see if their observations still hold.}

We have revised our analysis and figures to account for changes in population sizes across the semester. See below for a detailed response.
We also present testing volumes over time and show that they are reasonably stable across each semester.

\revtext{The second is that I wasn’t necessarily convinced that the authors could conclude with certainty that the discrepancies during the Omicron period were driven by superspreading events because the way that this is created in the model is by seeding new infections, not by adjusting contact structure per se. Some more thoughts here: (1) I think the vaccine effectiveness estimates stated for BA.1/BA.2 were overly optimistic, so that would be something to explore further; and (2) the model (in theory) should allow for some superspreading as a consequence of the overdispersion parameter used for the negative binomial distribution of infectious contacts. I’d be curious to see if the model can replicate that change more mechanistically by adjusting the mean and overdispersion of the binomial distribution.}

We recognize that our writing was unclear before.
But we know from the observation that the large outbreak during the Omicron period was driven by superspreading event related to a campus event.
We could not describe the event in detail due to privacy reasons, which led to unclear writing. 
We have tried to make the causality clearer throughout (see response to Reviewer 1's comment 9).
We also account for the possibility of shorter infectious periods and lower vaccine effectiveness.
See below for detailed responses for each point.

\revtext{Main comments:}

\revtext{Fig. 1- you might consider converting case counts for Princeton and Mercer County into cases per 100,000 per day (or some other denominator quantity). These plots are comparing cases for populations of ~15,000 for Princeton and ~360,000 for Mercer County, which makes Fig. 1 panels I-L a bit difficult to parse at first—by controlling the denominator you could more easily compare to a 1:1 slope for panels E-J and a 1:1 ratio for panels I-L. This approach would also account for the changing student body size at Princeton across semesters with reopening, etc.}

Thank you for the suggestion. 
We have changed the main figure and corresponding supplementary figures for New York and Philadelphia to show weekly cases per 100,000.
We have revised the writing accordingly and added a new Supplementary Figure comparing cases per 1000 for each population group on Princeton campus (Figure S1).
We decided to keep the raw cases for comparisons with model projections since we are no longer comparing populations of two different sizes.

\revtext{Lines 165-169- a key piece of this story seems to be variant predominance—you could consider adding a new figure with variant proportions through time or dashed vertical lines to Fig. 1 time courses indicating when a new variant reached dominance.}

We have variant data but we feel analyzing variant data is beyond the scope of this paper.
We decided not to add the variant data in this paper also because it involves a different group of individuals analysing the data.
And so we prefer to keep that analysis as a separate paper.

\revtext{Lines 214-226: the assumption that testing patterns remain the same on campus and off campus throughout the entire investigation period doesn’t seem realistic. In particular for the beginning of the Omicron period, the Princeton university campus continued with surveillance screening, while the general population likely experienced significantly decreased ascertainment rates in part due to the increased availability and use of at home tests which are not mandated reporting (e.g., one estimate of a change of ascertainment from 50\% pre-Omicron to 20-29\% during BA.1; \url{https://www.medrxiv.org/content/10.1101/2022.04.22.22274198v3.full.pdf}). Therefore, I would hesitate to conclude that the lack of correlation was indicative of a lack of community transmission—changes in testing patterns could really be driving that pattern for the final semester.}

Thank you for pointing this out. 
We meant that the testing patterns remained constant for each semester, rather than the entire investigation period.
We also meant that the case ratio would stay the same if testing patterns and transmission conditions were to remain constant, but we are not necessarily assuming that the testing patterns remained constant.
We have tried to make the writing of this paragraph clearer while acknowledging changes in testing patterns over the investigation period.

Also, we know that the outbreak during the final semester was associated with a specific campus event among students on one specific day, which we unfortunately cannot described in detail due to privacy reasons. 
So we know that the lack of correlation was likely driven by a lack of community transmission in this case. 
We have tried to clarified this point.

\revtext{Lines 243-245- Infectious periods changed with variants, e.g., Omicron << Delta. Was that component explored at all?}

We now provide simulations that assume shorter infectious and latent period during the Omicron outbreak:

``We considered the possibility that the Omicron variant can have shorter latent and infectious periods by decreasing the mean duration of latent, pre-symptomatic, and (a)symptomatic stages of infection by 0.5 days (therefore a total of 1.5 reduction in the duration of infection).
In this case, a shorter generation interval can lead to faster growth rate given the same values of $\Rc$ \citep{wallinga2007generation}.
However, we find that the effects of shorter infection has small effects on the overall dynamics (Supplementary Figure S13).''

\revtext{Lines 296-305- While reading through this the first time, I was wondering about differences in vaccine effectiveness during Omicron emergence however—maybe hint to the reader here that you address that later or separately?}

We have added the following sentence:

``We note that these assumptions are specific to the Delta variant---we discuss vaccine effectiveness against the Omicron variant later on.''

\revtext{Lines 453-454- I think your vaccine effectiveness estimates against Omicron are still quite optimistic, which could explain why ramping up your R-value didn’t really do much. In practice, we’ve seen closer to 30-50\% effectiveness against infection for example vs. 70\% pre-waning (\url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9398552/})}

We now provide simulations that assume 30\% vaccine effectiveness as a supplementary figure with the following text in the main manuscript:

``We also considered the possibility that the vaccine effectiveness against the Omicron variant might be lower by repeating the same analysis with 30\% effectiveness against infection \citep{tan2023vaccine}.
When the baseline $\Rc$ is low ($\Rc=2$), increasing $\Rc$ still does not increase the number of cases sufficiently.
When we assume an intermediate value of $\Rc=4$, the model does a better job at capturing the dynamics but it does so by overestimating the trough before the policy change and underestimating the peak after the policy change.
When we assume a high value of $\Rc=6$, the model overestimates both the trough and the peak.''

\revtext{Fig. 3-4- Apologies if you’ve stated this elsewhere and I’ve missed it, but have you taken a look at your testing volume through time? I know that the PU policy was x \# of tests per week based on vaccination status, but how was that enforced? I’m wondering if there is as much variation in the y-axis if you look at test positivity vs. raw number of cases? Another thing to consider, since I’m not sure about testing cadence and frequency would be to do some rolling averages to help smooth out differences across days/weeks.}

We now show testing volume through time in Supplementary Figure S1.
Everyone was required to be tested in order to be present on campus. 
And as we see in Supplementary Figure S1, there was very good compliance and stable testing volumes within each semester, except during the end of fall semester of the 2021--2022 academic year (reflecting Omicron circulations) and the beginning of spring semester of the 2021--2022 academic year (reflecting students returning to school).
Even then, these changes do not correlate with increase in cases---for example, the increase in testing volume at the end of fall semester of the 2021--2022 academic year is very sudden (doubling in a week), but the changes in cases at the same period is much smaller.

We agree there is variation in cases within a week but testing patterns were fairly stable across weeks, which is why we chose to look at weekly data, instead of daily data.
We feel that showing raw variation in weekly cases (rather than rolling averages) would be more accurate because we are also able to identify the sources of most variations (e.g., school holidays and classes ending).

\revtext{Lines 456-457- instead of artificially introducing new infections, what happens if you decrease $k$ your overdispersion parameter to increase the potential variance (i.e., allow for bigger gatherings?)}

Artificially introducing new infection is meant to emulate the increase in heterogeneity.
Decreasing $k$ alone would certainly increase variance but we have no control over when large outbreaks will happen, meaning that most of the simulations will not be able to match the data.

\revtext{Minor comments:}

\revtext{Line 40- double check consistency of variant name capitalization throughout (e.g., “omicron” vs. “Omicron”)}

Fixed.

\revtext{Lines 78-81- infrastructure including ventilation would also play a key role here}

We have added ventilation infrastructure to this sentence.

\revtext{Line 192- “counties” instead of “countries”?}

Fixed.

\revtext{Lines 270-271- in reality, those contacts are likely much more structured leading to a more modular network (e.g. dormmates would have repeated contacts, students would be more likely to have classes with others in that same major); how would you see less mixing affecting your model results and interpretation?}

We have added the following paragraph:

``In reality, the contact structure among the campus population is likely more structured, exhibiting strong assortativity.
For example, undergraduate students are more likely to mix with other undergraduate students, rather than graduate students or faculty and staff members.
Even among undergraduate students, students are more likely to mix with their close friend group than with other students.
On one hand, assortative mixing may lead to faster epidemic growth within certain population groups;
on the other hand, it can also make the disease more difficult to spread among other groups that have lower contact rates.
Therefore, predicting the impact of structured contact network requires more detailed information about whether the majority of cases were infected at random or from certain groups of the campus population.
For simplicity, we assume random mixing throughout the paper---nonetheless, allowing for overdispersion in transmission is expected to emulate variability in epidemic growth rates driven by complex contact structures \citep{lloyd2005superspreading}.''

\revtext{Line 278- it would be helpful to label $\theta$ as your “community transmission” parameter on first use, since you refer to it that way later, but not stated explicitly here.}

Based on suggestions by Reviewer 1, we now normalize $\theta$ by population size $N$ and provide a clear definition the first time it is used>

\revtext{Line 387- “cannot” instead of “can”?}

``can'' is correct. We have revised this sentence to make the point clearer:

``These simulations suggest that an increase in the number of cases in November can be explained by a combination of waning immunity alone without requiring additional changes in transmission dynamics (note we do not allow $\theta$ or $\Rc$ to vary over time)---we see that extending the simulation beyond November 26th still captures the increase in cases.''

\revtext{Lines 162-164- state date of policy implementation to make link clearer for Figs. 3 \& 4}

Done.

\bibliography{university-covid}

\end{document}

### 2023-01-22 doc/university-covid.tex
final edits

@@
 Daniel A. Notterman\textsuperscript{5}
-C. Jessica E. Metcalf\textsuperscript{1}
+C. Jessica E. Metcalf\textsuperscript{1,6}
 Bryan T.\ Grenfell\textsuperscript{1,6}
@@
 
-Mathematical models have played a crucial role in guiding pandemic responses. 
+Mathematical models have played a crucial role in exploring and guiding pandemic responses. 
 University campuses present a particularly well-documented case for institutional outbreaks, thereby providing a unique opportunity to understand detailed patterns of pathogen spread.
@@
 These findings are corroborated by our model predictions, which indicate that the amount of on-campus transmission was likely limited during much of the wider pandemic until the end of 2021.
-Finally, we find that a super-spreading event likely played a major role in driving the recent Omicron variant outbreak on the Princeton University campus during the spring semester of the 2021--2022 academic year.
+Finally, we find that a super-spreading event likely played a major role in driving the Omicron variant outbreak on the Princeton University campus during the spring semester of the 2021--2022 academic year.
 Despite large numbers of cases on campus in this period, case levels in surrounding communities remained low, suggesting that there was little spillover transmission from campus to the local community.
@@
 
-Mathematical models have played a significant role in guiding these pandemic responses and devising control strategies \citep{cobey2020modeling,holmdahl2020wrong,metcalf2020mathematical,koelle2022changing}.
+Mathematical models have played a significant role in guiding these pandemic responses and exploring control strategies \citep{cobey2020modeling,holmdahl2020wrong,metcalf2020mathematical,koelle2022changing}.
 Models can help monitor key parameters that govern epidemic dynamics \citep{kraemer2021monitoring} and retrospectively estimate the impact of intervention measures in reducing transmission \citep{flaxman2020estimating}.
@@
 
-We begin with a descriptive analysis of the PU outbreak (\fref{princeton}), and present modeling analyses of the individual epidemics during 2020--2022.
+We begin with a descriptive analysis of the PU outbreak (\fref{princeton}), then present modeling analyses of the individual epidemics during 2020--2022.
 PU is located in Mercer County, New Jersey, USA;
-the population comprises of 5267 undergraduate students, 2946 graduate students, and around 7000 faculty and staff members.
+the population comprises 5267 undergraduate students, 2946 graduate students, and around 7000 faculty and staff members.
 For simplicity, we divided the epidemic into four time periods representing four semesters across two academic years: Fall 2020--2021 (August 24, 2020--January 1, 2021; \fref{princeton}A), Spring 2020--2021 (January 16, 2021--May 14, 2021; \fref{princeton}B), Fall 2021--2022 (August 14, 2021--December 31, 2021; \fref{princeton}C), and Spring 2021--2022 (January 1, 2022--March 18, 2022; \fref{princeton}D).
@@
 The highest number of cases was reported among faculty and staff members ($=169$), followed by graduate students ($=41$), and undergraduate students ($=4$).
-Even when we control for the differing population sizes among these groups, we find that a considerably larger amount of cases were reported among faculty and staff members (84.5 cases per 1000) than from graduate students (41 cases per 1000) (Supplementary Figure S2A)---we exclude undergraduate student population from this comparison due to a very low number of undergraduate students present on campus during this period.
+Even when we control for the differing population sizes among these groups, we find that a considerably larger amount of cases were reported among faculty and staff members (84.5 cases per 1000) than from graduate students (41 cases per 1000) (Supplementary Figure S2A)---we exclude the undergraduate student population from this comparison due to a very low number of undergraduate students present on campus during this period.
 
@@
 
-For the \textbf{spring semester of the 2021--2022 academic year}, all eligible students and faculty and staff members were required to be boosted.
+For the \textbf{spring semester of the 2021--2022 academic year}, all eligible students and faculty and staff members were required to have obtained booster vaccination.
 By the beginning of the semesters, 65\% of undergraduate students, 71\% of graduate students, and 82\% of faculty and staff members were boosted.
@@
 Following the policy change, a large gathering event was held on campus, which resulted in an outbreak with high case numbers persisting until Spring Break (March 5th, 2022).
-The timing of this outbreak also coincided with a rapid turnover of the Omicron subvariant BA.2---the proportion of BA.2 subvariant reached 93.5\% (372/398) compared to 26.9\% (14/52) from the previous week.
+The timing of this outbreak also coincided with a rapid increases in the Omicron subvariant BA.2 cases---the proportion of BA.2 subvariant reached 93.5\% (372/398) compared to 26.9\% (14/52) from the previous week.
 On March 7, 2022, mask mandates were lifted and testing frequency was reduced to once a month.
@@
 
-Similar picture emerges for Philadelphia except for spring 2020: fall 2020--2021 ($\rho = 0.87$ (95\% CI: 0.68--0.95; $p < 0.001$), Supplementary Figure S6A); spring 2020--2021 ($\rho = 0.27$ (95\% CI: $-0.24$--0.66; $p = 0.30$), Supplementary Figure S6B); fall 2021--2022 ($\rho = 0.89$ (95\% CI: 0.74--0.96; $p < 0.001$), Supplementary Figure S6C);
+A similar picture emerges for Philadelphia except for spring 2020: fall 2020--2021 ($\rho = 0.87$ (95\% CI: 0.68--0.95; $p < 0.001$), Supplementary Figure S6A); spring 2020--2021 ($\rho = 0.27$ (95\% CI: $-0.24$--0.66; $p = 0.30$), Supplementary Figure S6B); fall 2021--2022 ($\rho = 0.89$ (95\% CI: 0.74--0.96; $p < 0.001$), Supplementary Figure S6C);
 and spring 2021--2022 ($\rho=0.46$ (95\% CI: $-0.19$--0.83; $p=0.15$), Supplementary Figure S6D).
@@
 Instead, we find that the ratio generally hovers above 1 during the fall semester of the 2020--2021 academic year even though there was little to no documented transmission on campus (\fref{princeton}I).
-This pattern likely reflects higher testing rate on campus, thereby resulting in a higher case ascertainment rate.
+This pattern likely reflects a higher testing rate on campus, thereby resulting in a higher case ascertainment rate.
 For the most part of the spring semester of the 2020--2021 academic year and the fall semester of the 2021--2022 academic year, the case ratios hover around 1 (\fref{princeton}J--K).
@@
 Therefore, the number of actual infections may be smaller than the number of contacts, especially since contacts can also land on non-susceptible individuals.
-We also note that the contact reproduction number implicitly accounts for all intervention measures that we do not model explicitly, such as social distancing and contact tracing---this is different from $\RR_0$, which typically does not include the effects of intervention measures.
-However, the contact reproduction number does not include asymptomatic testing or vaccination, which are modeled separately.
+We also note that the contact reproduction number implicitly accounts for all intervention measures that we do not model explicitly, such as social distancing and contact tracing---therefore, $\Rc$ is similar to the effective reproduction number, which typically accounts for intervention efforts.
+However, our contact reproduction number does not account for the effects of asymptomatic testing or vaccination, which are modeled separately.
 We further decompose $\Rc$ into pre-symptomatic $\mathcal R_p = \beta_p D_p$ and (a)symptomatic $\mathcal R_s = \beta_s D_s$ reproduction numbers, where $\beta_p$ and $\beta_s$ represent the corresponding infectious contact rates during pre-symptomatic and (a)symptomatic stages, respectively.
@@
 
-In reality, the contact structure among the campus population is likely more structured and exhibit strong assortativity.
+In reality, the contact structure among the campus population is likely more structured, exhibiting strong assortativity.
 For example, undergraduate students are more likely to mix with other undergraduate students, rather than graduate students or faculty and staff members.
@@
 On the other hand, it can also make the disease more difficult to spread among other groups that have lower contact rates.
-Therefore, predicting the impact of structured contact network requires a more detailed information about whether the majority of cases were infected at random or from certain groups of the campus population.
+Therefore, predicting the impact of structured contact network requires more detailed information about whether the majority of cases were infected at random or from certain groups of the campus population.
 For simplicity, we assume random mixing throughout the paper---nonetheless, allowing for overdispersion in transmission is expected to emulate variability in epidemic growth rates driven by complex contact structures \citep{lloyd2005superspreading}.
@@
 We further assume that symptomatic individuals are isolated immediately when they submit their samples until they receive negative results.
-All individuals who test positive are required to isolate (following the same isolation rule as described earlier) and are exempt from asymptomatic testing for 90 days---while this assumption reflects the isolation policy in Princeton University during the investigation period, it may be inapplicable in studying other institutional outbreaks.
+All individuals who test positive are required to isolate (following the same isolation rule as described earlier) and are exempt from asymptomatic testing for 90 days---while this assumption reflects the isolation policy in Princeton University during the investigation period, it may be inapplicable in studying institutional outbreaks in general.
 Isolated individuals are assumed to no longer transmit infections.
@@
 As most students, as well as faculty and staff members, had received two doses of vaccination in the beginning of fall 2021, we do not distinguish the first and second doses.
-Instead, we assume that all vaccinated individuals have 90\% reduced susceptibility and 20\% reduced transmissibility at the beginning of the semester---these assumptions are consistent with recent estimates by \cite{prunas2022vaccination} that vaccination with BNT162b2 reduces susceptibility by 89.4\% (95\% CI: 88.7\%--90.0\%) and infectiousness by 23.0\% (95\% CI: $-11.3\%$--46.7\%) against the Delta strain.
+Instead, we assume that all vaccinated individuals have 90\% reduced susceptibility and 20\% reduced transmissibility at the beginning of the semester---these assumptions are consistent with recent estimates by \cite{prunas2022vaccination} that vaccination with BNT162b2 reduces susceptibility by 89.4\% (95\% CI: 88.7\%--90.0\%) and infectiousness by 23.0\% (95\% CI: $-11.3\%$--46.7\%) against the Delta variant.
 Based on \cite{tartof2021effectiveness}, vaccine efficacy against susceptibility is allowed to exponentially wane from 90\% to 50\% in 20 weeks (and continues to wane at the same rate) for each vaccinated individual;
@@
 Epidemiological conditions and intervention measures changed throughout the spring semester of the 2021--2022 academic year.
-We therefore extend to model to account for these alterations and focus on the outbreak patterns among undergraduate students.
-First, based on \citep{ferguson2021report}, we assume that two and three doses of vaccines reduce susceptibility against the Omicron variant by 10\% and 70\%, respectively. 
+We therefore extend the model to account for these alterations and focus on the outbreak patterns among undergraduate students.
+First, based on \citep{ferguson2021report}, we assume that two and three doses of vaccines reduce susceptibility against the early Omicron variant by 10\% and 70\%, respectively. 
 We also assume that the transmissibility of Omicron is reduced proportionally following the previously assumed 90--to--20 ratio for the Delta variant;
@@
 
-We also considered the possibility that the vaccine effectiveness against the Omicron variant might be lower and run the same analysis with 30\% effectiveness against infection \citep{tan2023vaccine}.
+We also considered the possibility that the vaccine effectiveness against the Omicron variant might be lower by repeating the same analysis with 30\% effectiveness against infection \citep{tan2023vaccine}.
 When the baseline $\Rc$ is low ($\Rc=2$), increasing $\Rc$ still does not increase the number of cases sufficiently.
@@
 An increase in the number of cases by the end of November 2021 is consistent with the increase in the levels of community cases and waning immunity.
-Finally, our analysis highlights the role of super-spreading events in driving the spread of the Omicron variant on the PU campus.
+Finally, our analysis highlights the potential role of super-spreading events in driving the spread of the Omicron variant on the PU campus.
 
@@
 The degree to which community coupling affects campus transmission also depends on the campus. 
-Although Princeton University is located in a small town (Mercer County) with a population of ~390,000 (\url{www.census.gov}), it is located near large cities, such as New York City and Philadelphia, which can drive infections in smaller cities nearby \citep{grenfell1997meta}.
+Although Princeton University is located in a small county (Mercer County) with a population of ~390,000 (\url{www.census.gov}), it is located near large cities, such as New York City and Philadelphia, which can drive infections in smaller cities nearby \citep{grenfell1997meta}.
 For example, contact tracing efforts from Boston University, which is located in a large metropolitan area, found that more than 50\% of infections among Boston University affiliates with known exposures could be attributed to sources outside of the university campus \citep{hamer2021assessment}.
@@
 Our analysis also suggests that comparing the ratios between the cases on university campuses and neighboring communities can also provide a useful measure for how well a university campus is controlling the epidemic;
-however, that this ratio needs to be interpreted with caution as it is sensitive to changes in testing patterns as well as the numbers of students on campus.
+however, that this ratio needs to be interpreted with caution, as it is sensitive to changes in testing patterns as well as the numbers of students on campus.
 For example, the ratios of cases can suddenly change during holidays when students are away from campus.
@@
 
-Despite the simplicity of the analysis, our study provides important lessons for controlling SARS-CoV-2 outbreaks on university campuses in general.
+Despite the simplicity of the analysis, our study provides important lessons for controlling SARS-CoV-2 and similar outbreaks on university campuses in general.
 First, our analysis highlights the power of mass asymptomatic testing for epidemic measurement and planning---even if PCR testing may have lower sensitivity than what we assumed here \citep{hellewell2021estimating}, mass asymptomatic testing can still help track ongoing epidemic dynamics in real time.
@@
 \begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_testing.pdf}
+\includegraphics[width=0.9\textwidth]{../figure_princeton_new/figure_princeton_testing.pdf}
 \caption{

### 2023-02-22 doc/response.tex
am ready to resubmit this

@@
 Hellewell et al estimated that the probability of detecting an infection from a PCR peaks at 77\% (54–88\%) 4 days after infection, and decreases to 50\% (38–65\%) by 10 days after infection.
-But this estimate implicitly accounts for latency, whereas we assume 95\% sensitivity after sensitivity.
+But this estimate implicitly accounts for latency, whereas we assume 95\% sensitivity during the infectious period after 0\% sensitivity during latency.
 Therefore, while our sensitivity peaks at 92\% 5 days after infection, it decreases to 20\% by 10 days after infection (due to recovery from infections), which is considerably lower than what was estimated by Hellewell et al.
@@
 
-``Thanksgiving---a national holiday in the US during which many students''
+``Thanksgiving---a national holiday in the US during which many students travel off campus''
 
@@
 We have revised our analysis and figures to account for changes in population sizes across the semester. See below for a detailed response.
-We also present testing volumes over time and show that they are reasonably stable across each semester.
+We also present testing volumes over time on campus and show that they are reasonably stable across each semester.
+We were not able to find information on testing at the county level and therefore were not able to explore test positivity as an additional metric.
+However, we feel that test positivity can be even more sensitive to testing behavior in our cases.
+For example, testing frequencies were doubled for undergraduate students during the initial Omicron break, but this change did not cause a doubling of case numbers, meaning that the ascertainment rate was already high.
+It seems more likely that the increased numbers of cases during this period is reflective of real outbreak dynamics, rather than testing behavior---on the other hand, positivity would suddenly decrease during this period due to changes in testing behavior.
+Finally, strong cross correlations in case patterns across other counties in New Jersey and other large cities (New York City and Philadelphia) demonstrate the robustness of our observation.
 

### 2023-02-22 doc/university-covid.tex
am ready to resubmit this

@@
 Researchers from various institutions in the US---including Cornell \citep{frazier2022modeling}, Emory \citep{lopman2020model}, Georgia Institute of Technology \citep{gibson2021surveillance}, and UC Berkeley \citep{brook2021optimizing}---modeled the feasibility of controlling the epidemic on their campuses and considered mass asymptomatic testing as their main intervention.
+Mathematical models also played crucial roles in helping to make decisions about reopening, ensuring that infrastructure (e.g., isolation methods and spaces, supporting isolation and quarantine with food, etc.) and levels of mitigation (masks, ventilation, testing frequency, etc.) were adequate and appropriate.
 These modeling efforts helped identify key parameters for control, such as the testing turnaround time, and provided support for implementing similar measures at other institutions.
@@
 Asymptomatic individuals submitted self-collected saliva samples, from which the presence of SARS-CoV-2 was tested using Reverse Transcription Polymerase Chain Reaction (RT-PCR).  
-Those who tested positive were required to isolate for at least 10 days after symptom onset or test date (whichever was longer) and were released when they had been at least 48 hours with improving or resolving symptoms as per New Jersey Department of Health guidance.
+Those who tested positive were required to isolate for at least 10 days after symptom onset or test date (whichever was longer) and were released when they had been at least 48 hours with improving or resolving symptoms as per New Jersey Department of Health guidance;
+the isolation duration switched to 5 days on January 14, 2022, for all vaccinated people.
 PCR positives were exempt from asymptomatic testing for 90 days.
@@
 This in turn likely reduced the accuracy of surveillance; therefore, we chose to focus on epidemic patterns before this change was implemented.
-In Supplementary Figure S1, we show the testing volumes over time.
-Throughout the study period, contact tracing was also performed for positive cases to alert their close contacts to either quarantine or test more frequently according to the close contacts' vaccination status, and to gather data that could help uncover clusters of transmission or superspreader events.
+In Supplementary Figure S1, we show the testing volumes over time---the testing volumes remained roughly constant within each semester, except when testing frequency changed, which we note in the paper.
+However, changes in testing frequency were not associated with sudden changes in cases, meaning that the patterns in case trajectories likely reflect patterns of spread, rather than testing behavior.
+Throughout the study period, contact tracing was also performed for positive cases to alert their close contacts to test more frequently and quarantine as applicable, according to the close contacts’ vaccination status, and to gather data that could help uncover clusters of transmission or superspreader events.
 Changes in testing frequency and other intervention measures throughout the study period reflected various factors, including the impact of COVID-19 cases on continuity of operations or continuity of teaching; on severity of disease on campus; the capacity of testing and the healthcare system; and hospitalization rates on campus and in the area.
@@
 The number of cases remained relatively low throughout the semester with a peak occurring in early December, coinciding with the epidemic trajectory in Mercer County (\fref{princeton}A).  
-A sudden decrease in the number of cases around Thanksgiving---a national holiday in the US during which many students---partly reflects the reduced number of tests (3852 and 2972 asymptomatic tests performed on the week ending November 20th and 27th, respectively).
+A sudden decrease in the number of cases around Thanksgiving---a national holiday in the US during which many students travel off campus---partly reflects the reduced number of tests (3852 and 2972 asymptomatic tests performed on the week ending November 20th and 27th, respectively).
 The highest number of cases was reported among faculty and staff members ($=169$), followed by graduate students ($=41$), and undergraduate students ($=4$).
@@
 In the beginning of the \textbf{spring semester of the 2020--2021 academic year}, the number of cases suddenly increased before classes started (\fref{princeton}B), reflecting $\approx 3000$ undergraduate students returning to campus.
-Returning students were required to be tested and quarantine for 7 days regardless of their returning location.
+Returning students were required to be tested upon arrival and quarantine for 7 days regardless of their returning location.
 Most classes remained virtual, and the testing protocol did not change (twice a week for undergraduate and graduate students, and once a week for faculty and staff members).
@@
 
-For the \textbf{fall semester of the 2021--2022 academic year}, all students and faculty and staff members were required to be vaccinated, with very few medical and religious exemptions.
+For the \textbf{fall semester of the 2021--2022 academic year}, all students and faculty and staff members were required to be vaccinated with the COVID vaccine primary series, with very few medical and religious exemptions.
 By the beginning of the semesters, 97\% of undergraduate students, 96\% of graduate students, and 94\% of faculty and staff members were vaccinated.
-Vaccinees were required to be tested once a week, while unvaccinated individuals were required to be tested twice a week.
-In-person classes and social events fully resumed on campus, though all individuals were required to wear masks indoors with a few exceptions (e.g., when eating or drinking, or when teaching a small class).
+Vaccinees were required to be tested once a week, while unvaccinated and partially vaccinated individuals were required to be tested twice a week.
+In-person classes and social events fully resumed on campus, though all individuals were required to wear masks indoors with a few exceptions (e.g., when eating or drinking, or when teaching any size of class if social distancing can be maintained at all times).
 The number of cases remained similar to previous semesters until November when cases began to increase, primarily among undergraduate students around Thanksgiving (\fref{princeton}C).  
@@
 The number of cases remained high before classes began but decreased over time, following epidemic patterns in Mercer County (\fref{princeton}D).
-Coinciding with the decrease in the campus and local cases numbers, the gathering policy was updated on February 8, 2022 to allow food in events were no longer limited to 20 people;
-in addition, the testing frequency was reduced to once a week.
+Coinciding with the decrease in the campus and local cases numbers, the gathering policy was updated on February 8, 2022 to allow food in events and event sizes were no longer limited to 20 people;
+in addition, the testing frequency was reduced to once a week (but not for unvaccinated individuals).
 Following the policy change, a large gathering event was held on campus, which resulted in an outbreak with high case numbers persisting until Spring Break (March 5th, 2022).
@@
 Recovered individuals are assumed to be immune to reinfections throughout a semester.
-Presymptomatic, symptomatic, and asymptomatic infection stages are further divided into two subcompartments to allow for more realistic and narrower distributions than the exponential distribution \citep{brett2020transmission}.
+Presymptomatic, symptomatic, and asymptomatic infection stages are further divided into two subcompartments.
+Dividing each infection stage into subcompartments allows for the duration of infection to have narrower (and more realistic) distributions than the exponential distribution \citep{brett2020transmission}.
 Transitions between each (sub)compartments are modeled using a Bernoulli process with probabilities that match the assumed means \citep{he2010plug}:
@@
 Our assumption leads to a much lower PCR sensitivity 10 days after infection because we only model the PCR sensitivity during the infectious period.
-In reality, PCR can detect infections even after a person stops shedding infectious virus---we did not include this component in our model because it would not affect the effectiveness of isolation strategy in reducing transmission.
+In reality, PCR can also detect viral nucleic acids even after a person stops shedding infectious virus, but since these nucleic acids cannot contribute to person-to-person transmission or affect the effectiveness of the isolation strategy, we did not include this component in our model.
 
@@
 The population size and testing frequencies (with twice weekly testing modeled as testing every 3 days) are set to reflect realistic campus settings.
-Although we account for heterogeneity in the number of individuals in each population group on campus (i.e., undergraduate students, graduate students, and faculty and staff members) and their respective testing patterns (e.g., twice a week for undergraduate and graduate students and once a week for faculty and staff members during fall and spring, 2020), we assume, for simplicity, that all other parameters are equal across different groups
+Although we account for heterogeneity in the number of individuals in each population group on campus (i.e., undergraduate students, graduate students, and faculty and staff members) and their respective testing patterns (e.g., twice a week for undergraduate and graduate students and once a week for faculty and staff members during fall and spring, 2020), we assume, for simplicity, that all other parameters are equal across different groups.
 We further assume that the population mixes homogeneously.
@@
 We further assume that 4000 individuals (3000 undergraduate students and 1000 graduate students) returned to campus over 14 days (January 16, 2021--January 29, 2021).
-In the beginning of the semester, all returning students were required to quarantine in their rooms for 14 and tested upon returning by the university---in our model, this was implemented by preventing returning students from getting infected or infecting other individuals.
+In the beginning of the semester, all returning students were required to quarantine in their rooms for 14 days and tested upon returning by the university---in our model, this was implemented by preventing returning students from getting infected or infecting other individuals.
 Finally, to match the initial influx of cases, we assume that 1\% of both returning and on-campus populations are infected at the beginning of simulation (January 16, 2021).
@@
 We therefore extend the model to account for these alterations and focus on the outbreak patterns among undergraduate students.
-First, based on \citep{ferguson2021report}, we assume that two and three doses of vaccines reduce susceptibility against the early Omicron variant by 10\% and 70\%, respectively. 
+First, based on \cite{ferguson2021report}, we assume that two and three doses of vaccines reduce susceptibility against the early Omicron variant by 10\% and 70\%, respectively. 
 We also assume that the transmissibility of Omicron is reduced proportionally following the previously assumed 90--to--20 ratio for the Delta variant;
 in other words, two and three doses of vaccines reduce transmissibility by 2.2\% and 15.6\%, respectively.
-The immunity from the third dose is assumed to take 7 days to develop \citep{moreira2022safety} and wane at the same rate as before (in this case, 70\% to 39\% in 20 weeks).
-Finally, the isolation period is reduced to 5 days.
+The immunity from the third dose is assumed to take 7 days to develop in our model \citep{moreira2022safety} and wane at the same rate as before (in this case, 70\% to 39\% in 20 weeks).
+Finally, the isolation period was reduced to 5 days---the actual change was implemented on January 14, 2022 but we keep the 5 day isolation period throughout our simulations, which begins on January 7, 2022, for simplicity.
 
@@
 First, our analysis highlights the power of mass asymptomatic testing for epidemic measurement and planning---even if PCR testing may have lower sensitivity than what we assumed here \citep{hellewell2021estimating}, mass asymptomatic testing can still help track ongoing epidemic dynamics in real time.
-Combining other interventions measures, such as social distancing, mask wearing, and vaccination, can help provide a safe means of reopening university campuses---but the extent to which these interventions are implemented will necessarily depend on resource availability.
-Second, we expect immune waning and superspreading to continue to play important roles in driving campus transmission---keeping vaccine statuses up-to-date within the campus community will be critical moving forward.
+Combining other interventions measures, such as social distancing, mask wearing, and vaccination, can help provide some measures to consider for restoring operations on campuses---but the extent to which these interventions are implemented will necessarily depend on resource availability.
+Second, we expect immunity waning and superspreading to continue to play important roles in driving campus transmission---keeping vaccine statuses up-to-date within the campus community will be critical moving forward.
 Third, the safe reopening of a university campus must consider the spread of SARS-CoV-2 within the surrounding community as they can both potentially drive transmission in each other---however, the degree to which infections spread from campus to community remains uncertain.
 Finally, intervention measures placed on campuses must continue to adapt and change to reflect changes in epidemiological conditions.
+We note that the generality of our conclusions will necessarily depend on specific campus settings.
 
 The emergence of new variants---in particular, their ability to evade prior immunity and transmit better---continues to add uncertainty to the future controllability of the ongoing SARS-CoV-2 pandemic.
-Nonetheless, as population-level immunity increases (either due to infection or vaccination), we are (hopefully) transitioning to an endemic phase, during which COVID-19 is expected to become less severe \citep{lavine2021immunological}.
-Many institutions have so far sought to minimize infections on their campuses early in the pandemic, thereby implementing as many interventions as possible---but such measures can be difficult to maintain for prolonged periods both economically and societally.
-As the Omicron variant began to spread, many campuses---including the PU campus---opted in for less intense interventions, reflecting difficulties in controlling the spread and a lack of severe cases among the majority of vaccinated students.
-As we continue to transition to future phases of the pandemic, the expectations for reopening campuses (e.g., whether to minimize infections on campuses) need to be re-evaluated, accounting not only for changes in epidemic dynamics but also for our perception of the pathogen.
-The answers to these questions ultimately depend on the landscape of SARS-CoV-2 immunity and its future evolutionary dynamics \citep{saad2020immune,baker2021limits}.
+Nonetheless, as population-level immunity increases (either due to infection or vaccination), we are transitioning to an endemic phase, during which COVID-19 is expected to become less severe \citep{lavine2021immunological}.
+Understanding the the landscape of SARS-CoV-2 immunity and its impact on evolutionary dynamics will be critical to predicting future outbreak dynamics \citep{saad2020immune,baker2021limits}.
 

### 2023-02-27 doc/university-covid.tex
finalize

@@
 \begin{figure}[!th]
-\includegraphics[width=0.95\textwidth]{../figure_princeton_new/figure_princeton_new_per_1000.pdf}
+\includegraphics[width=0.95\textwidth]{figure_princeton_new_per_1000.pdf}
 \caption{
@@
 \begin{figure}[!th]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation.pdf}
+\includegraphics[width=\textwidth]{figure_princeton_simulation.pdf}
 \caption{
@@
 \begin{figure}[!ht]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_omicron_spring.pdf}
+\includegraphics[width=\textwidth]{figure_princeton_simulation_omicron_spring.pdf}
 \caption{
@@
 \begin{figure}[!thp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_omicron_outbreak.pdf}
+\includegraphics[width=\textwidth]{figure_princeton_simulation_omicron_outbreak.pdf}
 \caption{
@@
 
-\bibliography{university-covid}
+\section*{Author contribution}
 
-\pagebreak
+All authors contributed equally to the research design, data collection, formal analysis, interpretation of the results, and writing of the paper.
 
-\section*{Supplementary Figures}
+\begin{thebibliography}{}
 
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
+\bibitem[\protect\citeauthoryear{Baker, Mahmud, Wagner, Yang, Pitzer, Viboud,
+  Vecchi, Metcalf, and Grenfell}{Baker et~al.}{2019}]{baker2019epidemic}
+Baker, R.~E., A.~S. Mahmud, C.~E. Wagner, W.~Yang, V.~E. Pitzer, C.~Viboud,
+  G.~A. Vecchi, C.~J.~E. Metcalf, and B.~T. Grenfell (2019).
+\newblock Epidemic dynamics of respiratory syncytial virus in current and
+  future climates.
+\newblock {\em Nature communications\/}~{\em 10\/}(1), 1--8.
 
-\begin{figure}[!htp]
-\includegraphics[width=0.9\textwidth]{../figure_princeton_new/figure_princeton_testing.pdf}
-\caption{
-\textbf{The weekly number of PCR tests stratified by subpopulations.}
-}
-\end{figure}
+\bibitem[\protect\citeauthoryear{Baker, Park, Wagner, and Metcalf}{Baker
+  et~al.}{2021}]{baker2021limits}
+Baker, R.~E., S.~W. Park, C.~E. Wagner, and C.~J.~E. Metcalf (2021).
+\newblock {The limits of SARS-CoV-2 predictability}.
+\newblock {\em Nature Ecology \& Evolution\/}~{\em 5\/}(8), 1052--1054.
 
-\pagebreak
+\bibitem[\protect\citeauthoryear{Brett and Rohani}{Brett and
+  Rohani}{2020}]{brett2020transmission}
+Brett, T.~S. and P.~Rohani (2020).
+\newblock {Transmission dynamics reveal the impracticality of COVID-19 herd
+  immunity strategies}.
+\newblock {\em Proceedings of the National Academy of Sciences\/}~{\em
+  117\/}(41), 25897--25903.
 
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_new_per_1000_bypop.pdf}
-\caption{
-\textbf{The weekly number of cases per 1000 stratified by subpopulations.}
-The case trajectory for undergraduate students was omitted for the fall semester of 2020--2021 academic year due to a very low number of undergraduate students present on campus.
-}
-\end{figure}
+\bibitem[\protect\citeauthoryear{Brook, Northrup, Ehrenberg, Doudna, Boots,
+  Consortium, et~al.}{Brook et~al.}{2021}]{brook2021optimizing}
+Brook, C.~E., G.~R. Northrup, A.~J. Ehrenberg, J.~A. Doudna, M.~Boots,
+  I.~S.-C.-.~T. Consortium, et~al. (2021).
+\newblock {Optimizing COVID-19 control with asymptomatic surveillance testing
+  in a university environment}.
+\newblock {\em Epidemics\/}~{\em 37}, 100527.
 
-\pagebreak
+\bibitem[\protect\citeauthoryear{Cobey}{Cobey}{2020}]{cobey2020modeling}
+Cobey, S. (2020).
+\newblock Modeling infectious disease dynamics.
+\newblock {\em Science\/}~{\em 368\/}(6492), 713--714.
 
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_correlation_per_1000.pdf}
-\caption{
-\textbf{Correlations between the weekly number of cases in PU and in Mercer County stratified by subpopulations.}
-Points represent the number of reported cases.
-Solid lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
-}
-\end{figure}
+\bibitem[\protect\citeauthoryear{Currie, Moreno, Delahoy, Pray, Jovaag, Braun,
+  Cole, Shechter, Fajardo, Griggs, et~al.}{Currie
+  et~al.}{2021}]{currie2021interventions}
+Currie, D.~W., G.~K. Moreno, M.~J. Delahoy, I.~W. Pray, A.~Jovaag, K.~M. Braun,
+  D.~Cole, T.~Shechter, G.~C. Fajardo, C.~Griggs, et~al. (2021).
+\newblock {Interventions to Disrupt Coronavirus Disease Transmission at a
+  University, Wisconsin, USA, August--October 2020}.
+\newblock {\em Emerging infectious diseases\/}~{\em 27\/}(11), 2776.
 
-\pagebreak
+\bibitem[\protect\citeauthoryear{{Duke University}}{{Duke
+  University}}{2021}]{dukeoutbreak}
+{Duke University} (2021).
+\newblock {Important COVID updates for undergrads, Aug. 30}.
 
-\begin{figure}[!htp]
-\begin{center}
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_map_new.pdf}
-\end{center}
-\caption{
-\textbf{Correlations between the weekly number of cases in PU and in counties in New Jersey.}
-(A--C) Map of correlations between the weekly number of cases in PU and in counties in New Jersey.
-(D--F) Relationship between case correlations and distance from Mercer County.
-Points represent the estimated correlation coefficients.
-Solid lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
-Black points and lines represent correlations based on all cases in PU.
-Red points and lines represent correlations based on cases among faculty and staff members in PU.
-}
-\end{figure}
+\bibitem[\protect\citeauthoryear{Elderd, Dukic, and Dwyer}{Elderd
+  et~al.}{2006}]{elderd2006uncertainty}
+Elderd, B.~D., V.~M. Dukic, and G.~Dwyer (2006).
+\newblock Uncertainty in predictions of disease spread and public health
+  responses to bioterrorism and emerging diseases.
+\newblock {\em Proceedings of the National Academy of Sciences\/}~{\em
+  103\/}(42), 15693--15697.
 
-\pagebreak
+\bibitem[\protect\citeauthoryear{Endo, {Centre for the Mathematical Modelling
+  of Infectious Diseases COVID-19 Working Group}, Abbott, Kucharski, and
+  Funk}{Endo et~al.}{2020}]{10.12688/wellcomeopenres.15842.3}
+Endo, A., {Centre for the Mathematical Modelling of Infectious Diseases
+  COVID-19 Working Group}, S.~Abbott, A.~Kucharski, and S.~Funk (2020).
+\newblock {Estimating the overdispersion in COVID-19 transmission using
+  outbreak sizes outside China [version 3; peer review: 2 approved]}.
+\newblock {\em Wellcome Open Research\/}~{\em 5\/}(67).
 
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_newyork_new.pdf}
-\caption{
-\textbf{Dynamics of SARS-CoV-2 outbreaks in PU and New York City.}
-(A--C) Epidemic trajectories across three semesters: Fall 2020 (A), Spring 2020 (B), and Fall 20201 (C).
-Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in PU.
-Red lines represent the weekly number of cases in New York City.
-(D--F) Correlations between the weekly number of cases in PU and in New York City.
-Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
-}
-\end{figure}
+\bibitem[\protect\citeauthoryear{Ferguson, Ghani, Cori, Hogan, Hinsley, and
+  Volz}{Ferguson et~al.}{2021}]{ferguson2021report}
+Ferguson, N., A.~Ghani, A.~Cori, A.~Hogan, W.~Hinsley, and E.~Volz (2021).
+\newblock {Report 49: Growth, population distribution and immune escape of
+  Omicron in England}.
 
-\pagebreak
+\bibitem[\protect\citeauthoryear{Flaxman, Mishra, Gandy, Unwin, Mellan,
+  Coupland, Whittaker, Zhu, Berah, Eaton, et~al.}{Flaxman
+  et~al.}{2020}]{flaxman2020estimating}
+Flaxman, S., S.~Mishra, A.~Gandy, H.~J.~T. Unwin, T.~A. Mellan, H.~Coupland,
+  C.~Whittaker, H.~Zhu, T.~Berah, J.~W. Eaton, et~al. (2020).
+\newblock {Estimating the effects of non-pharmaceutical interventions on
+  COVID-19 in Europe}.
+\newblock {\em Nature\/}~{\em 584\/}(7820), 257--261.
 
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_philly_new.pdf}
-\caption{
-\textbf{Dynamics of SARS-CoV-2 outbreaks in PU and Philadelphia.}
-(A--C) Epidemic trajectories across three semesters: Fall 2020 (A), Spring 2020 (B), and Fall 20201 (C).
-Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in PU.
-Red lines represent the weekly number of cases in Philadelphia.
-(D--F) Correlations between the weekly number of cases in PU and in Philadelphia.
-Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
-}
-\end{figure}
+\bibitem[\protect\citeauthoryear{Fox, Bailey, Seamon, and Miranda}{Fox
+  et~al.}{2021}]{fox2021response}
+Fox, M.~D., D.~C. Bailey, M.~D. Seamon, and M.~L. Miranda (2021).
+\newblock {Response to a COVID-19 outbreak on a university campus—Indiana,
+  August 2020}.
+\newblock {\em Morbidity and Mortality Weekly Report\/}~{\em 70\/}(4), 118.
 
-\pagebreak
+\bibitem[\protect\citeauthoryear{Frazier, Cashore, Duan, Henderson, Janmohamed,
+  Liu, Shmoys, Wan, and Zhang}{Frazier et~al.}{2022}]{frazier2022modeling}
+Frazier, P.~I., J.~M. Cashore, N.~Duan, S.~G. Henderson, A.~Janmohamed, B.~Liu,
+  D.~B. Shmoys, J.~Wan, and Y.~Zhang (2022).
+\newblock {Modeling for COVID-19 college reopening decisions: Cornell, a case
+  study}.
+\newblock {\em Proceedings of the National Academy of Sciences\/}~{\em
+  119\/}(2).
 
-\begin{figure}[!htp]
-\includegraphics[width=0.8\textwidth]{../figure_princeton_new/figure_princeton_map_extended.pdf}
-\caption{
-\textbf{Correlations between the weekly number of cases in PU and in counties in New Jersey, Pennsylvania, and New York State.}
-(A--C) Map of correlations between the weekly number of cases in PU and in counties in New Jersey, Pennsylvania, and New York State.
-(D--F) Relationship between case correlations and distance from Mercer County.
-Points represent the estimated correlation coefficients.
-Solid lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
-Black points and lines represent correlations based on all cases in PU.
-Red points and lines represent correlations based on cases among faculty and staff members in PU.
-New York City is excluded from this analysis as the data provided by New York Times are not further stratified by county levels.
-}
-\end{figure}
+\bibitem[\protect\citeauthoryear{Galanti, Pei, Yamana, Angulo, Charos,
+  Swerdlow, and Shaman}{Galanti et~al.}{2021}]{galanti2021social}
+Galanti, M., S.~Pei, T.~K. Yamana, F.~J. Angulo, A.~Charos, D.~L. Swerdlow, and
+  J.~Shaman (2021).
+\newblock Social distancing remains key during vaccinations.
+\newblock {\em Science\/}~{\em 371\/}(6528), 473--474.
 
-\pagebreak
+\bibitem[\protect\citeauthoryear{Gibson, Weitz, Shannon, Holton, Bryksin, Liu,
+  Sieglinger, Coenen, Zhao, Beckett, et~al.}{Gibson
+  et~al.}{2021}]{gibson2021surveillance}
+Gibson, G., J.~S. Weitz, M.~P. Shannon, B.~Holton, A.~Bryksin, B.~Liu,
+  M.~Sieglinger, A.~R. Coenen, C.~Zhao, S.~J. Beckett, et~al. (2021).
+\newblock {Surveillance-to-Diagnostic Testing Program for Asymptomatic
+  SARS-CoV-2 Infections on a Large, Urban Campus in Fall 2020}.
+\newblock {\em Epidemiology\/}~{\em 33\/}(2), 209--216.
 
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_diagram/diagram.pdf}
-\caption{
-\textbf{Compartmental diagram of the individual-based model.}
-Each compartment represents a stage of infection: susceptible $S$, exposed $S$, pre-symptomatic $I_p$, symptomatic $I_s$, asymptomatic $I_a$, and recovered $R$.
-Pre-symptomatic, symptomatic, and asymptomatic stages are further divided into two subcompartments.
-Individuals in pre-symptomatic, symptomatic, and asymptomatic stages can test positive with 95\% sensitivity.
-}
-\end{figure}
+\bibitem[\protect\citeauthoryear{Grenfell and Harwood}{Grenfell and
+  Harwood}{1997}]{grenfell1997meta}
+Grenfell, B. and J.~Harwood (1997).
+\newblock (meta) population dynamics of infectious diseases.
+\newblock {\em Trends in ecology \& evolution\/}~{\em 12\/}(10), 395--399.
 
-\pagebreak
+\bibitem[\protect\citeauthoryear{Grenfell, Bj{\o}rnstad, and Kappey}{Grenfell
+  et~al.}{2001}]{grenfell2001travelling}
+Grenfell, B.~T., O.~N. Bj{\o}rnstad, and J.~Kappey (2001).
+\newblock Travelling waves and spatial hierarchies in measles epidemics.
+\newblock {\em Nature\/}~{\em 414\/}(6865), 716--723.
 
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_fall_2020_all.pdf}
-\caption{
-\textbf{Comparisons between model predictions and the observed numbers of cases for fall 2020.}
-Points represent the weekly number of reported cases in PU.
-Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
-See figure 2 in the main text for details.
-}
-\end{figure}
+\bibitem[\protect\citeauthoryear{Hamer, White, Jenkins, Gill, Landsberg,
+  Klapperich, Bulekova, Platt, Decarie, Gilmore, et~al.}{Hamer
+  et~al.}{2021}]{hamer2021assessment}
+Hamer, D.~H., L.~F. White, H.~E. Jenkins, C.~J. Gill, H.~E. Landsberg,
+  C.~Klapperich, K.~Bulekova, J.~Platt, L.~Decarie, W.~Gilmore, et~al. (2021).
+\newblock {Assessment of a COVID-19 control plan on an urban university campus
+  during a second wave of the pandemic}.
+\newblock {\em JAMA Network Open\/}~{\em 4\/}(6), e2116425--e2116425.
 
-\pagebreak
+\bibitem[\protect\citeauthoryear{{Harvard University}}{{Harvard
+  University}}{2021}]{harvardoutbreak}
+{Harvard University} (2021).
+\newblock {Increase in COVID-19 cases, Take Steps to Protect Yourself \& Our
+  Community}.
 
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_spring_2020_all.pdf}
-\caption{
-\textbf{Comparisons between model predictions and the observed numbers of cases for spring 2020.}
-Points represent the weekly number of reported cases in PU.
-Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
-See figure 2 in the main text for details.
-}
-\end{figure}
+\bibitem[\protect\citeauthoryear{He, Ionides, and King}{He
+  et~al.}{2010}]{he2010plug}
+He, D., E.~L. Ionides, and A.~A. King (2010).
+\newblock Plug-and-play inference for disease dynamics: measles in large and
+  small populations as a case study.
+\newblock {\em Journal of the Royal Society Interface\/}~{\em 7\/}(43),
+  271--283.
 
-\pagebreak
+\bibitem[\protect\citeauthoryear{Hellewell, Abbott, Gimma, Bosse, Jarvis,
+  Russell, Munday, Kucharski, Edmunds, Sun, et~al.}{Hellewell
+  et~al.}{2020}]{hellewell2020feasibility}
+Hellewell, J., S.~Abbott, A.~Gimma, N.~I. Bosse, C.~I. Jarvis, T.~W. Russell,
+  J.~D. Munday, A.~J. Kucharski, W.~J. Edmunds, F.~Sun, et~al. (2020).
+\newblock {Feasibility of controlling COVID-19 outbreaks by isolation of cases
+  and contacts}.
+\newblock {\em The Lancet Global Health\/}~{\em 8\/}(4), e488--e496.
 
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_fall_2021_all.pdf}
-\caption{
-\textbf{Comparisons between model predictions and the observed numbers of cases for fall 2021.}
-Points represent the weekly number of reported cases in PU.
-Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
-See figure 2 in the main text for details.
-}
-\end{figure}
+\bibitem[\protect\citeauthoryear{Hellewell, Russell, Beale, Kelly, Houlihan,
+  Nastouli, and Kucharski}{Hellewell et~al.}{2021}]{hellewell2021estimating}
+Hellewell, J., T.~W. Russell, R.~Beale, G.~Kelly, C.~Houlihan, E.~Nastouli, and
+  A.~J. Kucharski (2021).
+\newblock {Estimating the effectiveness of routine asymptomatic PCR testing at
+  different frequencies for the detection of SARS-CoV-2 infections}.
+\newblock {\em BMC medicine\/}~{\em 19\/}(1), 1--10.
 
-\pagebreak
+\bibitem[\protect\citeauthoryear{Holmdahl and Buckee}{Holmdahl and
+  Buckee}{2020}]{holmdahl2020wrong}
+Holmdahl, I. and C.~Buckee (2020).
+\newblock {Wrong but useful—what COVID-19 epidemiologic models can and cannot
+  tell us}.
+\newblock {\em New England Journal of Medicine\/}~{\em 383\/}(4), 303--305.
 
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_nowaning.pdf}
-\caption{
-\textbf{Comparisons between model predictions and the observed numbers of cases for fall 2021 without immunity waning.}
-Points represent the weekly number of reported cases in PU.
-Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
-See figure 2 in the main text for details.
-}
-\end{figure}
+\bibitem[\protect\citeauthoryear{Kissler, Tedijanto, Goldstein, Grad, and
+  Lipsitch}{Kissler et~al.}{2020}]{kissler2020projecting}
+Kissler, S.~M., C.~Tedijanto, E.~Goldstein, Y.~H. Grad, and M.~Lipsitch (2020).
+\newblock {Projecting the transmission dynamics of SARS-CoV-2 through the
+  postpandemic period}.
+\newblock {\em Science\/}~{\em 368\/}(6493), 860--868.
 
-\pagebreak
+\bibitem[\protect\citeauthoryear{Koelle, Martin, Antia, Lopman, and
+  Dean}{Koelle et~al.}{2022}]{koelle2022changing}
+Koelle, K., M.~A. Martin, R.~Antia, B.~Lopman, and N.~E. Dean (2022).
+\newblock {The changing epidemiology of SARS-CoV-2}.
+\newblock {\em Science\/}~{\em 375\/}(6585), 1116--1121.
 
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_omicron_spring_shortI.pdf}
-\caption{
-\textbf{The impact of changes in testing frequency and an increased reproduction number on the spread of the Omicron variant, assuming a shorter latent and infectious periods.}
-Solid lines represent median predictions.
-Shaded areas represent 90\% quantiles across 100 simulations.
-Points represent the observed data.
-Vertical dashed lines represent the week including February 8, 2022, when distancing and testing policies were updated on PU campus.
-For each row, we assume a fixed value of baseline contact reproduction number $\Rc$ ranging from 2 to 6 across rows.
-Then, we simulate increase in $\Rc$ at the time of policy change (indicated by column labels).
-Here, the assumed mean latent, presymptomatic, and asymptomatic stages of infection are 0.5 days shorter than the assumed values for simulations presented in Figure 3 in the main text.
-}
-\end{figure}
+\bibitem[\protect\citeauthoryear{Kraemer, Pybus, Fraser, Cauchemez, Rambaut,
+  and Cowling}{Kraemer et~al.}{2021}]{kraemer2021monitoring}
+Kraemer, M.~U., O.~G. Pybus, C.~Fraser, S.~Cauchemez, A.~Rambaut, and B.~J.
+  Cowling (2021).
+\newblock {Monitoring key epidemiological parameters of SARS-CoV-2
+  transmission}.
+\newblock {\em Nature medicine\/}~{\em 27\/}(11), 1854--1855.
 
-\pagebreak
+\bibitem[\protect\citeauthoryear{Lavezzo, Franchin, Ciavarella,
+  Cuomo-Dannenburg, Barzon, Del~Vecchio, Rossi, Manganelli, Loregian, Navarin,
+  et~al.}{Lavezzo et~al.}{2020}]{lavezzo2020suppression}
+Lavezzo, E., E.~Franchin, C.~Ciavarella, G.~Cuomo-Dannenburg, L.~Barzon,
+  C.~Del~Vecchio, L.~Rossi, R.~Manganelli, A.~Loregian, N.~Navarin, et~al.
+  (2020).
+\newblock {Suppression of a SARS-CoV-2 outbreak in the Italian municipality of
+  Vo’}.
+\newblock {\em Nature\/}~{\em 584\/}(7821), 425--429.
 
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_omicron_spring_shortI.pdf}
-\caption{
-\textbf{The impact of changes in testing frequency and an increased reproduction number on the spread of the Omicron variant, assuming a lower vaccine effectiveness.}
-Solid lines represent median predictions.
-Shaded areas represent 90\% quantiles across 100 simulations.
-Points represent the observed data.
-Vertical dashed lines represent the week including February 8, 2022, when distancing and testing policies were updated on PU campus.
-For each row, we assume a fixed value of baseline contact reproduction number $\Rc$ ranging from 2 to 6 across rows.
-Then, we simulate increase in $\Rc$ at the time of policy change (indicated by column labels).
-Here, we assume 30\% vaccine effectiveness against infection.
-}
-\end{figure}
+\bibitem[\protect\citeauthoryear{Lavine, Bjornstad, and Antia}{Lavine
+  et~al.}{2021}]{lavine2021immunological}
+Lavine, J.~S., O.~N. Bjornstad, and R.~Antia (2021).
+\newblock {Immunological characteristics govern the transition of COVID-19 to
+  endemicity}.
+\newblock {\em Science\/}~{\em 371\/}(6530), 741--745.
+
+\bibitem[\protect\citeauthoryear{Lloyd-Smith, Schreiber, Kopp, and
+  Getz}{Lloyd-Smith et~al.}{2005}]{lloyd2005superspreading}
+Lloyd-Smith, J.~O., S.~J. Schreiber, P.~E. Kopp, and W.~M. Getz (2005).
+\newblock Superspreading and the effect of individual variation on disease
+  emergence.
+\newblock {\em Nature\/}~{\em 438\/}(7066), 355--359.
+
+\bibitem[\protect\citeauthoryear{Lopman, Liu, Le~Guillou, Lash, Isakov, and
+  Jenness}{Lopman et~al.}{2020}]{lopman2020model}
+Lopman, B., C.~Y. Liu, A.~Le~Guillou, T.~L. Lash, A.~P. Isakov, and S.~M.
+  Jenness (2020).
+\newblock {A model of COVID-19 transmission and control on university
+  campuses}.
+\newblock {\em MedRxiv\/}.
+
+\bibitem[\protect\citeauthoryear{Metcalf, Morris, and Park}{Metcalf
+  et~al.}{2020}]{metcalf2020mathematical}
+Metcalf, C. J.~E., D.~H. Morris, and S.~W. Park (2020).
+\newblock Mathematical models to guide pandemic response.
+\newblock {\em Science\/}~{\em 369\/}(6502), 368--369.
+
+\bibitem[\protect\citeauthoryear{Moreira~Jr, Kitchin, Xu, Dychter, Lockhart,
+  Gurtman, Perez, Zerbini, Dever, Jennings, et~al.}{Moreira~Jr
+  et~al.}{2022}]{moreira2022safety}
+Moreira~Jr, E.~D., N.~Kitchin, X.~Xu, S.~S. Dychter, S.~Lockhart, A.~Gurtman,
+  J.~L. Perez, C.~Zerbini, M.~E. Dever, T.~W. Jennings, et~al. (2022).
+\newblock Safety and efficacy of a third dose of {BNT162b2} {COVID-19} vaccine.
+\newblock {\em New England Journal of Medicine\/}.
+
+\bibitem[\protect\citeauthoryear{Prunas, Warren, Crawford, Gazit, Patalon,
+  Weinberger, and Pitzer}{Prunas et~al.}{2022}]{prunas2022vaccination}
+Prunas, O., J.~L. Warren, F.~W. Crawford, S.~Gazit, T.~Patalon, D.~M.
+  Weinberger, and V.~E. Pitzer (2022).
+\newblock {Vaccination with BNT162b2 reduces transmission of SARS-CoV-2 to
+  household contacts in Israel}.
+\newblock {\em Science\/}, eabl4292.
+
+\bibitem[\protect\citeauthoryear{Saad-Roy, Morris, Metcalf, Mina, Baker,
+  Farrar, Holmes, Pybus, Graham, Levin, et~al.}{Saad-Roy
+  et~al.}{2021}]{saad2021epidemiological}
+Saad-Roy, C.~M., S.~E. Morris, C.~J.~E. Metcalf, M.~J. Mina, R.~E. Baker,
+  J.~Farrar, E.~C. Holmes, O.~G. Pybus, A.~L. Graham, S.~A. Levin, et~al.
+  (2021).
+\newblock {Epidemiological and evolutionary considerations of SARS-CoV-2
+  vaccine dosing regimes}.
+\newblock {\em Science\/}~{\em 372\/}(6540), 363--370.
+
+\bibitem[\protect\citeauthoryear{Saad-Roy, Wagner, Baker, Morris, Farrar,
+  Graham, Levin, Mina, Metcalf, and Grenfell}{Saad-Roy
+  et~al.}{2020}]{saad2020immune}
+Saad-Roy, C.~M., C.~E. Wagner, R.~E. Baker, S.~E. Morris, J.~Farrar, A.~L.
+  Graham, S.~A. Levin, M.~J. Mina, C.~J.~E. Metcalf, and B.~T. Grenfell (2020).
+\newblock {Immune life history, vaccination, and the dynamics of SARS-CoV-2
+  over the next 5 years}.
+\newblock {\em Science\/}~{\em 370\/}(6518), 811--818.
+
+\bibitem[\protect\citeauthoryear{Tan, Chiew, Pang, Lee, Ong, Lye, and Tan}{Tan
+  et~al.}{2023}]{tan2023vaccine}
+Tan, C.~Y., C.~J. Chiew, D.~Pang, V.~J. Lee, B.~Ong, D.~C. Lye, and K.~B. Tan
+  (2023).
+\newblock {Vaccine effectiveness against Delta, Omicron BA. 1, and BA. 2 in a
+  highly vaccinated Asian setting: a test-negative design study}.
+\newblock {\em Clinical Microbiology and Infection\/}~{\em 29\/}(1), 101--106.
+
+\bibitem[\protect\citeauthoryear{Tartof, Slezak, Fischer, Hong, Ackerson,
+  Ranasinghe, Frankland, Ogun, Zamparo, Gray, et~al.}{Tartof
+  et~al.}{2021}]{tartof2021effectiveness}
+Tartof, S.~Y., J.~M. Slezak, H.~Fischer, V.~Hong, B.~K. Ackerson, O.~N.
+  Ranasinghe, T.~B. Frankland, O.~A. Ogun, J.~M. Zamparo, S.~Gray, et~al.
+  (2021).
+\newblock {Effectiveness of mRNA BNT162b2 COVID-19 vaccine up to 6 months in a
+  large integrated health system in the USA: a retrospective cohort study}.
+\newblock {\em The Lancet\/}~{\em 398\/}(10309), 1407--1416.
+
+\bibitem[\protect\citeauthoryear{Viboud, Bj{\o}rnstad, Smith, Simonsen, Miller,
+  and Grenfell}{Viboud et~al.}{2006}]{viboud2006synchrony}
+Viboud, C., O.~N. Bj{\o}rnstad, D.~L. Smith, L.~Simonsen, M.~A. Miller, and
+  B.~T. Grenfell (2006).
+\newblock Synchrony, waves, and spatial hierarchies in the spread of influenza.
+\newblock {\em science\/}~{\em 312\/}(5772), 447--451.
+
+\bibitem[\protect\citeauthoryear{Wallinga and Lipsitch}{Wallinga and
+  Lipsitch}{2007}]{wallinga2007generation}
+Wallinga, J. and M.~Lipsitch (2007).
+\newblock How generation intervals shape the relationship between growth rates
+  and reproductive numbers.
+\newblock {\em Proceedings of the Royal Society B: Biological Sciences\/}~{\em
+  274\/}(1609), 599--604.
+
+\bibitem[\protect\citeauthoryear{Wilson, Donovan, Campbell, Chai, Pittman,
+  Se{\~n}a, Pettifor, Weber, Mallick, Cope, et~al.}{Wilson
+  et~al.}{2020}]{wilson2020multiple}
+Wilson, E., C.~V. Donovan, M.~Campbell, T.~Chai, K.~Pittman, A.~C. Se{\~n}a,
+  A.~Pettifor, D.~J. Weber, A.~Mallick, A.~Cope, et~al. (2020).
+\newblock {Multiple COVID-19 clusters on a university campus—North Carolina,
+  August 2020}.
+\newblock {\em Morbidity and Mortality Weekly Report\/}~{\em 69\/}(39), 1416.
+
+\end{thebibliography}
 

### 2023-04-21 doc/response.tex
campus done

--- 2023-01-22 doc/response.tex
@@
 \usepackage{color}
+\usepackage{graphicx}
 
@@
 Thank you for the chance to revise and resubmit our manuscript.
-We have made major revisions to our manuscript to address reviewers' concerns about the methods and assumptions of our work.
-We now compare normalized cases divided by the population size---this allows us to directly compare case patterns on campus and those around the surrounding community.
-We have reparameterized the community contact rate in our model to account for variation in population sizes across semesters and refitted the model.
-We have also added sensitivity analyses and additional figures to show the robustness of our conclusion.
+We have tried to highlight the limitation of our correlation analysis and avoid linking its results to any causal statements.
+We have also tried to clarify that the causal conclusions are supported by the modeling analysis, rather than the correlation analysis.
 Below please find our detailed responses to reviewers.
-
-\rev{Editor}
-
-\revtext{Associate Editor Comments: }
-
-\revtext{This paper has turned out to be extremely difficult to find sufficient expert reviewers who were willing to review.
-We have now received two reviews.
-The first reviewer \#1 is more supportive but the second reviewer \#2 has major concerns that need to be addressed.
-Please revise the paper to address the concerns and comments of the two reviewers.
-The revised paper will require further review.}
 
@@
 
-\revtext{The authors present a retrospective analysis of the COVID-19 epidemic at Princeton University and provide an extensive analysis that generally supports the hypothesis that a superspreading event led to a large uptick in case numbers after the introduction of the Omicron variant.
-Overall I like the paper.
-It's well written, and provide a detailed description of the application of a detailed computational model for policy support at an institutional level.
-It is therefore potentially a valuable contribution and I would support publication after some minor revisions.}
-
-Thank you.
-
-\revtext{Chief among these is the sensitivity of model results (i.e., calibration, and ultimate policy commentary) on the assumptions regarding test sensitivity.
-The authors claim "Our analysis highlights the power of mass asymptomatic testing" but the assumption is that a PCR test is 95\% sensitive after latency, and that case detection leads to full case isolation.
-These are both extremely strong assumptions, that are not conservative with respect to the claim.
-I recommend examining the results of Hellewell et al ("Estimating the effectiveness of routine asymptomatic PCR testing...") for strong evidence that sensitivity is much lower than claimed here.
-Also see Zachreson et al ("COVID-19 in low-tolerance border quarantine systems...") for a detailed individual-based implementation of dynamic test sensitivity.}
-
-We agree that 95\% sensitivity seem too strong.
-Hellewell et al estimated that the probability of detecting an infection from a PCR peaks at 77\% (54–88\%) 4 days after infection, and decreases to 50\% (38–65\%) by 10 days after infection.
-But this estimate implicitly accounts for latency, whereas we assume 95\% sensitivity after sensitivity.
-Therefore, while our sensitivity peaks at 92\% 5 days after infection, it decreases to 20\% by 10 days after infection (due to recovery from infections), which is considerably lower than what was estimated by Hellewell et al.
-Therefore, while the individual-level sensitivity is not dynamic (a step function changing from 0 to 0.95 and back down to 0), the population-level sensitivity is dynamic.
-We have added the following paragraph to our main text:
-
-``The 95\% sensitivity assumption may seem too high. 
-For example, \cite{hellewell2021estimating} estimated that the probability of detecting an infection from a PCR peaks at 77\% (54–88\%) 4 days after infection, and decreases to 50\% (38–65\%) by 10 days after infection.
-These estimates are considerably lower than our assumption because their estimates implicitly account for the latent period.
-At the individual level, we assume that an infected individual has no detectable infection (0\% sensitivity) during their latent period and 95\% sensitivity during their various stages of infectious periods.
-At the population level, this assumption translates to a peak sensitivity of 92\% by 4 days after infection, decreasing to 20\% sensitivity by 10 days after infection.
-Our assumption leads to a much lower PCR sensitivity 10 days after infection because we only model the PCR sensitivity during the infectious period.
-In reality, PCR can detect infections even after a person stops shedding infectious virus---we did not include this component in our model because it would not affect the effectiveness of isolation strategy in reducing transmission.''
-
-We further note that the full case isolation was what was happening at Princeton University during the investigation of the study period.
-However, we agree that this might be feasible in some institutions. We have added the following phrase:
-
-``while this assumption reflects the isolation policy in Princeton University during the investigation period, it may be inapplicable in studying institutional outbreaks in general.''
-
-\revtext{There are a number of other issues that are less important to my overall recommendation that I list below by section:}
-
-\revtext{Results:}
-
-\revtext{1) Figure 2 (B, E, H) - I would personally prefer to see a representative trajectory (i.e., one that is central in the ensemble based on summary statistics) than a median, since it's being compared to a single real-world trajectory, the 90\% quantiles can stay though.
-This would give a better illustration of the model dynamics (w.r.t.
-temporal fluctuations).
-This figure could also benefit from a clearer illustration of which parameters from the grids in (A, D, G) correspond to the ensembles in B, E, H.}
-
-We now show a simulation with the best goodness of fit (least sum of squared errors) alongside a media. 
-We agree that showing an actual simulation is important for capturing the variability in the case trajectory.
-We also feel that the median is useful for understanding the general trend and decided to keep it in place.
-We also added a circle to panels A, D, and G to show the best fitting parameter.
-
-\revtext{2) There is a minor contradiction between Fig 2.
-A and the statement: "We find that a low level of contacts R contact = 0.5 and a small amount of community transmission $\theta = 0.015$ is most consistent with the observed epidemic dynamics in fall 2020 (Fig.
-2A)." in Fig 2A, the minimum sits at Theta = 0.02, and is not definitive over the range > 0.01.}
-
-This was a typo on our side. We have now fixed this.
-
-\revtext{3) it looks like lower values of R could have been explored - what is the reason for the lower bound of 0.5?}
-
-We did this for computationally efficiency. We now explore R as low as 0.25.
-We could explore even lower values of R but we feel that lower values may be unrealistic.
-
-\revtext{4) Please explain "Thanksgiving" to non-Americans. This should be indicated visually in Fig. 2B}
-
-We now explain Thanksgiving the first time it is mentioned:
-
-``Thanksgiving---a national holiday in the US during which many students''
-
-\revtext{5) "all returning individuals are assumed to be quarantined for 14 days and tested upon returning." please elaborate (how is quarantine implemented?)}
-
-We have revised this sentence:
-
-``In the beginning of the semester, all returning students were required to quarantine in their rooms for 14 and tested upon returning by the university---in our model, this was implemented by preventing returning students from getting infected or infecting other individuals.''
-
-\revtext{6) Regarding Theta: The model should be refactored so that theta is per-capita on-campus population, the decision not to do so leads to re-scaling of the parameter to campus population fractions in the final section of the results and is confusing.
-This would probably also help explain why the values of theta are so much higher for the Delta period after vaccination (more people on campus).
-Finally, the parameter is variously referred to as 'contact rate' and 'transmission rate', this should be clarified in the methods description (they seem to mean the same thing for unvaccinated populations, but the definition gets fuzzier once the transmission event itself is subject to a Bernoulli trial given "infectious contact").}
-
-Thank you for pointing this out. 
-We have reparameterized the parameter theta and reran our simulations.
-We have also added the following description .
-
-``In particular, we assume that infectious contacts from local or regional community can be made at random to anyone on campus.
-These contacts are modeled using a Poisson distribution with a time-varying mean, which is calculated by multiplying the daily number of cases by the community contact rate $\theta$ and the population size on campus $N$.
-More precisely, $\theta$ is the probability that an infected individual from the community makes an infectious contact with an individual on campus per capita campus population. 
-By further multiplying this probability with the population size $N$, we are essentially assuming a density dependent contact, where a higher population size on campus leads to more infections from the community.''
-
-\revtext{7) The Omicron wave part has a lot of ad-hoc assumptions about policies (i.e., 70 boosters per day), it would be good to know which of these are guessed, and which are supported through consultation etc.}
-
-We have revised this section to make our assumptions clearer:
-
-``Based on known vaccination statuses, we assume that 99\% of students are vaccinated with 60\% of them being boosted as of January 1, 2022.
-Since all students were required to receive booster shots before returning to campus, we assumed that 70 booster shots were given on each day---this assumption allows all students to be boosted in 28 days.
-To match the high numbers of cases on the week ending January 7, 2022, we assume 14\% of the students present on campus are infected as of January 7, 2022 (roughly 100/700).
-To account for students who were infected with the Omicron variant during the fall semester, we assume that 100 students are already immune to Omicron infection at the beginning of the spring semester---this roughly corresponds to the number of PU cases that were reported in December.''
-
-\revtext{8) Figure 3: do the 9 plots on the right have different reproductive ratios?
-This should be clearly stated (Rcontact is given in the row label, but the column label gives a reproductive ratio multiplier, so it's unclear how assumed transmissibility varies between panels).}
-
-Thank you for pointing this out. The baseline Rcontact does not change within each row; only the reproductive ratio multiplier changes. 
-We have added the following sentences to clarify this:
-
-``
-For each row, we assume a fixed value of baseline contact reproduction number $\Rc$ ranging from 2 to 6 across rows.
-Then, we simulate increase in $\Rc$ at the time of policy change (indicated by column labels).
-''
-
-\revtext{9) Please be more clear about the policy change (dashed line in Fig 3).
-Was this strictly a change to gathering restrictions?
-Are you implying causation with the super-spreading event?
-I think the messaging around the relationship between the observed dynamics and the enacted policies needs to be stated more clearly and explicitly.}
-
-There was a change to gathering restriction and testing frequency. This is explained in the Descriptive analysis section:
-
-``Coinciding with the decrease in the campus and local cases numbers, the gathering policy was updated on February 8, 2022 to allow food in events were no longer limited to 20 people;
-in addition, the testing frequency was reduced to once a week.''
-
-We have also rephrased the following sentence to make the causality clearer:
-
-``Following the policy change, a large gathering event was held on campus, which resulted in an outbreak with high case numbers persisting until Spring Break (March 5th, 2022).''
-
-\revtext{Discussion:}
-
-\revtext{10) Please explain the phrase "our homogeneous mixing assumption is conservatively pessimistic." One could argue that assuming homogeneity is not 'conservatively pessimistic', rather, it assumes the highest rate of case growth possible for a given natural transmissibility and secondary case dispersion (because transmission is unconstrained by group size beyond the size of the total population), this means the calibrated natural transmission rates (i.e., Rcontact) will be biased down, for the same case incidence.
-Lower estimates of natural transmission rates (vis.
-lower viral load of an index case) could lead to liberal conclusions regarding the required stringency of interventions, so it needs to be made clear in what policy decision context the statement 'conservatively pessimistic' applies.}
-
-We meant conservatively pessimistic from a dynamical perspective, allowing for rapid growth.
-We agree that this terminology is confusing in a policy decision context. 
-We have tried to clarify the consequences our assumption and decided to remove the term ``pessimistic'' to avoid confusion.
-
-``For example, we assume conservatively that the entire university populations mix homogeneously and have identical campus and community contact rates (captured by $\Rc$ and $\theta$, respectively).
-This assumption can lead to the fastest epidemic growth rates because transmission is not limited by the size of the contact network---in other words, our estimates of the reproduction will be necessarily low, making the epidemic easier to control.
-In reality, increases in cases were often associated with specific transmission clusters, suggesting heterogeneity in transmission patterns.
-Contact levels also likely differ between different groups:
-for example, faculty and staff members are more likely to interact with community members than undergraduate students and would be at a higher risk for community infections \citep{frazier2022modeling}.''
-
-\revtext{11) The paper claims not to model behaviour.
-While this is strictly true, because the model does not incorporate behaviour explicitly, it is a bit misleading because the model does implicitly model behaviour by attaching case importation rates to the community case numbers (which are a function of behaviour).
-A better way of explaining the model would be (in my opinion) that it implicitly accounts for changes in community behaviour but does not explicitly simulate behavioural change in the campus environment.}
-
-Thank you for pointing this out. We have reworded this sentence as the following:
-
-``We also do not account for explicit changes in behavior on campus and assume constant $\Rc$ throughout each semester.
-Instead, we implicitly account for behavioral changes in the community by modeling community transmission to campus as a function of community case numbers.
-While we cannot rule out the possibility that behavioral changes on campus could have contributed to various epidemics (e.g., the Omicron wave beginning in the fall semester of the 2021--2022 academic year), we were able to capture the majority of epidemic patterns without modeling them---when the majority of transmission is caused by imported cases from the community, we expect behavioral changes on campus to have relatively weaker effects on overall transmission dynamics.''
-
-\revtext{12) "Our analysis highlights the power of mass asymptomatic testing" (see above main comment)}
-
-We have rephrased this sentence:
-
-``First, our analysis highlights the power of mass asymptomatic testing for epidemic measurement and planning---even if PCR testing may have lower sensitivity than what we assumed here \citep{hellewell2021estimating}, mass asymptomatic testing can still help track ongoing epidemic dynamics in real time''
-
-\revtext{13) "preventing large gatherings can help prevent large superspreading events in the midst of a rising epidemic" The model does not simulate gatherings, so this claim should be connected to the results explicitly or it should be removed.}
-
-This was a comment from our Princeton experience rather than simulation. We agree that our model does not simulate gathering and so we have removed this comment.
-
-\revtext{14) "intervention measures placed on campuses must continue to adapt and change" is there a clear statement in this work about the policy changes that led up to the Omicron outbreak?
-The paper implies such claims, but I do not see them concretely stated (which is frustrating - this is related to point (9)).}
-
-Please see our response to point 9.
-
-\revtext{Methods:}
-
-\revtext{15) Regarding Rcontact, is there a difference between this and the generally-used R0? The definition seems the same... if not, it would be good to know the precise difference.}
-
-Rcontact and R0 are largely similar.
-But we wanted to use a different terminology to emphasize their subtle differences because Rcontact is a parameter in our individual-based model, which makes a specific set of assumptions that are different from continuous time SIR models using ordinary differential equations.
-We have tried to clarify the definition of Rcontact.
-
-``We note that the definition of the contact reproduction number $\Rc$ is similar to the standard definition of basic reproduction number $\RR_0$.
-The main difference is that the contact reproduction number models the number of total contacts, rather than infections. 
-Since infected individuals make their contacts at random with replacement, the same susceptible person could be contacted multiple times by the same or different infected individual during a time step---all these overlapping contacts will result in one infection.
-Therefore, the number of actual infections may be smaller than the number of contacts, especially since contacts can also land on non-susceptible individuals.
-We also note that the contact reproduction number implicitly accounts for all intervention measures that we do not model explicitly, such as social distancing and contact tracing---therefore, $\Rc$ is similar to the effective reproduction number, which typically accounts for intervention efforts.
-However, our contact reproduction number does not account for the effects of asymptomatic testing or vaccination, which are modeled separately.''
-
-\revtext{Overall, I like the paper and appreciate having the opportunity to review it.
-Understand that my critiques are intended to help improve the quality of the work prior to publication, which I generally support.}
+\revtext{The authors have done a good job responding to my previous set of comments and have made appropriate revisions to their manuscript.}
 
@@
 
+\revtext{Examining the updated timeseries results, my remaining concern regards the use of Pearson correlation of weekly case incidence in the analysis of dynamic coupling between community and campus, which features strongly in the significance statement and generally as a contribution of the paper.}
+
+\revtext{I'm not sure it's appropriate to give p values based on a null model of independent identically distributed samples, when timeseries data like this is autocorrelated and non-stationary.
+If possible this should be reviewed by someone with a background in analysis of coupled timeseries data.
+At minimum, the methods used to compute correlation should be given in more detail and the text should be examined to ensure that causal inference is not being overstated.}
+
+Thank you for pointing this out.
+We recognize that the causal language has been too strong.
+In particular, the causal inference is primarily supported by the mathematical modeling analysis, rather than the correlation analysis.
+We have revised the paper throughout to make the distinction clear and not overstate the implications of the correlations.
+Overall, we have tried to avoid linking correlations to causal statements.
+
+We have also added the following paragraph in the beginning of the results section to make the limitation of the method clearer:
+
+``
+In order to compare the transmission dynamics on PU campus and in nearby communities, we calculate the Pearson correlation between the weekly numbers of cases from PU and those from nearby communities.
+Confidence intervals are calculated using Fisher's Z transformation, and the corresponding significance is calculated using asymptomatic t approximation \citep{hollander2013nonparametric}.
+The standard method of calculating Perason correlations and the associated significance assumes that the samples are independent and identically distributed---these assumptions generally do not hold for epidemic time series data, which are autocorrelated and non-stationary.
+Nonetheless, the Pearson correlation has been the standard measure in epidemic time series analyses and can provide useful information about the strength of spatial coupling \citep{rohani1999opposite,grenfell2001travelling,viboud2006synchrony,baker2019epidemic}.
+We note correlation analyses do not provide direct evidence for causality and therefore should be interpreted with care---we later investigate the role of community transmission using a mathematical model.''
+
+\revtext{For example, it is stated that: "we are not able to infer the direction of causality—", but I am concerned this may be an understatement of the limitations with respect to making causal inference of any kind from the correlation results.}
+
+Thank you for pointing this out.
+We recognize that our phrasing was unclear. 
+This sentence was meant to point at the limitation of our modeling analysis, rather than the correlation analysis.
+We have tried to highlight the limitation of our correlation results clearer while making sure that we don't conflate conclusions from correlation analyses and modeling analyses:
+
+``There are several limitations to our analysis.
+While we demonstrate strong spatiotemporal correlation in the spread of SARS-CoV-2, correlations do not provide causal link for community transmission.
+For example, even if there was no mixing between campus and community populations, two populations can exhibit correlated epidemic dynamics if both populations are changing behavior in similar ways, reflecting changes in intervention measures in the greater New Jersey area.
+However, it seems biologically implausible that there would be no mixing between the two populatins---other studies have also identified community transmission to be an important source of SARS-CoV-2 infections on campus \citep{fox2021response,hamer2021assessment}.
+Our mathematical modeling illustrates that community transmission can explain the observed dynamics on PU campus, but we are not able to infer the direction of causality''
+
 \revtext{Reviewer \#2 Comments for the Author:}
 
-\revtext{In this manuscript, Park et al.
-retrospectively examine the transmission patterns of COVID-19 at Princeton University and across the wider community in Mercer County, NJ.
-The authors explore correlations between university and community case counts and also employ a mechanistic modeling framework to see if patterns in transmission can be explained by changes in R-effective and transmission from the broader community.
-Retrospective analyses like this are important, and valuable evidence generation for decision makers.}
+\revtext{Thank you for the detailed point-by-point response from the authors.
+I think the revisions really help with the clarity of the manuscript and allow for a more reliable/realistic comparison between community and campus dynamics.
+I also appreciate the extended sensitivity analysis that accounts for some of the additional epidemiological changes during Omicron.
+I am happy to recommend this work for publication.}
 
-Thank you.
-
-\revtext{I had two main concerns about the conclusions as presented that I hope the authors can address.
-The first is that while Princeton is an ideal ecosystem because of its asymptomatic screening protocols, the surrounding community is not necessarily one—this raises concerns about how differences in testing practices could be controlled for/accounted for in the correlation analysis.}
-
-\revtext{Most estimates seem to a huge reduction in case ascertainment during the Omicron surge, which makes comparisons between the broader community and Princeton tricky for that time point.
-At the very least, I would encourage the authors to control for differences in denominator (e.g., converting to per 100K/day) and/or exploring test positivity as an additional metric to see if their observations still hold.}
-
-We have revised our analysis and figures to account for changes in population sizes across the semester. See below for a detailed response.
-We also present testing volumes over time and show that they are reasonably stable across each semester.
-
-\revtext{The second is that I wasn’t necessarily convinced that the authors could conclude with certainty that the discrepancies during the Omicron period were driven by superspreading events because the way that this is created in the model is by seeding new infections, not by adjusting contact structure per se.
-Some more thoughts here: (1) I think the vaccine effectiveness estimates stated for BA.1/BA.2 were overly optimistic, so that would be something to explore further; and (2) the model (in theory) should allow for some superspreading as a consequence of the overdispersion parameter used for the negative binomial distribution of infectious contacts.
-I’d be curious to see if the model can replicate that change more mechanistically by adjusting the mean and overdispersion of the binomial distribution.}
-
-We recognize that our writing was unclear before.
-But we know from the observation that the large outbreak during the Omicron period was driven by superspreading event related to a campus event.
-We could not describe the event in detail due to privacy reasons, which led to unclear writing. 
-We have tried to make the causality clearer throughout (see response to Reviewer 1's comment 9).
-We also account for the possibility of shorter infectious periods and lower vaccine effectiveness.
-See below for detailed responses for each point.
-
-\revtext{Main comments:}
-
-\revtext{Fig.
-1- you might consider converting case counts for Princeton and Mercer County into cases per 100,000 per day (or some other denominator quantity).
-These plots are comparing cases for populations of ~15,000 for Princeton and ~360,000 for Mercer County, which makes Fig.
-1 panels I-L a bit difficult to parse at first—by controlling the denominator you could more easily compare to a 1:1 slope for panels E-J and a 1:1 ratio for panels I-L.
-This approach would also account for the changing student body size at Princeton across semesters with reopening, etc.}
-
-Thank you for the suggestion. 
-We have changed the main figure and corresponding supplementary figures for New York and Philadelphia to show weekly cases per 100,000.
-We have revised the writing accordingly and added a new Supplementary Figure comparing cases per 1000 for each population group on Princeton campus (Figure S1).
-We decided to keep the raw cases for comparisons with model projections since we are no longer comparing populations of two different sizes.
-
-\revtext{Lines 165-169- a key piece of this story seems to be variant predominance—you could consider adding a new figure with variant proportions through time or dashed vertical lines to Fig.
-1 time courses indicating when a new variant reached dominance.}
-
-We have variant data but we feel analyzing variant data is beyond the scope of this paper.
-We decided not to add the variant data in this paper also because it involves a different group of individuals analysing the data.
-And so we prefer to keep that analysis as a separate paper.
-
-\revtext{Lines 214-226: the assumption that testing patterns remain the same on campus and off campus throughout the entire investigation period doesn’t seem realistic.
-In particular for the beginning of the Omicron period, the Princeton university campus continued with surveillance screening, while the general population likely experienced significantly decreased ascertainment rates in part due to the increased availability and use of at home tests which are not mandated reporting (e.g., one estimate of a change of ascertainment from 50\% pre-Omicron to 20-29\% during BA.1; \url{https://www.medrxiv.org/content/10.1101/2022.04.22.22274198v3.full.pdf}).
-Therefore, I would hesitate to conclude that the lack of correlation was indicative of a lack of community transmission—changes in testing patterns could really be driving that pattern for the final semester.}
-
-Thank you for pointing this out. 
-We meant that the testing patterns remained constant for each semester, rather than the entire investigation period.
-We also meant that the case ratio would stay the same if testing patterns and transmission conditions were to remain constant, but we are not necessarily assuming that the testing patterns remained constant.
-We have tried to make the writing of this paragraph clearer while acknowledging changes in testing patterns over the investigation period.
-
-Also, we know that the outbreak during the final semester was associated with a specific campus event among students on one specific day, which we unfortunately cannot described in detail due to privacy reasons.
-So we know that the lack of correlation was likely driven by a lack of community transmission in this case. 
-We have tried to clarified this point.
-
-\revtext{Lines 243-245- Infectious periods changed with variants, e.g., Omicron << Delta. Was that component explored at all?}
-
-We now provide simulations that assume shorter infectious and latent period during the Omicron outbreak:
-
-``We considered the possibility that the Omicron variant can have shorter latent and infectious periods by decreasing the mean duration of latent, pre-symptomatic, and (a)symptomatic stages of infection by 0.5 days (therefore a total of 1.5 reduction in the duration of infection).
-In this case, a shorter generation interval can lead to faster growth rate given the same values of $\Rc$ \citep{wallinga2007generation}.
-However, we find that the effects of shorter infection has small effects on the overall dynamics (Supplementary Figure S13).''
-
-\revtext{Lines 296-305- While reading through this the first time, I was wondering about differences in vaccine effectiveness during Omicron emergence however—maybe hint to the reader here that you address that later or separately?}
-
-We have added the following sentence:
-
-``We note that these assumptions are specific to the Delta variant---we discuss vaccine effectiveness against the Omicron variant later on.''
-
-\revtext{Lines 453-454- I think your vaccine effectiveness estimates against Omicron are still quite optimistic, which could explain why ramping up your R-value didn’t really do much.
-In practice, we’ve seen closer to 30-50\% effectiveness against infection for example vs.
-70\% pre-waning (\url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9398552/})}
-
-We now provide simulations that assume 30\% vaccine effectiveness as a supplementary figure with the following text in the main manuscript:
-
-``We also considered the possibility that the vaccine effectiveness against the Omicron variant might be lower by repeating the same analysis with 30\% effectiveness against infection \citep{tan2023vaccine}.
-When the baseline $\Rc$ is low ($\Rc=2$), increasing $\Rc$ still does not increase the number of cases sufficiently.
-When we assume an intermediate value of $\Rc=4$, the model does a better job at capturing the dynamics but it does so by overestimating the trough before the policy change and underestimating the peak after the policy change.
-When we assume a high value of $\Rc=6$, the model overestimates both the trough and the peak.''
-
-\revtext{Fig.
-3-4- Apologies if you’ve stated this elsewhere and I’ve missed it, but have you taken a look at your testing volume through time?
-I know that the PU policy was x \# of tests per week based on vaccination status, but how was that enforced?
-I’m wondering if there is as much variation in the y-axis if you look at test positivity vs.
-raw number of cases?
-Another thing to consider, since I’m not sure about testing cadence and frequency would be to do some rolling averages to help smooth out differences across days/weeks.}
-
-We now show testing volume through time in Supplementary Figure S1.
-Everyone was required to be tested in order to be present on campus. 
-And as we see in Supplementary Figure S1, there was very good compliance and stable testing volumes within each semester, except during the end of fall semester of the 2021--2022 academic year (reflecting Omicron circulations) and the beginning of spring semester of the 2021--2022 academic year (reflecting students returning to school).
-Even then, these changes do not correlate with increase in cases---for example, the increase in testing volume at the end of fall semester of the 2021--2022 academic year is very sudden (doubling in a week), but the changes in cases at the same period is much smaller.
-
-We agree there is variation in cases within a week but testing patterns were fairly stable across weeks, which is why we chose to look at weekly data, instead of daily data.
-We feel that showing raw variation in weekly cases (rather than rolling averages) would be more accurate because we are also able to identify the sources of most variations (e.g., school holidays and classes ending).
-
-\revtext{Lines 456-457- instead of artificially introducing new infections, what happens if you decrease $k$ your overdispersion parameter to increase the potential variance (i.e., allow for bigger gatherings?)}
-
-Artificially introducing new infection is meant to emulate the increase in heterogeneity.
-Decreasing $k$ alone would certainly increase variance but we have no control over when large outbreaks will happen, meaning that most of the simulations will not be able to match the data.
-
-\revtext{Minor comments:}
-
-\revtext{Line 40- double check consistency of variant name capitalization throughout (e.g., “omicron” vs. “Omicron”)}
-
-Fixed.
-
-\revtext{Lines 78-81- infrastructure including ventilation would also play a key role here}
-
-We have added ventilation infrastructure to this sentence.
-
-\revtext{Line 192- “counties” instead of “countries”?}
-
-Fixed.
-
-\revtext{Lines 270-271- in reality, those contacts are likely much more structured leading to a more modular network (e.g.
-dormmates would have repeated contacts, students would be more likely to have classes with others in that same major); how would you see less mixing affecting your model results and interpretation?}
-
-We have added the following paragraph:
-
-``In reality, the contact structure among the campus population is likely more structured, exhibiting strong assortativity.
-For example, undergraduate students are more likely to mix with other undergraduate students, rather than graduate students or faculty and staff members.
-Even among undergraduate students, students are more likely to mix with their close friend group than with other students.
-On one hand, assortative mixing may lead to faster epidemic growth within certain population groups;
-on the other hand, it can also make the disease more difficult to spread among other groups that have lower contact rates.
-Therefore, predicting the impact of structured contact network requires more detailed information about whether the majority of cases were infected at random or from certain groups of the campus population.
-For simplicity, we assume random mixing throughout the paper---nonetheless, allowing for overdispersion in transmission is expected to emulate variability in epidemic growth rates driven by complex contact structures \citep{lloyd2005superspreading}.''
-
-\revtext{Line 278- it would be helpful to label $\theta$ as your “community transmission” parameter on first use, since you refer to it that way later, but not stated explicitly here.}
-
-Based on suggestions by Reviewer 1, we now normalize $\theta$ by population size $N$ and provide a clear definition the first time it is used>
-
-\revtext{Line 387- “cannot” instead of “can”?}
-
-``can'' is correct. We have revised this sentence to make the point clearer:
-
-``These simulations suggest that an increase in the number of cases in November can be explained by a combination of waning immunity alone without requiring additional changes in transmission dynamics (note we do not allow $\theta$ or $\Rc$ to vary over time)---we see that extending the simulation beyond November 26th still captures the increase in cases.''
-
-\revtext{Lines 162-164- state date of policy implementation to make link clearer for Figs. 3 \& 4}
-
-Done.
+Thank you for your review.
 

### 2023-04-21 doc/university-covid.tex
campus done

--- 2023-02-22 doc/university-covid.tex
@@
 Here, we present descriptive and modeling analyses of SARS-CoV-2 transmission on the Princeton University campus---this model was used throughout the pandemic to inform policy decisions and operational guidelines for the university campus.
-We demonstrate strong spatiotemporal correlations in epidemic patterns between the university campus and surrounding communities.
-These findings are corroborated by our model predictions, which indicate that the amount of on-campus transmission was likely limited during much of the wider pandemic until the end of 2021.
+Epidemic patterns between the university campus and surrounding communities exhibit strong spatiotemporal correlations.
+Mathematical modeling analysis further suggests that the amount of on-campus transmission was likely limited during much of the wider pandemic until the end of 2021.
 Finally, we find that a super-spreading event likely played a major role in driving the Omicron variant outbreak on the Princeton University campus during the spring semester of the 2021--2022 academic year.
@@
 Here, we show that this model can parsimoniously capture observed outbreak patterns on campus during different eras of control.
-Our analysis reveals that strong coupling between epidemic dynamics on campus and in local communities drives the early epidemic.
+Our modeling analysis further reveals that strong coupling between epidemic dynamics on campus and in local communities drives the early epidemic.
 Subsequently, especially in the Omicron era, superspreading events came to dominate transmission on campus, thereby weakening the dynamical coupling of campus and community outbreaks.
@@
 
+In order to compare the transmission dynamics on PU campus and in nearby communities, we calculate the Pearson correlation between the weekly numbers of cases from PU and those from nearby communities.
+Confidence intervals are calculated using Fisher's Z transformation, and the corresponding significance is calculated using asymptomatic t approximation \citep{hollander2013nonparametric}.
+The standard method of calculating Perason correlations and the associated significance assumes that the samples are independent and identically distributed---these assumptions generally do not hold for epidemic time series data, which are autocorrelated and non-stationary.
+Nonetheless, the Pearson correlation has been the standard measure in epidemic time series analyses and can provide useful information about the strength of spatial coupling \citep{rohani1999opposite,grenfell2001travelling,viboud2006synchrony,baker2019epidemic}.
+We note correlation analyses do not provide direct evidence for causality and therefore should be interpreted with care---we later investigate the role of community transmission using a mathematical model.
+
 Across the first three semesters, we find strong and significant correlations between the weekly numbers of cases from PU and those from Mercer county:
@@
 However, the case patterns in PU were decoupled from those in Mercer county for the spring semester of the 2021--2022 academic year with unclear correlations (\fref{princeton}H): $\rho = 0.47$ (95\% CI: $-0.18$--0.83; $p = 0.15$).
-Stratifying cases by subpopulation shows that case patterns in graduate students and faculty and staff members were still strongly correlated with case patterns in Mercer county, meaning that the campus transmission was limited to undergraduate students (Supplementary Figure S3).
+Stratifying cases by subpopulation shows that case patterns in graduate students and faculty and staff members were still strongly correlated with case patterns in Mercer county (Supplementary Figure S3).
 We also find strong correlations between the weekly logged numbers of cases from PU and those from other counties in New Jersey (Supplementary Figure S4)---these correlations significantly decreased with distance from Mercer county in both spring 2020--2021 ($\rho=-0.48$ (95\% CI: $-0.75$--$-0.06$; $p = 0.03$) for all cases and $\rho=-0.51$ (95\% CI: $-0.77$--$-0.10$; $p = 0.02$) for faculty and staff cases) and fall 2021--2022 ($\rho=-0.68$ (95\% CI: $-0.86$--$-0.35$, $p < 0.001$) for all cases and $\rho=-0.74$ (95\% CI: $-0.89$--$-0.46$, $p < 0.001$) for faculty and staff cases).
 Across the first three semesters, both the total cases and faculty and staff cases showed similar levels of correlations with local cases.
-For the spring semester of the 2021--2022 academic year, we still find high correlations between faculty and staff cases and local cases throughout other counties ($\rho > 0.8$ across all counties in New Jersey); however, the total cases exhibit considerably weaker correlations due to student-to-student transmission on campus (Supplementary Figure S4).
+For the spring semester of the 2021--2022 academic year, we still find high correlations between faculty and staff cases and local cases throughout other counties ($\rho > 0.8$ across all counties in New Jersey); however, the total cases exhibit considerably weaker correlations (Supplementary Figure S4).
 
@@
 We demonstrate strong spatiotemporal correlations between the patterns of spread of SARS-CoV-2 on campus and those from surrounding communities.
-These correlations decreased with distance from Mercer County in fall 2021--2022, likely reflecting contact and commuting patterns as the university campus reopened.
+These correlations decreased with distance from Mercer County in fall 2021--2022, indicating weaker spatial coupling in the dynamics.
 Mathematical modeling further suggests limited transmission between the university population during fall and spring semesters of the 2020--2021 academic year and an increased frequency of infective community contacts during the fall semester of the 2021--2022 academic year, compared to previous semesters.
@@
 There are several limitations to our analysis.
-While we demonstrate strong spatiotemporal correlation in the spread of SARS-CoV-2, we are not able to infer the direction of causality---that is, our analysis does not rule out the possibility that transmission on campus drove infections in nearby communities (as opposed to community transmission driving on-campus infections).
+While we demonstrate strong spatiotemporal correlation in the spread of SARS-CoV-2, correlations do not provide causal link for community transmission.
+For example, even if there was no mixing between campus and community populations, two populations can exhibit correlated epidemic dynamics if both populations are changing behavior in similar ways, reflecting changes in intervention measures in the greater New Jersey area.
+However, it seems biologically implausible that there would be no mixing between the two populatins---other studies have also identified community transmission to be an important source of SARS-CoV-2 infections on campus \citep{fox2021response,hamer2021assessment}.
+Our mathematical modeling illustrates that community transmission can explain the observed dynamics on PU campus, but we are not able to infer the direction of causality---that is, our analysis does not rule out the possibility that transmission on campus drove infections in nearby communities (as opposed to community transmission driving on-campus infections).
 However, seeding from campus is unlikely: 
@@
 
+\section*{Author contribution}
+
+All authors contributed equally to the research design, data collection, formal analysis, interpretation of the results, and writing of the paper.
+
 \bibliography{university-covid}
 
-\pagebreak
-
-\section*{Supplementary Figures}
-
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-
-\begin{figure}[!htp]
-\includegraphics[width=0.9\textwidth]{../figure_princeton_new/figure_princeton_testing.pdf}
-\caption{
-\textbf{The weekly number of PCR tests stratified by subpopulations.}
-}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_new_per_1000_bypop.pdf}
-\caption{
-\textbf{The weekly number of cases per 1000 stratified by subpopulations.}
-The case trajectory for undergraduate students was omitted for the fall semester of 2020--2021 academic year due to a very low number of undergraduate students present on campus.
-}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_correlation_per_1000.pdf}
-\caption{
-\textbf{Correlations between the weekly number of cases in PU and in Mercer County stratified by subpopulations.}
-Points represent the number of reported cases.
-Solid lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
-}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!htp]
-\begin{center}
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_map_new.pdf}
-\end{center}
-\caption{
-\textbf{Correlations between the weekly number of cases in PU and in counties in New Jersey.}
-(A--C) Map of correlations between the weekly number of cases in PU and in counties in New Jersey.
-(D--F) Relationship between case correlations and distance from Mercer County.
-Points represent the estimated correlation coefficients.
-Solid lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
-Black points and lines represent correlations based on all cases in PU.
-Red points and lines represent correlations based on cases among faculty and staff members in PU.
-}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_newyork_new.pdf}
-\caption{
-\textbf{Dynamics of SARS-CoV-2 outbreaks in PU and New York City.}
-(A--C) Epidemic trajectories across three semesters: Fall 2020 (A), Spring 2020 (B), and Fall 20201 (C).
-Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in PU.
-Red lines represent the weekly number of cases in New York City.
-(D--F) Correlations between the weekly number of cases in PU and in New York City.
-Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
-}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_philly_new.pdf}
-\caption{
-\textbf{Dynamics of SARS-CoV-2 outbreaks in PU and Philadelphia.}
-(A--C) Epidemic trajectories across three semesters: Fall 2020 (A), Spring 2020 (B), and Fall 20201 (C).
-Colored bar plots represent the weekly number of cases from both asymptomatic and symptomatic testing in PU.
-Red lines represent the weekly number of cases in Philadelphia.
-(D--F) Correlations between the weekly number of cases in PU and in Philadelphia.
-Solid lines and shaded areas represent the estimated linear regression lines and the associated 95\% CIs.
-}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!htp]
-\includegraphics[width=0.8\textwidth]{../figure_princeton_new/figure_princeton_map_extended.pdf}
-\caption{
-\textbf{Correlations between the weekly number of cases in PU and in counties in New Jersey, Pennsylvania, and New York State.}
-(A--C) Map of correlations between the weekly number of cases in PU and in counties in New Jersey, Pennsylvania, and New York State.
-(D--F) Relationship between case correlations and distance from Mercer County.
-Points represent the estimated correlation coefficients.
-Solid lines and shaded areas represent the regression line and the associated 95\% confidence intervals.
-Black points and lines represent correlations based on all cases in PU.
-Red points and lines represent correlations based on cases among faculty and staff members in PU.
-New York City is excluded from this analysis as the data provided by New York Times are not further stratified by county levels.
-}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_diagram/diagram.pdf}
-\caption{
-\textbf{Compartmental diagram of the individual-based model.}
-Each compartment represents a stage of infection: susceptible $S$, exposed $S$, pre-symptomatic $I_p$, symptomatic $I_s$, asymptomatic $I_a$, and recovered $R$.
-Pre-symptomatic, symptomatic, and asymptomatic stages are further divided into two subcompartments.
-Individuals in pre-symptomatic, symptomatic, and asymptomatic stages can test positive with 95\% sensitivity.
-}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_fall_2020_all.pdf}
-\caption{
-\textbf{Comparisons between model predictions and the observed numbers of cases for fall 2020.}
-Points represent the weekly number of reported cases in PU.
-Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
-See figure 2 in the main text for details.
-}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_spring_2020_all.pdf}
-\caption{
-\textbf{Comparisons between model predictions and the observed numbers of cases for spring 2020.}
-Points represent the weekly number of reported cases in PU.
-Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
-See figure 2 in the main text for details.
-}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_fall_2021_all.pdf}
-\caption{
-\textbf{Comparisons between model predictions and the observed numbers of cases for fall 2021.}
-Points represent the weekly number of reported cases in PU.
-Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
-See figure 2 in the main text for details.
-}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_nowaning.pdf}
-\caption{
-\textbf{Comparisons between model predictions and the observed numbers of cases for fall 2021 without immunity waning.}
-Points represent the weekly number of reported cases in PU.
-Red lines and shaded areas represent median model predictions and 90\% quantiles across 100 simulations.
-See figure 2 in the main text for details.
-}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_omicron_spring_shortI.pdf}
-\caption{
-\textbf{The impact of changes in testing frequency and an increased reproduction number on the spread of the Omicron variant, assuming a shorter latent and infectious periods.}
-Solid lines represent median predictions.
-Shaded areas represent 90\% quantiles across 100 simulations.
-Points represent the observed data.
-Vertical dashed lines represent the week including February 8, 2022, when distancing and testing policies were updated on PU campus.
-For each row, we assume a fixed value of baseline contact reproduction number $\Rc$ ranging from 2 to 6 across rows.
-Then, we simulate increase in $\Rc$ at the time of policy change (indicated by column labels).
-Here, the assumed mean latent, presymptomatic, and asymptomatic stages of infection are 0.5 days shorter than the assumed values for simulations presented in Figure 3 in the main text.
-}
-\end{figure}
-
-\pagebreak
-
-\begin{figure}[!htp]
-\includegraphics[width=\textwidth]{../figure_princeton_new/figure_princeton_simulation_omicron_spring_shortI.pdf}
-\caption{
-\textbf{The impact of changes in testing frequency and an increased reproduction number on the spread of the Omicron variant, assuming a lower vaccine effectiveness.}
-Solid lines represent median predictions.
-Shaded areas represent 90\% quantiles across 100 simulations.
-Points represent the observed data.
-Vertical dashed lines represent the week including February 8, 2022, when distancing and testing policies were updated on PU campus.
-For each row, we assume a fixed value of baseline contact reproduction number $\Rc$ ranging from 2 to 6 across rows.
-Then, we simulate increase in $\Rc$ at the time of policy change (indicated by column labels).
-Here, we assume 30\% vaccine effectiveness against infection.
-}
-\end{figure}
-
 \end{document}
